---
title: "P.evermanni GFF reformatting"
author: "Sam White"
date: "2025-01-17"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

# INTRO

This notebook reformats the original _P.evermanni_ GFF ([Porites_evermanni_v1.annot.gff](../data/Porites_evermanni_v1.annot.gff)), which is not compliant with [the GFF standard](https://github.com/the-sequence-ontology/specifications/blob/master/gff3.md) (GitHub page). Additionally, the GFF is lacking the `gene` feature, which may (or may not) be needed/useful for downstream processing. This notebook adds a `gene` feature.

Finally, despite the naming convention, there aren't any actual annotations in that GFF, beyond the feature designations (i.e. no gene ontology, no SwissProt IDs, gene names, etc.). This notebook does _not_ address those shortcomings.

::: {.callout-note}
Unlike other scripts, this will output to [E-Peve/data](../data), instead of an output directory in `../output`.
:::

## Software requirements

Requires [genometools](https://github.com/genometools/genometools) (GitHub repo) to be installed and in the system `$PATH`.

## Inputs

- [Porites_evermanni_v1.annot.gff](../data/Porites_evermanni_v1.annot.gff)

## Outputs

- [Porites_evermanni_validated.gff](../data/Porites_evermanni_validated.gff)

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```


# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""

echo "# Data directories"
echo 'export repo_dir=~/gitrepos/urol-e5/deep-dive-expression/'
echo 'export data_dir=${repo_dir}/E-Peve/data'
echo ""

echo "# Input files"
echo 'export original_gff="Porites_evermanni_v1.annot.gff"'
echo 'export intermediate_gff="intermediate.gff"'
echo 'export validated_gff="Porites_evermanni_validated.gff"'
echo ""

echo "# Print formatting"
echo 'export line="--------------------------------------------------------"'
echo ""
} > .bashvars

cat .bashvars
```


# Peak at original GFF
```{r view-original-GFF, engine='bash', eval=TRUE}
source .bashvars

head "${data_dir}/${original_gff}"
```

## Check features
```{r features-original-GFF, engine='bash', eval=TRUE}
source .bashvars

awk '{print $3}' "${data_dir}/${original_gff}" | sort --unique
```

# Fix GFF


```{r fix-gff, engine='bash', eval=TRUE}
source .bashvars

awk '
BEGIN { OFS="\t"; mrna_count = 0; utr_count = 0; gene_count = 0; cds_count = 0 }
{
    if ($3 == "mRNA") {
        split($9, attributes, ";")
        for (i in attributes) {
            if (attributes[i] ~ /^ID=/) {
                original_id = substr(attributes[i], 4)
                gene_id = "ID=gene-" original_id
                parent_id = "Parent=gene-" original_id
                break
            }
        }
        
        # Increment the global mRNA counter
        mrna_count++
        
        new_mrna_id = "ID=mrna-" sprintf("%05d", mrna_count)
        
        # Store the mapping of original mRNA ID to new mRNA ID
        mrna_map[original_id] = "mrna-" sprintf("%05d", mrna_count)
        
        # Replace the old ID with the new mRNA ID
        for (i in attributes) {
            if (attributes[i] ~ /^ID=/) {
                attributes[i] = new_mrna_id
                break
            }
        }
        $9 = attributes[1]
        for (i = 2; i <= length(attributes); i++) {
            $9 = $9 ";" attributes[i]
        }
        $9 = $9 ";" parent_id
        print $1, $2, "gene", $4, $5, ".", $7, $8, gene_id
        
        # Increment the gene counter and reset the CDS counter for each new gene
        gene_count++
        
        cds_count = gene_count
    } else if ($3 == "UTR") {
        utr_count++
        new_utr_id = "ID=utr-" sprintf("%06d", utr_count)
        split($9, attributes, ";")
        for (i in attributes) {
            if (attributes[i] ~ /^Parent=/) {
                original_parent_id = substr(attributes[i], 8)
                if (original_parent_id in mrna_map) {
                    attributes[i] = "Parent=" mrna_map[original_parent_id]
                }
                break
            }
        }
        $9 = new_utr_id
        for (i = 1; i <= length(attributes); i++) {
            $9 = $9 ";" attributes[i]
        }
    } else if ($3 == "CDS") {
        new_cds_id = "ID=cds-" sprintf("%06d", cds_count)
        split($9, attributes, ";")
        for (i in attributes) {
            if (attributes[i] ~ /^Parent=/) {
                original_parent_id = substr(attributes[i], 8)
                if (original_parent_id in mrna_map) {
                    attributes[i] = "Parent=" mrna_map[original_parent_id]
                }
                break
            }
        }
        $9 = new_cds_id
        for (i = 1; i <= length(attributes); i++) {
            $9 = $9 ";" attributes[i]
        }
    } else {
        split($9, attributes, ";")
        for (i in attributes) {
            if (attributes[i] ~ /^Parent=/) {
                original_parent_id = substr(attributes[i], 8)
                if (original_parent_id in mrna_map) {
                    attributes[i] = "Parent=" mrna_map[original_parent_id]
                }
                break
            }
        }
        $9 = attributes[1]
        for (i = 2; i <= length(attributes); i++) {
            $9 = $9 ";" attributes[i]
        }
    }
    print $0
}' "${data_dir}"/"${original_gff}" > "${data_dir}"/"${intermediate_gff}"
```

## Inspect intermediate GFF
```{r intermediate-gff, engine='bash', eval=TRUE}
source .bashvars

head "${data_dir}"/"${intermediate_gff}"
```

# Validate GFF

Validate GFF using genometools `gff3`.

- `-tidy`: Attempts to clean/fix any potential issues.

- `-checkids`: Checks IDs.

- `-retainids`: Retains IDs from input GFF instead of assigning new ones.

```{r validate-gff, engine='bash', eval=FALSE}
source .bashvars

gt gff3 \
-tidy \
-checkids \
-retainids \
"${data_dir}"/"${intermediate_gff}" \
> "${data_dir}"/"${validated_gff}" \
2> "${data_dir}"/gt_gff3_validation_errors.log
```

## Check for error(s) in validation

Process would stop if error occurred, so only need to check end of file.

::: {.callout-note}
Warnings are expected, since file was not formatted correctly. They indicate changes being made to bring format into compliance.
:::

```{r validate-gff-error-check, engine='bash', eval=TRUE}
source .bashvars

tail "${data_dir}"/gt_gff3_validation_errors.log
```

## Inspect Validated GFF
```{r validated-gff, engine='bash', eval=TRUE}
source .bashvars

head "${data_dir}"/"${validated_gff}"
```

# Remove intermediate GFF
```{r rm-intermediate-gff, engine='bash', eval=FALSE}
source .bashvars

rm "${data_dir}"/"${intermediate_gff}"
```