---
title: "mRNA-miRNA interactions"
author: "Jill Ashey"
date: "2025-02-06"
output: html_document
---

This script will investigate mRNA-miRNA interactions using miranda predictions. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#BiocManager::install("SPONGE")

library(tidyverse)
#library(mirTarRnaSeq)
library(reshape2)
#library(SPONGE)
library(pheatmap)
library(energy)
library(parallel)
library(ggraph)
library(tidygraph)
library(igraph)
library(genefilter)
library(gridExtra)
```

Read in mRNA data 
```{r}
mRNA_counts <- read_csv("../../E-Peve/output/06.2-Peve-Hisat/Peve-gene_count_matrix.csv")
mRNA_counts <- as.data.frame(mRNA_counts)
rownames(mRNA_counts) <- mRNA_counts[,1] #set first column that contains gene names as rownames
mRNA_counts <- mRNA_counts[,-1] # remove column w/ gene names 

# Remove any genes with 0 counts across samples 
mRNA_counts<-mRNA_counts %>%
     mutate(Total = rowSums(.[, 1:5]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)
```

Read in miRNA data
```{r}
miRNA_counts <- read.delim("../../E-Peve/output/03.1-Peve-sRNA-summary/Peve_counts_miRNA_normalized.txt")
head(miRNA_counts)

# Remove sample150 column from mirna counts and remove any miRNAs with 0 for all samples 
miRNA_counts <- miRNA_counts %>%
  #select(-sample150) %>%
     mutate(Total = rowSums(.[, 1:3]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)

# Rename gene count cols to match miRNA count cols
#colnames(mRNA_counts) <- c("sample140", "sample145", "sample150", "sample173", "sample178")
```

Only 3 miRNA samples successfully sequenced for this species. Remove the samples that did not work from the mRNA df and rename columns to match miRNA df
```{r}
mRNA_counts <- mRNA_counts %>%
  dplyr::select("RNA-POR-73", "RNA-POR-79", "RNA-POR-82") %>%
  dplyr::rename("sample73" = "RNA-POR-73", "sample79" = "RNA-POR-79", "sample82" = "RNA-POR-82")
```

Read in miranda data 
```{r}
miranda_peve <- read.delim("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_peve_updated.txt", header = F)
colnames(miranda_peve) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_peve$miRNA <- sub("^>", "", miranda_peve$miRNA)  # Remove leading ">"
miranda_peve$miRNA <- sub("\\..*", "", miranda_peve$miRNA)  # Remove everything from the first period onwards
miranda_peve$mRNA <- sub(";.*", "", miranda_peve$mRNA)  # Remove everything from ";" onwards
miranda_peve$mRNA <- sub("ID=", "", miranda_peve$mRNA)  # Remove ID=

length(unique(miranda_peve$miRNA))
length(unique(miranda_peve$mRNA))
dim(miranda_peve)
```

Summarize how many mRNAs each miRNA are predicted to interact with. Summarize how many miRNAs each mRNA are predicted to interact with. 
```{r}
sum_mirna <- miranda_peve %>%
  group_by(miRNA) %>%
  summarise(mRNA_count = n_distinct(mRNA)) %>% 
  arrange(desc(mRNA_count))

ggplot(sum_mirna %>% arrange(desc(mRNA_count)), 
       aes(x = forcats::fct_reorder(miRNA, mRNA_count), y = mRNA_count)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "miRNA", y = "mRNA Count", title = "miRNA Regulation of mRNAs") #+
  #coord_flip()  # Optional: horizontal bar plot for better readability
ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miRNA_number_of_targets.png", last_plot(), width = 12, height = 10, dpi = 300)
ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miRNA_number_of_targets.pdf", last_plot(), width = 12, height = 10, dpi = 300)

# mRNA
sum_mrna <- miranda_peve %>%
  group_by(mRNA) %>%
  summarise(miRNA_count = n_distinct(miRNA))
```

Normalize counts
```{r}
# Function to normalize counts (simple RPM normalization)
normalize_counts <- function(counts) {
  rpm <- t(t(counts) / colSums(counts)) * 1e6
  return(rpm)
}

# Normalize miRNA and mRNA counts
miRNA_norm <- normalize_counts(miRNA_counts)
#miRNA_norm <- as.matrix(miRNA_counts_filt)

mRNA_norm <- normalize_counts(mRNA_counts)
#mRNA_norm <- as.matrix(mRNA_counts_filt)
```

In my [Apul mRNA-miRNA interactions code](https://github.com/urol-e5/deep-dive-expression/blob/main/D-Apul/code/09-Apul-mRNA-miRNA-interactions.Rmd), I calculated the distance correlation coefficient using the energy package. I am not going to do that now, but may come back to it later. 

Calculate PCC  
```{r}
# Function to calculate PCC and p-value for a pair of vectors
calc_pcc <- function(x, y) {
  result <- cor.test(x, y, method = "pearson")
  return(c(PCC = result$estimate, p_value = result$p.value))
}

# Create a data frame of all miRNA-mRNA pairs
pairs <- expand.grid(miRNA = rownames(miRNA_norm), mRNA = rownames(mRNA_norm))

# Calculate PCC and p-value for each pair
pcc_results <- pairs %>%
  rowwise() %>%
  mutate(
    pcc_stats = list(calc_pcc(miRNA_norm[miRNA,], mRNA_norm[mRNA,]))
  ) %>%
  unnest_wider(pcc_stats)

# Adjust p-values for FDR
pcc_results <- pcc_results %>%
  mutate(adjusted_p_value = p.adjust(p_value, method = "fdr"))

# Remove 'gene-' from mRNA names 
pcc_results$mRNA <- gsub("gene-", "", pcc_results$mRNA)

# Save as csv
write.csv(pcc_results, "../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-PCC_miRNA_mRNA.csv")
```

Merge with miranda data 
```{r}
combined_data_pcc <- pcc_results %>%
  inner_join(miranda_peve, by = c("miRNA", "mRNA")) %>%
  na.omit()
dim(combined_data_pcc)
length(unique(combined_data_pcc$miRNA))
length(unique(combined_data_pcc$mRNA))

# Save as csv
write.csv(combined_data_pcc, "../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miranda_PCC_miRNA_mRNA.csv")
```

Plot number of positive and negative interactions for each miRNA
```{r}
# Summarize the data
summary_data <- combined_data_pcc %>%
  group_by(miRNA) %>%
  summarise(
    positive_count = sum(PCC.cor > 0),
    negative_count = sum(PCC.cor < 0)
  ) %>%
  mutate(total_count = positive_count + negative_count) %>%
  arrange(desc(total_count))
sum(summary_data$positive_count)
sum(summary_data$negative_count)

# Assess how many miRNAs have more positive correlations vs more negative ones 
positive_dominant <- sum(summary_data$positive_count > summary_data$negative_count)
negative_dominant <- sum(summary_data$negative_count > summary_data$positive_count)
equal <- sum(summary_data$positive_count == summary_data$negative_count)

cat("miRNAs with more positive correlations:", positive_dominant, "\n")
cat("miRNAs with more negative correlations:", negative_dominant, "\n")
cat("miRNAs with equal positive and negative correlations:", equal, "\n")

# Reshape the data for plotting
plot_data <- summary_data %>%
  tidyr::pivot_longer(cols = c(positive_count, negative_count),
                      names_to = "correlation_type",
                      values_to = "count")

# Create the stacked bar plot
ggplot(plot_data, aes(x = reorder(miRNA, -total_count), y = count, fill = correlation_type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(x = "miRNA", y = "Count", fill = "Correlation Type",
       title = "Positive and Negative PCC Correlations per miRNA")
ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miRNA_positive_negative_correlations.png", last_plot(), width = 12, height = 10, dpi = 300)
ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miRNA_positive_negative_correlations.pdf", last_plot(), width = 12, height = 10, dpi = 300)
```

Investigate data 
```{r}
# How many p-values are < 0.05 or < 0.1?
pvalue_summary_pcc <- combined_data_pcc %>%
  summarise(
    pvalue_0.05 = sum(p_value < 0.05),
    pvalue_0.1 = sum(p_value < 0.1),
  )
print(pvalue_summary_pcc)

# How many pairs have a PCC correlation > |0.5|?
corr_0.5 <- sum(abs(combined_data_pcc$PCC.cor) > 0.5)
cat("\nPairs with Pearson Correlation > 0.5:", corr_0.5, "\n")

# Are there any pairs that have a PCC correlation > |0.5| and a p-value < 0.05?
pairs_of_interest_pcc <- combined_data_pcc %>%
  filter(abs(PCC.cor) > 0.5 & p_value < 0.05 )
cat("PCC correlation > |0.5| and a p-value < 0.05:", nrow(pairs_of_interest_pcc), "\n")

# How many unique miRNAs and mRNAs have a PCC correlation > |0.5| and a p-value < 0.05?
length(unique(pairs_of_interest_pcc$miRNA))
length(unique(pairs_of_interest_pcc$mRNA))

# Save pairs of interest df as csv 
write.csv(pairs_of_interest_pcc, "../../D-Apul/output/09-Apul-mRNA-miRNA-interactions/Apul-miranda_PCC_sig_miRNA_mRNA.csv")
```

Plot number of SIGNIFICANT positive and negative interactions for each miRNA
```{r}
# Summarize the data
summary_data <- pairs_of_interest_pcc %>%
  group_by(miRNA) %>%
  summarise(
    positive_count = sum(PCC.cor > 0),
    negative_count = sum(PCC.cor < 0)
  ) %>%
  mutate(total_count = positive_count + negative_count) %>%
  arrange(desc(total_count))
sum(summary_data$positive_count)
sum(summary_data$negative_count)

# Assess how many miRNAs have more positive correlations vs more negative ones 
positive_dominant <- sum(summary_data$positive_count > summary_data$negative_count)
negative_dominant <- sum(summary_data$negative_count > summary_data$positive_count)
equal <- sum(summary_data$positive_count == summary_data$negative_count)

cat("miRNAs with more positive correlations:", positive_dominant, "\n")
cat("miRNAs with more negative correlations:", negative_dominant, "\n")
cat("miRNAs with equal positive and negative correlations:", equal, "\n")

# Reshape the data for plotting
plot_data <- summary_data %>%
  tidyr::pivot_longer(cols = c(positive_count, negative_count),
                      names_to = "correlation_type",
                      values_to = "count")

# Create the stacked bar plot
ggplot(plot_data, aes(x = reorder(miRNA, -total_count), y = count, fill = correlation_type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(x = "miRNA", y = "Count", fill = "Correlation Type",
       title = "Significant Positive and Negative PCC Correlations per miRNA")

ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miRNA_significant_positive_negative_correlations.png", last_plot(), width = 12, height = 10, dpi = 300)
ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miRNA_significant_positive_negative_correlations.pdf", last_plot(), width = 12, height = 10, dpi = 300)
```

Plot using igraph 
```{r}
# Create the graph
g <- graph_from_data_frame(pairs_of_interest_pcc, directed = FALSE)

# Add edge attributes
E(g)$weight <- abs(E(g)$PCC.cor)  # Use absolute PCC for edge weight
E(g)$color <- ifelse(E(g)$PCC.cor > 0, "blue", "red")  # Blue for positive, red for negative correlations

# Add node attributes
V(g)$type <- ifelse(V(g)$name %in% pairs_of_interest_pcc$miRNA, "miRNA", "mRNA")

# Convert to tbl_graph for ggraph
g_tbl <- as_tbl_graph(g)

# Create the plot
p <- ggraph(g_tbl, layout = "fr") +
  geom_edge_link(aes(edge_width = weight, color = color), alpha = 0.6) +
  geom_node_point(aes(color = type), size = 5) +
  #geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_width(range = c(0.5, 3)) +
  scale_color_manual(values = c("miRNA" = "lightblue", "mRNA" = "lightgreen", "Positive correlation" = "blue", "Negative correlation" = "red")) +
  theme_graph() +
  labs(title = "miRNA-mRNA Interaction Network",
       subtitle = "Edge width represents |PCC|, color represents correlation direction");p
ggsave("../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-significant_miRNA_mRNA_network.png", p, width = 12, height = 10, dpi = 300)
```
