---
title: "12-Peve-WGBS-bismark.Rmd"
author: "Zoe Dellaert"
date: "2025-04-10"
output: 
  github_document:
    toc: true
    number_sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output
)
```

## This is the downstream methylation analysis of the WGBS data for *Porites evermanni*

Reads were trimmed and QC'd in [this code](https://github.com/urol-e5/deep-dive-expression/blob/main/E-Peve/code/01.00-E-Peve-WGBS-trimming-cutadapt-FastQC-MultiQC.md)

## Important file locations:

1. [Trimmed WGBS Reads](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/01.00-E-Peve-WGBS-trimming-cutadapt-FastQC-MultiQC/)
2. [Bismark Genome](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/data/Bisulfite_Genome/)
3. [All Bismark output files (BAMs, .cov files, .bedgraph files)](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/)
4. [Bismark and Qualimap MultiQC](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/multiqc_report.html) and [base bismark report](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/bismark_summary_report.html)

### Note: Most of this code is based on the [E5 Time Series Molecular](https://github.com/urol-e5/timeseries_molecular) code by Steven Roberts [here](https://github.com/urol-e5/timeseries_molecular/blob/main/D-Apul/code/15.5-Apul-bismark.qmd)

## Generate Bismark Bisulfite Genome

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --nodes=1 --ntasks-per-node=20
#SBATCH --mem=200GB
#SBATCH -t 24:00:00
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails
#SBATCH --error=scripts/outs_errs/"%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output=scripts/outs_errs/"%x_output.%j" #once your job is completed, any final job report comments will be put in this file

# load modules needed
module load uri/main
module load Bismark/0.23.1-foss-2021b
module load bowtie2/2.5.2

cd ../data

bismark_genome_preparation --verbose --parallel 10 ./
```

### output:

```{bash, eval=FALSE}
Using 10 threads for the top and bottom strand indexing processes each, so using 20 cores in total
Writing bisulfite genomes out into a single MFA (multi FastA) file

Bisulfite Genome Indexer version v0.23.1 (last modified: 27 Jan 2021)

Step I - Prepare genome folders - completed



Step II - Genome bisulfite conversions - completed


Bismark Genome Preparation - Step III: Launching the Bowtie 2 indexer
Preparing indexing of CT converted genome in /scratch3/workspace/zdellaert_uri_edu-deep_dive/deep-dive-expression/E-Peve/data/Bisulfite_Genome/CT_conversion/
Building a SMALL index
Preparing indexing of GA converted genome in /scratch3/workspace/zdellaert_uri_edu-deep_dive/deep-dive-expression/E-Peve/data/Bisulfite_Genome/GA_conversion/
Building a SMALL index
Renaming BS_GA.3.bt2.tmp to BS_GA.3.bt2
Renaming BS_GA.4.bt2.tmp to BS_GA.4.bt2
Renaming BS_GA.1.bt2.tmp to BS_GA.1.bt2
Renaming BS_GA.2.bt2.tmp to BS_GA.2.bt2
Renaming BS_GA.rev.1.bt2.tmp to BS_GA.rev.1.bt2
Renaming BS_GA.rev.2.bt2.tmp to BS_GA.rev.2.bt2
Renaming BS_CT.3.bt2.tmp to BS_CT.3.bt2
Renaming BS_CT.4.bt2.tmp to BS_CT.4.bt2
Renaming BS_CT.1.bt2.tmp to BS_CT.1.bt2
Renaming BS_CT.2.bt2.tmp to BS_CT.2.bt2
Renaming BS_CT.rev.1.bt2.tmp to BS_CT.rev.1.bt2
Renaming BS_CT.rev.2.bt2.tmp to BS_CT.rev.2.bt2
```

### Compress and generate md5

```{bash, eval=FALSE}
cd ../data
tar -czvf Bisulfite_Genome.tar.gz Bisulfite_Genome
md5sum Bisulfite_Genome.tar.gz | tee Bisulfite_Genome.tar.gz.md5
```

```{bash, eval=FALSE}
885c84c233ca313f23b8670903a18db6  Bisulfite_Genome.tar.gz
```

### Output file location: [Bismark Genome](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/data/Bisulfite_Genome/)

## Test parameters 

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH --ntasks=1 --cpus-per-task=30 #split one task over multiple CPU
#SBATCH --array=0-4 #for 5 samples
#SBATCH --mem=100GB
#SBATCH -t 00:30:00
#SBATCH --mail-type=END,FAIL,TIME_LIMIT_80 #email you when job stops and/or fails or is nearing its time limit
#SBATCH --error=scripts/outs_errs/"%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output=scripts/outs_errs/"%x_output.%j" #once your job is completed, any final job report comments will be put in this file

# load modules needed
module load uri/main
module load Bismark/0.23.1-foss-2021b
module load bowtie2/2.5.2

# Set directories and files
reads_dir="../output/01.00-E-Peve-WGBS-trimming-cutadapt-FastQC-MultiQC/"
genome_folder="../data/"

output_dir="../output/12-Peve-WGBS/bismark_paramtest_cutadapt"
checkpoint_file="${output_dir}/completed_samples.log"

# make output directory
mkdir -p ${output_dir}

# Create the checkpoint file if it doesn't exist
touch ${checkpoint_file}

# Get the list of sample files and corresponding sample names
files=(${reads_dir}*_R1_001.fastq.gz)
file="${files[$SLURM_ARRAY_TASK_ID]}"
sample_name=$(basename "$file" "_R1_001.fastq.gz")

# Check if the sample has already been processed
if grep -q "^${sample_name}$" ${checkpoint_file}; then
    echo "Sample ${sample_name} already processed. Skipping..."
    exit 0
fi

# Define log files for stdout and stderr
stdout_log="${output_dir}/${sample_name}_stdout.log"
stderr_log="${output_dir}/${sample_name}_stderr.log"

# Define the array of score_min parameters to test
score_min_params=(
    "L,0,-0.4"
    "L,0,-0.6"
    "L,0,-0.8"
    "L,0,-1.0"
    "L,-1,-0.6"
)

# Loop through each score_min parameter
for score_min in "${score_min_params[@]}"; do
    echo "Running Bismark for sample ${sample_name} with score_min ${score_min}"
    
    # Create a subdirectory for this parameter
    param_output_dir="${output_dir}/${sample_name}_score_${score_min//,/}"
    mkdir -p ${param_output_dir}

    # Run Bismark alignment
    bismark \
        -genome ${genome_folder} \
        -p 8 \
        -u 10000 \
        -score_min ${score_min} \
        --non_directional \
        -1 ${reads_dir}${sample_name}_R1_001.fastq.gz \
        -2 ${reads_dir}${sample_name}_R2_001.fastq.gz \
        -o ${param_output_dir} \
        --basename ${sample_name}_${score_min//,/} \
        2> "${param_output_dir}/${sample_name}-bismark_summary.txt"

    # Check if the command was successful
    if [ $? -eq 0 ]; then
        echo "Sample ${sample_name} with score_min ${score_min} processed successfully."
    else
        echo "Sample ${sample_name} with score_min ${score_min} failed. Check ${stderr_log} for details."
    fi
done

# Mark the sample as completed in the checkpoint file
if [ $? -eq 0 ]; then
    echo ${sample_name} >> ${checkpoint_file}
    echo "All tests for sample ${sample_name} completed."
else
    echo "Sample ${sample_name} encountered errors. Check logs for details."
fi

# Define directories
summary_file="${output_dir}/parameter_comparison_summary.csv"

# Initialize summary file
echo "Sample,Score_Min,Alignment_Rate" > ${summary_file}

# Loop through parameter output directories
for dir in ${output_dir}/*_score_*; do
    if [ -d "$dir" ]; then
        # Extract sample name and score_min parameter from directory name
        sample_name=$(basename "$dir" | cut -d '_' -f1-3)
        score_min=$(basename "$dir" | grep -o "score_.*" | sed 's/score_//; s/_/,/g')

        # Locate the summary file
        summary_file_path="${dir}/${sample_name}_${score_min}_PE_report.txt"

        # Extract metrics
        mapping=$(grep "Mapping efficiency:" ${summary_file_path} | awk '{print "mapping efficiency ", $3}')
        
        # Append to the summary file
        echo "${sample_name},${score_min},${mapping}" >> ${summary_file}
    fi
done
```

### Results from parameter tests:
 
| Sample          | Score_Min | Alignment_Rate |
|-----------------|-----------|----------------|
| trimmed_421_S8  | L0-0.4    | 29.20%         |
| trimmed_421_S8  | L0-0.6    | 36.00%         |
| trimmed_421_S8  | L0-0.8    | 41.00%         |
| trimmed_421_S8  | L0-1.0    | 47.30%         |
| trimmed_421_S8  | L-1-0.6   | 36.30%         |
| trimmed_471_S6  | L0-0.4    | 35.90%         |
| trimmed_471_S6  | L0-0.6    | 42.80%         |
| trimmed_471_S6  | L0-0.8    | 48.40%         |
| trimmed_471_S6  | L0-1.0    | 54.70%         |
| trimmed_471_S6  | L-1-0.6   | 43.00%         |
| trimmed_487_S9  | L0-0.4    | 34.60%         |
| trimmed_487_S9  | L0-0.6    | 41.40%         |
| trimmed_487_S9  | L0-0.8    | 47.50%         |
| trimmed_487_S9  | L0-1.0    | 54.20%         |
| trimmed_487_S9  | L-1-0.6   | 41.80%         |
| trimmed_489_S10 | L0-0.4    | 32.50%         |
| trimmed_489_S10 | L0-0.6    | 40.30%         |
| trimmed_489_S10 | L0-0.8    | 47.00%         |
| trimmed_489_S10 | L0-1.0    | 54.20%         |
| trimmed_489_S10 | L-1-0.6   | 40.70%         |
| trimmed_491_S7  | L0-0.4    | 42.00%         |
| trimmed_491_S7  | L0-0.6    | 50.50%         |
| trimmed_491_S7  | L0-0.8    | 56.60%         |
| trimmed_491_S7  | L0-1.0    | 62.10%         |
| trimmed_491_S7  | L-1-0.6   | 50.80%         |

I ran the parameter testing before fixing the file sample names. For reference:

- 421 = POR-82
- 471 = POR-76
- 487 = POR-71
- 489 = POR-79
- 491 = POR-73

## Align to genome

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH --ntasks=1 --cpus-per-task=48 #split one task over multiple CPU
#SBATCH --array=0-4 #for 5 samples
#SBATCH --mem=400GB
#SBATCH -t 48:00:00
#SBATCH --mail-type=END,FAIL,TIME_LIMIT_80 #email you when job stops and/or fails or is nearing its time limit
#SBATCH --error=scripts/outs_errs/"%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output=scripts/outs_errs/"%x_output.%j" #once your job is completed, any final job report comments will be put in this file

# load modules needed
module load uri/main
module load Bismark/0.23.1-foss-2021b
module load bowtie2/2.5.2

# Set directories and files
reads_dir="../output/01.00-E-Peve-WGBS-trimming-cutadapt-FastQC-MultiQC/"
genome_folder="../data/"

output_dir="../output/12-Peve-WGBS/bismark_cutadapt"
checkpoint_file="${output_dir}/completed_samples.log"

# make output directory
mkdir -p ${output_dir}

# Create the checkpoint file if it doesn't exist
touch ${checkpoint_file}

# Get the list of sample files and corresponding sample names
files=(${reads_dir}*_R1_001.fastq.gz)
file="${files[$SLURM_ARRAY_TASK_ID]}"
sample_name=$(basename "$file" "_R1_001.fastq.gz")

# Check if the sample has already been processed
if grep -q "^${sample_name}$" ${checkpoint_file}; then
    echo "Sample ${sample_name} already processed. Skipping..."
    exit 0
fi

# Define log files for stdout and stderr
stdout_log="${output_dir}/${sample_name}_stdout.log"
stderr_log="${output_dir}/${sample_name}_stderr.log"

    # Run Bismark alignment
    bismark \
        -genome ${genome_folder} \
        -p 48 \
        -score_min L,0,-1.0 \
        --non_directional \
        -1 ${reads_dir}${sample_name}_R1_001.fastq.gz \
        -2 ${reads_dir}${sample_name}_R2_001.fastq.gz \
        -o ${output_dir} \
        --basename ${sample_name} \
        2> "${output_dir}/${sample_name}-bismark_summary.txt"

# Check if the command was successful
if [ $? -eq 0 ]; then
    # Append the sample name to the checkpoint file
    echo ${sample_name} >> ${checkpoint_file}
    echo "Sample ${sample_name} processed successfully."
else
    echo "Sample ${sample_name} failed. Check ${stderr_log} for details."
fi

# Define directories
summary_file="${output_dir}/alignment_summary.csv"

# Initialize summary file
echo "Sample,Score_Min,Alignment_Rate" > ${summary_file}

# Loop through parameter output directories
for file in ${output_dir}/*_report.txt; do
    # Extract sample name and score_min parameter from directory name
    sample_name=$(basename "$file" | cut -d'_' -f1-3)
    score_min="L0-1.0"

    # Locate the summary file
    summary_file_path="${output_dir}/${sample_name}_PE_report.txt"

    # Extract metrics
    mapping=$(grep "Mapping efficiency:" ${summary_file_path} | awk '{gsub("%", "", $3); print $3}')

    # Append to the summary file
    echo "${sample_name},${score_min},${mapping}" >> ${summary_file}
done
```

### Output file location: [All Bismark output files](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/)

## Post-alignment code is based once again on [Steven's code](https://github.com/urol-e5/timeseries_molecular/blob/main/D-Apul/code/15.5-Apul-bismark.qmd)

### Deduplication, Sorting, and methylation extraction & calling

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --nodes=1 --ntasks-per-node=24
#SBATCH --mem=250GB
#SBATCH -t 24:00:00
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails
#SBATCH --error=scripts/outs_errs/"%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output=scripts/outs_errs/"%x_output.%j" #once your job is completed, any final job report comments will be put in this file

# load modules needed
module load uri/main
module load Bismark/0.23.1-foss-2021b
module load parallel/20240822

# set directories
bismark_dir="../output/12-Peve-WGBS/bismark_cutadapt/"
genome_folder="../data/"

### deduplicate bams
find ${bismark_dir}*.bam | \
xargs -n 1 basename | \
sed 's/^trimmed_//' | sed 's/_pe.bam$//' | \
parallel -j 8 deduplicate_bismark \
--bam \
--paired \
--output_dir ${bismark_dir} \
${bismark_dir}trimmed_{}_pe.bam

### methylation extraction

find ${bismark_dir}*deduplicated.bam | xargs -n 1 -I{} \
bismark_methylation_extractor --bedGraph --counts --comprehensive --merge_non_CpG \
--multicore 24 --buffer_size 75% --output ${bismark_dir} "{}"

### methylation call

find ${bismark_dir}*deduplicated.bismark.cov.gz | \
xargs -n 1 basename | \
sed 's/^trimmed_//' | sed 's/_pe.deduplicated.bismark.cov.gz$//' | \
parallel -j 24 coverage2cytosine \
--genome_folder ${genome_folder} \
-o ${bismark_dir}{} \
--merge_CpG \
--zero_based \
${bismark_dir}trimmed_{}_pe.deduplicated.bismark.cov.gz

### sort bams

# change modules

module purge
module load samtools/1.19.2

find ${bismark_dir}*deduplicated.bam | \
xargs -n 1 basename | \
sed 's/^trimmed_//' | sed 's/_pe.deduplicated.bam$//' | \
xargs -I{} samtools \
sort --threads 24 \
${bismark_dir}trimmed_{}_pe.deduplicated.bam \
-o ${bismark_dir}{}.sorted.bam
```

This took just  over 6 hours with max memory used per node as 249.99GiB.

### View output

```{bash, eval=FALSE}
head ${bismark_dir}*evidence.cov
```


### Make summary reports 

```{bash, eval=FALSE}
#!/usr/bin/env bash
#SBATCH --export=NONE
#SBATCH --nodes=1 --ntasks-per-node=8
#SBATCH --mem=250GB
#SBATCH -t 08:00:00
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails
#SBATCH --error=scripts/outs_errs/"%x_error.%j" #if your job fails, the error report will be put in this file
#SBATCH --output=scripts/outs_errs/"%x_output.%j" #once your job is completed, any final job report comments will be put in this file

module load uri/main
module load Bismark/0.23.1-foss-2021b
module load all/MultiQC/1.12-foss-2021b
module load qualimap/2.2.1

cd ../output/12-Peve-WGBS/bismark_cutadapt/

bam2nuc --genome_folder ../../../data/ *_pe.deduplicated.bam

mkdir -p qualimap/bamqc

for bamFile in *sorted.bam; do
	prefix=$(basename $bamFile .bam)

	qualimap \
    --java-mem-size=29491M \
    bamqc \
     \
    -bam ${bamFile}  \
     \
    -p non-strand-specific \
    --collect-overlap-pairs \
    -outdir qualimap/bamqc/${prefix} \
    -nt 6
done

bismark2report
bismark2summary *pe.bam

multiqc .
```

#### Output file location: [Bismark and Qualimap MultiQC](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/multiqc_report.html) and [base bismark report](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/bismark_summary_report.html)

### Sorting cov files and filtering for coverage and gene intersection

```{bash, eval=FALSE}
# salloc -p cpu -c 8 --mem 16G

# load modules needed
module load bedtools2/2.31.1

# set directories and files
root_dir="/scratch3/workspace/zdellaert_uri_edu-deep_dive/deep-dive-expression/E-Peve/"
bismark_dir="${root_dir}/output/12-Peve-WGBS/bismark_cutadapt/"
genome_folder="${root_dir}/data/"
gtf_name="Porites_evermanni_validated"

# Sort .cov files
cd ${bismark_dir}
for file in *merged_CpG_evidence.cov
do
  sample=$(basename "${file}" .CpG_report.merged_CpG_evidence.cov)
  bedtools sort -i "${file}" \
  > "${sample}"_sorted.cov
done

# Create bedgraphs for 5X coverage 
for file in *_sorted.cov
do
  sample=$(basename "${file}" _sorted.cov)
  cat "${file}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
  > "${sample}"_5x_sorted.bedgraph
done

# Create .tab files for 5x coverage
for file in *_sorted.cov
do
  sample=$(basename "${file}" _sorted.cov)
  cat "${file}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
  > "${sample}"_5x_sorted.tab
done

# Intersection of all samples
multiIntersectBed -i *_5x_sorted.tab > CpG.all.samps.5x_sorted.bed

## change number after == to your number of samples
cat CpG.all.samps.5x_sorted.bed | awk '$4 ==5' > CpG.filt.all.samps.5x_sorted.bed 

# Intersection of all samples with gene bodies:
awk '{if ($3 == "transcript") {print}}' "${genome_folder}/${gtf_name}.gtf"  > "${genome_folder}/${gtf_name}_transcripts.gtf"

for i in *5x_sorted.tab
do
  intersectBed \
  -wb \
  -a ${i} \
  -b "${genome_folder}/${gtf_name}_transcripts.gtf" \
  > ${i}_gene
done

# Keep only loci intersecting with genes found in all samples
for i in *_5x_sorted.tab_gene
do
  intersectBed \
  -a ${i} \
  -b CpG.filt.all.samps.5x_sorted.bed \
  > ${i}_CpG_5x_enrichment.bed
done

# Global methylation levels 
for file in *_sorted.cov; do
  sample=$(basename "$file" _sorted.cov)
  awk '{methylated+=$5; unmethylated+=$6} END {print "'$sample'", methylated/(methylated+unmethylated)}' "$file"
done > global_methylation_levels.txt
```

### Output file location: [All Bismark output files](https://gannet.fish.washington.edu/gitrepos/urol-e5/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/)


## Methylkit

```{r}
# Load methylKit and other packages
library("methylKit")
library("tidyverse")
library("parallel")

file_list <- list.files("/scratch3/workspace/zdellaert_uri_edu-deep_dive_exp/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/",pattern = ".merged_CpG_evidence.cov$",  full.names = TRUE, include.dirs = FALSE)
sample <- gsub("_S\\d{1,2}.CpG_report.merged_CpG_evidence.cov", "", basename(file_list))

file_list <- list.files("/scratch3/workspace/zdellaert_uri_edu-deep_dive_exp/deep-dive-expression/E-Peve/output/12-Peve-WGBS/bismark_cutadapt/",pattern = "_pe.deduplicated.bismark.cov.gz",  full.names = TRUE, include.dirs = FALSE)
sample <- gsub("trimmed_", "", basename(file_list))
sample <- gsub("_S\\d{1,2}_pe.deduplicated.bismark.cov.gz", "", sample)

file_list <- as.list(file_list)
treatment <- c(0,0,0,0,0)

sample
file_list

# methyl_list <- mclapply(seq_along(file_list), function(i) {
#   methRead(
#     file_list[[i]],
#     assembly = "Peve",
#     treatment = treatment[i],
#     sample.id = sample[[i]],
#     context = "CpG",
#     mincov = 10,
#     pipeline = "bismarkCoverage"
#   )
# }, mc.cores = 4) 
# 
# methylObj <- new("methylRawList", methyl_list)
# methylObj@treatment <- c(0, 0, 0, 0,0)
# 
# save(methylObj, file = "../output/12-Peve-WGBS/methylkit/MethylObj.RData")
```

```{r}
load("../output/12-Peve-WGBS/methylkit/MethylObj.RData")

getMethylationStats(methylObj[[1]],plot=FALSE,both.strands=FALSE)
getMethylationStats(methylObj[[1]], plot = TRUE,both.strands=FALSE)
getCoverageStats(methylObj[[1]],plot=TRUE,both.strands=FALSE)
```

```{r,eval=FALSE}
filtered_methylObj=filterByCoverage(methylObj,lo.count=5,lo.perc=NULL,
                                      hi.count=NULL,hi.perc=99.9)

filtered_methylObj_norm <-  methylKit::normalizeCoverage(filtered_methylObj)
meth_filter <- methylKit::unite(filtered_methylObj_norm,destrand=FALSE, min.per.group = 4L)

save(meth_filter, file = "../output/12-Peve-WGBS/methylkit/MethylObj_filtered.RData")
```

```{r}
load("../output/12-Peve-WGBS/methylkit/MethylObj_filtered.RData")

PCASamples(meth_filter)

clusterSamples(meth_filter, dist = "correlation", method = "ward", plot = TRUE)

getCorrelation(meth_filter)
```

```{r}
library("genomationData")
library("genomation")
library("GenomicRanges")

gff_df <- read.delim("../data/Porites_evermanni_v1.annot.gff", header = FALSE, sep = "\t", comment.char = "#",
                     col.names = c("seqid", "source", "type", "start", "end", "score", "strand", "phase", "attributes"))


gff_gr <- GRanges(
  seqnames = gff_df$seqid,
  ranges = IRanges(start = gff_df$start, end = gff_df$end),
  strand = gff_df$strand,
  type = gff_df$type,
  source = gff_df$source,
  score = gff_df$score,
  phase = gff_df$phase,
  attributes = gff_df$attributes
)

unique(gff_df$type)

# Get only exon entries
transcripts <- gff_gr[gff_gr$type == "mRNA"]
```


