---
title: "17-Peve-lncRNA"
output: html_document
date: "2025-06-21"
author: Steven Roberts
---

how good is <https://sr320.github.io/tumbling-oysters/posts/39-lncRNA/>

```{r}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Avoid automatic execution
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set default plot width
  fig.height = 4,      # Set default plot height
  fig.align = "center", # Center align plots
  comment = ""         # Prevents '##' in output
)
```

Some

```{r}
# Global R options
knitr::opts_chunk$set(echo = TRUE)

# Define key paths and tool directories
 
DATA_DIR <- "../data/17-Peve-lncRNA"
OUTPUT_DIR <- "../output/17-Peve-lncRNA"
THREADS <- "48"
  
FASTQ_SOURCE <- "https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/P_evermanni/trimmed/"
FASTQ_SUFFIX <- "fastq.gz"
GENOME_SOURCE <- "https://gannet.fish.washington.edu/seashell/snaps/Porites_evermanni_v1.fa"


GTF_SOURCE <- "https://raw.githubusercontent.com/urol-e5/deep-dive-expression/b0290e08af4eaeed30d74a758965debef6111801/E-Peve/data/Porites_evermanni_validated.gtf"
GFF_SOURCE <- "https://raw.githubusercontent.com/urol-e5/deep-dive-expression/b0290e08af4eaeed30d74a758965debef6111801/E-Peve/data/Porites_evermanni_validated.gff3"

GFFPATTERN <- 'class_code "u"|class_code "x"|class_code "o"|class_code "i"'

#RAVEN
# HISAT2_DIR <- "/home/shared/hisat2-2.2.1/"
# SAMTOOLS_DIR <- "/home/shared/samtools-1.12/"
# STRINGTIE_DIR <- "/home/shared/stringtie-2.2.1.Linux_x86_64"
# GFFCOMPARE_DIR <- "/home/shared/gffcompare-0.12.6.Linux_x86_64"
# BEDTOOLS_DIR <- "/home/shared/bedtools2/bin"
# CPC2_DIR <- "/home/shared/CPC2_standalone-1.0.1"
# CONDA_PATH <- "/opt/anaconda/anaconda3/bin/conda"

#KLONE
HISAT2_DIR <- ""
SAMTOOLS_DIR <- ""
STRINGTIE_DIR <- ""
GFFCOMPARE_DIR <- "/srlab/programs/gffcompare-0.12.6.Linux_x86_64/"
BEDTOOLS_DIR <- ""
CPC2_DIR <- "/srlab/programs/CPC2_standalone-1.0.1/bin/"
CONDA_PATH <- "/mmfs1/gscratch/srlab/nextflow/bin/miniforge/bin/conda"

GENOME_FASTA <- file.path(DATA_DIR, "genome.fasta")
GENOME_GTF <- file.path(DATA_DIR, "genome.gtf")
GENOME_GFF <- file.path(DATA_DIR, "genome.gff")
FASTQ_DIR <- file.path(DATA_DIR, "fastq")
GENOME_INDEX <- file.path(OUTPUT_DIR, "genome.index")

# Export these as environment variables for bash chunks.
Sys.setenv(
  THREADS = THREADS,
  DATA_DIR = DATA_DIR,
  FASTQ_SOURCE = FASTQ_SOURCE,
  FASTQ_SUFFIX = FASTQ_SUFFIX,
  OUTPUT_DIR = OUTPUT_DIR,
  GENOME_SOURCE = GENOME_SOURCE,
  GTF_SOURCE = GTF_SOURCE,
  GFF_SOURCE = GFF_SOURCE,
  HISAT2_DIR = HISAT2_DIR,
  SAMTOOLS_DIR = SAMTOOLS_DIR,
  STRINGTIE_DIR = STRINGTIE_DIR,
  GFFCOMPARE_DIR = GFFCOMPARE_DIR,
  BEDTOOLS_DIR = BEDTOOLS_DIR,
  CPC2_DIR = CPC2_DIR,
  CONDA_PATH = CONDA_PATH,
  GENOME_FASTA = GENOME_FASTA,
  GENOME_GTF = GENOME_GTF,
  GENOME_GFF = GENOME_GFF,
  FASTQ_DIR = FASTQ_DIR,
  GENOME_INDEX = GENOME_INDEX,
  GFFPATTERN = GFFPATTERN
)
```


```{r,engine='bash'}
mkdir -p "${DATA_DIR}"
mkdir -p "${OUTPUT_DIR}"
```

```{r, engine='bash'}
wget -nv -r \
--no-directories --no-parent \
-P ${FASTQ_DIR} \
-A "*${FASTQ_SUFFIX}" ${FASTQ_SOURCE}
```

```{r ls, engine='bash'}
ls ${FASTQ_DIR}
```

```{r fasta, engine='bash'}

curl -o "${GENOME_FASTA}" "${GENOME_SOURCE}"
```

```{r gtf, engine='bash'}


curl -o "${GENOME_GTF}" "${GTF_SOURCE}"
```

```{r gff, engine='bash'}


curl -o "${GENOME_GFF}" "${GFF_SOURCE}"
```

```{r file check, engine='bash'}
output_fasta=$(head -1 "${GENOME_FASTA}")
output_gff=$(tail -11 "${GENOME_GFF}")
output_gtf=$(head -1 "${GENOME_GTF}")

if [[ "$output_fasta" == *html* || "$output_gff" == *html* || "$output_gtf" == *html* ]]; then
    echo "FAIL - FFS you downloaded a HTML not and genome feature file!"
else
    echo "$output_fasta"
    echo "$output_gff"
    echo "$output_gtf"
fi
```

# HISAT

```{r, engine='bash'}
"${HISAT2_DIR}hisat2_extract_exons.py" "${GENOME_GTF}" > "${OUTPUT_DIR}/exon.txt"

"${HISAT2_DIR}hisat2_extract_splice_sites.py" "${GENOME_GTF}" > "${OUTPUT_DIR}/splice_sites.txt"

```

```{r, engine='bash'}

"${HISAT2_DIR}hisat2-build" \
  -p "${THREADS}" \
  "${GENOME_FASTA}" \
  "${GENOME_INDEX}" \
  --exon "${OUTPUT_DIR}/exon.txt" \
  --ss "${OUTPUT_DIR}/splice_sites.txt" \
  2> "${OUTPUT_DIR}/hisat2-build_stats.txt"

```

```{bash}
# Loop over every file ending in .fastq.gz that contains "_R2_"
for r2 in "${FASTQ_DIR}"/*_R2_*."${FASTQ_SUFFIX}"; do
    # Get the basename (filename without path)
    base=$(basename "$r2")
    
    # Derive a sample name by taking everything before "_R2_"
    sample="${base%%_R2_*}"
    
    # Construct the corresponding R1 filename by replacing "_R2_" with "_R1_"
    r1="${r2/_R2_/_R1_}"
    
    # Define the output SAM file name using the sample name
    output="${OUTPUT_DIR}/${sample}.sam"
    
    # Run hisat2 with the paired-end files
    "${HISAT2_DIR}hisat2" \
      -x "${GENOME_INDEX}" \
      -p "${THREADS}" \
      -1 "$r1" \
      -2 "$r2" \
      -S "$output"
done
```

## convert SAM to BAM

```{r convertSAM, engine='bash'}
for samfile in "${OUTPUT_DIR}/${sample}"*.sam; do
  bamfile="${samfile%.sam}.bam"
  sorted_bamfile="${samfile%.sam}.sorted.bam"
  
  # Convert SAM to BAM
  "${SAMTOOLS_DIR}samtools" view -bS -@ "${THREADS}" "$samfile" > "$bamfile"
  
  # Sort BAM
  "${SAMTOOLS_DIR}samtools" sort -@ "${THREADS}" "$bamfile" -o "$sorted_bamfile"
  
  # Index sorted BAM
  "${SAMTOOLS_DIR}samtools" index -@ "${THREADS}" "$sorted_bamfile"
done
```

# StringTie

StringTie uses the sorted BAM files to assemble transcripts for each sample, outputting them as GTF (Gene Transfer Format) files. And then merges all individual GTF assemblies into a single merged GTF file. This step extracts transcript information and merges GTFs from all samples–an important step in creating a canonical list of lncRNAs across all samples included in the pipeline.

```{r stringtie, engine='bash'}
find "${OUTPUT_DIR}" -name "*sorted.bam" \
| xargs -n 1 basename -s .sorted.bam | xargs -I{} \
"${STRINGTIE_DIR}stringtie" \
-p "${THREADS}" \
-G "${GENOME_GFF}" \
-o "${OUTPUT_DIR}/{}.gtf" \
"${OUTPUT_DIR}/{}.sorted.bam"
```

```{r, engine='bash'}
head ${OUTPUT_DIR}/*.gtf
```

```{r, engine='bash'}
wc -l ${OUTPUT_DIR}/*.gtf
```



```{r, engine='bash'}
echo "${STRINGTIE_DIR}stringtie"
echo "${GENOME_GFF}"

```

```{bash}
"${STRINGTIE_DIR}stringtie" -h
```





```{r merge, engine='bash'}
"${STRINGTIE_DIR}stringtie" \
--merge \
-G ../data/17-Peve-lncRNA/genome.gff \
-o ../output/17-Peve-lncRNA/stringtie_merged.gtf \
../output/17-Peve-lncRNA/RNA-POR*.gtf
```

```{r, engine='bash'}
wc -l ${OUTPUT_DIR}/stringtie_merged.gtf
```

```{r, engine='bash'}
head ${OUTPUT_DIR}/stringtie_merged.gtf
```

#GFFCOMPARE

![](http://gannet.fish.washington.edu/seashell/snaps/Monosnap_687474703a2f2f67616e6e65742e666973682e77617368696e67746f6e2e6564752f7365617368656c6c2f736e6170732f323032332d31312d30335f30392d3_2024-12-20_04-02-37.png)




```{r gffcomp, engine='bash'}
"${GFFCOMPARE_DIR}gffcompare" \
-r "${GENOME_GFF}" \
-o "${OUTPUT_DIR}/gffcompare_merged" \
"${OUTPUT_DIR}/stringtie_merged.gtf"
```

```{r viewgff, engine='bash'}
head -4 "${OUTPUT_DIR}"/gffcompare_merged*
```

```{r viewgff, engine='bash'}
wc -l "${OUTPUT_DIR}"/gffcompare_merged*
```


```{r, engine='bash'}
echo "${GFFPATTERN}"
echo "${OUTPUT_DIR}"
```



```{r gff filter, engine='bash'}
awk '$3 == "transcript" && $1 !~ /^#/' "${OUTPUT_DIR}/gffcompare_merged.annotated.gtf" | \
grep -E "${GFFPATTERN}" | \
awk '($5 - $4 > 199) || ($4 - $5 > 199)' > "${OUTPUT_DIR}/lncRNA_candidates.gtf"
```


```{r, engine='bash'}
head ${OUTPUT_DIR}/lncRNA_candidates.gtf
```

```{r, engine='bash'}
wc -l ${OUTPUT_DIR}/lncRNA_candidates.gtf
```

```{r, engine='bash'}
sort ../output/17-Ptuh-lncRNA/lncRNA_candidates.gtf | uniq -d
```


# Bedtools

```{r fasta2, engine='bash'}
"${BEDTOOLS_DIR}"bedtools getfasta \
-fi "${GENOME_FASTA}" \
-bed "${OUTPUT_DIR}/lncRNA_candidates.gtf" \
-fo "${OUTPUT_DIR}/lncRNA_candidates.fasta" \
-name -split
```

```{r, engine='bash'}
head ${OUTPUT_DIR}/lncRNA_candidates.fasta
```



#CPC2





```{r, engine='bash'}
eval "$(/mmfs1/gscratch/srlab/nextflow/bin/miniforge/bin/conda shell.bash hook)"
conda activate /mmfs1/gscratch/srlab/nextflow/bin/miniforge/envs/nextflow  # replace with your actual env name
python /mmfs1/gscratch/scrubbed/sr320/github/deep-dive-expression/F-Ptuh/code/CPC2_standalone-1.0.1/bin/CPC2.py \
  -i "${OUTPUT_DIR}/lncRNA_candidates.fasta" \
  -o "${OUTPUT_DIR}/CPC2"
```



Filter

```{r filCPC, engine='bash'}
awk '$8 == "noncoding" {print $1}' "${OUTPUT_DIR}/CPC2.txt" > "${OUTPUT_DIR}/noncoding_transcripts_ids.txt"
```

```{r, engine='bash'}
head "${OUTPUT_DIR}/CPC2.txt"
wc -l "${OUTPUT_DIR}/CPC2.txt"
```

```{r, engine='bash'}
head "${OUTPUT_DIR}/noncoding_transcripts_ids.txt"
wc -l "${OUTPUT_DIR}/noncoding_transcripts_ids.txt"
```

Subsetting fasta

```{r ssfasta, engine='bash'}
"${SAMTOOLS_DIR}samtools" faidx "${OUTPUT_DIR}/lncRNA_candidates.fasta" \
-r "${OUTPUT_DIR}/noncoding_transcripts_ids.txt" \
> "${OUTPUT_DIR}/lncRNA.fasta"
```

```{r, engine='bash'}
head -2 "${OUTPUT_DIR}/lncRNA.fasta"
grep -c ">" "${OUTPUT_DIR}/lncRNA.fasta"
```






```{r lncRNAgtf, engine='bash'}
# Define input and output file paths using the OUTPUT_DIR variable
input="${OUTPUT_DIR}/noncoding_transcripts_ids.txt"
output="${OUTPUT_DIR}/lncRNA.bed"

# Process each line of the input file
while IFS= read -r line; do
    # Remove "transcript::" from the line
    line="${line//transcript::/}"
    
    # Split the line by ':' to get the chromosome and position string
    IFS=':' read -r chromosome pos <<< "$line"
    
    # Split the position string by '-' to separate start and end positions
    IFS='-' read -r start end <<< "$pos"
    
    # Convert the start position to 0-based by subtracting 1
    start=$((start - 1))
    
    # Write the chromosome, updated start, and end positions to the output file (tab-separated)
    printf "%s\t%s\t%s\n" "$chromosome" "$start" "$end"
done < "$input" > "$output"
```

```{r, engine='bash'}
head -1 "${OUTPUT_DIR}/lncRNA.bed"
```

```{r renamegtf, engine='bash'}
awk 'BEGIN{OFS="\t"; count=1} {printf "%s\t.\tlncRNA\t%d\t%d\t.\t+\t.\tgene_id \"lncRNA_%03d\";\n", $1, $2, $3, count++;}' "${OUTPUT_DIR}/lncRNA.bed" \
> "${OUTPUT_DIR}/lncRNA.gtf"
```

```{r, engine='bas}
head "${OUTPUT_DIR}/lncRNA.gtf"
wc -l "${OUTPUT_DIR}/lncRNA.gtf"
```



# Summary Table

```{r summary, engine='bash'}
tf_file="${OUTPUT_DIR}/lncRNA.gtf"

awk '
BEGIN {
    total_entries = 0;
    min_length = 1e9;
    max_length = 0;
    sum_length = 0;
}
# Skip comment lines
/^#/ { next }
{
    if (NF < 9) next;
    total_entries++;
    start = $4;
    end = $5;
    gene_length = end - start + 1;
    if (gene_length < min_length) min_length = gene_length;
    if (gene_length > max_length) max_length = gene_length;
    sum_length += gene_length;
    feature[$3]++;
    chrom[$1]++;
    # Use two-argument match() and then extract the gene_id manually.
    if (match($9, /gene_id "[^"]+"/)) {
        gene_str = substr($9, RSTART, RLENGTH);
        # Remove the "gene_id " prefix and the quotes.
        gsub(/gene_id "/, "", gene_str);
        gsub(/"/, "", gene_str);
        genes[gene_str] = 1;
    }
}
END {
    avg_length = (total_entries > 0) ? sum_length / total_entries : 0;
    unique_gene_count = 0;
    for (g in genes)
        unique_gene_count++;
    print "Basic GTF File Statistics:";
    print "--------------------------";
    print "Total entries:      " total_entries;
    print "Unique genes:       " unique_gene_count;
    print "Min gene length:    " min_length;
    print "Max gene length:    " max_length;
    printf("Average gene length: %.2f\n", avg_length);
    print "\nFeature counts:";
    for (f in feature) {
        print "  " f ": " feature[f];
    }
    print "\nChromosome counts:";
    for (c in chrom) {
        print "  " c ": " chrom[c];
    }
}
' "$tf_file"
```
