---
title: "31-Peve-miRNA-mRNA-lncRNA-network"
author: "Kathleen Durkin"
date: "2025-07-16"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: ../../references.bib
link-citations: true
---

Reran 10/14/2025 to incorporate changes due to updated lncRNA counts matrix

```{r load packages}
library(dplyr)
library(ggplot2)
library(tidyr)
library(igraph)
library(tidygraph)
library(ggraph)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE         # Evaluate code chunks
)
```

I want to generate an interaction network plot showing the sum of miRNA-mRNA-lncRNA interactions.

This will include all putative miRNA-mRNA interactions (miRanda + significant PCC) and all putative miRNA-lncRNA interactions (miRanda + significant PCC). However, it will *NOT* include mRNA-lncRNA links. This is for two reasons: 

1. First, that dataset is simply too large -- there were tens of millions of mRNA-lncRNA pairs with significantly correlated expression. 

2. Second, the only method we have to investigate potential mRNA-lncRNA links is expression correlation, which doesn't provide any indication of the direction of action and is generally inappropriate to use as a sole predictor of causative relationships. In contrast, the miRNA-mRNA and miRNA-lncRNA putative interactions are primarily supported by binding predictions (through miRanda), and only use expression correlation as a method of validating interactions and evaluating their "polarity" (positive or negative relationship). When making a summary plot, I think it's only appropriate to include putative interactions which we can predict with similar degrees of confidence. Including mRNA-lncRNA interactions in the final plot may imply we are as confident in those interactions truly existing as we are in the miRNA-mRNA and miRNA-lncRNA interactions.

Because I'm excluding the mRNA-lncRNA links, I think this will functionally end up being a putative ceRNA (competative endogenous) network.


For both `igraph` and `Cytoscape` I need two inputs, an "Edges" dataframe and a "Nodes" dataframe (both should be saved as `.csv` files for export to `Cytoscape`.

The "Edges" file should associate each node with all other nodes it connects to. It should also contain edge-specific metadata. For example:

| source   | target | correlation | correlation magnitude | correlation direction | correlation pval | binding pval |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| miR-100  | Peve001 | -0.9        | 0.9                   | -1                    | 0.001            | 0.02         |
| miR-100  | Peve002 | 0.85        | 0.85                  | 1                     | 0.02             | 0.03         |
| lncRNA01 | Peve001 | -0.95       | 0.95                  | -1                    | 0.01             | 0.01         |

Note that there may be duplicates in both the "source" and "target" columns, and there may be duplicate source-target combinations if they have different attributes (e.g. representing different predicted binding sites of the same pair).However, the rows should be unique.

The "Nodes" file contains metadata for every node included in the plot. Importantly, the set of nodes listed in the "Nodes" file should match exactly the set of nodes included in the "Edges" document. For example:

| id       | type   |
|----------|--------|
| Peve001   | gene   |
| Peve002   | gene   |
| miR-100  | miRNA  |
| lncRNA01 | lncRNA |

I'll need the following files to compile the Cytoscape inputs:

-   miRNA-mRNA interaction files (contains binding and coexpression information for miRNA-gene pairs):

- miRNA-3UTR: `deep-dive-expression/E-Peve/output/10-Peve-mRNA-miRNA-interactions/miranda_PCC_miRNA_mRNA.csv`

- miRNA-CDS: `~/deep-dive-expression/E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv`

- miRNA-5UTR: `~/deep-dive-expression/E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv`

-   miRNA-lncRNA interaction file (contains binding and coexpression information for miRNA-lncRNA pairs): 

- `~/deep-dive-expression/E-Peve/output/15-Peve-miRNA-lncRNA-PCC/miranda_PCC_miRNA-lncRNA.csv`

Load packages:

```{r}
library(dplyr)
library(tidyr)
library(igraph)
```

Load and format files:

```{r}

# Load the three miRNA-mRNA files, and format so they have the same column contents and names

miRNA_3UTR <- read.csv("../output/10-Peve-mRNA-miRNA-interactions/Peve-miranda_PCC_miRNA_mRNA.csv") %>% dplyr::select(-X.1, -X)
# Add label for binding region
miRNA_3UTR$region <- "3UTR"


miRNA_CDS <- read.csv("../output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv") %>% dplyr::select(-X)
colnames(miRNA_CDS) <- c("miRNA", "mRNA_coord", "score", "energy", "query_start", "query_end", "subject_start", "subject_end", "total_bp_shared", "query_similar", "subject_similar", "mRNA", "PCC.cor", "p_value", "adjusted_p_value")
miRNA_CDS$query_start_end <- paste0(miRNA_CDS$query_start, " ", miRNA_CDS$query_end)
miRNA_CDS$subject_start_end <- paste0(miRNA_CDS$subject_start, " ", miRNA_CDS$subject_end)
# Add label for binding region
miRNA_CDS$region <- "CDS"
# Match column order of miRNA_3UTR
miRNA_CDS <- miRNA_CDS %>% dplyr::select(colnames(miRNA_3UTR))


miRNA_5UTR <- read.csv("../output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv") %>% dplyr::select(-X)
colnames(miRNA_5UTR) <- c("miRNA", "mRNA_coord", "score", "energy", "query_start", "query_end", "subject_start", "subject_end", "total_bp_shared", "query_similar", "subject_similar", "mRNA", "PCC.cor", "p_value", "adjusted_p_value")
miRNA_5UTR$query_start_end <- paste0(miRNA_5UTR$query_start, " ", miRNA_5UTR$query_end)
miRNA_5UTR$subject_start_end <- paste0(miRNA_5UTR$subject_start, " ", miRNA_5UTR$subject_end)
# Add label for binding region
miRNA_5UTR$region <- "5UTR"
# Match column order of miRNA_3UTR
miRNA_5UTR <- miRNA_5UTR %>% dplyr::select(colnames(miRNA_3UTR))


miRNA_gene <- rbind(miRNA_3UTR, miRNA_CDS, miRNA_5UTR)
# Remove NA rows
miRNA_gene <- miRNA_gene %>% filter(!is.na(miRNA))

# Load and format the miRNA-lncRNA file
miRNA_lncRNA <- read.csv("../output/15-Peve-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv") %>% dplyr::select(-X.1, -X)
# Add label for binding region
miRNA_lncRNA$region <- "lncRNA"
```


For miRNA, annotate with assigned names (e.g., Peve-miR-100, Peve-novel-7)
```{r}
Peve_names <- read.csv("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)

miRNA_gene <- left_join(miRNA_gene, Peve_names, by = c("miRNA" = "Name"))
miRNA_lncRNA <- left_join(miRNA_lncRNA, Peve_names, by = c("miRNA" = "Name"))
```



# All interactions

Format "Edges" file:

```{r}
# Add correlation magnitude and direction columns
miRNA_gene$PCC_magnitude <- abs(miRNA_gene$PCC.cor)
miRNA_gene$PCC_direction <- sign(miRNA_gene$PCC.cor)
miRNA_gene$Alignment <- paste0(miRNA_gene$mRNA, ";", miRNA_gene$query_start_end, ";", miRNA_gene$subject_start_end)
# Select columns I want to keep in Edges file
miRNA_gene_edges <- miRNA_gene %>% dplyr::select(given_miRNA_name, mRNA, region, Alignment, energy, total_bp_shared, query_similar, subject_similar, PCC.cor, PCC_magnitude, PCC_direction, p_value)
# rename columns
miRNA_gene_edges <- miRNA_gene_edges %>% rename(source = given_miRNA_name, target = mRNA)

# Add correlation magnitude and direction columns
miRNA_lncRNA$PCC_magnitude <- abs(miRNA_lncRNA$PCC.cor)
miRNA_lncRNA$PCC_direction <- sign(miRNA_lncRNA$PCC.cor)
miRNA_lncRNA$Alignment <- paste0(miRNA_lncRNA$lncRNA, ";", miRNA_lncRNA$query_start_end, ";", miRNA_lncRNA$subject_start_end)
# Select columns I want to keep in Edges file (ensure in same order as in the miRNA_gene_edges file)
miRNA_lncRNA_edges <- miRNA_lncRNA %>% dplyr::select(given_miRNA_name, lncRNA, region, Alignment, energy, total_bp_shared, query_similar, subject_similar, PCC.cor, PCC_magnitude, PCC_direction, p_value)
# rename columns
miRNA_lncRNA_edges <- miRNA_lncRNA_edges %>% rename(source = given_miRNA_name, target = lncRNA)

# Combine miRNA-gene edges and miRNA-lncRNA edges
edges <- rbind(miRNA_gene_edges, miRNA_lncRNA_edges)

# Ensure we have no duplicate rows
nrow(edges)
nrow(edges %>% distinct())

# Check formatting/contents
head(edges)
```

Format Nodes file:

```{r}
# Make a df that contains all miRNA, genes, and lncRNA listed in the `source` and `target` columns of `edges`
nodes <- data.frame(
  # The `unique` argument ensures we remove duplicates
  id = unique(unname(unlist(edges[, c("source", "target")])))
)

# Add column identifying the type of each node (miRNA, lncRNA, or gene)
nodes <- nodes %>%
  mutate(type = case_when(
    grepl("mir", id) ~ "miRNA",
    grepl("Peve", id) ~ "gene",
    grepl("lncRNA", id) ~ "lncRNA",
    TRUE ~ "other"
  ))

# Check formatting/contents
head(nodes)
```

# pval \< 0.05

Edges:

```{r}
edges_pval_0.05 <- edges %>% filter(p_value < 0.05)
nrow(edges_pval_0.05)
```

Nodes:

```{r}
# Make a df that contains all miRNA, genes, and lncRNA listed in the `source` and `target` columns of `edges`
nodes_pval_0.05 <- data.frame(
  # The `unique` argument ensures we remove duplicates
  id = unique(unname(unlist(edges_pval_0.05[, c("source", "target")])))
)

# Add column identifying the type of each node (miRNA, lncRNA, or gene)
nodes_pval_0.05 <- nodes_pval_0.05 %>%
  mutate(type = case_when(
    grepl("mir", id) ~ "miRNA",
    grepl("Peve", id) ~ "gene",
    grepl("lncRNA", id) ~ "lncRNA",
    TRUE ~ "other"
  ))

nrow(nodes_pval_0.05)
nrow(filter(nodes_pval_0.05, type == "miRNA"))
nrow(filter(nodes_pval_0.05, type == "gene"))
nrow(filter(nodes_pval_0.05, type == "lncRNA"))
```
When using a PCC significance level of 0.05, there are 1442 edges and 1293 nodes. In other words, 1293 features (42 miRNA, 1091 mRNA, and 160 lncRNA) form 1442 pairwise interactions based on both putative binding and expression correlation.

Some more summary stats:
Keep in mind that there may be multiple predicted interactions between a single feature pair (e.g. a single miRNA-mRNA pair), since there may be multiple viable binding sites. One must consider both interaction-level and unique feature level (e.g. An miRNA may target 15 unique mRNAs via 20 putative interactions)

```{r}

# Helper function to summarize counts
summarize_counts <- function(df, group_col) {
  df %>%
    count({{ group_col }}) %>%
    summarise(
      mean = signif(mean(n), 3),
      median = signif(median(n), 3),
      min = min(n),
      max = max(n)
    )
}

network_summary_stats <- function(edges_df) {

  # Separate mRNA and lncRNA edges
  mRNA_edges <- edges_df %>% filter(region %in% c("3UTR", "5UTR", "CDS"))
  lncRNA_edges <- edges_df %>% filter(region == "lncRNA")

  # ---- For miRNAs ----
  # mRNA interactions (all and unique)
  mRNA_int_all <- summarize_counts(mRNA_edges, source)
  mRNA_int_unique <- summarize_counts(distinct(mRNA_edges, source, target), source)

  # lncRNA interactions (all and unique)
  lncRNA_int_all <- summarize_counts(lncRNA_edges, source)
  lncRNA_int_unique <- summarize_counts(distinct(lncRNA_edges, source, target), source)

  # ---- For targets ----
  # miRNA interactions on mRNA (all and unique)
  miRNA_int_mRNA_all <- summarize_counts(mRNA_edges, target)
  miRNA_int_mRNA_unique <- summarize_counts(distinct(mRNA_edges, source, target), target)

  # miRNA interactions on lncRNA (all and unique)
  miRNA_int_lncRNA_all <- summarize_counts(lncRNA_edges, target)
  miRNA_int_lncRNA_unique <- summarize_counts(distinct(lncRNA_edges, source, target), target)

  # ---- Proportion of mRNA targets per miRNA ----
 
  # Unique target level
 miRNA_target_mRNA_prop_unique <- edges_df %>%
  mutate(type = ifelse(region %in% c("CDS", "3UTR", "5UTR"), "mRNA", "lncRNA")) %>%
  distinct(source, target, type) %>%                     # Keep unique miRNA–mRNA or miRNA-lncRNA pairs
  group_by(source, type) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(source) %>%
  mutate(prop = count / sum(count)) %>%
  filter(type == "mRNA") %>%                               # Keep only mRNA proportions
  ungroup() %>%
  summarise(
    mean = signif(mean(prop), 3),
    median = signif(median(prop), 3),
    min = signif(min(prop), 3),
    max = signif(max(prop), 3)
  )

  # ---- Combine into summary table ----
summary_stats <- bind_rows(
  mRNA_int_all %>% mutate(Metric = "mRNA interactions per miRNA (all)"),
  mRNA_int_unique %>% mutate(Metric = "mRNA targets per miRNA (unique)"),
  lncRNA_int_all %>% mutate(Metric = "lncRNA interactions per miRNA (all)"),
  lncRNA_int_unique %>% mutate(Metric = "lncRNA targets per miRNA (unique)"),
  miRNA_int_mRNA_all %>% mutate(Metric = "miRNA interactions per mRNA (all)"),
  miRNA_int_mRNA_unique %>% mutate(Metric = "miRNAs per mRNA (unique)"),
  miRNA_int_lncRNA_all %>% mutate(Metric = "miRNA interactions per lncRNA (all)"),
  miRNA_int_lncRNA_unique %>% mutate(Metric = "miRNAs per lncRNA (unique)"),
  miRNA_target_mRNA_prop_unique %>% mutate(Metric = "Prop. mRNA targets per miRNA (unique)")
) %>%
  select(Metric, everything())

  return(summary_stats)
}

network_summary_stats(edges_pval_0.05)

```

**At the interaction-level (which allows for multiple interactions between a single feature pair):**
Each miRNA has 30.9 interactions with mRNA and 6.0 interactions with lncRNA, on average, though there is quite a bit of variability. Each mRNA has an average of 1.2 interactions with miRNA, and each lncRNA has an average of 1.1 interactions with miRNA.


**At the unique-feature level:**
On average, each miRNA has 27.7 unique mRNA targets and 5.7 unique lncRNA targets, though again there is a lot of variation. Both mRNA and lncRNA are targeted by 1.0 unique miRNA, on average.

Additionally, there is variation in the relative proportion of miRNA targets that are mRNA. Relative mRNA proportion ranges from 33.3% (lncRNA-dominant) to 100% (mRNA-dominant), with an average of 85.0% (mRNA-dominant)

# pval \< 0.01

Edges:

```{r}
edges_pval_0.01 <- edges %>% filter(p_value < 0.01)
nrow(edges_pval_0.01)
```

Nodes:

```{r}
# Make a df that contains all miRNA, genes, and lncRNA listed in the `source` and `target` columns of `edges`
nodes_pval_0.01 <- data.frame(
  # The `unique` argument ensures we remove duplicates
  id = unique(unname(unlist(edges_pval_0.01[, c("source", "target")])))
)

# Add column identifying the type of each node (miRNA, lncRNA, or gene)
nodes_pval_0.01 <- nodes_pval_0.01 %>%
  mutate(type = case_when(
    grepl("mir", id) ~ "miRNA",
    grepl("Peve", id) ~ "gene",
    grepl("lncRNA", id) ~ "lncRNA",
    TRUE ~ "other"
  ))

nrow(nodes_pval_0.01)
nrow(filter(nodes_pval_0.01, type == "miRNA"))
nrow(filter(nodes_pval_0.01, type == "gene"))
nrow(filter(nodes_pval_0.01, type == "lncRNA"))
```

When using a PCC significance level of 0.01, there are 455 edges and 444 nodes. In other words, 444 features (31 miRNA, 381 mRNA, and 32 lncRNA) form 455 pairwise interactions based on both putative binding and expression correlation.

Some more summary stats:
Keep in mind that there may be multiple predicted interactions between a single feature pair (e.g. a single miRNA-mRNA pair), since there may be multiple viable binding sites. One must consider both interaction-level and unique feature level (e.g. An miRNA may target 15 unique mRNAs via 20 putative interactions)

```{r}
network_summary_stats(edges_pval_0.01)
```

**At the interaction-level (which allows for multiple interactions between a single feature pair):**
Each miRNA has 14.0 interactions with mRNA and 2.3 interactions with lncRNA, on average, though there is quite a bit of variability. Both mRNA and lncRNA have an average of 1.1 interactions with miRNA.


**At the unique-feature level:**
On average, each miRNA has 12.8 unique mRNA targets and 2.2 unique lncRNA targets, though again there is a lot of variation. Both mRNA and lncRNA are targeted by 1.0 unique miRNA, on average.

Additionally, there is some variation in the relative proportion of miRNA targets that are mRNA, and this variation is notably lower than our Peve network at the 0.05 significance level. All modules are mRNA-dominant, with relative mRNA proportion ranges from 62.5% to 100%, with an average of 92.8%.


# Save

Save files

**At the interaction-level (which allows for multiple interactions between a single feature pair):**
```{r, eval=TRUE}
write.csv(edges_pval_0.05, "../output/31-Peve-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.05.csv", quote = FALSE)
write.csv(nodes_pval_0.05, "../output/31-Peve-miRNA-mRNA-lncRNA-network/nodes_miRNA_mRNA_lncRNA_network_p0.05.csv", quote = FALSE)
write.csv(edges_pval_0.01, "../output/31-Peve-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.01.csv", quote = FALSE)
write.csv(nodes_pval_0.01, "../output/31-Peve-miRNA-mRNA-lncRNA-network/nodes_miRNA_mRNA_lncRNA_network_p0.01.csv", quote = FALSE)

```

**At the unique-feature level:**
```{r, eval=TRUE}
edges_pval_0.05_unique <- edges_pval_0.05 %>% dplyr::select(-region, -Alignment, -energy, -total_bp_shared, -query_similar, -subject_similar) %>% distinct()
edges_pval_0.01_unique <- edges_pval_0.01 %>% dplyr::select(-region, -Alignment, -energy, -total_bp_shared, -query_similar, -subject_similar) %>% distinct()

write.csv(edges_pval_0.05_unique, "../output/31-Peve-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.05_unique.csv", quote = FALSE)
write.csv(edges_pval_0.01_unique, "../output/31-Peve-miRNA-mRNA-lncRNA-network/edges_miRNA_mRNA_lncRNA_network_p0.01_unique.csv", quote = FALSE)

```

# Using Cytoscape

To load a network into Cytoscape:

1.  Open Cytoscape and select File \> Import \> Network from File...

2.  Select "Edges" file. Ensure The source and target columns are appropriately identified before loading the file.

3.  To load "Nodes" file, select File \> Import \> Table from File...

# Plotting with igraph

## p < 0.05

```{r}
# Rename columns for igraph
colnames(edges_pval_0.05)[1:2] <- c("from", "to")         # For igraph edge input
colnames(nodes_pval_0.05)[1] <- "name"                    # For igraph vertex input; must match nodes in edges

# Build graph
g <- graph_from_data_frame(d = edges_pval_0.05, vertices = nodes_pval_0.05, directed = FALSE)

# Edge attributes
E(g)$edge_color <- ifelse(E(g)$PCC_direction > 0, "positive", "negative")

# Convert to tbl_graph
g_tbl <- as_tbl_graph(g)


p <- ggraph(g_tbl, layout = "fr") +
  geom_edge_link(aes(edge_width = PCC_magnitude, color = edge_color), alpha = 0.6) +
  geom_node_point(aes(color = type), size = 2) +
  scale_edge_width(range = c(0.5, 3)) +
  
  # Split color scales
  scale_edge_color_manual(
    values = c("positive" = "green3", "negative" = "red"),
    name = "Correlation Direction"
  ) +
  scale_color_manual(
    values = c("miRNA" = "orange", "gene" = "lightblue", "lncRNA" = "steelblue4"),
    name = "type"
  ) +

  theme_graph() +
  labs(title = "miRNA-lncRNA-mRNA Interaction Network")

print(p)


```

## p < 0.01

```{r}
# Rename columns for igraph
colnames(edges_pval_0.01)[1:2] <- c("from", "to")         # For igraph edge input
colnames(nodes_pval_0.01)[1] <- "name"                    # For igraph vertex input; must match nodes in edges

# Build graph
g <- graph_from_data_frame(d = edges_pval_0.01, vertices = nodes_pval_0.01, directed = FALSE)

# Edge attributes
E(g)$edge_color <- ifelse(E(g)$PCC_direction > 0, "positive", "negative")

# Convert to tbl_graph
g_tbl <- as_tbl_graph(g)


p <- ggraph(g_tbl, layout = "fr") +
  geom_edge_link(aes(edge_width = PCC_magnitude, color = edge_color), alpha = 0.6) +
  geom_node_point(aes(color = type), size = 2) +
  scale_edge_width(range = c(0.5, 3)) +
  
  # Split color scales
  scale_edge_color_manual(
    values = c("positive" = "green3", "negative" = "red"),
    name = "Correlation Direction"
  ) +
  scale_color_manual(
    values = c("miRNA" = "orange", "gene" = "lightblue", "lncRNA" = "steelblue4"),
    name = "type"
  ) +

  theme_graph() +
  labs(title = "miRNA-lncRNA-mRNA Interaction Network")

print(p)


```


