---
title: "04.1-miRNA-comparison-expression"
author: "Kathleen Durkin"
date: "2025-06-24"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

We hypothesize that miRNA which are conserved among species are more highly expressed than species-specific miRNA. This document will compare miRNA expression accross species and levels of conservation

Load packages
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(edgeR)
```


Load and format counts files

Note: will be normalizing using CPM (counts per million, normalized for library size) to describe expression level in miRNA. We don't use TPM (transcripts per milion), which is normalized for both transcript length and library size, for miRNA because they are, by definition, all the same size (~22nt)

```{r}
# mature miRNA count matrices
Apul_mature <- read.table("../../D-Apul/output/03.1-Apul-sRNA-summary/Apul_miRNA_ShortStack_counts_formatted.txt")
Apul_mature <- data.frame(cpm(Apul_mature)) # CPM normalization
Apul_mature$mean <- rowMeans(Apul_mature) # Average expression across samples
Apul_mature$miRNA <- rownames(Apul_mature)
Apul_mature$species <- "Apul"

Peve_mature <- read.table("../../E-Peve/output/03.1-Peve-sRNA-summary/Peve_miRNA_ShortStack_counts_formatted.txt")
Peve_mature <- data.frame(cpm(Peve_mature)) # CPM normalization
Peve_mature$mean <- rowMeans(Peve_mature) # Average expression across samples
Peve_mature$miRNA <- rownames(Peve_mature)
Peve_mature$species <- "Peve"

Ptuh_mature <- read.table("../../F-Ptuh/output/03.1-Ptuh-sRNA-summary/Ptuh_miRNA_ShortStack_counts_formatted.txt")
Ptuh_mature <- data.frame(cpm(Ptuh_mature)) # CPM normalization
Ptuh_mature$mean <- rowMeans(Ptuh_mature) # Average expression across samples
Ptuh_mature$miRNA <- rownames(Ptuh_mature)
Ptuh_mature$species <- "Ptuh"

```


Load in assigned miRNA names
```{r}
Apul_names <- read.csv("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Apul_Results_mature_named_miRNAs.csv") %>% select(Name, given_miRNA_name)

Peve_names <- read.csv("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv") %>% select(Name, given_miRNA_name)

Ptuh_names <- read.csv("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Ptuh_Results_mature_named_miRNAs.csv") %>% select(Name, given_miRNA_name)
```

Annotate miRNA dfs with given names
```{r}
Apul_mature_df <- left_join(Apul_mature, Apul_names, by = c("miRNA" = "Name"))
Peve_mature_df <- left_join(Peve_mature, Peve_names, by = c("miRNA" = "Name"))
Ptuh_mature_df <- left_join(Ptuh_mature, Ptuh_names, by = c("miRNA" = "Name"))
```

Separate conserved miRNA (present in all 3 species) from the rest. In `04-miRNA-comparison`, identified the miRNA conserved among all 3 species to be: miR-100, miR-2023, miR-2025, and miR-2036
```{r}
Apul_mature_conserved <- Apul_mature_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Apul_mature_unconserved <- Apul_mature_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

Peve_mature_conserved <- Peve_mature_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Peve_mature_unconserved <- Peve_mature_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

Ptuh_mature_conserved <- Ptuh_mature_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Ptuh_mature_unconserved <- Ptuh_mature_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

# Also annotate the full dfs
Apul_mature_df$conservation <- ifelse(Apul_mature_df$given_miRNA_name %in% Apul_mature_conserved$given_miRNA_name, "conserved", "unconserved")
Peve_mature_df$conservation <- ifelse(Peve_mature_df$given_miRNA_name %in% Peve_mature_conserved$given_miRNA_name, "conserved", "unconserved")
Ptuh_mature_df$conservation <- ifelse(Ptuh_mature_df$given_miRNA_name %in% Ptuh_mature_conserved$given_miRNA_name, "conserved", "unconserved")
```


Expression summary stats (already normalized using CPM):
```{r}
Apul_mature_counts <- as.matrix(Apul_mature[,1:5])
Peve_mature_counts <- as.matrix(Peve_mature[,1:3])
Ptuh_mature_counts <- as.matrix(Ptuh_mature[,1:5])

Apul_mature_conserved_counts <- as.matrix(Apul_mature_conserved[,1:5])
Peve_mature_conserved_counts <- as.matrix(Peve_mature_conserved[,1:3])
Ptuh_mature_conserved_counts <- as.matrix(Ptuh_mature_conserved[,1:5])

Apul_mature_unconserved_counts <- as.matrix(Apul_mature_unconserved[,1:5])
Peve_mature_unconserved_counts <- as.matrix(Peve_mature_unconserved[,1:3])
Ptuh_mature_unconserved_counts <- as.matrix(Ptuh_mature_unconserved[,1:5])
```

```{r}
cat("Mean expression of all miRNA in A.pulchra, P.evermanni, and P.tuahiniensis:  ")
cat(mean(Apul_mature_counts), ", ", mean(Peve_mature_counts), ", ", mean(Ptuh_mature_counts))
cat("\n")
cat("Median expression of all miRNA in A.pulchra, P.evermanni, and P.tuahiniensis:  ")
cat(median(Apul_mature_counts), ", ", median(Peve_mature_counts), ", ", median(Ptuh_mature_counts))

cat("\n")
cat("\n")

cat("Mean expression of *conserved* miRNA in A.pulchra, P.evermanni, and P.tuahiniensis:  ")
cat(mean(Apul_mature_conserved_counts), ", ", mean(Peve_mature_conserved_counts), ", ", mean(Ptuh_mature_conserved_counts))
cat("\n")
cat("Median expression of *conserved* miRNA in A.pulchra, P.evermanni, and P.tuahiniensis:  ")
cat(median(Apul_mature_conserved_counts), ", ", median(Peve_mature_conserved_counts), ", ", median(Ptuh_mature_conserved_counts))

cat("\n")
cat("\n")

cat("Mean expression of *unconserved* miRNA in A.pulchra, P.evermanni, and P.tuahiniensis:  ")
cat(mean(Apul_mature_unconserved_counts), ", ", mean(Peve_mature_unconserved_counts), ", ", mean(Ptuh_mature_unconserved_counts))
cat("\n")
cat("Median expression of *unconserved* miRNA in A.pulchra, P.evermanni, and P.tuahiniensis:  ")
cat(median(Apul_mature_unconserved_counts), ", ", median(Peve_mature_unconserved_counts), ", ", median(Ptuh_mature_unconserved_counts))

```

Plot
```{r}
### Apul ###
# long format
Apul_mature_long <- Apul_mature_df %>%
  pivot_longer(cols = starts_with("sample"), 
               names_to = "sample", 
               values_to = "expression")

# plot
ggplot(Apul_mature_long, aes(x=sample, y=expression, group=given_miRNA_name, color=given_miRNA_name)) +
  geom_point() + 
  geom_line() +
  #geom_smooth(method = "loess") +
  #geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~conservation)

### Peve ###
# long format
Peve_mature_long <- Peve_mature_df %>%
  pivot_longer(cols = starts_with("sample"), 
               names_to = "sample", 
               values_to = "expression")

# plot
ggplot(Peve_mature_long, aes(x=sample, y=expression, group=given_miRNA_name, color=given_miRNA_name)) +
  geom_point() + 
  geom_line() +
  #geom_smooth(method = "loess") +
  #geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~conservation)

### Ptuh ###
# long format
Ptuh_mature_long <- Ptuh_mature_df %>%
  pivot_longer(cols = starts_with("sample"), 
               names_to = "sample", 
               values_to = "expression")

# plot
ggplot(Ptuh_mature_long, aes(x=sample, y=expression, group=given_miRNA_name, color=given_miRNA_name)) +
  geom_point() + 
  geom_line() +
  #geom_smooth(method = "loess") +
  #geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~conservation)

```

```{r}
all_long <- rbind(Apul_mature_long, 
                  Peve_mature_long, 
                  Ptuh_mature_long)

ggplot(all_long, aes(x=sample, y=expression, group=given_miRNA_name, color=conservation)) +
  geom_point() + 
  geom_line(size=0.3) +
  scale_y_log10() +
  facet_wrap(~species, scales="free_x") +
  theme(axis.text.x = element_text(angle=45, hjust=1))

```


```{r}
all_conserved_long <- rbind(Apul_mature_long[Apul_mature_long$conservation == "conserved",], 
                            Peve_mature_long[Peve_mature_long$conservation == "conserved",], 
                            Ptuh_mature_long[Ptuh_mature_long$conservation == "conserved",])

# Remove the species-specific prefixes from miRNA names for plotting
all_conserved_long$given_miRNA_name <- substr(all_conserved_long$given_miRNA_name, 
                                              6, nchar(all_conserved_long$given_miRNA_name))

ggplot(all_conserved_long, aes(x=sample, y=expression, group=given_miRNA_name, color=given_miRNA_name)) +
  geom_point() + 
  geom_line() +
  scale_y_log10() +
  facet_wrap(~species, scales="free_x") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```


```{r}
### Apul ###
Apul_conserved_sampleMeans <- colMeans(Apul_mature_conserved[,1:5])
Apul_unconserved_sampleMeans <- colMeans(Apul_mature_unconserved[,1:5])

# Test whether expression of conserved miRNA significantly differs from unconserved
# For both groups, averaged miRNA expression in each sample
# Small sample size (n=5), so use non-parametric test
wilcox.test(x=Apul_conserved_sampleMeans, y=Apul_unconserved_sampleMeans, paired=TRUE)


### Peve ###
Peve_conserved_sampleMeans <- colMeans(Peve_mature_conserved[,1:3])
Peve_unconserved_sampleMeans <- colMeans(Peve_mature_unconserved[,1:3])

# Test whether expression of conserved miRNA significantly differs from unconserved
# For both groups, averaged miRNA expression in each sample
# Small sample size (n=5), so use non-parametric test
wilcox.test(x=Peve_conserved_sampleMeans, y=Peve_unconserved_sampleMeans, paired=TRUE)


### Ptuh ###
Ptuh_conserved_sampleMeans <- colMeans(Ptuh_mature_conserved[,1:5])
Ptuh_unconserved_sampleMeans <- colMeans(Ptuh_mature_unconserved[,1:5])

# Test whether expression of conserved miRNA significantly differs from unconserved
# For both groups, averaged miRNA expression in each sample
# Small sample size (n=5), so use non-parametric test
wilcox.test(x=Ptuh_conserved_sampleMeans, y=Ptuh_unconserved_sampleMeans, paired=TRUE)
```

