---
title: "04.1-miRNA-comparison-expression"
author: "Kathleen Durkin"
date: "2025-07-10"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

Load packages
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
```


Hypothesize that conserved miRNA present in all three species will also have targets enriched for the same processes across species.

## Functionally enriched processes

Load 3UTR FE tables:

```{r}
Apul_3UTR_FE <- read.csv("../../D-Apul/output/09.1-Apul-mRNA-miRNA-interactions-functional-enrichment/miRNA_sig_cor_targets_topGO_FE.csv")

Peve_3UTR_FE <- read.csv("../../E-Peve/output/10.13-Peve-mRNA-miRNA-interactions-FE-3UTR/miRNA_3UTR_sig_cor_targets_topGO_FE.csv")
  
Ptuh_3UTR_FE <- read.csv("../../F-Ptuh/output/11.13-Ptuh-mRNA-miRNA-interactions-FE-3UTR/miRNA_3UTR_sig_cor_targets_topGO_FE.csv")
```

Load in assigned miRNA names
```{r}
Apul_names <- read.csv("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Apul_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)

Peve_names <- read.csv("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)

Ptuh_names <- read.csv("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Ptuh_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)
```

Annotate miRNA dfs with given names
```{r}
Apul_3UTR_FE_df <- left_join(Apul_3UTR_FE, Apul_names, by = c("miRNA" = "Name"))
Peve_3UTR_FE_df <- left_join(Peve_3UTR_FE, Peve_names, by = c("miRNA" = "Name"))
Ptuh_3UTR_FE_df <- left_join(Ptuh_3UTR_FE, Ptuh_names, by = c("miRNA" = "Name"))
```

Separate conserved miRNA (present in all 3 species) from the rest. In `04-miRNA-comparison`, identified the miRNA conserved among all 3 species to be: miR-100, miR-2023, miR-2025, and miR-2036
```{r}
Apul_3UTR_FE_conserved <- Apul_3UTR_FE_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Apul_3UTR_FE_unconserved <- Apul_3UTR_FE_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

Peve_3UTR_FE_conserved <- Peve_3UTR_FE_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Peve_3UTR_FE_unconserved <- Peve_3UTR_FE_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

Ptuh_3UTR_FE_conserved <- Ptuh_3UTR_FE_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Ptuh_3UTR_FE_unconserved <- Ptuh_3UTR_FE_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

# Also annotate the full dfs
Apul_3UTR_FE_df$conservation <- ifelse(Apul_3UTR_FE_df$given_miRNA_name %in% Apul_3UTR_FE_conserved$given_miRNA_name, "conserved", "unconserved")
Peve_3UTR_FE_df$conservation <- ifelse(Peve_3UTR_FE_df$given_miRNA_name %in% Peve_3UTR_FE_conserved$given_miRNA_name, "conserved", "unconserved")
Ptuh_3UTR_FE_df$conservation <- ifelse(Ptuh_3UTR_FE_df$given_miRNA_name %in% Ptuh_3UTR_FE_conserved$given_miRNA_name, "conserved", "unconserved")
```

Take a look
```{r}
print(Apul_3UTR_FE_conserved)
print(Peve_3UTR_FE_conserved)
print(Ptuh_3UTR_FE_conserved)
```
Very few functionally enriched processes in conserved miRNA, and in general, likely due to a small smaple size and only a subset of all genes being annotated. 

Let's instead look at the annotations themseleves, regardless of whether the processes were statistically overrepresented

## Functionally annotated

Load 3UTR FA tables:

```{r}
Apul_3UTR_FA <- read.csv("../../D-Apul/output/09.1-Apul-mRNA-miRNA-interactions-functional-enrichment/miRNA_sig_cor_targets_FA.csv")

Peve_3UTR_FA <- read.csv("../../E-Peve/output/10.13-Peve-mRNA-miRNA-interactions-FE-3UTR/miRNA_3UTR_sig_cor_targets_FA.csv")
  
Ptuh_3UTR_FA <- read.csv("../../F-Ptuh/output/11.13-Ptuh-mRNA-miRNA-interactions-FE-3UTR/miRNA_3UTR_sig_cor_targets_FA.csv")
```

Annotate miRNA dfs with given names
```{r}
Apul_3UTR_FA_df <- left_join(Apul_3UTR_FA, Apul_names, by = c("miRNA" = "Name"))
Peve_3UTR_FA_df <- left_join(Peve_3UTR_FA, Peve_names, by = c("miRNA" = "Name"))
Ptuh_3UTR_FA_df <- left_join(Ptuh_3UTR_FA, Ptuh_names, by = c("miRNA" = "Name"))
```

Separate conserved miRNA (present in all 3 species) from the rest. In `04-miRNA-comparison`, identified the miRNA conserved among all 3 species to be: miR-100, miR-2023, miR-2025, and miR-2036
```{r}
Apul_3UTR_FA_conserved <- Apul_3UTR_FA_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Apul_3UTR_FA_unconserved <- Apul_3UTR_FA_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

Peve_3UTR_FA_conserved <- Peve_3UTR_FA_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Peve_3UTR_FA_unconserved <- Peve_3UTR_FA_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

Ptuh_3UTR_FA_conserved <- Ptuh_3UTR_FA_df %>% filter(str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))
Ptuh_3UTR_FA_unconserved <- Ptuh_3UTR_FA_df %>% filter(!str_detect(given_miRNA_name, "mir-100|mir-2023|mir-2025|mir-2036"))

# Also annotate the full dfs
Apul_3UTR_FA_df$conservation <- ifelse(Apul_3UTR_FA_df$given_miRNA_name %in% Apul_3UTR_FA_conserved$given_miRNA_name, "conserved", "unconserved")
Peve_3UTR_FA_df$conservation <- ifelse(Peve_3UTR_FA_df$given_miRNA_name %in% Peve_3UTR_FA_conserved$given_miRNA_name, "conserved", "unconserved")
Ptuh_3UTR_FA_df$conservation <- ifelse(Ptuh_3UTR_FA_df$given_miRNA_name %in% Ptuh_3UTR_FA_conserved$given_miRNA_name, "conserved", "unconserved")
```

Take a look
```{r}
print(Apul_3UTR_FA_conserved %>% dplyr::select(given_miRNA_name, Protein.names,  Gene.Ontology..biological.process.) %>% filter(!is.na(Protein.names)))
print(Peve_3UTR_FA_conserved %>% dplyr::select(given_miRNA_name, Protein.names, Gene.Ontology..biological.process.) %>% filter(!is.na(Protein.names)))
print(Ptuh_3UTR_FA_conserved %>% dplyr::select(given_miRNA_name, Protein.names, Gene.Ontology..biological.process.) %>% filter(!is.na(Protein.names)))
```

The only miRNA that is conserved accross the 3 species and has at least one functionally annotated target (miRanda + significant PCC) is miR-100. This isn't surprising in A.pulchra and P.evermanni, it is also the conserved miRNA with the highest number of putative targets (see below). miR-100 is functionally conserved across the three species in cell cycle regulation and early development. The other conserved miRNA don't have enough annotated targets in all species to determine whether functional conservation exists. Also keep in mind that some important functions (e.g. reproduction, development, immune response) make not appear in this dataset, since deep-dive samples were collected only at a single site and timepoint, and are limited in number.

```{r}
# Apul
# Count number of unique mRNA targets per miRNA
Apul_target_counts <- Apul_3UTR_FA_df %>%
  group_by(given_miRNA_name) %>%
  summarise(num_targets = n_distinct(mRNA)) %>%
  arrange(desc(num_targets))

# Plot
ggplot(Apul_target_counts, aes(x = reorder(given_miRNA_name, -num_targets), y = num_targets)) +
  geom_col(fill = "#408EC6") +
  labs(x = "miRNA", y = "Number of unique mRNA targets", title = "Apul: Target Count per miRNA") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Peve
# Count number of unique mRNA targets per miRNA
Peve_target_counts <- Peve_3UTR_FA_df %>%
  group_by(given_miRNA_name) %>%
  summarise(num_targets = n_distinct(mRNA)) %>%
  arrange(desc(num_targets))

# Plot
ggplot(Peve_target_counts, aes(x = reorder(given_miRNA_name, -num_targets), y = num_targets)) +
  geom_col(fill = "#1E2761") +
  labs(x = "miRNA", y = "Number of unique mRNA targets", title = "Peve: Target Count per miRNA") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Ptuh
# Count number of unique mRNA targets per miRNA
Ptuh_target_counts <- Ptuh_3UTR_FA_df %>%
  group_by(given_miRNA_name) %>%
  summarise(num_targets = n_distinct(mRNA)) %>%
  arrange(desc(num_targets))

# Plot
ggplot(Ptuh_target_counts, aes(x = reorder(given_miRNA_name, -num_targets), y = num_targets)) +
  geom_col(fill = "#7A2048") +
  labs(x = "miRNA", y = "Number of unique mRNA targets", title = "Ptuh: Target Count per miRNA") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

