---
title: "mRNA GO enrichment comparison"
author: "Jill Ashey"
date: "2025-06-09"
output: html_document
---

This code will look at the overlap between the GO terms that were enriched in the expressed mRNAs between species. GO enrichment code for expressed mRNAs for each species can be found at the following links: 

- [ACR](https://github.com/urol-e5/deep-dive-expression/blob/main/D-Apul/code/29-Apul-mRNA-GO-enrichment.Rmd)
- [POC](https://github.com/urol-e5/deep-dive-expression/blob/main/F-Ptuh/code/13-Ptuh-mRNA-GO-enrichment.Rmd)
- [POR](https://github.com/urol-e5/deep-dive-expression/blob/main/E-Peve/code/13-Peve-mRNA-GO-enrichment.Rmd) 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rrvgo)
library(org.Hs.eg.db)
library(UpSetR)
#library(org.Ce.eg.db)
```

Read in data 
```{r}
apul_go <- read.csv("../../D-Apul/output/29-Apul-mRNA-GO-enrichment/Apul-GO_en_sig_expressed_genes.csv")
ptuh_go <- read.csv("../../F-Ptuh/output/13-Ptuh-mRNA-GO-enrichment/Ptuh-GO_en_sig_expressed_genes.csv")
peve_go <- read.csv("../../E-Peve/output/13-Peve-mRNA-GO-enrichment/Peve-GO_en_sig_expressed_genes.csv")

# How many unique GO terms?
length(unique(apul_go$GO))
length(unique(ptuh_go$GO))
length(unique(peve_go$GO))
```

256 unique GO terms for Apul, 297 unique GO terms for Ptuh, and 299 unique GO terms for Peve. 

Examine overlap in GO enrichment terms (ALL ontologies)
```{r}
apul_go_terms <- unique(apul_go$GO)
ptuh_go_terms <- unique(ptuh_go$GO)
peve_go_terms <- unique(peve_go$GO)

# Overlap between all three
go_all_three <- Reduce(intersect, list(apul_go_terms, ptuh_go_terms, peve_go_terms))

# Overlap only between apul and ptuh (not peve)
go_apul_ptuh_only <- setdiff(intersect(apul_go_terms, ptuh_go_terms), peve_go_terms)

# Overlap only between apul and peve (not ptuh)
go_apul_peve_only <- setdiff(intersect(apul_go_terms, peve_go_terms), ptuh_go_terms)

# Overlap only between ptuh and peve (not apul)
go_ptuh_peve_only <- setdiff(intersect(ptuh_go_terms, peve_go_terms), apul_go_terms)

# Unique to apul
go_unique_apul <- setdiff(apul_go_terms, union(ptuh_go_terms, peve_go_terms))

# Unique to ptuh
go_unique_ptuh <- setdiff(ptuh_go_terms, union(apul_go_terms, peve_go_terms))

# Unique to peve
go_unique_peve <- setdiff(peve_go_terms, union(apul_go_terms, ptuh_go_terms))

# To check the counts:
length(go_all_three)
length(go_apul_ptuh_only)
length(go_apul_peve_only)
length(go_ptuh_peve_only)
length(go_unique_apul)
length(go_unique_ptuh)
length(go_unique_peve)

```

Plot with upset plot 
```{r}
go_list <- list(
  apul = apul_go_terms,
  ptuh = ptuh_go_terms,
  peve = peve_go_terms
)
upset(fromList(go_list), order.by = "freq")
```

Plot only BP terms with upset plot 
```{r}
# Filter by BP
apul_go_BP <- apul_go %>%
  filter(ontology == "Biological Processes")
ptuh_go_BP <- ptuh_go %>%
  filter(ontology == "Biological Processes")
peve_go_BP <- peve_go %>%
  filter(ontology == "Biological Processes")

# Unique terms 
apul_go_BP_terms <- unique(apul_go_BP$GO)
ptuh_go_BP_terms <- unique(ptuh_go_BP$GO)
peve_go_BP_terms <- unique(peve_go_BP$GO)

# Plot
go_BP_list <- list(
  apul = apul_go_BP_terms,
  ptuh = ptuh_go_BP_terms,
  peve = peve_go_BP_terms
)
upset(fromList(go_BP_list), order.by = "freq")
```

Use [rrvgo](https://bioconductor.org/packages/devel/bioc/vignettes/rrvgo/inst/doc/rrvgo.html#calculating-the-similarity-matrix-and-reducing-go-terms) to visualize general trends for GO terms shared and unique. ONLY BP below, as rrvgo appears to be defaulting to just doing BP ontology. 

Overlap of all three
```{r}
# Similarity of terms 
simMatrix_all <- calculateSimMatrix(go_all_three,
                                orgdb="org.Hs.eg.db",
                                ont = c("BP", "MF", "CC"),
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_all <- reduceSimMatrix(simMatrix_all,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_all,
            reducedTerms_all,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_all, reducedTerms_all)

pdf("../output/05-expressed-mRNA-GO-enrichment/all_overlap_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_all, force.print.labels=T)
dev.off()
```

Overlap of Apul and Ptuh
```{r}
# Similarity of terms 
simMatrix_apul_ptuh <- calculateSimMatrix(go_apul_ptuh_only,
                                orgdb="org.Hs.eg.db",
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_apul_ptuh <- reduceSimMatrix(simMatrix_apul_ptuh,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_apul_ptuh,
            reducedTerms_apul_ptuh,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_apul_ptuh, reducedTerms_apul_ptuh)

pdf("../output/05-expressed-mRNA-GO-enrichment/apul_ptuh_overlap_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_apul_ptuh, force.print.labels=T)
dev.off()
```

Overlap of Apul and Peve
```{r}
# Similarity of terms 
simMatrix_apul_peve <- calculateSimMatrix(go_apul_peve_only,
                                orgdb="org.Hs.eg.db",
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_apul_peve <- reduceSimMatrix(simMatrix_apul_peve,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_apul_peve,
            reducedTerms_apul_peve,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_apul_peve, reducedTerms_apul_peve)

pdf("../output/05-expressed-mRNA-GO-enrichment/apul_peve_overlap_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_apul_peve, force.print.labels=T)
dev.off()
```

Overlap of Ptuh and Peve
```{r}
# Similarity of terms 
simMatrix_ptuh_peve <- calculateSimMatrix(go_ptuh_peve_only,
                                orgdb="org.Hs.eg.db",
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_ptuh_peve <- reduceSimMatrix(simMatrix_ptuh_peve,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_ptuh_peve,
            reducedTerms_ptuh_peve,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_ptuh_peve, reducedTerms_ptuh_peve)

pdf("../output/05-expressed-mRNA-GO-enrichment/ptuh_peve_overlap_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_ptuh_peve, force.print.labels=T)
dev.off()
```

Just Apul
```{r}
# Similarity of terms 
simMatrix_apul <- calculateSimMatrix(go_unique_apul,
                                orgdb="org.Hs.eg.db",
                                ont = c("BP", "MF", "CC"),
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_apul <- reduceSimMatrix(simMatrix_apul,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_apul,
            reducedTerms_apul,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_apul, reducedTerms_apul)

pdf("../output/05-expressed-mRNA-GO-enrichment/apul_only_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_apul, force.print.labels=T)
dev.off()
```

Just Ptuh
```{r}
# Similarity of terms 
simMatrix_ptuh <- calculateSimMatrix(go_unique_ptuh,
                                orgdb="org.Hs.eg.db",
                                ont = c("BP", "MF", "CC"),
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_ptuh <- reduceSimMatrix(simMatrix_ptuh,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_ptuh,
            reducedTerms_ptuh,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_ptuh, reducedTerms_ptuh)

pdf("../output/05-expressed-mRNA-GO-enrichment/ptuh_only_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_ptuh, force.print.labels=T)
dev.off()
```

Just Peve
```{r}
# Similarity of terms 
simMatrix_peve <- calculateSimMatrix(go_unique_peve,
                                orgdb="org.Hs.eg.db",
                                ont = c("BP", "MF", "CC"),
                                method="Rel")

# reducing sig enriched terms to higher order terms and grouping them together 
reducedTerms_peve <- reduceSimMatrix(simMatrix_peve,
                                scores = "uniqueness",
                                threshold=0.9,
                                orgdb="org.Hs.eg.db")
## Plot
heatmapPlot(simMatrix_peve,
            reducedTerms_peve,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

scatterPlot(simMatrix_peve, reducedTerms_peve)

pdf("../output/05-expressed-mRNA-GO-enrichment/peve_only_GO_BP_semsim.pdf", width = 18, height = 8)
treemapPlot(reducedTerms_peve, force.print.labels=T)
dev.off()
```