---
title: "09.1-epimachinery-ncRNA-protein-expression"
author: "Kathleen Durkin"
date: "2026-04-28"
output: 
  github_document:
    toc: true
    number_sections: false
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(stringr)
library(purrr)
```

Read in ncRNA/epimachinery protein related info 
```{r}
Apul_ncRNA_df <- read.csv("../../D-Apul/output/34-Apul-ncRNA-machinery-BLAST/ncRNAmach-blastp-Apul_annotated_reduced.csv")
Apul_ncRNA_df$Species <- "Apul"
Peve_ncRNA_df <- read.csv("../../E-Peve/output/32-Peve-ncRNA-machinery-BLAST/ncRNAmach-blastp-Peve_annotated_reduced.csv")
Peve_ncRNA_df$Species <- "Peve"
Ptuh_ncRNA_df <- read.csv("../../F-Ptuh/output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_annotated_reduced.csv")
Ptuh_ncRNA_df$Species <- "Ptuh"

ncRNA_df <- rbind(Apul_ncRNA_df, Peve_ncRNA_df, Ptuh_ncRNA_df)
ncRNA_df$gene_name <- ncRNA_df$std_gene_name
ncRNA_df <- ncRNA_df %>% dplyr::select(-std_gene_name)

Apul_epi_df <- read.csv("../../D-Apul/output/35-Apul-epi-machinery-BLAST/Mach-blastp-Apul_reduced.csv")
Apul_epi_df$Species <- "Apul"
Peve_epi_df <- read.csv("../../E-Peve/output/33-Peve-epi-machinery-BLAST/Mach-blastp-Peve_reduced.csv")
Peve_epi_df$Species <- "Peve"
Ptuh_epi_df <- read.csv("../../F-Ptuh/output/33-Ptuh-epi-machinery-BLAST/Mach-blastp-Ptuh_reduced.csv")
Ptuh_epi_df$Species <- "Ptuh"

epi_df <- rbind(Apul_epi_df, Peve_epi_df, Ptuh_epi_df)
# The epimachinery proteins originally sourced form Hollie's set are not always fully capitalized (e.g., Dnmt1 instead of DNMT1). To appropriately match my curated ncRNA machinery gene names, capitalize everything
epi_df$gene_name <- toupper(epi_df$gene_name)

prot_df <- rbind(ncRNA_df, epi_df) %>% distinct() # some epimachinery was present in both curated sets
# Rename
prot_df$Protein_ID <- prot_df$target
prot_df$Protein_of_interest <- prot_df$gene_name
prot_df <- prot_df %>% dplyr::select(-target, -gene_name)


# Remove -T* from Apul protein names 
prot_df <- prot_df %>%
  mutate(Protein_ID = if_else(
    Species == "Apul",
    sub("-T\\d+$", "", Protein_ID),
    Protein_ID
  ))
```

## Clean up machinery list

There are a few instances of single coral genes matching more than one ncRNA/epimachinery gene. Let's clean that up before proceeding

```{r}
# Collapse rows where one Protein_ID maps to multiple genes (within a species)
prot_df <- prot_df %>%
  group_by(Species, Protein_ID) %>%
  summarise(
    Protein_of_interest = paste(sort(unique(Protein_of_interest)), collapse = "; "),
    .groups = "drop"
  )
```

Also want a single, more "generic" epimachinery name for instances of multiple matches (e.g., if a gene matches DICER1 and DICER2, assign it the name DICER)

```{r}

# Manual lookup for biologically meaningful collapses that don't share a clean prefix
manual_lookup <- c(
  "NONO; PSPC1; SFPQ"                              = "DBHS family",
  "KMT3B-NSD1; WHSC1-NSD2; WHSC1L-NSD3"            = "NSD",
  "KMT1C-EHMT2; KMT1D-EHMT1"                       = "EHMT",
  "KMT2H-ASH1L; KMT3A-SETD2"                       = "H3K36 KMT (ASH1L/SETD2)",
  "HDAC; HDAC4; HDAC5; HDAC7; HDAC9"               = "HDAC class IIa",
  "HDAC; HDAC10; HDAC6"                            = "HDAC class IIb",
  "MAPK8; MAPK8B; MAPK9"                           = "MAPK8/9 (JNK)",
  "EZH1; EZH2; KMT6-EZH2"                          = "EZH",
  # histone lysine methyltransferases
  "BAZ1B; SUV39H1-KMT1A; SUV39H2-KMT1B"            = "H3K9 KMT (SUV39H)",
  "KMT2H-ASH1L; KMT3A-SETD2; WHSC1-NSD2"           = "H3K36 KMT",
  "KMT3B-NSD1; WHSC1L-NSD3"                        = "NSD",
  "KMT3C-SMYD2; SMYD3"                             = "SMYD",
  "KMT3C-SMYD2; SMYD1; SMYD3"                      = "SMYD",
  "KMT2E; SETD5"                                   = "SETD5/KMT2E",
  # Histone demethylases
  "KDM6A; UTY"                                     = "KDM6A/UTY",
  "KDM6A; KDM6B; UTY"                              = "KDM6",
  "KDM7A; PHF2; PHF8"                              = "KDM7",
  # PRDM family (zinc-finger H3K methyltransferases)
  "CTCF; PRDM15; PRDM4; PRDM9"                     = "PRDM (+CTCF)",
  "CTCF; PRDM5; PRDM9"                             = "PRDM (+CTCF)",
  
  "EWSR1; FUS"                                     = "FET family",
  # Histone variants
  "H1-2; MACROH2A1"                                = "Histone variants",
  "H1-2; H2AJ; H2AW; MACROH2A1"                    = "Histone variants",
  "H2AX; HDAC1; HDAC2"                             = "FLAG: H2AX+HDAC1/2 (review)",
  # Chromatin readers
  "BAZ2B; MBD6"                                    = "BAZ2B/MBD6",
  # Mixed / potentially artifactual
  "H2AX; HDAC; HDAC1; HDAC2"                       = "FLAG: H2AX+HDAC1/2 (review)",
  "ALG13; OTUD4"                                   = "FLAG: ALG13+OTUD4 (review)",
  "DNMT3L; USP9Y"                                  = "FLAG: DNMT3L+USP9Y (review)"
  )


# Helper: longest common alphabetical prefix
common_prefix <- function(x) {
  if (length(x) <= 1) return(x[1] %||% "")
  splits <- strsplit(x, "")
  min_len <- min(lengths(splits))
  if (min_len == 0) return("")
  out <- character(0)
  for (i in seq_len(min_len)) {
    chars <- map_chr(splits, i)
    if (length(unique(chars)) == 1) out <- c(out, chars[1]) else break
  }
  paste(out, collapse = "")
}

# Main simplifier
simplify_genes <- function(gene_str, lookup = manual_lookup, min_prefix = 3) {
  # Normalize the input string for lookup matching
  genes <- str_split(gene_str, "\\s*;\\s*")[[1]] |> unique() |> sort()
  norm_str <- paste(genes, collapse = "; ")
  # Manual lookup wins
  if (norm_str %in% names(lookup)) return(unname(lookup[norm_str]))
  # Single gene: return as-is
  if (length(genes) == 1) return(genes)
  # Drop generic placeholder if a paralog of it is present
  #    (e.g., "HDAC" alongside "HDAC4" → drop the bare "HDAC")
  is_generic <- str_detect(genes, "^[A-Z]+$")
  if (any(is_generic)) {
    keep <- map_lgl(genes, function(g) {
      if (!str_detect(g, "^[A-Z]+$")) return(TRUE)
      !any(str_detect(setdiff(genes, g), paste0("^", g, "\\d")))
    })
    genes <- genes[keep]
    if (length(genes) == 1) return(genes)
  }
  # Try common prefix — keep trailing letters OR digits, only strip non-alphanumerics
  prefix <- common_prefix(genes) |> str_remove("[^A-Z0-9]+$")
  if (nchar(prefix) >= min_prefix) return(prefix)
  # Fall back to the (cleaned) semicolon list
  paste(genes, collapse = "; ")
}

prot_df <- prot_df %>%
  mutate(Protein_of_interest_simplified = map_chr(Protein_of_interest, simplify_genes))

# Check
head(prot_df)
```

Save master table, in case want to use it in the future
```{r}
write.csv(prot_df, "../output/09.1-epimachinery-ncRNA-protein-expression/ncRNAepimachinery_gene_db_spec.csv", row.names = FALSE)
```

Read in Apul, Peve and Ptua mRNA counts 
```{r}
peve <- read.csv("../../E-Peve/output/06.2-Peve-Hisat/Peve-gene_count_matrix.csv")
peve$gene_id <- gsub("^gene-", "", peve$gene_id)

ptua <- read.csv("../../F-Ptuh/output/06.2-Ptuh-Hisat/Ptuh-gene_count_matrix.csv")
ptua$gene_id <- gsub("^gene-", "", ptua$gene_id)

apul <- read.csv("../../D-Apul/output/07-Apul-Hisat/Apul-gene_count_matrix.csv")
```

### Apul

Join prot_df with Peve counts matrix 
```{r}
apul_prots <- prot_df %>%
  inner_join(apul, by = c("Protein_ID" = "gene_id"))
```

Normalize counts 
```{r}
# Function to normalize counts (simple RPM normalization)
normalize_counts <- function(counts) {
  rpm <- t(t(counts) / colSums(counts)) * 1e6
  return(rpm)
}

# Get only the RNA count columns
counts <- apul_prots %>% select(starts_with("RNA."))

# Normalize to RPM
rpm <- normalize_counts(counts)

# Add the normalized RPM back to the original dataframe
apul_prots_rpm <- apul_prots %>%
  select(Species, Protein_of_interest_simplified, Protein_ID) %>%
  bind_cols(rpm)
```

Summarize for Apul 
```{r}
apul_summary_stats <- apul_prots_rpm %>%
  rowwise() %>%
  mutate(mean_rpm = mean(c_across(starts_with("RNA."))),
         sd_rpm   = sd(c_across(starts_with("RNA.")))) %>%
  ungroup() %>%
  group_by(Species, Protein_of_interest_simplified) %>%
  summarise(avg_rpm = mean(mean_rpm),
            std_rpm = mean(sd_rpm),
            n = n_distinct(Protein_ID),
            .groups = "drop")

write.csv(apul_summary_stats, "../output/09.1-epimachinery-ncRNA-protein-expression/apul_epi_ncRNA_protein_transcript_rpm.csv")
```

### Peve 

Join prot_df with Peve counts matrix 
```{r}
peve_prots <- prot_df %>%
  inner_join(peve, by = c("Protein_ID" = "gene_id"))
```

Normalize counts 
```{r}
# Function to normalize counts (simple RPM normalization)
normalize_counts <- function(counts) {
  rpm <- t(t(counts) / colSums(counts)) * 1e6
  return(rpm)
}

# Get only the RNA count columns
counts <- peve_prots %>% select(starts_with("RNA."))

# Normalize to RPM
rpm <- normalize_counts(counts)

# Add the normalized RPM back to the original dataframe
peve_prots_rpm <- peve_prots %>%
  select(Species, Protein_of_interest_simplified, Protein_ID) %>%
  bind_cols(rpm)
```

Summarize for Peve 
```{r}
peve_summary_stats <- peve_prots_rpm %>%
  rowwise() %>%
  mutate(mean_rpm = mean(c_across(starts_with("RNA."))),
         sd_rpm   = sd(c_across(starts_with("RNA.")))) %>%
  ungroup() %>%
  group_by(Species, Protein_of_interest_simplified) %>%
  summarise(avg_rpm = mean(mean_rpm),
            std_rpm = mean(sd_rpm),
            n = n_distinct(Protein_ID),
            .groups = "drop")

write.csv(peve_summary_stats, "../output/09.1-epimachinery-ncRNA-protein-expression/peve_epi_ncRNA_protein_transcript_rpm.csv")
```

### Ptua

Join prot_df with Ptua counts matrix 
```{r}
ptua_prots <- prot_df %>%
  inner_join(ptua, by = c("Protein_ID" = "gene_id"))
```

Normalize counts 
```{r}
# Function to normalize counts (simple RPM normalization)
normalize_counts <- function(counts) {
  rpm <- t(t(counts) / colSums(counts)) * 1e6
  return(rpm)
}

# Get only the RNA count columns
counts <- ptua_prots %>% select(starts_with("RNA."))

# Normalize to RPM
rpm <- normalize_counts(counts)

# Add the normalized RPM back to the original dataframe
ptua_prots_rpm <- ptua_prots %>%
  select(Species, Protein_of_interest_simplified, Protein_ID) %>%
  bind_cols(rpm)
```

Summarize for Ptua 
```{r}
ptua_summary_stats <- ptua_prots_rpm %>%
  rowwise() %>%
  mutate(mean_rpm = mean(c_across(starts_with("RNA."))),
         sd_rpm   = sd(c_across(starts_with("RNA.")))) %>%
  ungroup() %>%
  group_by(Species, Protein_of_interest_simplified) %>%
  summarise(avg_rpm = mean(mean_rpm),
            std_rpm = mean(sd_rpm),
            n = n_distinct(Protein_ID),
            .groups = "drop")

write.csv(ptua_summary_stats, "../output/09.1-epimachinery-ncRNA-protein-expression/ptuh_epi_ncRNA_protein_transcript_rpm.csv")
```

### All species 

```{r}
all_spp <- rbind(apul_summary_stats, peve_summary_stats, ptua_summary_stats)
```


## Plot 

```{r}
species_labels <- c(
  Apul = expression(italic("Acropora pulchra")),
  Peve = expression(italic("Porites evermanni")),
  Ptuh = expression(italic("Pocillopora tuahiniensis"))
)

# Plotting function
plot_set <- function(set){
  
  df <- all_spp %>% filter(Protein_of_interest_simplified %in% set)
  
  ggplot(df, aes(x = Species, y = avg_rpm, fill = Species)) +
    geom_col(color = "black", width = 0.7, size = 1.5) +
    geom_errorbar(aes(ymin = avg_rpm - std_rpm, ymax = avg_rpm + std_rpm),
                  width = 0.2, color = "black", size = 1.2) +
    facet_wrap(~ Protein_of_interest_simplified, scales = "free_y") +
    scale_fill_manual(
      values = c("Apul" = "#408EC6", "Peve" = "#1E2761", "Ptuh" = "#7A2048"),
      labels = species_labels
    ) +
    scale_x_discrete(labels = species_labels) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = "", y = "Average RPM", fill = "Species") +
    theme_bw() +
    theme(
      axis.title = element_text(size = 36, face = "bold"),        # Y-axis title
      axis.text = element_text(size = 34, colour = "black"),      # Axis text
      axis.text.x = element_text(angle = 45, hjust = 1),          # X-axis angle
      legend.title = element_text(size = 34, face = "bold"),      # Legend title
      legend.position = "top",
      legend.text = element_text(size = 32),                      # Legend labels
      strip.text = element_text(size = 34, face = "bold"),        # Facet labels
      strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),  # Facet background
      panel.border = element_rect(color = "black", size = 1.2),   # Panel border
      panel.grid.major = element_line(size = 0.5, color = "gray"),# Internal grid lines
      panel.grid.minor = element_blank(),
      panel.spacing = unit(1, "lines"),
      axis.line = element_line(size = 1, colour = "black"),       # Axis lines
      axis.ticks = element_line(size = 1),                        # Axis ticks
      strip.placement = "outside"
    )
}
```

```{r}
ncrna_biogen <- c("DROSHA", "DGCR8", "DICER1", "AGO2", "AGO", "XPO5", "TARBP1", "INTS11", "NCBP1", "SRRT")
methyl <- c("DNMT1", "DNMT3A", "DNMT3B", "DNMT", "DNMT3", "UHRF1", "UHRF", "TET1", "TET2", "TET3", "TET", "MBD2")
hist_actyl <- c("KAT", "KAT1", "KAT1-HAT1", "KAT2", "KAT2A", "KAT2B", "KAT3", "KAT3A-CREBBP", "KAT3B-EP300", "KAT4", "KAT5", "KAT6", "KAT6A", "KAT7", "KAT8", "KAT9", "KAT9-ELP3")
hist_deactyl <- c("HDAC1", "HDAC2", "HDAC3", "HDAC6", "HDAC8", "HDAC", "SIRT1", "SIRT2", "SIRT3", "SIRT4", "SIRT5", "SIRT6", "SIRT7")
hist_methyl <- c("EZH2", "EZH", "KMT", "KMT2", "KMT2A", "KMT2D", "KMT4-DOT1L", "KMT3A-SETD2", "KMT5", "KMT1E-SETDB1", "SUV39H1-KMT1A", "PRMT5", "KMT5A-SETD8")
hist_demethyl <- c("KDM1A", "KDM1B", "KDM2", "KDM3", "KDM4", "KDM5", "KDM6", "KDM7", "KDM8", "KDM4A", "KDM5A", "KDM5B", "KDM6A", "KDM6B", "KDM2A", "KDM3A", "JMJD6")
chrom <- c("SMARCA2", "SMARCA4", "SMARCC2", "SMARCA", "ARID1A", "ARID1B", "ARID", "BRD4", "BRD3", "BRD", "HIRA", "BAZ1B")
hist_ubq <- c("RNF20", "RNF40", "RNF", "RNF168", "BAP1", "MYSM1", "UBE2A", "USP16", "USP22", "USP44")
rna_mod <- c("METTL14", "METTL16", "WTAP", "ALKBH5", "YTHDC1", "YTHDF2", "NSUN2", "DKC1", "PUS7")
```

```{r}
ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_ncrna_biogen.png", plot_set(ncrna_biogen), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_methyl.png", plot_set(methyl), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_hist_actyl.png", plot_set(hist_actyl), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_hist_deactyl.png", plot_set(hist_deactyl), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_hist_methyl.png", plot_set(hist_methyl), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_hist_demethyl.png", plot_set(hist_demethyl), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_chrom.png", plot_set(chrom), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_hist_ubq.png", plot_set(hist_ubq), width = 31, height = 20)

ggsave(filename = "../output/09.1-epimachinery-ncRNA-protein-expression/transcript_rpm_by_spp_rna_mod.png", plot_set(rna_mod), width = 31, height = 20)
```

