---
title: "12-miRNA-epimachinery"
author: "Kathleen Durkin"
date: "2026-04-08"
output: 
  github_document:
    toc: true
    number_sections: false
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

We've previously looked for instances of miRNA putatively binding to ncRNA and methylation machinery (a.g. AGO transcripts), but this search seemed to miss hits. Perhaps more importantly, the reference used was not comprehensive, because it didn't include hits for other epigenetic machinery, such as those involved in histone modifications. Let's redo and expand the epimachinery analysis.

Load packages
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)
library(GSEABase)
library(GO.db)
library(knitr)
library(tidyverse)
```

Load and format miRNA target FA tables

```{r}
# Apul
apul_FA <- read.csv("../../D-Apul/output/09.13-Apul-mRNA-miRNA-interactions-FE-pooled/miRNA_pooled_targets_FA.csv") %>% dplyr::select(miRNA, mRNA, PCC.cor, p_value) %>% distinct()

# Peve
peve_FA <- read.csv("../../E-Peve/output/10.14-Peve-mRNA-miRNA-interactions-FE-pooled/miRNA_pooled_targets_FA.csv") %>% dplyr::select(miRNA, mRNA, PCC.cor, p_value) %>% distinct()

# Ptuh
ptuh_FA <- read.csv("../../F-Ptuh/output/11.14-Ptuh-mRNA-miRNA-interactions-FE-pooled/miRNA_pooled_targets_FA.csv") %>% dplyr::select(miRNA, mRNA, PCC.cor, p_value) %>% distinct()
```


Load miRNA assigned names (e.g., "miR-100") to annotate, instead of using ShortStack's Cluster labels
```{r}
# Load
## Apul
apul_names <- read.csv("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Apul_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)

## Peve
peve_names <- read.csv("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)

## Ptuh
ptuh_names <- read.csv("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Ptuh_Results_mature_named_miRNAs.csv") %>% dplyr::select(Name, given_miRNA_name)
```

Combine with miRNA annotated target+coexpression tables
```{r}
apul_FA_annot <- left_join(apul_FA, apul_names, by = c("miRNA" = "Name"))
peve_FA_annot <- left_join(peve_FA, peve_names, by = c("miRNA" = "Name"))
ptuh_FA_annot <- left_join(ptuh_FA, ptuh_names, by = c("miRNA" = "Name"))
```


## Epimachinery 

Load epimachinery BLAST hits to the three coral genomes
```{r}
#Load
apul_epi <- read.table("../../D-Apul/output/35-Apul-epi-machinery-BLAST/Mach-blastp-Apul_out.tab") %>% dplyr::select(V1,V2)
# Remove transcript # suffixes ("-T1")
apul_epi$V2 <- gsub("-T[0-9]+$", "", apul_epi$V2)

# Rename columns
apul_epi$gene_name <- apul_epi$V1
apul_epi$target <- apul_epi$V2
apul_epi <- apul_epi %>% dplyr::select(-V1, -V2)


#Load
peve_epi <- read.table("../../E-Peve/output/33-Peve-epi-machinery-BLAST/Mach-blastp-Peve_out.tab") %>% dplyr::select(V1,V2)

# Rename columns
peve_epi$gene_name <- peve_epi$V1
peve_epi$target <- peve_epi$V2
peve_epi <- peve_epi %>% dplyr::select(-V1, -V2)


#Load
ptuh_epi <- read.table("../../F-Ptuh/output/33-Ptuh-epi-machinery-BLAST/Mach-blastp-Ptuh_out.tab") %>% dplyr::select(V1,V2)

# Rename columns
ptuh_epi$gene_name <- ptuh_epi$V1
ptuh_epi$target <- ptuh_epi$V2
ptuh_epi <- ptuh_epi %>% dplyr::select(-V1, -V2)

```


### Apul

```{r}
# Bind to annotate putative miRNA targets with epimachinery matches
apul_epi_miRNA <- left_join(apul_epi, apul_FA_annot, by = c("target" = "mRNA"))

# Filter to retain only putative miRNA targets that are significantly coexpressed 
apul_epi_miRNA <- apul_epi_miRNA %>% filter(p_value < 0.05)

# Summary
cat("\n# unique miRNAs:", apul_epi_miRNA$miRNA %>% unique() %>% length(), "\n")
cat("# unique mRNAs:", apul_epi_miRNA$target %>% unique() %>% length(), "\n")
cat("# unique epimachinery proteins:", apul_epi_miRNA$gene_name %>% unique() %>% length(), "\n")
```
In *A. pulchra*, 22 of the 39 miRNAs putatively bind to a total of 29 unique epigenetic protein machinery transcripts, based on both predicted binding and significant coexpression. These transcripts match 38 epimachinery annotations. Note that the # of unique transcripts is higher than the # of unique annotations because some of the coral genes match multiple variants of a single protein (e.g., FUN_001354, which matches both Parp5a-Tnks-201 and Parp5b-Tnks2-201)

Collapse to unique miRNA-mRNA pairs
```{r}
apul_epi_miRNA_collapsed <- apul_epi_miRNA %>%
  group_by(target, given_miRNA_name) %>%
  mutate(gene = paste(unique(gene_name), collapse = "; ")) %>%
  distinct(target, given_miRNA_name, .keep_all = TRUE) %>%
  ungroup() %>%
  dplyr::select(-gene_name, -miRNA)
```


### Peve

```{r}
# Bind to annotate putative miRNA targets with epimachinery matches
peve_epi_miRNA <- left_join(peve_epi, peve_FA_annot, by = c("target" = "mRNA"))

# Filter to retain only putative miRNA targets that are significantly coexpressed 
peve_epi_miRNA <- peve_epi_miRNA %>% filter(p_value < 0.05)


# Summary
cat("\n# unique miRNAs:", peve_epi_miRNA$miRNA %>% unique() %>% length(), "\n")
cat("# unique mRNAs:", peve_epi_miRNA$target %>% unique() %>% length(), "\n")
cat("# unique epimachinery proteins:", peve_epi_miRNA$gene_name %>% unique() %>% length(), "\n")
```
In *P. evermanni*, 7 of the 45 miRNAs putatively bind to a total of 11 unique epigenetic protein machinery transcripts, based on both predicted binding and significant. These transcripts match 14 epimachinery annotations. Again, note that the # of unique transcripts is higher than the # of unique annotations because some of the coral genes match multiple variants of a single protein.

Collapse to unique miRNA-mRNA pairs
```{r}
peve_epi_miRNA_collapsed <- peve_epi_miRNA %>%
  group_by(target, given_miRNA_name) %>%
  mutate(gene = paste(unique(gene_name), collapse = "; ")) %>%
  distinct(target, given_miRNA_name, .keep_all = TRUE) %>%
  ungroup() %>%
  dplyr::select(-gene_name, -miRNA)
```


### Ptuh

```{r}
# Bind to annotate putative miRNA targets with epimachinery matches
ptuh_epi_miRNA <- left_join(ptuh_epi, ptuh_FA_annot, by = c("target" = "mRNA"))

# Filter to retain only putative miRNA targets that are significantly coexpressed 
ptuh_epi_miRNA <- ptuh_epi_miRNA %>% filter(p_value < 0.05)


# Summary
cat("\n# unique miRNAs:", ptuh_epi_miRNA$miRNA %>% unique() %>% length(), "\n")
cat("# unique mRNAs:", ptuh_epi_miRNA$target %>% unique() %>% length(), "\n")
cat("# unique epimachinery proteins:", ptuh_epi_miRNA$gene_name %>% unique() %>% length(), "\n")
```

In *P. tuahiniensis*, 9 of the 37 miRNAs putatively bind to a total of 10 unique epigenetic protein machinery transcripts, based on both predicted binding and significant. These transcripts match 14 epimachinery annotations. Again, note that the # of unique transcripts is higher than the # of unique annotations because some of the coral genes match multiple variants of a single protein.

Collapse to unique miRNA-mRNA pairs
```{r}
ptuh_epi_miRNA_collapsed <- ptuh_epi_miRNA %>%
  group_by(target, given_miRNA_name) %>%
  mutate(gene = paste(unique(gene_name), collapse = "; ")) %>%
  distinct(target, given_miRNA_name, .keep_all = TRUE) %>%
  ungroup() %>%
  dplyr::select(-gene_name, -miRNA)
```


## Updated ncRNA machinery analysis

I manually curated a db of fasta sequences for protein machinery important to miRNA and lncRNA biogenesis and function (details in notebook, [here](https://shedurkin.github.io/Roberts-LabNotebook/posts/projects/E5_coral/2026_04_10_DDE_ncRNA_machinery.html) and [here](https://shedurkin.github.io/Roberts-LabNotebook/posts/projects/E5_coral/2026_04_13_DDE_ncRNA_machinery_fastas.html))

load BLAST hits for this ncRNA machinery set

```{r}
#Load
apul_ncRNA <- read.csv("../../D-Apul/output/34-Apul-ncRNA-machinery-BLAST/ncRNAmach-blastp-Apul_annotated_reduced.csv") 

# Remove transcript # suffixes ("-T1")
apul_ncRNA$target <- gsub("-T[0-9]+$", "", apul_ncRNA$target)

#Load
peve_ncRNA <- read.csv("../../E-Peve/output/32-Peve-ncRNA-machinery-BLAST/ncRNAmach-blastp-Peve_annotated_reduced.csv") 

#Load
ptuh_ncRNA <- read.csv("../../F-Ptuh/output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_annotated_reduced.csv") 

```


### Apul

```{r}
# Bind to annotate putative miRNA targets with ncRNA machinery matches
apul_ncRNA_miRNA <- left_join(apul_ncRNA, apul_FA_annot, by = c("target" = "mRNA")) %>% distinct()

# Filter to retain only putative miRNA targets that are significantly coexpressed 
apul_ncRNA_miRNA <- apul_ncRNA_miRNA %>% filter(p_value < 0.05)

# Summary
cat("\n# unique miRNAs:", apul_ncRNA_miRNA$miRNA %>% unique() %>% length(), "\n")
cat("# unique mRNAs:", apul_ncRNA_miRNA$target %>% unique() %>% length(), "\n")
cat("# unique ncRNA machinery genes:", apul_ncRNA_miRNA$std_gene_name %>% unique() %>% length(), "\n")
```

In *A. pulchra*, 8 of the 39 miRNAs putatively bind to a total of 8 unique ncRNA protein machinery transcripts, based on both predicted binding and significant coexpression. These transcripts match 12 ncRNA machinery annotations. Note that the # of unique annotations is higher than the # of unique transcripts because some of the coral genes match multiple similar proteins (e.g., FUN_032699, which matched both AGO1 and AGO2 in the original BLASTp search)

Collapse to unique miRNA-mRNA pairs
```{r}
apul_ncRNA_miRNA_collapsed <- apul_ncRNA_miRNA %>%
  group_by(target, given_miRNA_name) %>%
  mutate(gene = paste(unique(std_gene_name), collapse = "; ")) %>%
  distinct(target, given_miRNA_name, .keep_all = TRUE) %>%
  ungroup() %>%
  dplyr::select(-std_gene_name, -miRNA)
```


### Peve

```{r}
# Bind to annotate putative miRNA targets with ncRNA machinery matches
peve_ncRNA_miRNA <- left_join(peve_ncRNA, peve_FA_annot, by = c("target" = "mRNA")) %>% distinct()

# Filter to retain only putative miRNA targets that are significantly coexpressed 
peve_ncRNA_miRNA <- peve_ncRNA_miRNA %>% filter(p_value < 0.05)

# Summary
cat("\n# unique miRNAs:", peve_ncRNA_miRNA$miRNA %>% unique() %>% length(), "\n")
cat("# unique mRNAs:", peve_ncRNA_miRNA$target %>% unique() %>% length(), "\n")
cat("# unique ncRNA machinery genes:", peve_ncRNA_miRNA$std_gene_name %>% unique() %>% length(), "\n")
```

In *P.evermanni*, 6 of the 45 miRNAs putatively bind to a total of 8 unique ncRNA protein machinery transcripts, based on both predicted binding and significant coexpression. These transcripts match 9 ncRNA machinery annotations. Note that the # of unique annotations is higher than the # of unique transcripts because one of the coral genes matched multiple similar proteins (Peve_00038764, which matched both SARCA2 and SARCA4 in the original BLASTp search)

Collapse to unique miRNA-mRNA pairs
```{r}
peve_ncRNA_miRNA_collapsed <- peve_ncRNA_miRNA %>%
  group_by(target, given_miRNA_name) %>%
  mutate(gene = paste(unique(std_gene_name), collapse = "; ")) %>%
  distinct(target, given_miRNA_name, .keep_all = TRUE) %>%
  ungroup() %>%
  dplyr::select(-std_gene_name, -miRNA)
```


### Ptuh

```{r}
# Bind to annotate putative miRNA targets with ncRNA machinery matches
ptuh_ncRNA_miRNA <- left_join(ptuh_ncRNA, ptuh_FA_annot, by = c("target" = "mRNA")) %>% distinct()

# Filter to retain only putative miRNA targets that are significantly coexpressed 
ptuh_ncRNA_miRNA <- ptuh_ncRNA_miRNA %>% filter(p_value < 0.05)

# Summary
cat("\n# unique miRNAs:", ptuh_ncRNA_miRNA$miRNA %>% unique() %>% length(), "\n")
cat("# unique mRNAs:", ptuh_ncRNA_miRNA$target %>% unique() %>% length(), "\n")
cat("# unique ncRNA machinery genes:", ptuh_ncRNA_miRNA$std_gene_name %>% unique() %>% length(), "\n")
```

In *P.tuahiniensis*, 8 of the 37 miRNAs putatively bind to a total of 12 unique ncRNA protein machinery transcripts, based on both predicted binding and significant coexpression. These transcripts match 16 ncRNA machinery annotations. Note that the # of unique annotations is higher than the # of unique transcripts because some of the coral genes matched multiple similar proteins (e.g., 	
Pocillopora_meandrina_HIv1___RNAseq.g14104.t1, which matched both STAG1 and STAG2 in the original BLASTp search)

Collapse to unique miRNA-mRNA pairs
```{r}
ptuh_ncRNA_miRNA_collapsed <- ptuh_ncRNA_miRNA %>%
  group_by(target, given_miRNA_name) %>%
  mutate(gene = paste(unique(std_gene_name), collapse = "; ")) %>%
  distinct(target, given_miRNA_name, .keep_all = TRUE) %>%
  ungroup() %>%
  dplyr::select(-std_gene_name, -miRNA)
```


## Combine epimachinery and ncRNA machinery outputs to single results table for each species

### Apul

```{r}
# Label with source
apul_epi_miRNA_collapsed$type <- "epimachinery"
apul_ncRNA_miRNA_collapsed$type <- "ncRNA machinery"

# Ensure identical column names and order
colnames(apul_epi_miRNA_collapsed) == colnames(apul_ncRNA_miRNA_collapsed)

# Combine
apul_results <- rbind(apul_epi_miRNA_collapsed, apul_ncRNA_miRNA_collapsed)

# Save
write.csv(apul_results, "../output/12-miRNA-epimachinery/Apul_miRNAtargets_mach.csv", row.names = FALSE)
```

### Peve

```{r}
# Label with source
peve_epi_miRNA_collapsed$type <- "epimachinery"
peve_ncRNA_miRNA_collapsed$type <- "ncRNA machinery"

# Ensure identical column names and order
colnames(peve_epi_miRNA_collapsed) == colnames(peve_ncRNA_miRNA_collapsed)

# Combine
peve_results <- rbind(peve_epi_miRNA_collapsed, peve_ncRNA_miRNA_collapsed)

# Save
write.csv(peve_results, "../output/12-miRNA-epimachinery/Peve_miRNAtargets_mach.csv", row.names = FALSE)
```

### Ptuh

```{r}
# Label with source
ptuh_epi_miRNA_collapsed$type <- "epimachinery"
ptuh_ncRNA_miRNA_collapsed$type <- "ncRNA machinery"

# Ensure identical column names and order
colnames(ptuh_epi_miRNA_collapsed) == colnames(ptuh_ncRNA_miRNA_collapsed)

# Combine
ptuh_results <- rbind(ptuh_epi_miRNA_collapsed, ptuh_ncRNA_miRNA_collapsed)

# Save
write.csv(ptuh_results, "../output/12-miRNA-epimachinery/Ptuh_miRNAtargets_mach.csv", row.names = FALSE)
```


## Summary

Generally, the protein machinery that are putatively targeted by our coral miRNAs is extremely diverse, covering all major "branches" of epigenetic regulation:

#### Histone Methylation

Covering demethylases (KDM3, KDM6, RIOX1), methyltransferases (KMT5A, KMT2E), acetyltransferases (KAT6A/B, KAT14, KAT2A/B), and deacetylases (HDAC4/5/7/9, SIRT7), plus arginine methylation (PRMT7)

#### DNA Methylation & Reading

DNMT1 (maintenance methyltransferase), TET3 (active demethylation), MBD1/2/3 (methylation readers/NuRD recruitment), and PRDM14 (pluripotency-associated methylation regulator)

#### ADP-Ribosylation (form of chromatin regulation)

Tankyrases (PARP5a/b), PARP7/TIPARP, macro-PARPs (PARP9/14/15), and the eraser ARH1

#### Ubiquitin Signaling (form of protien modification)

Deubiquitinases (USPs, PSMD14, BAP1), E2/E3 ligases (UBE2D, UBE3A, CUL4A, DDB1, UBR2, RNF8, BRCA1)

#### RNA Modification (Epitranscriptomics)

tRNA modifiers (TRMT1, TRMT61A/B, PUS7L, DUS1L), demethylases (ALKBH3), m6A reader (HNRNPA2B1), and ADAR (A-to-I editing)

#### ncRNA Biogenesis & Silencing

Core RISC silencing complex (AGO1/2, TNRC6), biogenesis regulators (LIN28A, KHSRP, SRRT), Integrator complex (INTS1/4/6), RNA decay machinery (exosome, EDC4, CCR4-NOT, SMG1, PAN2), and chromatin/structural components (RBBP4, STAG1/2, SMARCA2/4, RDRP)

#### Other Regulators

Signaling kinases (PKN1, MAP3K12, MAP2K1, Chuk/IKKα), phosphatases (PP1), and protein quality control (COPS6, Hspbap1)


### Plot

```{r}
library(tidyverse)
apul_results_spec <- apul_results
apul_results_spec$species <- "A. pulchra"
peve_results_spec <- peve_results
peve_results_spec$species <- "P. evermanni"
ptuh_results_spec <- ptuh_results
ptuh_results_spec$species <- "P. tuahiniensis"

all_data <- bind_rows(apul_results_spec, peve_results_spec, ptuh_results_spec)

# function to classify genes into broad "type" of regulation
classify_gene <- function(gene_str) {
  gene <- toupper(gene_str)
  # Histone modification (methylation, acetylation, deacetylation, phosphorylation)
  histone_mod <- c("KDM3", "KDM6", "KMT5", "KMT2", "SETD8", "RIOX1", "KAT6", "KAT14", "KAT2", "HDAC", "SIRT7", "PRMT7", "PKN1", "HSPBAP1", "COPS6")
  # DNA methylation & reading
  dna_meth <- c("DNMT1", "TET3", "MBD1", "MBD2", "MBD3", "PRDM14", "MCM3AP")
  # ADP-ribosylation
  adprib <- c("PARP", "TNKS", "TIPARP", "ARH1", "ADPRH")
  # Ubiquitin signaling
  ubiquitin <- c("USP", "PSMD14", "BAP1", "UBE2D", "UBE3A", "UBE4A", "CUL4", "DDB1", "UBR2", "RNF8", "BRCA1", "VCPIP1", "JOSD")
  # RNA modification (epitranscriptomics)
  rna_mod <- c("TRMT1", "TRMT61", "ALKBH3", "PUS7L", "DUS1L", "HNRNPA2B1", "HNRNPA1", "ADAR", "RBMX")
  # ncRNA biogenesis & silencing
  ncrna <- c("AGO1", "AGO2", "TNRC6", "LIN28", "KHSRP", "SRRT", "RBBP4", "INTS", "SMARCA", "SMG1", "EXOSC", "EDC4", "DIS3L", "CNOT11", "EWSR1", "FUS(?!E)", "PAN2", "RDRP", "STAG1", "STAG2", "MTREX", "ZFC3H1")
  # Chromatin signaling
  signaling <- c("MAP3K12", "MAP2K1", "CHUK", "PPP1CB", "PPP1R7")

  categories <- list(
    "Histone modification" = histone_mod,
    "Chromatin signaling" = signaling,
    "DNA methylation & reading" = dna_meth,
    "ncRNA biogenesis & silencing" = ncrna,
    "RNA modification" = rna_mod,
    "ADP-ribosylation" = adprib,
    "Ubiquitin signaling" = ubiquitin
  )

  matched <- c()
  for (cat_name in names(categories)) {
    for (kw in categories[[cat_name]]) {
      if (grepl(kw, gene, perl = TRUE)) {
        matched <- c(matched, cat_name)
        break
      }
    }
  }
  if (length(matched) == 0) {
    return("Unclassified")
  }
  # return the first matched category
  return(matched[1])
}

# classify targets
all_data <- all_data %>% rowwise() %>% mutate(category = classify_gene(gene)) %>% ungroup()

# summarize
category_counts <- all_data %>% count(species, category, name = "n_targets")

# Set factor levels
cat_order <- c("Histone modification",
               "Chromatin signaling",
               "DNA methylation & reading",
               "ncRNA biogenesis & silencing",
               "RNA modification",
               "ADP-ribosylation",
               "Ubiquitin signaling")

species_order <- c("A. pulchra", "P. evermanni", "P. tuahiniensis")

category_counts <- category_counts %>%
  mutate(category = factor(category, levels = rev(cat_order)),
         species  = factor(species, levels = rev(species_order))
         )

# set colors
cat_colors <- c(
  "Histone modification" = "#378ADD",
  "Chromatin signaling" = "#888780",
  "DNA methylation & reading" = "#D4537E",
  "ncRNA biogenesis & silencing" = "#534AB7",
  "RNA modification" = "#D85A30",
  "ADP-ribosylation" = "#BA7517",
  "Ubiquitin signaling" = "#1D9E75"
)

# plot!
p <- ggplot(category_counts, aes(x = n_targets, y = species, fill = category)) +
  geom_bar(stat = "identity", position = "stack", width = 0.65) +
  scale_fill_manual(
    values = cat_colors,
    breaks = cat_order,
    name   = NULL
  ) +
  labs(
    x = "Number of miRNA target interactions",
    y = NULL
  ) +
  theme_minimal()

p
```
save
```{r}
ggsave("../output/12-miRNA-epimachinery/miRNA_target_categories_by_species.png", p, width = 9, height = 4.5, dpi = 300)
```
