---
title: "20-supplementary-files"
author: "Kathleen Durkin"
date: "2026-06-15"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(stringr)
```

Code for any compiling/formating of supplementary figures/tables for the DDE manuscript.


# Alignment summary tables

Compiling per-sample alignment and processing statistics for all three species

## RNA-seq

Set up the metadata
```{r}
species_meta <- tibble(
  species = c("D-Apul", "E-Peve", "F-Ptuh"),
  species_full = c("Acropora pulchra", "Porites evermanni",
                   "Pocillopora tuahiniensis"),
  genome = c("Apulcra-genome.fa", "Porites_evermanni_v1.fa",
             "Pocillopora_meandrina_HIv1.assembly.fasta"),
  rna_prefix = c("../../D-Apul", "../../E-Peve", "../../F-Ptuh")
)
species_meta
```

### 1.1 Raw reads

from MultiQC FastQC on raw reads
```{r}
raw_reads <- rbind(
  
read_tsv("../../D-Apul/output/01-Apul-RNA-trimming-FastQC/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../E-Peve/output/01-Peve-RNA-trimming-FastQC/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../F-Ptuh/output/01-Ptuh-RNA-trimming-FastQC/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences')
)

raw_reads
```

### 1.2 Trimmed reads

from MultiQC FastQC on trimmed reads
```{r}
trimmed_reads <- rbind(
  
read_tsv("../../D-Apul/output/01-Apul-RNA-trimming-FastQC/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../E-Peve/output/01-Peve-RNA-trimming-FastQC/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../F-Ptuh/output/01-Ptuh-RNA-trimming-FastQC/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences')
)

trimmed_reads
```

### 1.3 Retained after trimming (%)

```{r}
read_counts <- raw_reads %>%
  left_join(trimmed_reads, by = c("Sample")) %>%
  mutate(retained_pct = round(trimmed_reads$'Total Sequences' / raw_reads$'Total Sequences' * 100, 2))

read_counts
```

### 1.4 MISSING HISAT2 alignment stats (overall rate, unique, multi)

D-Apul has a consolidated log; E-Peve and F-Ptuh have per-sample stderr files.

```{r, eval = FALSE}
# Parse a single HISAT2 stderr block.
# Returns a one-row tibble with reads, overall_rate, unique_pct, multi_pct.
parse_hisat2_block <- function(lines) {
  text <- paste(lines, collapse = "\n")

  reads <- as.numeric(str_match(text, "^(\\d+) reads")[2])
  overall <- as.numeric(str_match(text, "(\\S+)% overall alignment rate")[2])

  # Concordant exactly 1 time  -> uniquely mapped
  unique_line <- str_match(text,
    "aligned concordantly exactly 1 time\n")[1]
  unique_pct <- as.numeric(str_extract(unique_line, "\\d+\\.\\d+"))

  # Concordant >1 times -> multi-mapped
  multi_line <- str_match(text,
    "aligned concordantly >1 times\n")[1]
  multi_pct <- as.numeric(str_extract(multi_line, "\\d+\\.\\d+"))

  tibble(
    hisat2_reads   = reads,
    overall_align  = overall,
    unique_mapped  = unique_pct,
    multi_mapped   = multi_pct
  )
}

# D-Apul: consolidated log, 5 blocks separated by blank lines
parse_hisat2_consolidated <- function(path, samples) {
  lines <- readLines(path)
  # Split into blocks at "N reads; of these:"
  block_starts <- which(str_detect(lines, "^\\d+ reads; of these:"))
  results <- list()
  for (i in seq_along(block_starts)) {
    end <- if (i < length(block_starts)) block_starts[i + 1] - 1 else length(lines)
    block <- lines[block_starts[i]:end]
    results[[i]] <- parse_hisat2_block(block) %>%
      mutate(sample = samples[i])
  }
  bind_rows(results)
}

# E-Peve / F-Ptuh: one stderr file per sample
parse_hisat2_per_file <- function(dir_path, sample_pattern, sample_replacement) {
  files <- list.files(dir_path, pattern = sample_pattern, full.names = TRUE)
  results <- list()
  for (f in files) {
    lines <- readLines(f)
    # Extract sample name from filename: RNA-POR-71_hisat.stderr -> POR-71
    sname <- basename(f) %>%
      str_remove("_hisat.stderr") %>%
      str_remove("^RNA-")
    results[[sname]] <- parse_hisat2_block(lines) %>%
      mutate(sample = sname)
  }
  bind_rows(results)
}

hisat2_stats <- bind_rows(
  parse_hisat2_consolidated(
    "../../D-Apul/output/07-Apul-Hisat/hisat.out",
    samples = c("ACR-140", "ACR-145", "ACR-150", "ACR-173", "ACR-178")
  ),
  parse_hisat2_per_file(
    "../../E-Peve/output/06-Peve-Hisat",
    pattern = "_hisat.stderr$"
  ),
  parse_hisat2_per_file(
    "../../F-Ptuh/output/06-Ptuh-Hisat",
    pattern = "_hisat.stderr$"
  )
) %>%
  mutate(species = case_when(
    str_detect(sample, "^ACR") ~ "D-Apul",
    str_detect(sample, "^POR") ~ "E-Peve",
    str_detect(sample, "^POC") ~ "F-Ptuh"
  )) %>%
  select(species, sample, everything())

hisat2_stats
```

### 1.5 MISSING Mapped reads (from featureCounts .summary)

```{r, eval = FALSE}
# featureCounts .summary: rows = Status (Assigned, Unassigned_Unmapped, etc.),
# columns = BAM paths. Mapped reads = total - Unassigned_Unmapped.
# Total reads per sample = sum of all rows.

read_featurecounts_summary <- function(path, species_code) {
  df <- read_tsv(path, show_col_types = FALSE)
  # Column names are BAM paths; extract sample names
  bam_cols <- colnames(df)[-1]  # first col is "Status"
  samples <- str_extract(bam_cols, "(ACR|POR|POC)-\\d+")

  assigned <- df %>% filter(Status == "Assigned") %>% select(-Status) %>% as.numeric()
  unmapped <- df %>% filter(Status == "Unassigned_Unmapped") %>% select(-Status) %>% as.numeric()

  # Total = sum of all status rows
  total <- colSums(df %>% select(-Status) %>% mutate(across(everything, as.numeric)))

  tibble(
    species = species_code,
    sample = samples,
    mapped_reads = total - unmapped,
    total_reads_fc = total
  )
}

mapped_reads <- bind_rows(
  read_featurecounts_summary("../../D-Apul/output/32-Apul-lncRNA-matrix/Apul-lncRNA-counts.txt.summary", "D-Apul"),
  read_featurecounts_summary("../../E-Peve/output/18-Peve-lncRNA-matrix/Peve-lncRNA-counts.txt.summary", "E-Peve"),
  read_featurecounts_summary("../../F-Ptuh/output/18-Ptuh-lncRNA-matrix/Ptuh-lncRNA-counts.txt.summary", "F-Ptuh")
)
mapped_reads
```

### 1.6 MISSING Mismatch rate (%)

```{r, eval=FALSE}
# NOTE: Mismatch rate is NOT available in this repository.
# There are no samtools stats / flagstat / idxstats or Picard outputs.
# The sorted BAMs are gitignored (available via the project wiki).
#
# To compute mismatch rate, run the following on the BAMs:
#
#   samtools stats <sample.sorted.bam> | grep "^SN" > <sample>.stats.txt
#
# Then parse the line:
#   SN	mismatch rate:	X.XXe-XX
# and convert to percentage.
#
# Placeholder:
mismatch_rate <- tibble(
  species = character(),
  sample = character(),
  mismatch_rate_pct = numeric()
)
# mismatch_rate  # leave empty — populate after running samtools stats on BAMs
```

### 1.7 Number of lncRNAs identified

```{r}
count_lncRNAs <- function(filepath) {
  
df <- read_tsv(filepath, col_names = TRUE)
sample_cols <- grep(".sorted.bam$", colnames(df), value = TRUE)
sample_names <- str_remove(sample_cols, ".*pipeline\\.RNA\\.") %>% str_remove("\\.sorted\\.bam$")

# Count non-zero entries per sample column
nonzero_counts <- sapply(sample_cols, function(col) {
  sum(df[[col]] > 0, na.rm = TRUE)
})

# Build a summary tibble
summary <- tibble(
  sample = sample_names,
  n_unique_lncRNA = nonzero_counts
)
print(summary)
}

lncRNA_counts <- rbind(
count_lncRNAs("../output/01.6-lncRNA-pipeline/Apul-lncRNA-counts-filtered.txt"),
count_lncRNAs("../output/01.6-lncRNA-pipeline/Peve-lncRNA-counts-filtered.txt"),
count_lncRNAs("../output/01.6-lncRNA-pipeline/Ptuh-lncRNA-counts-filtered.txt")
)
lncRNA_counts
```


## sRNA-seq alignment statistics

### 2.1 Species metadata (genome)

```{r}
# Genome files are referenced in the ShortStack Rmd scripts:
srna_genomes <- tibble(
  species = c("D-Apul", "E-Peve", "F-Ptuh"),
  genome = c("Apulcra-genome.fa", "Porites_evermanni_v1.fa",
             "Pocillopora_meandrina_HIv1.assembly.fasta")
)
srna_genomes
```

### 2.1 Raw reads

```{r}
s_raw_reads <- rbind(
  
read_tsv("~/deep-dive/D-Apul/output/08-Apul-sRNAseq-trimming/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("~/deep-dive/E-Peve/output/06-Peve-sRNAseq-trimming/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("~/deep-dive/F-Pmea/output/08-Pmea-sRNAseq-trimming/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences')
)

s_raw_reads$Sample <- str_remove(s_raw_reads$Sample, "_R1_001")

s_raw_reads
```

### 2.2 Trimmed reads

```{r}
s_trimmed_reads <- rbind(
  
read_tsv("~/deep-dive/D-Apul/output/08.2-Apul-sRNAseq-trimming-31bp-fastp-merged/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "sRNA-")) %>% select(Sample, 'Total Sequences'), 

read_tsv("~/deep-dive/E-Peve/output/06.2-Peve-sRNAseq-trimming-31bp-fastp-merged/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% select(Sample, 'Total Sequences') %>% mutate(Sample = paste0("sRNA-", Sample)), 

read_tsv("~/deep-dive/F-Pmea/output/08.2-Pmea-sRNAseq-trimming-31bp-fastp-merged/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "sRNA-")) %>% select(Sample, 'Total Sequences')
)

s_trimmed_reads$Sample <- str_remove(s_trimmed_reads$Sample, "-fastp-adapters-polyG-31bp-merged")

s_trimmed_reads
```

### 2.3 Retained after trimming (%)

```{r}
s_read_counts <- s_raw_reads %>%
  left_join(s_trimmed_reads, by = c("Sample")) %>%
  mutate(retained_pct = round(s_trimmed_reads$'Total Sequences' / s_raw_reads$'Total Sequences' * 100, 2))

s_read_counts
```

### 2.4 ShortStack alignment stats

from alignment_details.tsv

NOTE: the Github repo README.md states that the H tag means "H: Very highly multi-mapped read (>=50 hits)". However,the log from running ShortStack 4.1.0 (e.g., the file ~/deep-dive-expression/D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/shortstack.log) states that the tag represents "Very highly multi-mapped (>=20 hits)(H)". It looks like the >=50 threshold was used in previous versions of ShortStack (confirmed by looking at deep-dive ShortStack logs), but the updated 4.1.0 version switched to a >=20 threshold. The Github repo's README is out of date.
```{r}
# alignment_details.tsv columns:
#   readfile  mapping_type  sequence_length  sequence_count  read_count
# mapping_type (from https://github.com/MikeAxtell/ShortStack)
# U: Uniquely mapped (not a multimapper).
# P: Multimapper placed using the method set by option --mmap.
# R: Multimapper placed at random.
# H: Very highly multi-mapped read (>=20 hits).
# N: Unmapped reads.

read_alignment_details <- function(path, species_code) {
  df <- read_tsv(path, show_col_types = FALSE)

  # Extract sample name from readfile path
  df <- df %>%
    mutate(
      sample = str_extract(readfile, "(ACR|POR|POC)-\\d+"),
      species = species_code
    )

  # Aggregate by sample × mapping_type, sum sequence_count
  agg <- df %>%
    group_by(species, sample, mapping_type) %>%
    summarise(seq_count = sum(read_count), .groups = "drop")

  # Pivot to wide format
  agg_wide <- agg %>%
    pivot_wider(names_from = mapping_type, values_from = seq_count,
                values_fill = 0)

  # totals and percentages
  # U=unique, P=guided, R=random, H=very-high, N=not mapped
  agg_wide %>%
    mutate(
      total_reads = U + P + R + H + N,
      mapped_reads = U + P + R + H,
      overall_align_pct = round((mapped_reads) / total_reads * 100, 2),
      unique_pct = round(U / total_reads * 100, 2),
      guided_pct = round(P / total_reads * 100, 2),
      random_pct = round(R / total_reads * 100, 2),
      very_high_pct = round(H / total_reads * 100, 2),
    ) %>%
    select(
      species, sample, total_reads, overall_align_pct, unique_pct, guided_pct, 
      random_pct, very_high_pct, mapped_reads
    ) %>%
    rename(
      `Overall alignment (%)` = overall_align_pct,
      `Uniquely mapped reads (%)` = unique_pct,
      `Multi-mapped reads placed with guidance (%)` = guided_pct,
      `Multi-mapped reads placed randomly (%)` = random_pct,
      `Very high multi-mapping reads (>=50 hits; %)` = very_high_pct,
      `Mapped sRNA reads` = mapped_reads
    )
}

srna_align <- bind_rows(
  read_alignment_details("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/alignment_details.tsv", "D-Apul"),
  read_alignment_details("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/alignment_details.tsv", "E-Peve"),
  read_alignment_details("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/alignment_details.tsv", "F-Ptuh")
)
srna_align
```

### 2.5 Mapped miRNA reads

from ShortStack Counts.txt
```{r}
# Counts.txt is a TSV: Coords, Name, MIRNA, <sample1>_condensed, ...
# Filter rows where MIRNA == "Y", then sum each sample column.

read_mirna_counts <- function(path, species_code) {
  df <- read_tsv(path, show_col_types = FALSE)
  sample_cols <- colnames(df)[-(1:3)]  # skip Coords, Name, MIRNA

  # Extract clean sample names from column names
  # e.g. "sRNA-ACR-140-S1-TP2-..._condensed" -> "ACR-140"
  samples <- str_extract(sample_cols, "(ACR|POR|POC)-\\d+")

  mirna_only <- df %>% filter(MIRNA == "Y")

  # Sum each sample column across all miRNA rows
  mapped_mirna <- colSums(mirna_only[, sample_cols])

  tibble(
    species = species_code,
    sample = samples,
    mapped_mirna_reads = mapped_mirna
  )
}

srna_mirna_reads <- bind_rows(
  read_mirna_counts("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Counts.txt", "D-Apul"),
  read_mirna_counts("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt", "E-Peve"),
  read_mirna_counts("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt", "F-Ptuh")
)
srna_mirna_reads
```

### 2.6 Number of miRNAs identified

```{r}
count_expressed_mirnas <- function(path) {
  df <- read_tsv(path, show_col_types = FALSE)

  sample_cols <- grep("merged_condensed", colnames(df), value = TRUE)

  as.data.frame(t(df %>%
    filter(MIRNA == "Y") %>%
    summarise(across(all_of(sample_cols), ~ sum(.x > 0, na.rm = TRUE)))))
}

miRNA_counts <- rbind(
count_expressed_mirnas("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Counts.txt"),
count_expressed_mirnas("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt"),
count_expressed_mirnas("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt")
)
colnames(miRNA_counts) <- c("miRNA_count")
miRNA_counts$Sample <- rownames(miRNA_counts)
miRNA_counts <- miRNA_counts %>% select(Sample, miRNA_count) %>% remove_rownames()
miRNA_counts$Sample <- miRNA_counts$Sample %>% str_remove("-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed") %>% str_remove("sRNA-")

miRNA_counts
```











