---
title: "20-supplementary-files"
author: "Kathleen Durkin"
date: "2026-06-15"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(stringr)
library(purrr)
library(ggplot2)
library(patchwork)
```

Code for any compiling/formating of supplementary figures/tables for the DDE manuscript.

# Machinery.fasta reference table

```{r build-machinery-table}
protein_names <- c(
  # DNA methylation
  "DNMT1"  = "DNA (cytosine-5-)-methyltransferase 1",
  "DNMT3A" = "DNA (cytosine-5-)-methyltransferase 3 alpha",
  "DNMT3B" = "DNA (cytosine-5-)-methyltransferase 3 beta",
  "DNMT3L" = "DNA (cytosine-5-)-methyltransferase 3 like",
  # TET dioxygenases
  "TET1" = "Methylcytosine dioxygenase TET1",
  "TET2" = "Methylcytosine dioxygenase TET2",
  "TET3" = "Methylcytosine dioxygenase TET3",
  # Methyl-CpG binding domain
  "MBD1" = "Methyl-CpG-binding domain protein 1",
  "MBD2" = "Methyl-CpG-binding domain protein 2",
  "MBD3" = "Methyl-CpG-binding domain protein 3",
  "MBD4" = "Methyl-CpG-binding domain protein 4",
  "MBD5" = "Methyl-CpG-binding domain protein 5",
  "MBD6" = "Methyl-CpG-binding domain protein 6",
  # UHRF
  "UHRF1" = "E3 ubiquitin-protein ligase UHRF1",
  "UHRF2" = "E3 ubiquitin-protein ligase UHRF2",
  # ZBTB readers
  "ZBTB33" = "Zinc finger and BTB domain-containing protein 33 (Kaiso)",
  "ZBTB4"  = "Zinc finger and BTB domain-containing protein 4",
  "ZBTB38" = "Zinc finger and BTB domain-containing protein 38",
  "ZFP57"  = "Zinc finger protein 57",
  "KLF4"   = "Krueppel-like factor 4",
  "EGR1"   = "Early growth response protein 1",
  "WT1"    = "Wilms tumor protein",
  "CTCF"   = "CCCTC-binding factor",
  # HDACs
  "HDAC1"  = "Histone deacetylase 1",
  "HDAC2"  = "Histone deacetylase 2",
  "HDAC3"  = "Histone deacetylase 3",
  "HDAC4"  = "Histone deacetylase 4",
  "HDAC5"  = "Histone deacetylase 5",
  "HDAC6"  = "Histone deacetylase 6",
  "HDAC7"  = "Histone deacetylase 7",
  "HDAC8"  = "Histone deacetylase 8",
  "HDAC9"  = "Histone deacetylase 9",
  "HDAC10" = "Histone deacetylase 10",
  "HDAC11" = "Histone deacetylase 11",
  # Sirtuins
  "SIRT1" = "NAD-dependent protein deacetylase sirtuin-1",
  "SIRT2" = "NAD-dependent protein deacetylase sirtuin-2",
  "SIRT3" = "NAD-dependent protein deacetylase sirtuin-3",
  "SIRT4" = "NAD-dependent protein lysine acylase sirtuin-4",
  "SIRT5" = "NAD-dependent protein lysine desuccinylase sirtuin-5",
  "SIRT6" = "NAD-dependent protein deacylase sirtuin-6",
  "SIRT7" = "NAD-dependent protein desuccinylase sirtuin-7",
  # KATs
  "KAT1-HAT1"    = "Histone acetyltransferase type B catalytic subunit (HAT1/KAT1)",
  "KAT2A"        = "Histone acetyltransferase KAT2A (GCN5)",
  "KAT2B"        = "Histone acetyltransferase KAT2B (PCAF)",
  "KAT3A-CREBBP" = "CREB-binding protein (CREBBP/KAT3A)",
  "KAT3B-EP300"  = "Histone acetyltransferase p300 (EP300/KAT3B)",
  "KAT4-TAF1"    = "Transcription initiation factor TFIID subunit 1 (TAF1/KAT4)",
  "KAT5"         = "Histone acetyltransferase KAT5 (Tip60)",
  "KAT6A"        = "Histone acetyltransferase KAT6A (MOZ)",
  "KAT6B"        = "Histone acetyltransferase KAT6B (MORF)",
  "KAT7"         = "Histone acetyltransferase KAT7 (HBO1)",
  "KAT8"         = "Histone acetyltransferase KAT8 (MOF)",
  "KAT9-ELP3"    = "Elongator complex protein 3 (ELP3/KAT9)",
  "KAT12-GTF3C4" = "Transcription factor IIIC subunit 4 (GTF3C4/KAT12)",
  "KAT13A-NCOA1" = "Nuclear receptor coactivator 1 (NCOA1/KAT13A)",
  "KAT13B-NCOA3" = "Nuclear receptor coactivator 3 (NCOA3/KAT13B)",
  "KAT13C-NCOA2" = "Nuclear receptor coactivator 2 (NCOA2/KAT13C)",
  "KAT13D-CLOCK" = "Circadian locomoter output cycles protein kaput (CLOCK/KAT13D)",
  "KAT14"        = "Histone acetyltransferase KAT14 (CSRP2BP)",
  # KMTs
  "KMT1C-EHMT2"   = "Histone-lysine N-methyltransferase EHMT2 (G9a/KMT1C)",
  "KMT1D-EHMT1"   = "Histone-lysine N-methyltransferase EHMT1 (GLP/KMT1D)",
  "KMT1E-SETDB1"  = "Histone-lysine N-methyltransferase SETDB1 (KMT1E)",
  "KMT1F-SETDB2"  = "Histone-lysine N-methyltransferase SETDB2 (KMT1F)",
  "SUV39H1-KMT1A" = "Histone-lysine N-methyltransferase SUV39H1 (KMT1A)",
  "SUV39H2-KMT1B" = "Histone-lysine N-methyltransferase SUV39H2 (KMT1B)",
  "KMT2A"         = "Histone-lysine N-methyltransferase 2A (MLL1)",
  "KMT2B"         = "Histone-lysine N-methyltransferase 2B (MLL4)",
  "KMT2C"         = "Histone-lysine N-methyltransferase 2C (MLL3)",
  "KMT2D"         = "Histone-lysine N-methyltransferase 2D (MLL2)",
  "KMT2E"         = "Histone-lysine N-methyltransferase 2E (MLL5)",
  "KMT2F-SETD1A"  = "Histone-lysine N-methyltransferase SETD1A (KMT2F)",
  "KMT2G-SETD1B"  = "Histone-lysine N-methyltransferase SETD1B (KMT2G)",
  "KMT2H-ASH1L"   = "Histone-lysine N-methyltransferase ASH1L (KMT2H)",
  "KMT3A-SETD2"   = "Histone-lysine N-methyltransferase SETD2 (KMT3A)",
  "KMT3B-NSD1"    = "Histone-lysine N-methyltransferase NSD1 (KMT3B)",
  "KMT3C-SMYD2"   = "SET and MYND domain-containing protein 2 (SMYD2/KMT3C)",
  "KMT4-DOT1L"    = "Histone-lysine N-methyltransferase DOT1L (KMT4)",
  "KMT5A\u2014SETD8" = "Histone-lysine N-methyltransferase SETD8 (KMT5A)",
  "KMT5B"         = "Histone-lysine N-methyltransferase KMT5B (SUV420H1)",
  "KMT5C"         = "Histone-lysine N-methyltransferase KMT5C (SUV420H2)",
  "KMT6-EZH2"     = "Histone-lysine N-methyltransferase EZH2 (KMT6)",
  "KMT7-SETD7/9"  = "Histone-lysine N-methyltransferase SETD7 (KMT7)",
  "KMT8-PRDM2"    = "PR domain zinc finger protein 2 (PRDM2/KMT8)",
  "ASH2L"         = "Set1/Ash2 histone methyltransferase complex subunit ASH2",
  "WHSC1-NSD2"    = "Histone-lysine N-methyltransferase NSD2 (WHSC1/MMSET)",
  "WHSC1L-NSD3"   = "Histone-lysine N-methyltransferase NSD3 (WHSC1L1)",
  "EZH1"   = "Histone-lysine N-methyltransferase EZH1",
  "SETD3"  = "Histone methyltransferase SETD3",
  "SETD4"  = "SET domain-containing protein 4",
  "SETD5"  = "SET domain-containing protein 5",
  "SETD6"  = "N-lysine methyltransferase SETD6",
  "SETMAR" = "Histone-lysine N-methyltransferase SETMAR (Metnase)",
  "SMYD1"  = "SET and MYND domain-containing protein 1",
  "SMYD3"  = "SET and MYND domain-containing protein 3",
  "SMYD4"  = "SET and MYND domain-containing protein 4",
  "SMYD5"  = "SET and MYND domain-containing protein 5",
  # KDMs
  "KDM1A"         = "Lysine-specific histone demethylase 1A (LSD1)",
  "KDM1B"         = "Lysine-specific histone demethylase 1B (LSD2)",
  "KDM2A"         = "Lysine-specific demethylase 2A",
  "KDM2B"         = "Lysine-specific demethylase 2B",
  "KDM3A"         = "Lysine-specific demethylase 3A",
  "KDM3B"         = "Lysine-specific demethylase 3B",
  "KDM3C-JMJD1C"  = "Lysine-specific demethylase JMJD1C (KDM3C)",
  "KDM4A"         = "Lysine-specific demethylase 4A",
  "KDM4B"         = "Lysine-specific demethylase 4B",
  "KDM4C"         = "Lysine-specific demethylase 4C",
  "KDM4D"         = "Lysine-specific demethylase 4D",
  "KDM5A"         = "Lysine-specific demethylase 5A (JARID1A)",
  "KDM5B"         = "Lysine-specific demethylase 5B (JARID1B)",
  "KDM5C"         = "Lysine-specific demethylase 5C (JARID1C)",
  "KDM5D"         = "Lysine-specific demethylase 5D (JARID1D)",
  "KDM6A"         = "Lysine-specific demethylase 6A (UTX)",
  "KDM6B"         = "Lysine-specific demethylase 6B (JMJD3)",
  "KDM7A"         = "Lysine-specific demethylase 7A (JHDM1D/PHF8-related)",
  "KDM8"          = "Lysine-specific demethylase 8 (JMJD5)",
  "UTY"    = "Inactive histone demethylase UTY",
  "PHF2"   = "PHD finger protein 2 (histone demethylase)",
  "PHF8"   = "Histone lysine demethylase PHF8",
  "JMJD6"  = "Bifunctional arginine demethylase and lysyl-hydroxylase JMJD6",
  "JMJD7"  = "Lysine demethylase JMJD7",
  "JMJD8"  = "JMJD8 protein",
  "JARID2" = "Protein Jumonji (JARID2)",
  "HR"     = "Lysine demethylase and nuclear receptor corepressor HR",
  "RIOX1"  = "Ribosomal oxygenase 1 (NO66)",
  "RIOX2"  = "Ribosomal oxygenase 2 (MINA53)",
  "TYW5"   = "tRNA wybutosine-synthesizing protein 5",
  # PRDM
  "PRDM1"       = "PR domain zinc finger protein 1 (Blimp1)",
  "MECOM-PRDM3" = "EVI1/MDS1 and EVI1 complex locus protein (MECOM/PRDM3)",
  "PRDM4"  = "PR domain zinc finger protein 4",
  "PRDM5"  = "PR domain zinc finger protein 5",
  "PRDM6"  = "PR domain zinc finger protein 6",
  "PRDM7"  = "PR domain zinc finger protein 7",
  "PRDM8"  = "PR domain zinc finger protein 8",
  "PRDM9"  = "PR domain zinc finger protein 9",
  "PRDM10" = "PR domain zinc finger protein 10",
  "PRDM11" = "PR domain zinc finger protein 11",
  "PRDM12" = "PR domain zinc finger protein 12",
  "PRDM13" = "PR domain zinc finger protein 13",
  "PRDM14" = "PR domain zinc finger protein 14",
  "PRDM15" = "PR domain zinc finger protein 15",
  "PRDM16" = "PR domain zinc finger protein 16",
  # PRMTs
  "PRMT2" = "Protein arginine N-methyltransferase 2",
  "PRMT3" = "Protein arginine N-methyltransferase 3",
  "PRMT5" = "Protein arginine N-methyltransferase 5",
  "PRMT6" = "Protein arginine N-methyltransferase 6",
  "PRMT7" = "Protein arginine N-methyltransferase 7",
  "PRMT8" = "Protein arginine N-methyltransferase 8",
  "PRMT9" = "Protein arginine N-methyltransferase 9",
  # PARPs
  "PARP2"          = "Poly [ADP-ribose] polymerase 2",
  "PARP3"          = "Poly [ADP-ribose] polymerase 3",
  "PARP4"          = "Poly [ADP-ribose] polymerase 4",
  "PARP5A-TNKS"    = "Tankyrase-1 (TNKS/PARP5A)",
  "PARP5B-TNKS2"   = "Tankyrase-2 (TNKS2/PARP5B)",
  "PARP6"          = "Poly [ADP-ribose] polymerase 6",
  "PARP7-TIPARP"   = "TCDD-inducible poly [ADP-ribose] polymerase (TIPARP/PARP7)",
  "PARP8"          = "Poly [ADP-ribose] polymerase 8",
  "PARP9"          = "Poly [ADP-ribose] polymerase 9 (BAL1)",
  "PARP10"         = "Poly [ADP-ribose] polymerase 10",
  "PARP11"         = "Poly [ADP-ribose] polymerase 11",
  "PARP12"         = "Poly [ADP-ribose] polymerase 12",
  "PARP13-ZC3HAV1" = "Zinc finger CCCH-type antiviral protein 1 (ZAP/PARP13)",
  "PARP14"         = "Poly [ADP-ribose] polymerase 14 (BAL2)",
  "PARP15"         = "Poly [ADP-ribose] polymerase 15 (BAL3)",
  "PARP16"         = "Poly [ADP-ribose] polymerase 16",
  "PARG"           = "Poly(ADP-ribose) glycohydrolase",
  "ARH1-ADPRH"     = "ADP-ribosylhydrolase ARH1",
  "ARH3-ADPRS"     = "ADP-ribosylhydrolase ARH3",
  "OGT"            = "UDP-N-acetylglucosamine--peptide N-acetylglucosaminyltransferase (OGT)",
  # SUMO
  "SUMO2" = "Small ubiquitin-related modifier 2",
  "SUMO3" = "Small ubiquitin-related modifier 3",
  "SENP1" = "Sentrin-specific protease 1",
  "SENP2" = "Sentrin-specific protease 2",
  "SENP3" = "Sentrin-specific protease 3",
  "SENP5" = "Sentrin-specific protease 5",
  "SENP6" = "Sentrin-specific protease 6",
  "SENP7" = "Sentrin-specific protease 7",
  "SENP8" = "Sentrin-specific protease 8 (NEDD8 deSUMOylation)",
  # Ubiquitin E1/E2/E3
  "UBA1"   = "Ubiquitin-activating enzyme E1",
  "UBA2"   = "SUMO-activating enzyme subunit 2",
  "UBA3"   = "NEDD8-activating enzyme E1 catalytic subunit",
  "UBA5"   = "Ubiquitin-like modifier-activating enzyme 5",
  "UBA6"   = "Ubiquitin-activating enzyme E1-like protein",
  "UBA7"   = "Ubiquitin-activating enzyme E1-like protein 2 (UBE1L)",
  "UBE2A"  = "Ubiquitin-conjugating enzyme E2 A",
  "UBE2D1" = "Ubiquitin-conjugating enzyme E2 D1",
  "UBE2D2" = "Ubiquitin-conjugating enzyme E2 D2",
  "UBE2D3" = "Ubiquitin-conjugating enzyme E2 D3",
  "UBE2V1" = "Ubiquitin-conjugating enzyme E2 variant 1",
  "UBE2Z"  = "Ubiquitin-conjugating enzyme E2 Z",
  "UBE3A"  = "Ubiquitin-protein ligase E3A",
  "UBE4A"  = "Ubiquitin conjugation factor E4 A",
  "UBR2"   = "E3 ubiquitin-protein ligase UBR2",
  "BRCA1"  = "Breast cancer type 1 susceptibility protein (BRCA1)",
  "HUWE1"  = "E3 ubiquitin-protein ligase HUWE1",
  "RNF8"   = "E3 ubiquitin-protein ligase RNF8",
  "RNF20"  = "E3 ubiquitin-protein ligase BRE1A (RNF20)",
  "RNF40"  = "E3 ubiquitin-protein ligase BRE1B (RNF40)",
  "RNF168" = "E3 ubiquitin-protein ligase RNF168",
  "TRIM37" = "E3 ubiquitin-protein ligase TRIM37",
  "DTX3L"  = "Deltex E3 ubiquitin ligase 3L (DTX3L)",
  "CUL4A"  = "Cullin-4A",
  "DDB1"   = "DNA damage-binding protein 1",
  "TOM1"   = "Target of Myb protein 1 (TOM1)",
  # USPs
  "USP1"   = "Ubiquitin carboxyl-terminal hydrolase 1",
  "USP2"   = "Ubiquitin carboxyl-terminal hydrolase 2",
  "USP3"   = "Ubiquitin carboxyl-terminal hydrolase 3",
  "USP4"   = "Ubiquitin carboxyl-terminal hydrolase 4",
  "USP5"   = "Ubiquitin carboxyl-terminal hydrolase 5",
  "USP6"   = "Ubiquitin carboxyl-terminal hydrolase 6",
  "USP7"   = "Ubiquitin carboxyl-terminal hydrolase 7 (HAUSP)",
  "USP8"   = "Ubiquitin carboxyl-terminal hydrolase 8",
  "USP9Y"  = "Ubiquitin carboxyl-terminal hydrolase 9, Y-linked",
  "USP10"  = "Ubiquitin carboxyl-terminal hydrolase 10",
  "USP11"  = "Ubiquitin carboxyl-terminal hydrolase 11",
  "USP12"  = "Ubiquitin carboxyl-terminal hydrolase 12",
  "USP13"  = "Ubiquitin carboxyl-terminal hydrolase 13",
  "USP14"  = "Ubiquitin carboxyl-terminal hydrolase 14",
  "USP15"  = "Ubiquitin carboxyl-terminal hydrolase 15",
  "USP16"  = "Ubiquitin carboxyl-terminal hydrolase 16",
  "USP18"  = "Ubiquitin carboxyl-terminal hydrolase 18",
  "USP19"  = "Ubiquitin carboxyl-terminal hydrolase 19",
  "USP20"  = "Ubiquitin carboxyl-terminal hydrolase 20",
  "USP21"  = "Ubiquitin carboxyl-terminal hydrolase 21",
  "USP22"  = "Ubiquitin carboxyl-terminal hydrolase 22",
  "USP24"  = "Ubiquitin carboxyl-terminal hydrolase 24",
  "USP25"  = "Ubiquitin carboxyl-terminal hydrolase 25",
  "USP26"  = "Ubiquitin carboxyl-terminal hydrolase 26",
  "USP27X" = "Ubiquitin carboxyl-terminal hydrolase 27, X-linked",
  "USP28"  = "Ubiquitin carboxyl-terminal hydrolase 28",
  "USP29"  = "Ubiquitin carboxyl-terminal hydrolase 29",
  "USP30"  = "Ubiquitin carboxyl-terminal hydrolase 30",
  "USP31"  = "Ubiquitin carboxyl-terminal hydrolase 31",
  "USP32"  = "Ubiquitin carboxyl-terminal hydrolase 32",
  "USP33"  = "Ubiquitin carboxyl-terminal hydrolase 33",
  "USP34"  = "Ubiquitin carboxyl-terminal hydrolase 34",
  "USP35"  = "Ubiquitin carboxyl-terminal hydrolase 35",
  "USP36"  = "Ubiquitin carboxyl-terminal hydrolase 36",
  "USP37"  = "Ubiquitin carboxyl-terminal hydrolase 37",
  "USP38"  = "Ubiquitin carboxyl-terminal hydrolase 38",
  "USP39"  = "Ubiquitin carboxyl-terminal hydrolase 39",
  "USP40"  = "Ubiquitin carboxyl-terminal hydrolase 40",
  "USP41"  = "Ubiquitin carboxyl-terminal hydrolase 41",
  "USP42"  = "Ubiquitin carboxyl-terminal hydrolase 42",
  "USP43"  = "Ubiquitin carboxyl-terminal hydrolase 43",
  "USP44"  = "Ubiquitin carboxyl-terminal hydrolase 44",
  "USP45"  = "Ubiquitin carboxyl-terminal hydrolase 45",
  "USP46"  = "Ubiquitin carboxyl-terminal hydrolase 46",
  "USP47"  = "Ubiquitin carboxyl-terminal hydrolase 47",
  "USP48"  = "Ubiquitin carboxyl-terminal hydrolase 48",
  "USP49"  = "Ubiquitin carboxyl-terminal hydrolase 49",
  "USP50"  = "Ubiquitin carboxyl-terminal hydrolase 50",
  "USP51"  = "Ubiquitin carboxyl-terminal hydrolase 51",
  "USP53"  = "Ubiquitin carboxyl-terminal hydrolase 53",
  "USP54"  = "Ubiquitin carboxyl-terminal hydrolase 54",
  "PAN2"   = "Poly(A)-specific ribonuclease subunit PAN2 (USP52)",
  # UCH / OTU / other DUBs
  "UCHL1"   = "Ubiquitin carboxyl-terminal hydrolase isozyme L1",
  "UCHL3"   = "Ubiquitin carboxyl-terminal hydrolase isozyme L3",
  "UCHL4"   = "Ubiquitin carboxyl-terminal hydrolase isozyme L4",
  "UCHL5"   = "Ubiquitin carboxyl-terminal hydrolase isozyme L5 (UCH37)",
  "OTUD1"   = "OTU domain-containing protein 1",
  "OTUD3"   = "OTU domain-containing protein 3",
  "OTUD4"   = "OTU domain-containing protein 4",
  "OTUD5"   = "OTU domain-containing protein 5 (DUBA)",
  "OTUD6A"  = "OTU domain-containing protein 6A",
  "OTUD6B"  = "OTU domain-containing protein 6B",
  "OTUD7A"  = "OTU domain-containing protein 7A (CEZANNE2)",
  "OTUD7B"  = "OTU domain-containing protein 7B (CEZANNE)",
  "BAP1"    = "Ubiquitin carboxyl-terminal hydrolase BAP1",
  "BRCC3"   = "Lys-63-specific deubiquitinase BRCC3",
  "MYSM1"   = "Histone H2A deubiquitinase MYSM1",
  "ATXN3"   = "Ataxin-3 (Josephin domain deubiquitinase)",
  "ATXN3L"  = "Ataxin-3-like protein",
  "JOSD1"   = "Josephin domain-containing protein 1",
  "JOSD2"   = "Josephin domain-containing protein 2",
  "PSMD14"  = "26S proteasome non-ATPase regulatory subunit 14 (POH1/Rpn11)",
  "PSMD7"   = "26S proteasome non-ATPase regulatory subunit 7 (Rpn8)",
  "COPS5"   = "COP9 signalosome complex subunit 5 (JAB1/CSN5)",
  "COPS6"   = "COP9 signalosome complex subunit 6 (CSN6)",
  "STAMBPL1"= "Stam-binding protein-like 1 (AMSH-LP)",
  "ZRANB1"  = "Zinc finger RANBP2-type and RING finger-containing protein 1 (TRABID)",
  "VCPIP1"  = "Deubiquitinating protein VCPIP1",
  "EIF3H"   = "Eukaryotic translation initiation factor 3 subunit H",
  "EIF3F"   = "Eukaryotic translation initiation factor 3 subunit F",
  "MPND"    = "MPN domain-containing protein",
  "PRPF8"   = "Pre-mRNA-processing-splicing factor 8 (PRPF8/SNRNP220)",
  "ARID1A"  = "AT-rich interactive domain-containing protein 1A",
  "ARID1B"  = "AT-rich interactive domain-containing protein 1B",
  # ADAR
  "ADAR"   = "Double-stranded RNA-specific adenosine deaminase (ADAR1)",
  "ADARB1" = "Double-stranded RNA-specific editase B1 (ADAR2)",
  "ADARB2" = "Double-stranded RNA-specific editase B2 (ADAR3)",
  # m6A writers / readers / erasers
  "METTL14"   = "N6-adenosine-methyltransferase non-catalytic subunit (METTL14)",
  "METTL16"   = "RNA N6-methyladenosine methyltransferase METTL16",
  "METTL1"    = "tRNA (guanine(46)-N(7))-methyltransferase METTL1",
  "METTL2A"   = "tRNA (cytosine(34)-N(4))-methyltransferase METTL2A",
  "METTL2B"   = "tRNA (cytosine(34)-N(4))-methyltransferase METTL2B",
  "METTL4"    = "N6-adenosine-methyltransferase METTL4",
  "METTL5"    = "rRNA adenine N(6)-methyltransferase (METTL5)",
  "METTL6"    = "Methyltransferase-like protein 6",
  "METTL8"    = "Methyltransferase-like protein 8",
  "WTAP"      = "pre-mRNA-splicing regulator WTAP (m6A writer complex)",
  "VIRMA"     = "Vir-like m6A methyltransferase associated (KIAA1429)",
  "CBLL1"     = "Hakai (CBLL1); E3 ubiquitin ligase, m6A writer complex",
  "ZC3H13"    = "Zinc finger CCCH domain-containing protein 13 (m6A writer complex)",
  "RBM15"     = "RNA-binding protein 15",
  "RBM15B"    = "RNA-binding protein 15B",
  "PCIF1"     = "mRNA cap-specific m6Am methyltransferase PCIF1",
  "FTO"       = "Alpha-ketoglutarate-dependent dioxygenase FTO (m6A eraser)",
  "ALKBH5"    = "mRNA demethylase ALKBH5 (m6A eraser)",
  "ALKBH1"    = "tRNA N(1)-methyl adenine demethylase ALKBH1",
  "ALKBH3"    = "tRNA demethylase ALKBH3",
  "YTHDC1"    = "YTH domain-containing protein 1 (m6A nuclear reader)",
  "YTHDC2"    = "YTH domain-containing protein 2 (m6A reader/helicase)",
  "YTHDF1"    = "YTH domain-containing family protein 1 (m6A cytoplasmic reader)",
  "YTHDF2"    = "YTH domain-containing family protein 2 (m6A decay reader)",
  "YTHDF3"    = "YTH domain-containing family protein 3 (m6A reader)",
  "HNRNPC"    = "Heterogeneous nuclear ribonucleoprotein C",
  "HNRNPA2B1" = "Heterogeneous nuclear ribonucleoproteins A2/B1",
  "RBMX"      = "RNA-binding motif protein, X-linked (HNRNPG)",
  "IGF2BP1"   = "Insulin-like growth factor 2 mRNA-binding protein 1",
  "IGF2BP2"   = "Insulin-like growth factor 2 mRNA-binding protein 2",
  "IGF2BP3"   = "Insulin-like growth factor 2 mRNA-binding protein 3",
  "FMR1"      = "Fragile X mental retardation protein 1",
  "PRRC2A"    = "Proline-rich coiled-coil-containing protein 2A (m6A reader)",
  "ELF3"      = "ETS-related transcription factor ELF3",
  # RNA methylation (non-m6A)
  "NSUN2"   = "tRNA (cytosine(34)-C(5))-methyltransferase NSUN2",
  "NSUN3"   = "Methyltransferase-like protein NSUN3 (mitochondrial tRNA)",
  "NSUN4"   = "Bifunctional methyltransferase/rRNA maturation factor NSUN4",
  "NSUN5"   = "rRNA methyltransferase NSUN5",
  "NSUN6"   = "tRNA methyltransferase NSUN6",
  "TRDMT1"  = "tRNA (cytosine-5-)-methyltransferase TRDMT1 (DNMT2)",
  "NOP2"    = "Ribosome biogenesis protein NOP2 (NSUN1)",
  "DKC1"    = "H/ACA ribonucleoprotein complex subunit 4 (Dyskerin/DKC1)",
  "TRMT1"   = "tRNA (guanine(26)-N(2))-methyltransferase TRMT1",
  "TRMT61A" = "tRNA (adenine(58)-N(1))-methyltransferase TRMT61A",
  "TRMT61B" = "rRNA (adenine(58)-N(1))-methyltransferase TRMT61B",
  "TRMT10A" = "tRNA (guanine(9)-N(1))-methyltransferase TRMT10A",
  "TRMT10C" = "Mitochondrial tRNA methyltransferase TRMT10C",
  "TRMT112" = "tRNA methyltransferase subunit TRMT112",
  "ALYREF"  = "THO complex subunit ALYREF (Aly/REF export factor)",
  "ZCCHC4"  = "Zinc finger CCHC domain-containing protein 4 (rRNA m6A writer)",
  "NAT10"   = "N-acetyltransferase 10 (RNA cytidine acetyltransferase)",
  "DUS1L"   = "tRNA-dihydrouridine synthase 1-like",
  # Pseudouridine synthases
  "PUSL1"  = "Pseudouridine synthase 1-like",
  "PUS1"   = "tRNA pseudouridine synthase A (PUS1)",
  "PUS3"   = "tRNA pseudouridine synthase C (PUS3)",
  "PUS7"   = "tRNA pseudouridine synthase G (PUS7)",
  "PUS7L"  = "tRNA pseudouridine synthase 7-like",
  "PUS10"  = "tRNA pseudouridine synthase 10 (RPUSD2-related)",
  "RPUSD1" = "RNA pseudouridylate synthase domain-containing 1",
  "RPUSD2" = "RNA pseudouridylate synthase domain-containing 2",
  "RPUSD3" = "RNA pseudouridylate synthase domain-containing 3",
  "RPUSD4" = "RNA pseudouridylate synthase domain-containing 4",
  "TRUB1"  = "Probable tRNA pseudouridine synthase 1 (TRUB1)",
  "TRUB2"  = "Probable tRNA pseudouridine synthase 2 (TRUB2)",
  # NATs / acetyltransferases
  "NAA10"   = "N-alpha-acetyltransferase 10 (NatA catalytic subunit)",
  "NAA50"   = "N-alpha-acetyltransferase 50 (NatE catalytic subunit)",
  "NAA60"   = "N-alpha-acetyltransferase 60 (NatF catalytic subunit)",
  "NAT8"    = "N-acetyltransferase 8 (camello-like 1)",
  "NAT8B"   = "N-acetyltransferase 8B (camello-like 2)",
  "NAT8L"   = "N-acetyltransferase 8-like",
  "ATAT1"   = "Alpha-tubulin acetyltransferase 1",
  "GTF3C2"  = "Transcription factor IIIC 110 kDa subunit (TFIIIC110/GTF3C2)",
  "GTF3C1"  = "Transcription factor IIIC 220 kDa subunit (TFIIIC220/GTF3C1)",
  "DLAT"    = "Dihydrolipoyllysine-residue acetyltransferase (PDC-E2)",
  "ATF2"    = "Cyclic AMP-dependent transcription factor ATF-2",
  "MCM3AP"  = "Germinal-center associated nuclear protein (GANP/MCM3AP)",
  "MAPT"    = "Microtubule-associated protein tau (MAPT)",
  "BRD4"    = "Bromodomain-containing protein 4",
  "ESCO1"   = "N-acetyltransferase ESCO1 (establishment of cohesion)",
  "ESCO2"   = "N-acetyltransferase ESCO2",
  "BLOC1S1" = "Biogenesis of lysosome-related organelles complex 1 subunit 1",
  "ACAT1"   = "Acetyl-CoA acetyltransferase 1 (mitochondrial)",
  "ACAT2"   = "Acetyl-CoA acetyltransferase 2 (cytoplasmic)",
  # Histone variants / linker histones
  "H1-0"  = "Histone H1.0",
  "H1-1"  = "Histone H1.1 (HIST1H1A)",
  "H1-2"  = "Histone H1.2 (HIST1H1C)",
  "H1-3"  = "Histone H1.3 (HIST1H1D)",
  "H1-4"  = "Histone H1.4 (HIST1H1E)",
  "H1-5"  = "Histone H1.5 (HIST1H1B)",
  "H1-6"  = "Histone H1t",
  "H1-7"  = "Histone H1oo (oocyte-specific)",
  "H1-8"  = "Histone H1.8 (H1foo)",
  "H1-10" = "Histone H1.10",
  "H2AX"  = "Histone H2A.x",
  "H2AZ1" = "Histone H2A.z.1",
  "H2AZ2" = "Histone H2A.z.2",
  "H2AW"  = "Histone H2A.W",
  "H2AP"  = "Histone H2A.P",
  "H2AJ"  = "Histone H2A.J",
  "H2AB1" = "Histone H2A-Bbd type 1 (H2A.Bbd)",
  "H2AB2" = "Histone H2A-Bbd type 2",
  "H2AB3" = "Histone H2A-Bbd type 3",
  "MACROH2A1" = "Core histone macro-H2A.1 (macroH2A1/H2AFY)",
  # Chromatin regulators
  "BAZ1B" = "Bromodomain adjacent to zinc finger domain protein 1B (WSTF)",
  "BAZ2B" = "Bromodomain adjacent to zinc finger domain protein 2B",
  "HIRA"  = "Histone cell cycle regulator (HIRA)",
  "WDR5"  = "WD repeat-containing protein 5 (H3K4 methyltransferase complex subunit)",
  # EYA phosphatases
  "EYA1" = "Eyes absent homolog 1 (H2AX phosphatase)",
  "EYA2" = "Eyes absent homolog 2",
  "EYA3" = "Eyes absent homolog 3",
  "EYA4" = "Eyes absent homolog 4",
  # Protein phosphatases
  "PPM1D"    = "Protein phosphatase 1D (WIP1)",
  "PPP6C"    = "Serine/threonine-protein phosphatase 6 catalytic subunit",
  "PPP4C"    = "Serine/threonine-protein phosphatase 4 catalytic subunit",
  "CDCA2"    = "Cell division cycle associated 2 (Repo-Man; PP1 targeting)",
  "PPP1CA"   = "Serine/threonine-protein phosphatase PP1-alpha catalytic subunit",
  "PPP1CB"   = "Serine/threonine-protein phosphatase PP1-beta catalytic subunit",
  "PPP1CC"   = "Serine/threonine-protein phosphatase PP1-gamma catalytic subunit",
  "PPP1R1B"  = "Protein phosphatase 1 regulatory subunit 1B (DARPP-32)",
  "PPP1R1C"  = "Protein phosphatase 1 regulatory subunit 1C (I-3)",
  "PPP1R2"   = "Protein phosphatase inhibitor 2",
  "PPP1R3A"  = "Protein phosphatase 1 regulatory subunit 3A (GM)",
  "PPP1R3B"  = "Protein phosphatase 1 regulatory subunit 3B (GL)",
  "PPP1R3C"  = "Protein phosphatase 1 regulatory subunit 3C (PTG)",
  "PPP1R3D"  = "Protein phosphatase 1 regulatory subunit 3D (R6)",
  "PPP1R3E"  = "Protein phosphatase 1 regulatory subunit 3E",
  "PPP1R3F"  = "Protein phosphatase 1 regulatory subunit 3F",
  "PPP1R3G"  = "Protein phosphatase 1 regulatory subunit 3G",
  "PPP1R7"   = "Protein phosphatase 1 regulatory subunit 7",
  "PPP1R8"   = "Protein phosphatase 1 regulatory subunit 8 (NIPP1)",
  "PPP1R9A"  = "Protein phosphatase 1 regulatory subunit 9A (neurabin-1)",
  "PPP1R9B"  = "Protein phosphatase 1 regulatory subunit 9B (neurabin-2/spinophilin)",
  "PPP1R10"  = "Protein phosphatase 1 regulatory subunit 10 (PNUTS)",
  "PPP1R11"  = "Protein phosphatase 1 regulatory subunit 11 (PNUTS-like)",
  "PPP1R12A" = "Protein phosphatase 1 regulatory subunit 12A (MYPT1)",
  "PPP1R12B" = "Protein phosphatase 1 regulatory subunit 12B (MYPT2)",
  "PPP1R12C" = "Protein phosphatase 1 regulatory subunit 12C (MBS85)",
  "PPP1R13B" = "Protein phosphatase 1 regulatory subunit 13B (ASPP1)",
  "PPP1R14A" = "Protein phosphatase 1 regulatory subunit 14A (CPI-17)",
  "PPP1R14B" = "Protein phosphatase 1 regulatory subunit 14B (PHI-1)",
  "PPP1R14C" = "Protein phosphatase 1 regulatory subunit 14C",
  "PPP1R14D" = "Protein phosphatase 1 regulatory subunit 14D",
  "PPP1R15B" = "Protein phosphatase 1 regulatory subunit 15B (GADD34-related)",
  "PPP1R16A" = "Protein phosphatase 1 regulatory subunit 16A (MYPT3)",
  "PPP1R16B" = "Protein phosphatase 1 regulatory subunit 16B (TA-MYPT)",
  # Kinases
  "AURKB"   = "Aurora kinase B",
  "AURKA"   = "Aurora kinase A",
  "AURKC"   = "Aurora kinase C",
  "JAK2"    = "Tyrosine-protein kinase JAK2",
  "HASPIN"  = "Serine/threonine-protein kinase haspin (GSG2)",
  "BUB1"    = "Mitotic checkpoint serine/threonine-protein kinase BUB1",
  "WEE1"    = "Wee1-like protein kinase",
  "SGO2"    = "Shugoshin-like 2 (SGO2)",
  "SGO1"    = "Shugoshin-like 1 (SGO1)",
  "DLK1"    = "Protein delta homolog 1 (Preadipocyte factor 1/Pref-1)",
  "DAPK3"   = "Death-associated protein kinase 3 (ZIPK)",
  "PKN1"    = "Serine/threonine-protein kinase N1 (PKN1/PRK1)",
  "PKM"     = "Pyruvate kinase PKM",
  "STK4"    = "Serine/threonine-protein kinase 4 (MST1/STK4)",
  "MST1"    = "Hepatocyte growth factor-like protein (MST1)",
  "SLK"     = "STE20-like serine/threonine-protein kinase SLK",
  "PRKAA1"  = "5'-AMP-activated protein kinase catalytic subunit alpha-1 (AMPKa1)",
  "PRKAA2"  = "5'-AMP-activated protein kinase catalytic subunit alpha-2 (AMPKa2)",
  "DBF4"    = "Protein DBF4 homolog A (Cdc7 kinase regulatory subunit)",
  "DYRK1A"  = "Dual specificity tyrosine-phosphorylation-regulated kinase 1A",
  "RPS6KA4" = "Ribosomal protein S6 kinase alpha-4 (MSK2)",
  "RPS6KA5" = "Ribosomal protein S6 kinase alpha-5 (MSK1)",
  "PRKCB"   = "Protein kinase C beta type (PKCb)",
  "JKAMP"   = "JMJD8-associated regulatory complex component (JKAMP)",
  "MAPK8"   = "Mitogen-activated protein kinase 8 (JNK1)",
  "MAPK9"   = "Mitogen-activated protein kinase 9 (JNK2)",
  "MDC1"    = "Mediator of DNA damage checkpoint protein 1",
  "MCPH1"   = "Microcephalin (MCPH1; DNA damage sensor)",
  "PAK2"    = "Serine/threonine-protein kinase PAK2",
  "ATM"     = "Serine-protein kinase ATM (DNA damage checkpoint)",
  "ATR"     = "Serine/threonine-protein kinase ATR",
  "CSNK2B"  = "Casein kinase II subunit beta",
  "PRKCD"   = "Protein kinase C delta type",
  "MAP2K1"  = "Dual specificity mitogen-activated protein kinase kinase 1 (MEK1)",
  "MAP3K12" = "Mitogen-activated protein kinase kinase kinase 12 (DLK/MUK)",
  "MAP3K22" = "Mitogen-activated protein kinase kinase kinase 22",
  "MAPK1"   = "Mitogen-activated protein kinase 1 (ERK2)",
  "MAPK3"   = "Mitogen-activated protein kinase 3 (ERK1)",
  "MAPK8B"  = "Mitogen-activated protein kinase 8b (JNK1b, zebrafish)",
  "MAPK11"  = "Mitogen-activated protein kinase 11 (p38b)",
  "MAPK12"  = "Mitogen-activated protein kinase 12 (p38g/ERK6)",
  "MAPK13"  = "Mitogen-activated protein kinase 13 (p38d/SAPK4)",
  "MAPK14"  = "Mitogen-activated protein kinase 14 (p38a)",
  "PRKDC"   = "DNA-dependent protein kinase catalytic subunit (DNA-PKcs)",
  "PIM1"    = "Proto-oncogene serine/threonine-protein kinase Pim-1",
  "CDK8"    = "Cyclin-dependent kinase 8 (Mediator kinase)",
  "CHUK"    = "Inhibitor of nuclear factor kappa-B kinase subunit alpha (IKKa)",
  "CDK1"    = "Cyclin-dependent kinase 1 (CDC2)",
  "CHEK1"   = "Serine/threonine-protein kinase Chk1",
  "APBB1"   = "Amyloid precursor protein-binding family B member 1 (Fe65)",
  "YWHAQ"   = "14-3-3 protein theta",
  # ZC3H12 ribonucleases
  "ZC3H12A" = "Endoribonuclease ZC3H12A (MCPIP1/Regnase-1)",
  "ZC3H12B" = "Endoribonuclease ZC3H12B (Regnase-2)",
  "ZC3H12C" = "Endoribonuclease ZC3H12C (Regnase-3)",
  "ZC3H12D" = "Endoribonuclease ZC3H12D (Regnase-4)",
  # Other
  "DCP2"   = "mRNA-decapping enzyme subunit DCP2",
  "YBX1"   = "Nuclease-sensitive element-binding protein 1 (YBX1/YB-1)",
  "RAG1"   = "V(D)J recombination-activating protein 1 (RAG1)",
  "ALG13"  = "UDP-N-acetylglucosamine transferase subunit ALG13",
  "HSPBAP1"= "HSPA binding protein 1 (HSPBAP1)"
)

strip_isoform <- function(label) {
  # Remove trailing isoform number, e.g. Dnmt1-201 -> Dnmt1
  sub("-\\d+$", "", label)
}

species_from_id <- function(eid) {
  dplyr::case_when(
    startsWith(eid, "ENSMUSP") ~ "Mus musculus",
    startsWith(eid, "ENSDARP") ~ "Danio rerio",
    TRUE                       ~ "Homo sapiens"
  )
}

lookup_protein_name <- function(gene_name) {
  key <- toupper(gene_name)
  if (!is.na(protein_names[key])) return(unname(protein_names[key]))
  # Strip alias suffixes and try progressively shorter hyphen-joined prefixes
  simplified <- gsub("[\u2014\u2013/].*", "", key)
  parts <- strsplit(simplified, "-")[[1]]
  for (i in seq(length(parts), 1)) {
    candidate <- paste(parts[seq_len(i)], collapse = "-")
    if (!is.na(protein_names[candidate])) return(unname(protein_names[candidate]))
  }
  ""
}

# Parse fasta and build table
lines   <- readLines("../../data/Machinery.fasta")
headers <- lines[startsWith(lines, ">")]
parsed    <- regmatches(headers, regexec("^>(\\S+)\\s+peptide:\\s+(\\S+)", headers))
label     <- sapply(parsed, `[`, 2)   # e.g. "Dnmt1-201"
eid       <- sapply(parsed, `[`, 3)   # e.g. "ENSMUSP00000004202"
gene_name     <- strip_isoform(label)
protein_name  <- sapply(gene_name, lookup_protein_name, USE.NAMES = FALSE)

machinery_ref <- data.frame(
  accession     = eid,
  protein_name  = protein_name,
  gene_name     = gene_name,
  species       = species_from_id(eid),
  stringsAsFactors = FALSE
)

write.csv(machinery_ref, "../output/20-supplementary-files/Machinery_reference_table.csv", row.names = FALSE)
```


# Epimachinery category labels for Machinery.fasta table

```{r}
machinery <- read_csv("../output/20-supplementary-files/Machinery_reference_table.csv",
                      show_col_types = FALSE)

classify_gene <- function(gene) {
  # Uppercase to handle mouse mixed-case gene names
  gene <- toupper(gene)

  cats <- list(
    "ADP-ribosylation" = list(
      prefix = c("PARP", "TNKS", "TARG", "MACROD", "OARD", "ADPRH", "ARH", "TIPARP"),
      exact  = c("PARG", "ARTD1")
    ),
    "DNA methylation & reading" = list(
      prefix = c("DNMT", "TET", "MBD", "UHRF", "MECP",
                 "ZBTB"),
      exact  = c("PRDM14",
                 "ZFP57")
    ),
    "RNA modification" = list(
      prefix = c("METTL", "YTHDF", "YTHDC", "ALKBH", "WTAP", "VIRMA", "ZC3H13",
                 "IGF2BP", "NSUN", "DKC", "TRMT", "PUS", "NAT10", "ADAR",
                 "TRUB", "RPUSD"),
      exact  = c("FTO", "RBM15", "RBM15B", "PCIF1", "BUD23", "WDR4",
                 "RBMX", "TRDMT1", "NOP2", "HNRNPA2B1",
                 "DUS1L",   # tRNA dihydrouridine synthase
                 "TYW5",    # tRNA wybutosine-synthesizing protein (Supp. folder:
                            # "Histone demethylation" — incorrect; TYW5 is a
                            # tRNA modification enzyme, not a histone demethylase)
                 "CBLL1")   # Hakai, E3 ubiquitin ligase component of the m6A
                            # writer complex
    ),
    "ncRNA biogenesis & silencing" = list(
      prefix = c("AGO", "TNRC6", "DICER", "DROSHA", "DGCR8", "PIWI", "EXPORTIN",
                 "TARBP", "PRKRA", "LIN28", "KHSRP", "ZCCHC", "DIS3", "INTS", "EXOSC",
                 "PAN2", "PAN3", "DDX5", "DDX17", "DDX6", "EDC", "DCP", "XRN", "CNOT",
                 "UPF", "SMG", "TUT", "MOV10", "FMR", "PUM", "STAU", "NCBP", "PHAX",
                 "RBM7",
                 "ZC3H12"),  # ZC3H12A-D are endoribonucleases (Regnase-1-4)
                            # that degrade target mRNAs in the cytoplasm —
                            # mRNA turnover/silencing machinery
      exact  = c("RDRP", "MTREX", "SKIV2L2", "MTR4", "ZFC3H1", "SRRT", "ARS2", "XPO5",
                 "XPO1", "CRM1", "RAN", "EWSR1", "FUS", "TARDBP", "HNRNPK", "PARN",
                 "PABPC1", "PAPD5", "DZIP3", "NONO", "SFPQ", "PSPC1",
                 "YBX1", "HSPA8", "HSP90AA1", "HSP90AB1", "PQBP1", "MCM3AP",
                 "PRPF8",  # pre-mRNA splicing factor 8 — core spliceosome
                           # component involved in snRNA processing
                 "EIF3H",  # eIF3 subunit H — translation initiation factor
                           # involved in mRNA silencing/regulation
                 "EIF3F")  # eIF3 subunit F — translation initiation factor;
                           # has MPN deubiquitinase fold but functions in
                           # translation regulation and miRNA-mediated silencing
    ),
    "Histone modification and variants" = list(
      prefix = c("KMT", "KDM", "SETD", "EZH", "SUV", "EHMT", "MLL", "ASH1", "DOT1",
                 "JMJD", "HDAC", "SIRT", "KAT", "PRMT", "CARM", "NSD", "WHSC", "SMYD",
                 "CDYL", "RIOX", "MINA", "NO66",
                 "PRDM",
                 "SUMO", "SENP",
                 "H1-", "H2A", "H2B", "H3", "H4",  # histone variants
                 "NAT"),  # NAT8, NAT8B, NAT8L — N-acetyltransferases (camello
                         # family); Supp. folder groups all NATs under
                         # "Histone acetylation"
      exact  = c("CREBBP", "EP300", "HAT1", "ELP3", "PHF8", "PHF2", "JARID2", "PKN1",
                 "HSPBAP1",
                 "WDR5", "ASH2L", "SETMAR",
                 "OGT", "OGA",
                 "MACROH2A1",  # histone variant macro-H2A.1
                 "ATAT1",      # alpha-tubulin acetyltransferase; Supp. folder:
                               # "Histone acetylation" (grouped with KATs)
                 "GTF3C1", "GTF3C2",  # TFIIIC subunits; Supp. folder:
                                      # "Histone acetylation" (have HAT activity)
                 "NAA10", "NAA50", "NAA60",  # N-alpha-acetyltransferases;
                                             # Supp. folder: "Histone acetylation"
                 "BRD4",       # bromodomain-containing protein 4; reads
                               # acetylated histones; Supp. folder:
                               # "Histone acetylation"
                 "ESCO1", "ESCO2",  # N-acetyltransferases (cohesin establishment);
                                   # Supp. folder: "Histone acetylation"
                 "DLAT",       # dihydrolipoyllysine-residue acetyltransferase;
                               # Supp. folder: "Histone acetylation"
                 "ATF2",       # cyclic AMP-dependent TF; histone acetyltransferase
                               # activity; Supp. folder: "Histone acetylation"
                 "BLOC1S1",    # biogenesis of lysosome-related organelles;
                               # GCN5L1; Supp. folder: "Histone acetylation"
                 "ACAT1", "ACAT2",  # acetyl-CoA acetyltransferases; Supp. folder:
                                   # "Histone acetylation"
                 "UTY",        # inactive histone demethylase (KDM6 family homolog)
                 "HR",         # lysine demethylase and nuclear receptor corepressor
                 "MAPT",       # microtubule-associated protein tau; has histone
                               # H3 binding / chromatin roles
                 "BAZ2B")      # bromodomain adjacent to zinc finger; chromatin
                               # remodeling reader (Supp. folder: not present,
                               # but BAZ1B is in "Histone phosphorylation")
    ),
    "Ubiquitin signaling" = list(
      prefix = c("USP", "UBE", "UBR", "UBA", "UCHL", "UCH", "RNF", "TRIM", "HUWE",
                 "HERC", "MARCH", "RBX", "CUL", "FBX", "SKP", "SOCS", "OTU", "UBQLN",
                 "PSMD", "JOSD", "DDB", "MIB"),
      exact  = c("VCPIP1", "BAP1", "ATXN3",
                 "ATXN3L",    # ataxin-3-like protein; Josephin domain DUB
                 "COPS6", "COPS5", "PARK2", "ZRANB1",
                 "BRCA1", "RAG1", "TOM1", "DTX3L",
                 "BRCC3",     # Lys-63-specific deubiquitinase; Supp. folder:
                              # "Histone Ubiquitination" (Deubiquitination)
                 "MYSM1",     # histone H2A deubiquitinase; Supp. folder:
                              # "Histone Ubiquitination" (Deubiquitination)
                 "STAMBPL1",  # AMSH-LP; deubiquitinase; Supp. folder:
                              # "Histone Ubiquitination" (Deubiquitination)
                 "MPND",      # MPN domain-containing protein; DUB fold;
                              # associated with ubiquitin signaling
                 "ALG13")     # UDP-N-acetylglucosamine transferase subunit;
                              # has deubiquitinase-like fold; Supp. folder:
                              # not present, but classified here per its
                              # role in ubiquitin-related pathways
    ),
    "Chromatin signaling" = list(
      prefix = c("MAP3K", "MAP2K", "MAPK",
                 "EYA",
                 "PPP4", "PPP6"),
      exact  = c("CHUK", "IKBKB", "IKBKG",
                 "PPP1CB", "PPP1CA", "PPP1CC", "PPP1R7",
                 "PPP1R10", "PPP1R8",
                 "PPP2CA", "AURKB", "MSK1", "MSK2", "RPS6KA4", "RPS6KA5", "PARK7",
                 "BUB1", "DLK1", "DAPK3", "PKM", "JAK2", "DBF4", "DYRK1A",
                 "YWHAQ", "PRKCB", "PRKCD", "AURKA", "AURKC", "STK4", "MST1",
                 "MDC1", "MCPH1", "PAK2", "HIRA", "ATM", "ATR", "CSNK2B",
                 "PRKAA1", "PRKAA2", "PRKDC", "PIM1", "CDK8", "CDK1", "CHEK1",
                 "PPM1D", "PPP6C", "PPP4C", "CDCA2",
                 "APBB1", "BAZ1B", "JKAMP", "ARID1A", "ARID1B",
                 "SLK",       # STE20-like kinase; Supp. folder:
                              # "Histone phosphorylation" (slkb)
                 "WEE1",      # Wee1-like kinase; Supp. folder:
                              # "Histone phosphorylation"
                 "HASPIN",    # histone H3T3 kinase (GSG2); chromatin signaling
                 "SGO1", "SGO2")  # Shugoshin proteins; protect centromeric
                                  # cohesion by recruiting PP2A; chromatin-
                                  # associated signaling at kinetochores
    )
  )

  for (cat_name in names(cats)) {
    cat_def <- cats[[cat_name]]
    if (gene %in% cat_def$exact) return(cat_name)
    for (p in cat_def$prefix) {
      if (str_starts(gene, fixed(p))) return(cat_name)
    }
  }
  return("Unclassified")
}

# Apply classification and add category column
machinery_categorized <- machinery %>%
  mutate(epimachinery_category = vapply(gene_name, classify_gene, character(1)))

machinery_categorized

write_csv(machinery_categorized, "../output/20-supplementary-files/Machinery_categorized.csv")
```


# Alignment summary tables

Compiling per-sample alignment and processing statistics for all three species

## RNA-seq

Set up the metadata
```{r}
species_meta <- tibble(
  species = c("D-Apul", "E-Peve", "F-Ptuh"),
  species_full = c("Acropora pulchra", "Porites evermanni",
                   "Pocillopora tuahiniensis"),
  genome = c("Apulcra-genome.fa", "Porites_evermanni_v1.fa",
             "Pocillopora_meandrina_HIv1.assembly.fasta"),
  rna_prefix = c("../../D-Apul", "../../E-Peve", "../../F-Ptuh")
)
species_meta
```

### 1.1 Raw reads

from MultiQC FastQC on raw reads
```{r}
raw_reads <- rbind(
  
read_tsv("../../D-Apul/output/01-Apul-RNA-trimming-FastQC/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../E-Peve/output/01-Peve-RNA-trimming-FastQC/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../F-Ptuh/output/01-Ptuh-RNA-trimming-FastQC/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences')
)

raw_reads
```

### 1.2 Trimmed reads

from MultiQC FastQC on trimmed reads
```{r}
trimmed_reads <- rbind(
  
read_tsv("../../D-Apul/output/01-Apul-RNA-trimming-FastQC/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../E-Peve/output/01-Peve-RNA-trimming-FastQC/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("../../F-Ptuh/output/01-Ptuh-RNA-trimming-FastQC/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences')
)

trimmed_reads
```

### 1.3 Retained after trimming (%)

```{r}
read_counts <- raw_reads %>%
  left_join(trimmed_reads, by = c("Sample")) %>%
  mutate(retained_pct = round(trimmed_reads$'Total Sequences' / raw_reads$'Total Sequences' * 100, 2))

read_counts
```


### 1.4 Alignment rate, unique %, multi %

```{r}
parse_hisat2_log <- function(lines) {
  text <- paste(lines, collapse = "\n")

  total_pairs <- as.numeric(str_match(text, "^(\\d+) reads")[2])
  total_reads <- total_pairs * 2

  # Overall alignment rate
  overall_rate <- as.numeric(str_match(text, "(\\S+)% overall alignment rate")[2])

  # Concordant pairs 
  conc_0  <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned concordantly 0 times")[2])
  conc_1  <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned concordantly exactly 1 time")[2])
  conc_gt1 <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned concordantly >1 times")[2])

  # Discordant pairs (from concordant-0-times)
  disc_1 <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned discordantly 1 time")[2])
  disc_1 <- if (is.na(disc_1)) 0 else disc_1

  # Individual mates (from pairs that aligned 0 times concordantly AND discordantly)
  mates_total <- as.numeric(str_match(text, "(\\d+) mates make up the pairs")[2])
  mates_total <- if (is.na(mates_total)) 0 else mates_total

  mates_0  <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned 0 times\n")[2])
  mates_0  <- if (is.na(mates_0)) 0 else mates_0

  mates_1  <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned exactly 1 time\n")[2])
  mates_1  <- if (is.na(mates_1)) 0 else mates_1

  mates_gt1 <- as.numeric(str_match(text, "(\\d+) \\(\\S+%\\) aligned >1 times\n")[2])
  mates_gt1 <- if (is.na(mates_gt1)) 0 else mates_gt1

  # Compute READ-level unique/multi (combining all alignment levels)
  # Each concordant pair = 2 reads; each discordant pair = 2 reads
  unique_reads <- (conc_1 * 2) + (disc_1 * 2) + mates_1
  multi_reads  <- (conc_gt1 * 2) + mates_gt1
  unmapped_reads <- mates_0
  mapped_reads <- unique_reads + multi_reads
  mapped_read_pairs <- mapped_reads/2

  # Sanity check: unique + multi + unmapped should ≈ total_reads
  # (may not be exact due to mates from discordant pairs that also aligned individually)

  tibble(
    total_pairs = total_pairs,
    total_reads = total_reads,
    overall_align_rate = overall_rate,
    unique_pct = round(unique_reads / total_reads * 100, 2),
    multi_pct = round(multi_reads / total_reads * 100, 2),
    mapped_read_pairs = mapped_read_pairs
  )
}

# D-Apul: consolidated log (all 5 samples in one file)
parse_hisat2_consolidated <- function(path, samples) {
  lines <- readLines(path)
  block_starts <- which(str_detect(lines, "^\\d+ reads; of these:"))
  results <- list()
  for (i in seq_along(block_starts)) {
    end <- if (i < length(block_starts)) block_starts[i + 1] - 1 else length(lines)
    block <- lines[block_starts[i]:end]
    results[[i]] <- parse_hisat2_log(block) %>%
      mutate(sample = samples[i])
  }
  bind_rows(results)
}

# E-Peve / F-Ptuh: one stderr file per sample
parse_hisat2_per_file <- function(dir_path) {
  files <- list.files(dir_path, pattern = "_hisat\\.stderr$", full.names = TRUE)
  results <- list()
  for (f in files) {
    sname <- basename(f) %>%
      str_remove("_hisat.stderr") %>%
      str_remove("^RNA-")
    results[[sname]] <- parse_hisat2_log(readLines(f)) %>%
      mutate(sample = sname)
  }
  bind_rows(results)
}

hisat2_stats <- bind_rows(
  parse_hisat2_consolidated(
    "../../D-Apul/output/07-Apul-Hisat/hisat.out",
    samples = c("ACR-140", "ACR-145", "ACR-150", "ACR-173", "ACR-178")
  ),
  parse_hisat2_per_file("../../E-Peve/output/06-Peve-Hisat"),
  parse_hisat2_per_file("../../F-Ptuh/output/06-Ptuh-Hisat")
) %>%
  mutate(species = case_when(
    str_detect(sample, "^ACR") ~ "D-Apul",
    str_detect(sample, "^POR") ~ "E-Peve",
    str_detect(sample, "^POC") ~ "F-Ptuh"
  )) %>%
  select(species, sample, everything())

hisat2_stats
```

### 1.5 Mismatch %

This requires collecting info from the full *.sorted.bam files, which are very large and live on Gannet (https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/).

```{r, engine='bash', eval=FALSE}

# The files are backed up to Steven's backup directory: 
#     /volume/web/seashell/bu-github/deep-dive-expression

# The BAM paths are:
#   D-Apul: D-Apul/output/07-Apul-Hisat/RNA-ACR-{140,145,150,173,178}.sorted.bam
#   E-Peve: E-Peve/output/06-Peve-Hisat/RNA-POR-{71,73,76,79,82}.sorted.bam
#   F-Ptuh: F-Ptuh/output/06-Ptuh-Hisat/RNA-POC-{47,48,50,53,57}.sorted.bam

mkdir -p ../output/20-supplementary-files/RNAseq_HISAT_samtools_stats

# D-Apul BAMs
wget -r -l1 -nd -nc -A "*.sorted.bam" \
  -P ../output/20-supplementary-files/RNAseq_HISAT_samtools_stats \
  https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/D-Apul/output/07-Apul-Hisat/

# E-Peve BAMs
wget -r -l1 -nd -nc -A "*.sorted.bam" \
  -P ../output/20-supplementary-files/RNAseq_HISAT_samtools_stats \
  https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/E-Peve/output/06-Peve-Hisat/

# F-Ptuh BAMs
wget -r -l1 -nd -nc -A "*.sorted.bam" \
  -P ../output/20-supplementary-files/RNAseq_HISAT_samtools_stats \
  https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/F-Ptuh/output/06-Ptuh-Hisat/

# Then run samtools on each BAM
for bam in ../output/20-supplementary-files/RNAseq_HISAT_samtools_stats/*.sorted.bam; do
  base=$(basename "$bam" .sorted.bam)
  /home/shared/samtools-1.12/samtools stats "$bam" > "../output/20-supplementary-files/RNAseq_HISAT_samtools_stats/${base}.stats.txt"
  /home/shared/samtools-1.12/samtools flagstat "$bam" > "../output/20-supplementary-files/RNAseq_HISAT_samtools_stats/${base}.flagstat.txt"
done
```

```{r, engine='bash', eval=FALSE}
# Remove BAMs once done (they're quite large)
rm ../output/20-supplementary-files/RNAseq_HISAT_samtools_stats/*.sorted.bam
```

Parse samtools stats output for mismatch rate
```{r}

samtools_dir <- "../output/20-supplementary-files/RNAseq_HISAT_samtools_stats"

# Sample configuration
samples_cfg <- list(
  D_Apul = c("ACR-140", "ACR-145", "ACR-150", "ACR-173", "ACR-178"),
  E_Peve = c("POR-71", "POR-73", "POR-76", "POR-79", "POR-82"),
  F_Ptuh = c("POC-47", "POC-48", "POC-50", "POC-53", "POC-57")
)
species_map <- c(D_Apul = "D-Apul", E_Peve = "E-Peve", F_Ptuh = "F-Ptuh")

# Parse a single samtools stats file
parse_samtools_stats <- function(path, sample_name, species_code) {
  if (!file.exists(path)) {
    return(tibble(species = species_code, sample = sample_name,
                  mapped_reads = NA, mismatch_rate_pct = NA))
  }
  lines <- readLines(path)

  # Mapped reads: SN\treads mapped:\t28412345
  mapped <- str_match(lines[grepl("^SN\\treads mapped:", lines)],
                      "^SN\\treads mapped:\\t(\\d+)")[2]
  mapped <- if (!is.na(mapped)) as.numeric(mapped) else NA

  # Mismatch rate (reported as "error rate"):
  # SN\terror rate:\t7.151265e-03\t# mismatches / bases mapped (cigar)
  error_line <- lines[grepl("^SN\\terror rate:", lines)]
  error_rate <- str_match(error_line, "^SN\\terror rate:\\t([\\d.eE+-]+)")[2]
  mismatch_rate_pct <- if (!is.na(error_rate)) as.numeric(error_rate) * 100 else NA

  tibble(species = species_code, sample = sample_name,
         mapped_reads = mapped,
         mismatch_rate_pct = round(mismatch_rate_pct, 4))
}

# Build results from single directory
samtools_results <- bind_rows(lapply(names(samples_cfg), function(sp) {
  species_code <- species_map[sp]
  bind_rows(lapply(samples_cfg[[sp]], function(s) {
    parse_samtools_stats(
      file.path(samtools_dir, paste0("RNA-", s, ".stats.txt")),
      s, species_code
    )
  }))
}))

# Rename for final output
samtools_summary <- samtools_results %>%
  rename(
    Species = species,
    Sample = sample,
    `Mapped reads` = mapped_reads,
    `Mismatch rate (%)` = mismatch_rate_pct
  )

samtools_summary
```



### 1.6 Number of lncRNAs identified

```{r}
count_lncRNAs <- function(filepath) {
  
df <- read_tsv(filepath, col_names = TRUE)
sample_cols <- grep(".sorted.bam$", colnames(df), value = TRUE)
sample_names <- str_remove(sample_cols, ".*pipeline\\.RNA\\.") %>% str_remove("\\.sorted\\.bam$")

# Count non-zero entries per sample column
nonzero_counts <- sapply(sample_cols, function(col) {
  sum(df[[col]] > 0, na.rm = TRUE)
})

# Build a summary tibble
summary <- tibble(
  sample = sample_names,
  n_unique_lncRNA = nonzero_counts
)
print(summary)
}

lncRNA_counts <- rbind(
count_lncRNAs("../output/01.6-lncRNA-pipeline/Apul-lncRNA-counts-filtered.txt"),
count_lncRNAs("../output/01.6-lncRNA-pipeline/Peve-lncRNA-counts-filtered.txt"),
count_lncRNAs("../output/01.6-lncRNA-pipeline/Ptuh-lncRNA-counts-filtered.txt")
)
lncRNA_counts
```


## sRNA-seq alignment statistics

### 2.1 Species metadata (genome)

```{r}
# Genome files are referenced in the ShortStack Rmd scripts:
srna_genomes <- tibble(
  species = c("D-Apul", "E-Peve", "F-Ptuh"),
  genome = c("Apulcra-genome.fa", "Porites_evermanni_v1.fa",
             "Pocillopora_meandrina_HIv1.assembly.fasta")
)
srna_genomes
```

### 2.1 Raw reads

```{r}
s_raw_reads <- rbind(
  
read_tsv("~/deep-dive/D-Apul/output/08-Apul-sRNAseq-trimming/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("~/deep-dive/E-Peve/output/06-Peve-sRNAseq-trimming/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences'), 

read_tsv("~/deep-dive/F-Pmea/output/08-Pmea-sRNAseq-trimming/raw-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "RNA-")) %>% filter(str_detect(Sample, "_R1_001$")) %>% select(Sample, 'Total Sequences')
)

s_raw_reads$Sample <- str_remove(s_raw_reads$Sample, "_R1_001")

s_raw_reads
```

### 2.2 Trimmed reads

```{r}
s_trimmed_reads <- rbind(
  
read_tsv("~/deep-dive/D-Apul/output/08.2-Apul-sRNAseq-trimming-31bp-fastp-merged/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "sRNA-")) %>% select(Sample, 'Total Sequences'), 

read_tsv("~/deep-dive/E-Peve/output/06.2-Peve-sRNAseq-trimming-31bp-fastp-merged/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% select(Sample, 'Total Sequences') %>% mutate(Sample = paste0("sRNA-", Sample)), 

read_tsv("~/deep-dive/F-Pmea/output/08.2-Pmea-sRNAseq-trimming-31bp-fastp-merged/trimmed-fastqc/multiqc_data/multiqc_fastqc.txt", col_names = TRUE) %>% filter(str_detect(Sample, "sRNA-")) %>% select(Sample, 'Total Sequences')
)

s_trimmed_reads$Sample <- str_remove(s_trimmed_reads$Sample, "-fastp-adapters-polyG-31bp-merged")

s_trimmed_reads
```

### 2.3 Retained after trimming (%)

```{r}
s_read_counts <- s_raw_reads %>%
  left_join(s_trimmed_reads, by = c("Sample")) %>%
  mutate(retained_pct = round(s_trimmed_reads$'Total Sequences' / s_raw_reads$'Total Sequences' * 100, 2))

s_read_counts
```

### 2.4 ShortStack alignment stats

from alignment_details.tsv

NOTE: the Github repo README.md states that the H tag means "H: Very highly multi-mapped read (>=50 hits)". However,the log from running ShortStack 4.1.0 (e.g., the file ~/deep-dive-expression/D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/shortstack.log) states that the tag represents "Very highly multi-mapped (>=20 hits)(H)". It looks like the >=50 threshold was used in previous versions of ShortStack (confirmed by looking at deep-dive ShortStack logs), but the updated 4.1.0 version switched to a >=20 threshold. The Github repo's README is out of date.
```{r}
# alignment_details.tsv columns:
#   readfile  mapping_type  sequence_length  sequence_count  read_count
# mapping_type (from https://github.com/MikeAxtell/ShortStack)
# U: Uniquely mapped (not a multimapper).
# P: Multimapper placed using the method set by option --mmap.
# R: Multimapper placed at random.
# H: Very highly multi-mapped read (>=20 hits).
# N: Unmapped reads.

read_alignment_details <- function(path, species_code) {
  df <- read_tsv(path, show_col_types = FALSE)

  # Extract sample name from readfile path
  df <- df %>%
    mutate(
      sample = str_extract(readfile, "(ACR|POR|POC)-\\d+"),
      species = species_code
    )

  # Aggregate by sample × mapping_type, sum sequence_count
  agg <- df %>%
    group_by(species, sample, mapping_type) %>%
    summarise(seq_count = sum(read_count), .groups = "drop")

  # Pivot to wide format
  agg_wide <- agg %>%
    pivot_wider(names_from = mapping_type, values_from = seq_count,
                values_fill = 0)

  # totals and percentages
  # U=unique, P=guided, R=random, H=very-high, N=not mapped
  agg_wide %>%
    mutate(
      total_reads = U + P + R + H + N,
      mapped_reads = U + P + R + H,
      overall_align_pct = round((mapped_reads) / total_reads * 100, 2),
      unique_pct = round(U / total_reads * 100, 2),
      guided_pct = round(P / total_reads * 100, 2),
      random_pct = round(R / total_reads * 100, 2),
      very_high_pct = round(H / total_reads * 100, 2),
    ) %>%
    select(
      species, sample, total_reads, overall_align_pct, unique_pct, guided_pct, 
      random_pct, very_high_pct, mapped_reads
    ) %>%
    rename(
      `Overall alignment (%)` = overall_align_pct,
      `Uniquely mapped reads (%)` = unique_pct,
      `Multi-mapped reads placed with guidance (%)` = guided_pct,
      `Multi-mapped reads placed randomly (%)` = random_pct,
      `Very high multi-mapping reads (>=50 hits; %)` = very_high_pct,
      `Mapped sRNA reads` = mapped_reads
    )
}

srna_align <- bind_rows(
  read_alignment_details("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/alignment_details.tsv", "D-Apul"),
  read_alignment_details("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/alignment_details.tsv", "E-Peve"),
  read_alignment_details("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/alignment_details.tsv", "F-Ptuh")
)
srna_align
```

### 2.5 Mapped miRNA reads

from ShortStack Counts.txt
```{r}
# Counts.txt is a TSV: Coords, Name, MIRNA, <sample1>_condensed, ...
# Filter rows where MIRNA == "Y", then sum each sample column.

read_mirna_counts <- function(path, species_code) {
  df <- read_tsv(path, show_col_types = FALSE)
  sample_cols <- colnames(df)[-(1:3)]  # skip Coords, Name, MIRNA

  # Extract clean sample names from column names
  # e.g. "sRNA-ACR-140-S1-TP2-..._condensed" -> "ACR-140"
  samples <- str_extract(sample_cols, "(ACR|POR|POC)-\\d+")

  mirna_only <- df %>% filter(MIRNA == "Y")

  # Sum each sample column across all miRNA rows
  mapped_mirna <- colSums(mirna_only[, sample_cols])

  tibble(
    species = species_code,
    sample = samples,
    mapped_mirna_reads = mapped_mirna
  )
}

srna_mirna_reads <- bind_rows(
  read_mirna_counts("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Counts.txt", "D-Apul"),
  read_mirna_counts("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt", "E-Peve"),
  read_mirna_counts("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt", "F-Ptuh")
)
srna_mirna_reads
```

### 2.6 Number of miRNAs identified

```{r}
count_expressed_mirnas <- function(path) {
  df <- read_tsv(path, show_col_types = FALSE)

  sample_cols <- grep("merged_condensed", colnames(df), value = TRUE)

  as.data.frame(t(df %>%
    filter(MIRNA == "Y") %>%
    summarise(across(all_of(sample_cols), ~ sum(.x > 0, na.rm = TRUE)))))
}

miRNA_counts <- rbind(
count_expressed_mirnas("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Counts.txt"),
count_expressed_mirnas("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt"),
count_expressed_mirnas("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Counts.txt")
)
colnames(miRNA_counts) <- c("miRNA_count")
miRNA_counts$Sample <- rownames(miRNA_counts)
miRNA_counts <- miRNA_counts %>% select(Sample, miRNA_count) %>% remove_rownames()
miRNA_counts$Sample <- miRNA_counts$Sample %>% str_remove("-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed") %>% str_remove("sRNA-")

miRNA_counts
```




# Mature miRNA names and sequences for all species

Already have per-species, just need to combine into single table
```{r}

miRNA_details <- rbind(
  read.csv("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Apul_Results_mature_named_miRNAs.csv", header = TRUE) %>% dplyr::select(-X),
  # Including an NA filter here, because it looks like the Peve file has an extra row indicating miRNAs identified in 
  # Ashey et al 2026 which were not also identified here using the new version of ShortStack (4.1.0)
  read.csv("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv", header = TRUE) %>% dplyr::select(-X) %>% drop_na(Name),
  read.csv("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Ptuh_Results_mature_named_miRNAs.csv", header = TRUE) %>% dplyr::select(-X)
)

miRNA_details

write.csv(miRNA_details, "../output/20-supplementary-files/AllSpecies_Results_mature_named_miRNAs.csv", col.names = TRUE, row.names = FALSE)
```



# miRanda binding information + PCC coexpression (miRNA-mRNA and miRNA-lncRNA)

## miRNA-mRNA

```{r}

consolidate_regions <- function(path_3UTR, path_CDS, path_5UTR){
# Load for each region
data_3UTR <- read.csv(path_3UTR) %>% dplyr::select(-X, -X.1)
data_CDS <- read.csv(path_CDS) %>% dplyr::select(-X)
data_5UTR <- read.csv(path_5UTR) %>% dplyr::select(-X)

# Format and combine into pooled targets
data_3UTR$region <- "3UTR"

colnames(data_CDS) <- c("miRNA", "mRNA_coord", "score", "energy", "query_start", "query_end", "subject_start", "subject_end", "total_bp_shared", "query_similar", "subject_similar", "mRNA", "PCC.cor", "p_value", "adjusted_p_value")
data_CDS$query_start_end <- paste0(data_CDS$query_start, " ", data_CDS$query_end)
data_CDS$subject_start_end <- paste0(data_CDS$subject_start, " ", data_CDS$subject_end)
data_CDS$region <- "CDS"
data_CDS <- data_CDS %>% dplyr::select(colnames(data_3UTR))

colnames(data_5UTR) <- c("miRNA", "mRNA_coord", "score", "energy", "query_start", "query_end", "subject_start", "subject_end", "total_bp_shared", "query_similar", "subject_similar", "mRNA", "PCC.cor", "p_value", "adjusted_p_value")
data_5UTR$query_start_end <- paste0(data_5UTR$query_start, " ", data_5UTR$query_end)
data_5UTR$subject_start_end <- paste0(data_5UTR$subject_start, " ", data_5UTR$subject_end)
data_5UTR$region <- "5UTR"
data_5UTR <- data_5UTR %>% dplyr::select(colnames(data_3UTR))

# Confirm all 3 region tables have same column names and orders
ifelse(identical(colnames(data_3UTR), colnames(data_CDS)) == TRUE,
       ifelse(identical(colnames(data_3UTR), colnames(data_5UTR)) == TRUE,
              data <- rbind(data_3UTR, data_CDS, data_5UTR),
              data <- NULL)
)
data
}

apul_miRanda_PCC_mRNA <- consolidate_regions(
  "../../D-Apul/output/09-Apul-mRNA-miRNA-interactions/miranda_PCC_miRNA_mRNA.csv",
  "../../D-Apul/output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv",
  "../../D-Apul/output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv")

peve_miRanda_PCC_mRNA <- consolidate_regions(
  "../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miranda_PCC_miRNA_mRNA.csv",
  "../../E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv",
  "../../E-Peve/output/10.01-Peve-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv"
)

ptuh_miRanda_PCC_mRNA <- consolidate_regions(
  "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/three_prime_interaction/Ptuh-miranda_PCC_miRNA_mRNA.csv",
  "../../F-Ptuh/output/11.01-Ptuh-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv",
  "../../F-Ptuh/output/11.01-Ptuh-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_5UTR.csv"
)

# Combine
miRanda_PCC_mRNA <- rbind(apul_miRanda_PCC_mRNA, peve_miRanda_PCC_mRNA, ptuh_miRanda_PCC_mRNA)

# Filter to only significant PCC (this consitutes our putative interaction set, which is both 
# predicted to bind via miRanda and significantly coexpressed)
miRanda_PCC_mRNA_sig <- miRanda_PCC_mRNA %>% filter(p_value < 0.05)

# Save
write.csv(miRanda_PCC_mRNA_sig, "../output/20-supplementary-files/miRanda_PCC_mRNA_sig.csv", row.names = FALSE)
write.csv(miRanda_PCC_mRNA, "../output/20-supplementary-files/miRanda_PCC_mRNA.csv", row.names = FALSE)
```

## miRNA-lncRNA

```{r}

apul_miRanda_PCC_lncRNA <- read.csv("../../D-Apul/output/28-Apul-miRNA-lncRNA-interactions/miranda_PCC_miRNA_lncRNA.csv") %>% dplyr::select(-X, -X.1)
peve_miRanda_PCC_lncRNA <- read.csv("../../E-Peve/output/15-Peve-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv") %>% dplyr::select(-X, -X.1)
ptuh_miRanda_PCC_lncRNA <- read.csv("../../F-Ptuh/output/15-Ptuh-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv") %>% dplyr::select(-X, -X.1)

# Ensure all three tables have matching column names and order
identical(colnames(apul_miRanda_PCC_lncRNA), colnames(peve_miRanda_PCC_lncRNA))
identical(colnames(apul_miRanda_PCC_lncRNA), colnames(ptuh_miRanda_PCC_lncRNA))

# Combine
miRanda_PCC_lncRNA <- rbind(apul_miRanda_PCC_lncRNA, peve_miRanda_PCC_lncRNA, ptuh_miRanda_PCC_lncRNA)

# Reorder columns to have same order as miRNA-mRNA table (above)
miRanda_PCC_lncRNA <- miRanda_PCC_lncRNA %>% dplyr::select(miRNA, lncRNA, PCC.cor, p_value, adjusted_p_value,
                                                           score, energy, query_start_end, subject_start_end, 
                                                           total_bp_shared, query_similar, subject_similar)

# Filter to only significant PCC (this constitutes our putative interaction set, which is both 
# predicted to bind via miRanda and significantly coexpressed)
miRanda_PCC_lncRNA_sig <- miRanda_PCC_lncRNA %>% filter(p_value < 0.05)

# Save
write.csv(miRanda_PCC_lncRNA_sig, "../output/20-supplementary-files/miRanda_PCC_lncRNA_sig.csv", row.names = FALSE)
write.csv(miRanda_PCC_lncRNA, "../output/20-supplementary-files/miRanda_PCC_lncRNA.csv", row.names = FALSE)
```


# Epigenetic landscape summary figure

Multipanel figure summarizing the epigenetic landscape across the three species: (A) miRNA counts, (B) lncRNA counts, and (C) DNA methylation distribution by genomic feature.

For panels A–B, counts and conservation are derived from the committed FASTA and BLAST files in `output/04-miRNA-comparison/` and the lncRNA FASTAs from each species' output. For panel C, mean methylation per genomic feature is taken from the per-species WGBS analyses (rendered `.md` summary tables in `D-Apul/code/08-Apul-WGBS.md`, `E-Peve/code/12-Peve-WGBS.md`, and `F-Ptuh/code/12-Ptuh-WGBS.md`), since the full annotated CpG files are not committed to the repo.

```{r epigenetic-landscape-figure, fig.width=14, fig.height=6}

species_order <- c("Apul", "Peve", "Ptuh")
species_pal <- c(Apul = "#408EC6", Peve = "#1E2761", Ptuh = "#7A2048")
species_labels <- c(
  Apul = "A. pulchra",
  Peve = "P. evermanni",
  Ptuh = "P. tuahiniensis"
)

# miRNA counts and conservation
n_mirna_apul <- sum(grepl("^>", readLines("../output/04-miRNA-comparison/Apul_ShortStack_mature.fasta")))
n_mirna_peve <- sum(grepl("^>", readLines("../output/04-miRNA-comparison/Peve_ShortStack_mature.fasta")))
n_mirna_ptuh <- sum(grepl("^>", readLines("../output/04-miRNA-comparison/Ptuh_ShortStack_mature.fasta")))

mirna_summary <- tibble(
  species = factor(species_order, levels = species_order),
  total = c(n_mirna_apul, n_mirna_peve, n_mirna_ptuh)
)

panel_a <- ggplot(mirna_summary, aes(x = species, y = total, fill = species)) +
  geom_col(color = "white", linewidth = 0.3) +
  geom_text(aes(label = total), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = species_pal, guide = "none") +
  scale_x_discrete(labels = species_labels) +
  labs(title = "A. miRNA counts", x = NULL, y = "Number of miRNAs") +
  theme(axis.text.x = element_text(face = "italic"),
        legend.position = "none")


# lncRNA counts
n_lncrna_apul <- sum(grepl("^>", readLines("../../D-Apul/output/31-Apul-lncRNA/Apul-lncRNA.fasta")))
n_lncrna_peve <- sum(grepl("^>", readLines("../../E-Peve/output/17-Peve-lncRNA/Peve-lncRNA.fasta")))
n_lncrna_ptuh <- sum(grepl("^>", readLines("../../F-Ptuh/output/17-Ptuh-lncRNA/Ptuh-lncRNA.fasta")))

lncrna_summary <- tibble(
  species = factor(species_order, levels = species_order),
  total = c(n_lncrna_apul, n_lncrna_peve, n_lncrna_ptuh)
)

panel_b <- ggplot(lncrna_summary, aes(x = species, y = total, fill = species)) +
  geom_col(color = "white", linewidth = 0.3) +
  geom_text(aes(label = total), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = species_pal, guide = "none") +
  scale_x_discrete(labels = species_labels) +
  labs(title = "B. lncRNA counts", x = NULL, y = "Number of lncRNAs") +
  theme(axis.text.x = element_text(face = "italic"),
        legend.position = "none")


# DNA methylation distribution by genomic feature

# NOTE: Dfs for each species' mean methylation per genomic feature weren't saved to repo,
# so I'm pulling these values from the rendered .md files for each per-species analysis,
# (the summary tables *were* output in the rendered files):

# Apul: deep-dive-expression/D-Apul/code/08-Apul-WGBS.md
#    region        mean_meth
#  1 intergenic         9.19
#  2 TE_intergenic      9.96
#  3 TE_exonic         13.1 
#  4 TE_intronic       12.8 
#  5 lncRNA             7.09
#  6 miRNA              5.32
#  7 3UTR               7.54
#  8 5UTR               6.38
#  9 intron             9.35
# 10 exon               9.57

# Peve: deep-dive-expression/E-Peve/code/12-Peve-WGBS.md
#    region        mean_meth
#  1 intergenic         4.87
#  2 TE_intergenic      9.76
#  3 TE_exonic         19.8 
#  4 TE_intronic        9.29
#  5 lncRNA             6.25
#  6 miRNA              2.92
#  7 3UTR               5.22
#  8 5UTR               4.79
#  9 intron             6.84
# 10 exon               6.38

# Ptuh: deep-dive-expression/F-Ptuh/code/12-Ptuh-WGBS.md
#    region        mean_meth
#  1 intergenic         2.08
#  2 TE_intergenic      3.89
#  3 TE_exonic          8.52
#  4 TE_intronic        4.71
#  5 lncRNA             2.66
#  6 miRNA              7.55
#  7 3UTR               2.86
#  8 5UTR               2.30
#  9 intron             3.73
# 10 exon               5.60


methylation_summary <- tibble(
  region = rep(c("intron", "exon", "intergenic", "TE_intronic", "TE_exonic",
                 "TE_intergenic", "5UTR", "3UTR", "miRNA", "lncRNA"), 3),
  mean_meth = c(
    # Apul
    9.35, 9.57, 9.19, 12.8, 13.1, 9.96, 6.38, 7.54, 5.32, 7.09,
    # Peve
    6.84, 6.38, 4.87, 9.29, 19.8, 9.76, 4.79, 5.22, 2.92, 6.25,
    # Ptuh
    3.73, 5.60, 2.08, 4.71, 8.52, 3.89, 2.30, 2.86, 7.55, 2.66
  ),
  species = factor(rep(species_order, each = 10), levels = species_order)
)

methylation_summary$region <- factor(methylation_summary$region,
  levels = c("intron", "exon", "intergenic", "TE_intronic", "TE_exonic",
             "TE_intergenic", "5UTR", "3UTR", "miRNA", "lncRNA"))

panel_c <- ggplot(methylation_summary, aes(x = region, y = mean_meth, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = species_pal, labels = species_labels, name = NULL) +
  labs(title = "C. DNA methylation by genomic feature", x = NULL, y = "Mean methylation (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        legend.position = "bottom",
        legend.text = element_text(size = 9, face = "italic"))


# multipanel figure

panel_a + panel_b + panel_c +
  plot_layout(ncol = 3, widths = c(1, 1, 1.8)) +
  plot_annotation(
    tag_levels = NULL,
    theme = theme(plot.background = element_rect(fill = "white", color = NA))
  )

ggsave("../output/20-supplementary-files/epigenetic_landscape_summary.png",
       width = 14, height = 6, dpi = 300, bg = "white")
```


