---
title: "22-miRNA-methylation-features"
author: "Kathleen Durkin"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

To make the ncRNA-methylation integration more complete, I want to evaluate the methylation of ncRNA-encoding genomic regions. Let's start with miRNA, since there's the most literature (in model systems) supporting methylation-driven regulation of their production (in comparison to lncRNA).

For each species, we already have some basic methylation stats computed, including mean % methylation of mature miRNA coordinates (~22nt) (per-miRNA x per-sample; `CpG_miRNA_CountMat.csv`). Because mature miRNAs are so short, very few overlap a CpG. Here I want to recompute the overlap using **miRNA precursor (hairpin) coordinates** (~60–100 nt), which capture CpGs in the stem-loop region and proximal regulatory context. This yields methylation data for a much larger fraction of miRNAs, and also makes more biological sense, as this is the molocule originally transcribed from DNA before being processed down into the mature form.

I then want to test whether miRNA-precursor methylation state is associated with relevant miRNA features, including:

- **Expression level** (mean normalized RPM)
- **Number of predicted mRNA targets** (miRanda + PCC)
- **Number of predicted lncRNA targets** (miRanda + PCC)
- **miRNA length** (precursor length from ShortStack)
- **Known vs novel status** (whether the miRNA has a match in miRBase / published
  cnidarian miRNA annotations)

The precursor methylation matrices will be computed by downloading the Bismark `.cov.gz` files from Gannet (large file storage), applying the same coverage filters as the WGBS Rmds (≥ 5x coverage, ≤ 99.9th percentile, present in ≥ 4 of 5 samples), and overlapping with the `*_precursor.gff3` coordinates from ShortStack.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(data.table)

# Use same colors as throughout repo
species_palette <- c("A. pulchra" = "#408EC6",
                     "P. evermanni" = "#1E2761",
                     "P. tuahiniensis" = "#7A2048")
```

# Load data

## miRNA precursor methylation (per-miRNA × per-sample % CpG methylation)

```{r load-meth}
load_meth <- function(path, species) {
  df <- read.csv(path, stringsAsFactors = FALSE, row.names = 1)
  # Precursor CSV: row names are "Cluster_XXXX" (no .mature/.precursor suffix)
  df$miRNA_id <- rownames(df)
  df$species  <- species
  df
}

meth_apul <- load_meth("../../D-Apul/output/08-Apul-WGBS/CpG_miRNA_precursor_CountMat.csv", "A. pulchra")
meth_peve <- load_meth("../../E-Peve/output/12-Peve-WGBS/CpG_miRNA_precursor_CountMat.csv", "P. evermanni")
meth_ptuh <- load_meth("../../F-Ptuh/output/12-Ptuh-WGBS/CpG_miRNA_precursor_CountMat.csv", "P. tuahiniensis")

# Sample columns per species (Peve has 5 WGBS samples, but only 3 sRNA)
meth_samples <- list(
  "A. pulchra"        = c("ACR.140.TP2", "ACR.145.TP2", "ACR.150.TP2", "ACR.173.TP2", "ACR.178.TP2"),
  "P. evermanni"      = c("POR.71.TP2", "POR.73.TP2", "POR.76.TP2", "POR.79.TP2", "POR.82.TP2"),
  "P. tuahiniensis"   = c("POC.47.TP2", "POC.48.TP2", "POC.50.TP2", "POC.53.TP2", "POC.57.TP2")
)

# Long format
meth_long <- bind_rows(
  meth_apul, meth_peve, meth_ptuh
) %>%
  pivot_longer(cols = all_of(unique(unlist(meth_samples))),
               names_to = "sample", values_to = "perc_meth") %>%
  filter(!is.na(perc_meth))

# Per-miRNA mean methylation
meth_summary <- meth_long %>%
  group_by(species, miRNA_id) %>%
  summarise(mean_meth = mean(perc_meth, na.rm = TRUE),
            sd_meth   = sd(perc_meth, na.rm = TRUE),
            n_samples = n(),
            .groups = "drop")

cat("miRNAs with methylation data:\n")
meth_summary %>% count(species, name = "n_miRNAs") %>% knitr::kable()
```

So in Apul, Peve, and Ptuh, there are **27, 19, and 19 miRNAs**, respectively, for whom we have methylation data coverage at the miRNA precursors coordinates (this filtering still permits miRNAs with 0% mean methylation).

## miRNA expression (normalized RPM)

```{r load-expr}
load_expr <- function(path, species, sample_cols) {
  df <- read.delim(path)
  df$miRNA_id <- rownames(df)
  df$species  <- species
  df %>%
    pivot_longer(cols = all_of(sample_cols), names_to = "sample_sRNA", values_to = "rpm") %>%
    mutate(sample_sRNA = gsub("sample", "", sample_sRNA))
}

expr_apul <- load_expr("../../D-Apul/output/03.1-Apul-sRNA-summary/Apul_counts_miRNA_normalized.txt",
                        "A. pulchra", c("sample140","sample145","sample150","sample173","sample178"))
expr_peve <- load_expr("../../E-Peve/output/03.1-Peve-sRNA-summary/Peve_counts_miRNA_normalized.txt",
                        "P. evermanni", c("sample73","sample79","sample82"))
expr_ptuh <- load_expr("../../F-Ptuh/output/03.1-Ptuh-sRNA-summary/Ptuh_counts_miRNA_normalized.txt",
                        "P. tuahiniensis", c("sample47","sample48","sample50","sample53","sample57"))

expr_long <- bind_rows(expr_apul, expr_peve, expr_ptuh)

# Per-miRNA mean expression
expr_summary <- expr_long %>%
  group_by(species, miRNA_id) %>%
  summarise(mean_rpm = mean(rpm, na.rm = TRUE),
            sd_rpm   = sd(rpm, na.rm = TRUE),
            .groups = "drop")
```

## miRNA target counts (miRanda + PCC)

```{r load-targets}
# mRNA targets — all predicted binding pairs and p < 0.05 subset
load_targets <- function(all_path, sig_path, species) {
  all_df <- fread(all_path)
  sig_df <- if (file.exists(sig_path)) fread(sig_path) else data.table()

  all_counts <- all_df[, .(n_targets_all = uniqueN(mRNA)), by = .(miRNA)]

  if (nrow(sig_df) > 0) {
    sig_counts <- sig_df[, .(n_targets_sig = uniqueN(mRNA)), by = .(miRNA)]
    all_counts <- merge(all_counts, sig_counts, by = "miRNA", all = TRUE)
  } else {
    all_counts$n_targets_sig <- 0L
  }

  all_counts$species <- species
  all_counts[is.na(n_targets_sig), n_targets_sig := 0L]
  all_counts
}

targets_apul <- load_targets(
  "../../D-Apul/output/09-Apul-mRNA-miRNA-interactions/miranda_PCC_miRNA_mRNA.csv",
  "../../D-Apul/output/09-Apul-mRNA-miRNA-interactions/Apul-miranda_PCC_sig_miRNA_mRNA.csv",
  "A. pulchra"
)
targets_peve <- load_targets(
  "../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miranda_PCC_miRNA_mRNA.csv",
  "../../E-Peve/output/10-Peve-mRNA-miRNA-interactions/Peve-miranda_PCC_sig_miRNA_mRNA.csv",
  "P. evermanni"
)
targets_ptuh <- load_targets(
  "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/three_prime_interaction/Ptuh-miranda_PCC_miRNA_mRNA.csv",
  "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/three_prime_interaction/Ptuh-miranda_PCC_sig_miRNA_mRNA.csv",
  "P. tuahiniensis"
)

# lncRNA targets
load_lncrna_targets <- function(path, species) {
  df <- fread(path)
  df[, .(n_lncrna_targets = uniqueN(lncRNA)), by = .(miRNA = miRNA)][, species := species][]
}

lnc_targets_apul <- load_lncrna_targets("../../D-Apul/output/28-Apul-miRNA-lncRNA-interactions/miranda_PCC_miRNA_lncRNA.csv", "A. pulchra")
lnc_targets_peve <- load_lncrna_targets("../../E-Peve/output/15-Peve-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv", "P. evermanni")
lnc_targets_ptuh <- load_lncrna_targets("../../F-Ptuh/output/15-Ptuh-miRNA-lncRNA-PCC/miranda_PCC_miRNA_lncRNA.csv", "P. tuahiniensis")

targets_summary <- bind_rows(targets_apul, targets_peve, targets_ptuh) %>%
  left_join(bind_rows(lnc_targets_apul, lnc_targets_peve, lnc_targets_ptuh),
            by = c("miRNA", "species")) %>%
  mutate(n_lncrna_targets = replace_na(n_lncrna_targets, 0L))
```

## ShortStack features (length, known/novel, total reads)

```{r load-shortstack}
conserved_miRNAs <- c("apul-mir-100", "peve-mir-100", "ptuh-mir-100", 
              "apul-mir-2036", "peve-mir-2036", "ptuh-mir-2036",
              "apul-mir-2023", "peve-mir-2023", "ptuh-mir-2023",
              "apul-mir-2025", "peve-mir-2025", "ptuh-mir-2025")

load_shortstack <- function(path, species) {
  df <- read.csv(path, stringsAsFactors = FALSE)
  df %>%
    select(Name, Length, Reads, DicerCall, MIRNA, known_miRNAs, given_miRNA_name) %>%
    rename(miRNA_id = Name) %>%
    mutate(
      species = species,
      known   = !is.na(known_miRNAs) & known_miRNAs != "" & known_miRNAs != "NA",
      conserved = given_miRNA_name %in% conserved_miRNAs
    )
}

ss_apul <- load_shortstack("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Apul_Results_mature_named_miRNAs.csv", "A. pulchra")
ss_peve <- load_shortstack("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv", "P. evermanni")
ss_ptuh <- load_shortstack("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Ptuh_Results_mature_named_miRNAs.csv", "P. tuahiniensis")

ss_summary <- bind_rows(ss_apul, ss_peve, ss_ptuh)
```

# Merge into a single per-miRNA table

```{r merge}
# Only miRNAs with methylation data are retained (inner join on meth)
master <- meth_summary %>%
  left_join(expr_summary, by = c("species", "miRNA_id")) %>%
  left_join(
    targets_summary %>% rename(miRNA_id = miRNA),
    by = c("species", "miRNA_id")
  ) %>%
  left_join(ss_summary, by = c("species", "miRNA_id")) %>%
  mutate(
    n_targets_all    = replace_na(n_targets_all, 0L),
    n_targets_sig    = replace_na(n_targets_sig, 0L),
    n_lncrna_targets = replace_na(n_lncrna_targets, 0L),
    log2_rpm         = log2(mean_rpm + 1)
  )

fwrite(master, file.path("../output/22-miRNA-methylation-features", "miRNA_methylation_features_master.csv"))
knitr::kable(master %>% select(species, miRNA_id, mean_meth, mean_rpm, n_targets_all, n_targets_sig, n_lncrna_targets, Length, known, given_miRNA_name))
```

# Within-miRNA methylation–expression correlation (across samples)

For each miRNA that has methylation data in ≥ 3 samples, compute the
within-miRNA Pearson correlation between % methylation and RPM across samples.
This asks: does methylation fluctuate with expression *within* a given miRNA?

```{r within-mirna-corr}
# Build per-sample matched table
# Apul: sRNA samples 140,145,150,173,178 ↔ WGBS ACR.140,145,150,173,178
# Peve: sRNA samples 73,79,82 ↔ WGBS POR.73,79,82 (3 overlap → usable)
# Ptuh: sRNA samples 47,48,50,53,57 ↔ WGBS POC.47,48,50,53,57

build_matched <- function(meth_df, expr_df, meth_cols, expr_cols, species) {
  meth_sub <- meth_df %>%
    select(miRNA_id, all_of(meth_cols)) %>%
    pivot_longer(cols = all_of(meth_cols), names_to = "sample_meth", values_to = "perc_meth") %>%
    # Extract sample number from names like "ACR.140.TP2" → "140"
    mutate(sample_num = sub(".*\\.(\\d+)\\.TP\\d+$", "\\1", sample_meth))

  expr_sub <- expr_df %>%
    select(miRNA_id, all_of(expr_cols)) %>%
    pivot_longer(cols = all_of(expr_cols), names_to = "sample_expr", values_to = "rpm") %>%
    mutate(sample_num = sample_expr)

  inner_join(meth_sub, expr_sub, by = c("miRNA_id", "sample_num")) %>%
    mutate(species = species) %>%
    select(species, miRNA_id, sample_num, perc_meth, rpm)
}

matched_apul <- build_matched(
  meth_apul, expr_apul %>% select(-species) %>% pivot_wider(names_from = sample_sRNA, values_from = rpm) %>% mutate(miRNA_id = miRNA_id),
  meth_samples[["A. pulchra"]],
  c("140","145","150","173","178"),
  "A. pulchra"
)

matched_ptuh <- build_matched(
  meth_ptuh, expr_ptuh %>% select(-species) %>% pivot_wider(names_from = sample_sRNA, values_from = rpm) %>% mutate(miRNA_id = miRNA_id),
  meth_samples[["P. tuahiniensis"]],
  c("47","48","50","53","57"),
  "P. tuahiniensis"
)

# Peve: 3 overlapping samples (73, 79, 82) — just enough for correlation
matched_peve <- build_matched(
  meth_peve, expr_peve %>% select(-species) %>% pivot_wider(names_from = sample_sRNA, values_from = rpm) %>% mutate(miRNA_id = miRNA_id),
  c("POR.73.TP2", "POR.79.TP2", "POR.82.TP2"),
  c("73","79","82"),
  "P. evermanni"
)

matched_all <- bind_rows(matched_apul, matched_peve, matched_ptuh)

# Within-miRNA correlation (only for miRNAs with ≥ 3 matched samples)
within_corr <- matched_all %>%
  group_by(species, miRNA_id) %>%
  filter(n() >= 3) %>%
  summarise(
    r = tryCatch(cor(perc_meth, rpm, method = "pearson", use = "pairwise.complete.obs"),
                 error = function(e) NA_real_),
    p = tryCatch(cor.test(perc_meth, rpm, method = "pearson")$p.value,
                 error = function(e) NA_real_),
    n_samples = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(r)) %>%
  mutate(direction = ifelse(r > 0, "Positive", "Negative"))

fwrite(within_corr, file.path("../output/22-miRNA-methylation-features", "within_miRNA_meth_expr_corr.csv"))
knitr::kable(within_corr)
```

`r if (exists("within_corr") && nrow(within_corr) > 0) paste0("Within-miRNA correlations are available for ", nrow(within_corr), " miRNAs across all three species (Peve has 3 overlapping sRNA/WGBS samples).")`

# Cross-miRNA correlations (methylation vs features)

For each species, test whether mean miRNA-precursor methylation correlates with: mean expression, number of mRNA targets (all and significant), number of lncRNA targets, and miRNA length. Also test whether methylation differs between known and novel miRNAs.

```{r cross-mirna-corr}
# Helper: correlation test with tidy output
corr_test <- function(df, var_x, var_y) {
  if (nrow(df) < 4) return(data.frame(r = NA_real_, p = NA_real_, n = nrow(df)))
  t <- cor.test(df[[var_x]], df[[var_y]], method = "pearson")
  data.frame(r = unname(t$estimate), p = t$p.value, n = nrow(df))
}

features <- c("log2_rpm", "n_targets_all", "n_targets_sig", "n_lncrna_targets", "Length")
feature_labels <- c(log2_rpm = "log2(mean RPM + 1)",
                    n_targets_all = "# mRNA targets (all)",
                    n_targets_sig = "# mRNA targets (p < 0.05)",
                    n_lncrna_targets = "# lncRNA targets",
                    Length = "Precursor length (nt)")

corr_results <- list()
for (sp in unique(master$species)) {
  sp_df <- master %>% filter(species == sp)
  for (feat in features) {
    ct <- corr_test(sp_df, "mean_meth", feat)
    corr_results[[paste(sp, feat, sep = "__")]] <- data.frame(
      species = sp, feature = feat, feature_label = feature_labels[feat],
      r = ct$r, p = ct$p, n = ct$n
    )
  }
}
corr_results_df <- bind_rows(corr_results) %>%
  mutate(signif = ifelse(p < 0.05, "*", ""),
         r_round = round(r, 3),
         p_round = signif(p, 3))

fwrite(corr_results_df, file.path("../output/22-miRNA-methylation-features", "cross_miRNA_meth_feature_corr.csv"))
knitr::kable(corr_results_df)
```

# Figures

## Fig 1: Methylation vs expression (cross-miRNA, per species)

```{r fig-meth-expr, fig.width=10, fig.height=4}
p_expr <- ggplot(master, aes(x = mean_meth, y = log2_rpm, color = species)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.6) +
  facet_wrap(~ species, scales = "free") +
  scale_color_manual(values = species_palette) +
  labs(x = "Mean CpG % methylation at miRNA precursor",
       y = "log2(mean RPM + 1)",
       title = "miRNA-precursor methylation vs expression") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "italic"),
        legend.position = "none")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_meth_vs_expr.png"), p_expr, width = 10, height = 4, dpi = 300)
p_expr
```

## Fig 2: Methylation vs number of targets (cross-miRNA, per species)

```{r fig-meth-targets, fig.width=12, fig.height=8}
plot_target_data <- master %>%
  select(species, miRNA_id, mean_meth,
         `mRNA (all)` = n_targets_all,
         `mRNA (p<0.05)` = n_targets_sig,
         `lncRNA` = n_lncrna_targets) %>%
  pivot_longer(cols = c(`mRNA (all)`, `mRNA (p<0.05)`, `lncRNA`),
               names_to = "target_type", values_to = "n_targets")

p_targets <- ggplot(plot_target_data, aes(x = mean_meth, y = n_targets, color = species)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.6) +
  facet_grid(target_type ~ species, scales = "free_y") +
  scale_color_manual(values = species_palette) +
  labs(x = "Mean CpG % methylation at miRNA precursor",
       y = "Number of predicted targets",
       title = "miRNA-precursor methylation vs number of predicted targets") +
  theme_bw(base_size = 11) +
  theme(strip.text.y = element_text(face = "bold"),
        strip.text.x = element_text(face = "italic"),
        legend.position = "none")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_meth_vs_targets.png"), p_targets, width = 12, height = 8, dpi = 300)
p_targets
```

## Fig 3: Methylation vs miRNA length and known/novel status

```{r fig-meth-length-known, fig.width=10, fig.height=4}
p_length <- ggplot(master, aes(x = mean_meth, y = Length, color = species)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.6) +
  facet_wrap(~ species, scales = "free") +
  scale_color_manual(values = species_palette) +
  labs(x = "Mean CpG % methylation at miRNA precursor",
       y = "Precursor length (nt)",
       title = "miRNA-precursor methylation vs precursor length") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "italic"),
        legend.position = "none")

p_known <- ggplot(master, aes(x = known, y = mean_meth, fill = species)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2, aes(color = species)) +
  facet_wrap(~ species, scales = "free_x") +
  scale_fill_manual(values = species_palette) +
  scale_color_manual(values = species_palette) +
  labs(x = "Known (annotated) vs Novel miRNA",
       y = "Mean CpG % methylation",
       title = "miRNA-precursor methylation: known vs novel miRNAs") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "italic"),
        legend.position = "none")

p_conserved <- ggplot(master, aes(x = conserved, y = mean_meth, fill = species)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2, aes(color = species)) +
  facet_wrap(~ species, scales = "free_x") +
  scale_fill_manual(values = species_palette) +
  scale_color_manual(values = species_palette) +
  labs(x = "Conserved v species-specific miRNA",
       y = "Mean CpG % methylation",
       title = "miRNA-precursor methylation: conserved v species-specific miRNAs") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "italic"),
        legend.position = "none")

p_length_known_conserved <- p_length / p_known / p_conserved

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_meth_vs_length_known_conserved.png"), p_length_known_conserved, width = 10, height = 10, dpi = 300)
p_length_known_conserved
```

## Fig 4: Within-miRNA methylation–expression correlation (across samples)

```{r fig-within-corr, fig.width=8, fig.height=4}
if (exists("within_corr") && nrow(within_corr) > 0) {
  p_within <- ggplot(within_corr, aes(x = r, fill = species)) +
    geom_histogram(binwidth = 0.2, boundary = 0, color = "white", alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    facet_wrap(~ species) +
    scale_fill_manual(values = species_palette) +
    labs(x = "Within-miRNA Pearson r (methylation vs expression, across samples)",
         y = "Number of miRNAs",
         title = "Within-miRNA methylation–expression correlation") +
    theme_bw(base_size = 11) +
    theme(strip.text = element_text(face = "italic"),
          legend.position = "none")

  ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_within_miRNA_corr.png"), p_within, width = 8, height = 4, dpi = 300)
  p_within
} else {
  cat("No miRNAs with >= 3 matched samples for within-miRNA correlation.\n")
}
```

## Fig 5: Per-sample scatter (methylation vs expression, all miRNAs pooled)

This reproduces the plot from the WGBS Rmds but combines all three species.

```{r fig-per-sample, fig.width=10, fig.height=4}
# Build a combined per-sample table
matched_combined <- bind_rows(
  matched_apul %>% mutate(sample_label = paste0("ACR.", sample_num)),
  matched_peve %>% mutate(sample_label = paste0("POR.", sample_num)),
  matched_ptuh %>% mutate(sample_label = paste0("POC.", sample_num))
) %>%
  mutate(species = factor(species, levels = c("A. pulchra", "P. evermanni", "P. tuahiniensis")))

p_per_sample <- ggplot(matched_combined, aes(x = perc_meth, y = log2(rpm + 1), color = species)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  facet_wrap(~ species, scales = "free") +
  scale_color_manual(values = species_palette) +
  labs(x = "CpG % methylation at miRNA precursor (per sample)",
       y = "log2(RPM + 1)",
       title = "Per-sample miRNA methylation vs expression (all methylated miRNAs pooled)") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "italic"),
        legend.position = "none")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_per_sample_meth_expr.png"), p_per_sample, width = 10, height = 4, dpi = 300)
p_per_sample
```

# Methylation level of all miRNAs per species (mature vs precursor)

Dot plot and bar plot showing mean CpG % methylation for every miRNA with methylation data, faceted by species and feature type (mature vs precursor). Points are colored by conserved v species-specific status.

```{r fig-methylation-by-species, fig.width=12, fig.height=7}
# Load name annotations (already loaded as ss_summary, but rebuild for clarity)
load_names <- function(path, species) {
  df <- read.csv(path, stringsAsFactors = FALSE)
  df %>% select(Name, known_miRNAs, given_miRNA_name) %>%
    rename(miRNA_id = Name) %>%
    mutate(species = species,
           known = !is.na(known_miRNAs) & known_miRNAs != "" & known_miRNAs != "NA",
           conserved = given_miRNA_name %in% conserved_miRNAs)
}

all_names <- bind_rows(
  load_names("../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Apul_Results_mature_named_miRNAs.csv", "A. pulchra"),
  load_names("../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Peve_Results_mature_named_miRNAs.csv", "P. evermanni"),
  load_names("../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Ptuh_Results_mature_named_miRNAs.csv", "P. tuahiniensis")
)

# Load both mature and precursor methylation
load_meth_both <- function(mature_path, precursor_path, species) {
  # Mature
  mat <- read.csv(mature_path, stringsAsFactors = FALSE, header = TRUE) %>% dplyr::select(-X)
  mat$miRNA_id <- gsub("[.]mature$", "", mat$miRNA_id)
  mat$species <- species
  mat$feature_type <- "Mature"
  # Precursor
  prec <- read.csv(precursor_path, stringsAsFactors = FALSE, header = TRUE) 
  prec$miRNA_id <- prec$precursor_id
  prec <- prec %>% dplyr::select(-precursor_id)
  prec$species <- species
  prec$feature_type <- "Precursor"
  bind_rows(mat, prec)
}

meth_both <- bind_rows(
  load_meth_both("../../D-Apul/output/08-Apul-WGBS/CpG_miRNA_CountMat.csv",
                 "../../D-Apul/output/08-Apul-WGBS/CpG_miRNA_precursor_CountMat.csv", "A. pulchra"),
  load_meth_both("../../E-Peve/output/12-Peve-WGBS/CpG_miRNA_CountMat.csv",
                 "../../E-Peve/output/12-Peve-WGBS/CpG_miRNA_precursor_CountMat.csv", "P. evermanni"),
  load_meth_both("../../F-Ptuh/output/12-Ptuh-WGBS/CpG_miRNA_CountMat.csv",
                 "../../F-Ptuh/output/12-Ptuh-WGBS/CpG_miRNA_precursor_CountMat.csv", "P. tuahiniensis")
)

# Long format and mean per miRNA
meth_both_long <- meth_both %>%
  pivot_longer(cols = matches("\\.TP2$"), names_to = "sample", values_to = "perc_meth") %>%
  filter(!is.na(perc_meth)) %>%
  group_by(species, feature_type, miRNA_id) %>%
  summarise(mean_meth = mean(perc_meth, na.rm = TRUE), .groups = "drop") %>%
  left_join(all_names, by = c("species", "miRNA_id")) %>%
  mutate(
    known = ifelse(is.na(known), FALSE, known),
    display_name = ifelse(is.na(given_miRNA_name) | given_miRNA_name == "NA", miRNA_id,
                          sub("^[a-z]+-", "", given_miRNA_name)),
    display_name = sub(";.*", "", display_name),
    species = factor(species, levels = c("A. pulchra", "P. evermanni", "P. tuahiniensis")),
    feature_type = factor(feature_type, levels = c("Mature", "Precursor"))
  ) %>%
  arrange(species, feature_type, mean_meth) %>%
  group_by(species, feature_type) %>%
  mutate(rank = row_number()) %>%
  ungroup()

# Dot plot
p_meth <- ggplot(meth_both_long, aes(x = rank, y = mean_meth, color = conserved)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  facet_grid(feature_type ~ species, scales = "free_x") +
  scale_color_manual(values = c("TRUE" = "#1B9E77", "FALSE" = "#D6604D"),
                     labels = c("TRUE" = "Conserved", "FALSE" = "Species-specific"), name = NULL) +
  labs(x = "miRNA (ranked by methylation)", y = "Mean CpG % methylation",
       title = "miRNA methylation level per species",
       subtitle = "Each point is one miRNA; green = conserved , red = species-specific") +
  theme_bw(base_size = 11) +
  theme(strip.text.x = element_text(face = "italic"),
        strip.text.y = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_miRNA_methylation_by_species.png"), p_meth, width = 12, height = 7, dpi = 300)
p_meth
```

Interesting sort of bimodal pattern -- precursors are either ver lowly or very highly methylated. Also, all conserved miRNA are lowly methylated (both mature and precursor sequences).

```{r fig-methylation-by-species-barplot, fig.width=16, fig.height=8}
# Bar plot with miRNA names on x-axis
p_meth_bar <- ggplot(meth_both_long, aes(x = reorder(display_name, mean_meth), y = mean_meth, fill = conserved)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  facet_grid(feature_type ~ species, scales = "free_x") +
  scale_fill_manual(values = c("TRUE" = "#1B9E77", "FALSE" = "#D6604D"),
                    labels = c("TRUE" = "Conserved", "FALSE" = "Species-specific"), name = NULL) +
  labs(x = NULL, y = "Mean CpG % methylation",
       title = "miRNA methylation level per species",
       subtitle = "Green = conserved, red = species-specific") +
  theme_bw(base_size = 9) +
  theme(strip.text.x = element_text(face = "italic"),
        strip.text.y = element_text(face = "bold"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 6),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_miRNA_methylation_by_species_barplot.png"), p_meth_bar, width = 16, height = 8, dpi = 300)
p_meth_bar
```

# Methylation class (Low <50% vs High >=50%) vs expression

The methylation plots reveal a bimodal distribution, particularly in the precursor data: miRNAs cluster either near 0% or above 50% methylation. Here I want to test whether these two classes differ in expression level.

```{r fig-methylation-class-vs-expression, fig.width=12, fig.height=7}
# Build a combined table with both mature and precursor methylation + expression
load_meth_both <- function(mature_path, precursor_path, species) {
  # Mature: first col is row index, second col is miRNA_id
  mat <- read.csv(mature_path, stringsAsFactors = FALSE)
  mat$miRNA_id <- gsub("[.]mature$", "", mat$miRNA_id)
  mat <- mat %>% select(-1)  # drop row index
  mat$species <- species; mat$feature_type <- "Mature"
  # Precursor: first col is precursor_id (row names)
  prec <- read.csv(precursor_path, stringsAsFactors = FALSE, row.names = 1)
  prec$miRNA_id <- rownames(prec)
  prec$species <- species; prec$feature_type <- "Precursor"
  bind_rows(mat, prec)
}

meth_both <- bind_rows(
  load_meth_both("../../D-Apul/output/08-Apul-WGBS/CpG_miRNA_CountMat.csv",
                 "../../D-Apul/output/08-Apul-WGBS/CpG_miRNA_precursor_CountMat.csv", "A. pulchra"),
  load_meth_both("../../E-Peve/output/12-Peve-WGBS/CpG_miRNA_CountMat.csv",
                 "../../E-Peve/output/12-Peve-WGBS/CpG_miRNA_precursor_CountMat.csv", "P. evermanni"),
  load_meth_both("../../F-Ptuh/output/12-Ptuh-WGBS/CpG_miRNA_CountMat.csv",
                 "../../F-Ptuh/output/12-Ptuh-WGBS/CpG_miRNA_precursor_CountMat.csv", "P. tuahiniensis")
)

meth_both_summary <- meth_both %>%
  pivot_longer(cols = matches("[.]TP2$"), names_to = "sample", values_to = "perc_meth") %>%
  filter(!is.na(perc_meth)) %>%
  group_by(species, feature_type, miRNA_id) %>%
  summarise(mean_meth = mean(perc_meth, na.rm = TRUE), .groups = "drop") %>%
  left_join(all_names, by = c("species", "miRNA_id")) %>%
  mutate(known = ifelse(is.na(known), FALSE, known)) %>%
  # Join expression (already loaded as expr_summary earlier)
  left_join(expr_summary, by = c("species", "miRNA_id")) %>%
  mutate(log2_rpm = log2(mean_rpm + 1)) %>%
  mutate(
    meth_class = ifelse(mean_meth >= 50, "High (>=50%)", "Low (<50%)"),
    display_name = ifelse(is.na(given_miRNA_name) | given_miRNA_name == "NA", miRNA_id,
                          sub("^[a-z]+-", "", given_miRNA_name)),
    display_name = sub(";.*", "", display_name),
    species = factor(species, levels = c("A. pulchra", "P. evermanni", "P. tuahiniensis")),
    feature_type = factor(feature_type, levels = c("Mature", "Precursor"))
  )

# Summary table
cat("High vs Low methylation: expression comparison\n\n")
print(meth_both_summary %>% filter(!is.na(log2_rpm)) %>%
  group_by(species, feature_type, meth_class) %>%
  summarise(n = n(), mean_log2_rpm = round(mean(log2_rpm), 2),
            median_log2_rpm = round(median(log2_rpm), 2), .groups = "drop") %>%
  as.data.frame(), row.names = FALSE)

# Wilcoxon tests
cat("\nWilcoxon tests (High vs Low):\n\n")
for (sp in levels(meth_both_summary$species)) {
  for (ft in c("Mature", "Precursor")) {
    sub <- meth_both_summary %>% filter(species == sp, feature_type == ft, !is.na(log2_rpm))
    if (nrow(sub) >= 4 && length(unique(sub$meth_class)) == 2) {
      wt <- wilcox.test(log2_rpm ~ meth_class, data = sub)
      cat(sprintf("  %s %s: W=%.1f, p=%.4f (n_high=%d, n_low=%d)\n",
                  sp, ft, wt$statistic, wt$p.value,
                  sum(sub$meth_class == "High (>=50%)"), sum(sub$meth_class == "Low (<50%)")))
    } else {
      cat(sprintf("  %s %s: skipped (n=%d, groups=%d)\n", sp, ft, nrow(sub), length(unique(sub$meth_class))))
    }
  }
}

# Boxplot
p_class <- ggplot(meth_both_summary %>% filter(!is.na(log2_rpm)),
                  aes(x = meth_class, y = log2_rpm, fill = meth_class)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.8, size = 2, aes(color = conserved)) +
  facet_grid(feature_type ~ species, scales = "free_y") +
  scale_fill_manual(values = c("Low (<50%)" = "#85C1E9", "High (>=50%)" = "#E74C3C"),
                    name = "Methylation class") +
  scale_color_manual(values = c("TRUE" = "#1B9E77", "FALSE" = "#D6604D"),
                     labels = c("TRUE" = "Conserved", "FALSE" = "Species-specific"), name = NULL) +
  labs(x = NULL, y = "log2(mean RPM + 1)",
       title = "miRNA expression by methylation class (Low <50% vs High >=50%)",
       subtitle = "Boxplots + individual points; green/red points = conserved/species-specific") +
  theme_bw(base_size = 11) +
  theme(strip.text.x = element_text(face = "italic"),
        strip.text.y = element_text(face = "bold"),
        legend.position = "bottom")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_methylation_class_vs_expression.png"), p_class, width = 12, height = 7, dpi = 300)
p_class
```

```{r fig-methylation-vs-expression-scatter, fig.width=12, fig.height=7}
# Scatter with 50% threshold line
p_scatter <- ggplot(meth_both_summary %>% filter(!is.na(log2_rpm)),
                    aes(x = mean_meth, y = log2_rpm, color = conserved)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "grey50") +
  facet_grid(feature_type ~ species, scales = "free") +
  scale_color_manual(values = c("TRUE" = "#1B9E77", "FALSE" = "#D6604D"),
                     labels = c("TRUE" = "Conserved", "FALSE" = "Species-specific"), name = NULL) +
  labs(x = "Mean CpG % methylation", y = "log2(mean RPM + 1)",
       title = "miRNA methylation vs expression",
       subtitle = "Dashed line at 50% marks the Low/High split") +
  theme_bw(base_size = 11) +
  theme(strip.text.x = element_text(face = "italic"),
        strip.text.y = element_text(face = "bold"),
        legend.position = "bottom")

ggsave(file.path("../output/22-miRNA-methylation-features", "Fig_methylation_vs_expression_scatter.png"), p_scatter, width = 12, height = 7, dpi = 300)
p_scatter
```

# Summary

```{r summary-table}
# Count how many miRNAs have methylation data per species
n_table <- master %>%
  group_by(species) %>%
  summarise(
    n_with_meth = n(),
    n_conserved     = sum(conserved),
    n_spec     = sum(!conserved),
    mean_meth   = round(mean(mean_meth, na.rm = TRUE), 2),
    range_meth  = paste0(round(min(mean_meth, na.rm = TRUE), 1), "–", round(max(mean_meth, na.rm = TRUE), 1)),
    .groups = "drop"
  )

fwrite(n_table, file.path("../output/22-miRNA-methylation-features", "summary_methylated_miRNAs.csv"))
knitr::kable(n_table)
```

```{r corr-summary}
# Highlight any significant correlations
sig_corrs <- corr_results_df %>% filter(p < 0.05)
if (nrow(sig_corrs) > 0) {
  cat("Significant cross-miRNA correlations (p < 0.05):\n\n")
  knitr::kable(sig_corrs)
} else {
  cat("No significant cross-miRNA correlations (p < 0.05) detected.\n")
}
```

# Session info

```{r}
sessionInfo()
```
