---
title: "22.1-genomic-context-annotations"
author: "Kathleen Durkin"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

Here I've compiled a more complete genome annotation for each species (*A. pulchra*, *P. evermanni*, *P. tuahiniensis*) that brings together every genomic-context feature needed for downstream ncRNA-methylation analyses (miRNA (mature + precursor), lncRNA, 5'/3' UTRs, 
TEs (split by exonic/intronic/intergenic context), exons, introns, and intergenic regions).

The compiled annotations are also used here to (1) classify each miRNA as intergenic or intragenic based on whether its precursor overlaps a host gene, and (2) define promoter regions for every miRNA and lncRNA. Promoters are defined strand-aware as 1500 bp upstream and 400 bp downstream of the transcription start site (TSS). The TSS is defined as:

- lncRNA: the 5' end of the lncRNA transcript itself
- Intragenic miRNA: the 5' end of the host gene it sits inside
- Intergenic miRNA: the 5' end of the miRNA precursor (hairpin)

Note: in reality, the TSS for an intergenic miRNA would actually sit at the beginning of the primary miRNA sequence (pri-miRNA). However, these sequences are much longer than mature or precursor miRNA and are not captured in the sRNAseq data, nor are they identifiable using ShortStack. Since we don't have info on the length or start site of pri-miRNAs for our data, I'm going to be using the pre-miRNA start site as a proxy instead.

Note: while I'm going to define promoter regions for all of our ncRNA for posterity (given the promoter-based regulatory role of DNA methylation in vertebrate systems), I don't necessarily expect that information to be useful in our analyses. There's no evidence for methylation of gene promoters having relevance for expression or physiology outcomes, and there's no reason to think that will be different for the promoters of ncRNAs.

The context annotation logic mirrors the per-species WGBS Rmds (`D-Apul/code/08-Apul-WGBS.Rmd`, `E-Peve/code/12-Peve-WGBS.Rmd`, `F-Ptuh/code/12-Ptuh-WGBS.Rmd`) so that the region categories and priority scheme stay consistent with the rest of the repo. However, instead of using a heirarchical assignment strategy, I'll be using a proportional assignment strategy. Hierarchical assignment is perfectly sound logic when assigning individual cytosine nucleotides to a given context, but fails when applied to longer loci. Proportional assignment will more accurately reflect each feature's majority context.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(GenomicRanges)
library(data.table)

# Use same colors as throughout repo
species_palette <- c("A. pulchra" = "#408EC6",
                     "P. evermanni" = "#1E2761",
                     "P. tuahiniensis" = "#7A2048")

# Promoter definition (bp relative to TSS, strand-aware)
PROM_UPSTREAM   <- 1500   # bp upstream of TSS
PROM_DOWNSTREAM <- 400    # bp downstream of TSS
```

# Species metadata

Per-species paths to every input file used below
```{r species-meta}
species_meta <- tibble::tribble(
  ~species, ~display, ~genome_fai, ~gene_gtf, ~utr5_gff, ~utr3_gff, ~rmsk_out, ~mirna_mature_gff3, ~mirna_results_gff3, ~lncrna_gtf,
  "Apul", "A. pulchra",
  "../../D-Apul/data/Apulchra-genome.fa.fai",
  "../../D-Apul/data/Apulchra-genome.gtf",
  "../../D-Apul/output/15-Apul-annotate-UTRs/Apul.GFFannotation.5UTR_1kb_corrected.gff",
  "../../D-Apul/output/15-Apul-annotate-UTRs/Apul.GFFannotation.3UTR_1kb_corrected.gff",
  "../../D-Apul/data/apul.hifiasm.s55_pa.p_ctg.fa.k32.w100.z1000.ntLink.5rounds.fa.out",
  "../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/Apul_ShortStack_4.1.0_mature.gff3",
  "../../D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Results.gff3",
  "../../D-Apul/output/31-Apul-lncRNA/Apul-lncRNA.gtf",

  "Peve", "P. evermanni",
  "../../E-Peve/data/Porites_evermanni_v1.fa.fai",
  "../../E-Peve/data/Porites_evermanni_validated.gtf",
  "../../E-Peve/output/16-Peve-annotate-UTRs/Peve.GFFannotation.5UTR_1kb_corrected.gff",
  "../../E-Peve/output/16-Peve-annotate-UTRs/Peve.GFFannotation.3UTR_1kb_corrected.gff",
  "../../E-Peve/data/Porites_evermanni_v1.fa.out",
  "../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/Peve_ShortStack_4.1.0_mature.gff3",
  "../../E-Peve/output/05-Peve-sRNA-ShortStack_4.1.0/ShortStack_out/Results.gff3",
  "../../E-Peve/output/17-Peve-lncRNA/Peve-lncRNA.gtf",

  "Ptuh", "P. tuahiniensis",
  "../../F-Ptuh/data/Pocillopora_meandrina_HIv1.assembly.fasta.fai",
  "../../F-Ptuh/data/Pocillopora_meandrina_HIv1.genes-validated.gtf",
  "../../F-Ptuh/output/16-Ptuh-annotate-UTRs/Ptuh.GFFannotation.5UTR_1kb_corrected.gff",
  "../../F-Ptuh/output/16-Ptuh-annotate-UTRs/Ptuh.GFFannotation.3UTR_1kb_corrected.gff",
  "../../F-Ptuh/data/Pocillopora_meandrina_HIv1.assembly.fasta.out",
  "../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/Ptuh_ShortStack_4.1.0_mature.gff3",
  "../../F-Ptuh/output/05-Ptuh-sRNA-ShortStack_4.1.0/ShortStack_out/Results.gff3",
  "../../F-Ptuh/output/17-Ptuh-lncRNA/Ptuh-lncRNA.gtf"
)

species_meta
```

# Helper GTF/GFF reader

`genomation::gffToGRanges` failed on some of the Peve annotation files (the validated GTF has `##gtf-version` header lines and unquoted attributes that this built-in parser can't handle). Instead, I need to use a custom reader based on `read.delim` with `comment.char = "#"` to handle all three species uniformly.
```{r helper-read-gtf}
#' Read a GTF or GFF file into a GRanges object, optionally filtering by feature type
#'
#' Handles both GTF attributes (key "value"; or key value;) and GFF3 attributes
#' (key=value;). Skips comment/header lines starting with #.
#'
#' @param path path to the GTF/GFF file
#' @param filter optional character vector of feature types to keep (column 3)
#' @return GRanges with mcols: type, source, and any parsed attributes
#'   (gene_id, transcript_id, ID, Parent, etc.)
read_gtf_granges <- function(path, filter = NULL) {
  df <- read.delim(path, header = FALSE, sep = "\t", comment.char = "#",
                   stringsAsFactors = FALSE, quote = "")
  colnames(df)[1:9] <- c("seqid", "source", "type", "start", "end",
                         "score", "strand", "phase", "attributes")

  # Filter by feature type if requested
  if (!is.null(filter)) {
    df <- df[df$type %in% filter, ]
  }
  if (nrow(df) == 0) {
    return(GRanges())
  }

  # Parse attributes from column 9
  # Try GTF format first (key "value"; or key value;), then GFF3 (key=value;)
  parse_attrs <- function(attr_str) {
    # Split on semicolons
    pairs <- strsplit(attr_str, ";")[[1]]
    pairs <- trimws(pairs)
    pairs <- pairs[pairs != ""]
    result <- list()
    for (p in pairs) {
      # GFF3 style: key=value
      if (grepl("=", p)) {
        kv <- strsplit(p, "=", fixed = TRUE)[[1]]
        if (length(kv) >= 2) {
          result[[kv[1]]] <- paste(kv[-1], collapse = "=")
        }
      } else {
        # GTF style: key "value" or key value
        # Split on first whitespace only
        sp_pos <- regexpr("\\s", p)
        if (sp_pos > 0) {
          kv <- c(substr(p, 1, sp_pos - 1), substr(p, sp_pos + 1, nchar(p)))
          val <- gsub('^"|"$', "", kv[2])
          result[[kv[1]]] <- val
        }
      }
    }
    result
  }

  parsed <- lapply(df$attributes, parse_attrs)
  attr_names <- unique(unlist(lapply(parsed, names)))

  # Skip attribute names that conflict with reserved GRanges column names
  reserved <- c("seqnames", "ranges", "strand", "seqlevels", "seqlengths",
                "isCircular", "start", "end", "width", "element", "type",
                "source", "score", "phase")
  attr_names <- setdiff(attr_names, reserved)

  gr <- GRanges(
    seqnames = df$seqid,
    ranges   = IRanges(start = df$start, end = df$end),
    strand   = df$strand
  )
  mcols(gr)$type <- df$type
  mcols(gr)$source <- df$source

  # Add parsed attributes as mcols
  for (an in attr_names) {
    mcols(gr)[[an]] <- sapply(parsed, function(x) {
      v <- x[[an]]
      if (is.null(v)) NA_character_ else v
    })
  }

  gr
}
```

# Helper to build per-species annotation set

Trying to maintain annotation logic used in the original WGBS context annotations (e.g., in D-Apul/code/08...):

- Exons and introns are derived from the gene GTF
- UTRs are loaded from the corrected GFFs
- TEs from the RepeatMasker `.out`
- intergenic is computed as the genome minus all annotated features
- TEs are further split into `TE_exonic`, `TE_intronic`, and `TE_intergenic` by overlapping with the exon/intron/intergenic partitions

```{r helper-build-annotation}
build_annotation <- function(sp_row) {
  sp   <- sp_row$species
  disp <- sp_row$display
  message(">>> Building annotation for ", sp, " (", disp, ")")

  # Genome sizes
  chr_sizes <- read.table(sp_row$genome_fai, header = FALSE, stringsAsFactors = FALSE)[, 1:2]
  colnames(chr_sizes) <- c("seqnames", "seqlengths")
  genome_GR <- GRanges(seqnames = chr_sizes$seqnames,
                       ranges   = IRanges(start = 1, end = chr_sizes$seqlengths))

  # Gene models: exons, transcripts, introns
  # Use the custom reader (handles Peve GTF header lines that gffToGRanges can't)
  gtf.file <- sp_row$gene_gtf
  transcripts <- read_gtf_granges(gtf.file, filter = "transcript")
  exons       <- read_gtf_granges(gtf.file, filter = "exon")
  # Introns = transcript span minus exons (per WGBS convention)
  introns     <- GenomicRanges::setdiff(transcripts, exons)

  # 5' and 3' UTRs
  gff5UTR <- read_gtf_granges(sp_row$utr5_gff)
  gff3UTR <- read_gtf_granges(sp_row$utr3_gff)

  # Transposable elements / repeats (RepeatMasker .out)
  rmsk <- read.table(sp_row$rmsk_out, skip = 3, fill = TRUE,
                     stringsAsFactors = FALSE, quote = "") %>%
    tidyr::drop_na()
  colnames(rmsk)[c(5, 6, 7, 10)] <- c("seqname", "start", "end", "repeat_name")
  TE <- GRanges(seqnames = rmsk$seqname,
                ranges   = IRanges(start = as.numeric(rmsk$start),
                                   end   = as.numeric(rmsk$end)),
                strand   = "*",
                repeat_name = rmsk$repeat_name)

  # miRNAs: mature + precursor (hairpin)
  miRNA_mature <- read_gtf_granges(sp_row$mirna_mature_gff3)
  # Precursor hairpins come from ShortStack Results.gff3 (feature = MIRNA_hairpin)
  miRNA_precursor <- read_gtf_granges(sp_row$mirna_results_gff3,
                                      filter = "MIRNA_hairpin")

  # lncRNAs
  lncRNAs <- read_gtf_granges(sp_row$lncrna_gtf)

  # Intergenic = genome - all annotated features
  all_annotated <- suppressWarnings(GenomicRanges::reduce(c(gff5UTR, gff3UTR, exons, introns,
                                           lncRNAs, miRNA_mature, miRNA_precursor)))
  intergenic <- suppressWarnings(GenomicRanges::setdiff(genome_GR, all_annotated, ignore.strand = TRUE))

  # TE context split (exonic / intronic / intergenic)
  TE_exonic     <- IRanges::subsetByOverlaps(TE, exons,       ignore.strand = TRUE)
  TE_intronic   <- IRanges::subsetByOverlaps(TE, introns,     ignore.strand = TRUE)
  TE_intergenic <- IRanges::subsetByOverlaps(TE, intergenic,  ignore.strand = TRUE)

  # Bundle into a named list
  annot <- list(
    genome         = genome_GR,
    seqlengths     = setNames(chr_sizes$seqlengths, chr_sizes$seqnames),
    transcripts    = transcripts,
    exons          = exons,
    introns        = introns,
    `5UTR`         = gff5UTR,
    `3UTR`         = gff3UTR,
    TE             = TE,
    TE_exonic      = TE_exonic,
    TE_intronic    = TE_intronic,
    TE_intergenic  = TE_intergenic,
    miRNA_mature   = miRNA_mature,
    miRNA_precursor = miRNA_precursor,
    lncRNA         = lncRNAs,
    intergenic     = intergenic
  )

  # Report feature counts
  counts <- sapply(annot, function(x) length(x))
  message("  Feature counts:")
  for (nm in names(counts)) {
    message("    ", nm, ": ", counts[nm])
  }

  annot
}
```

# Build annotations for all three species

```{r build-all}
annotations <- list()
for (i in seq_len(nrow(species_meta))) {
  sp <- species_meta$species[i]
  annotations[[sp]] <- build_annotation(species_meta[i, ])
}

# Summary table of feature counts per species
count_summary <- do.call(rbind, lapply(names(annotations), function(sp) {
  counts <- sapply(annotations[[sp]], length)
  data.frame(species = sp, feature = names(counts), n = as.integer(counts),
             stringsAsFactors = FALSE)
}))

count_summary %>%
  tidyr::pivot_wider(names_from = feature, values_from = n) %>%
  knitr::kable()
```

# Save compiled annotations

Save each species' annotation set as an RDS (for direct reuse in downstream Rmds) and write each feature to a BED file (for inspection / bedtools use). (keep in mind, BED uses 0-based start coordinates)

```{r save-annotations}
gr_to_bed <- function(gr, path) {
  bed <- data.frame(chr   = as.character(seqnames(gr)),
                    start = start(gr) - 1,   # GRanges is 1-based; BED is 0-based
                    end   = end(gr),
                    name  = if (!is.null(gr$repeat_name)) gr$repeat_name else ".",
                    score = ".",
                    strand = as.character(strand(gr)),
                    stringsAsFactors = FALSE)
  # Use a stable name column where available
  if (!is.null(names(gr))) bed$name <- names(gr)
  data.table::fwrite(bed, path, sep = "\t", col.names = FALSE, quote = FALSE)
}

for (sp in names(annotations)) {
  sp_dir <- file.path("../output/22.1-genomic-context-annotations", sp)
  dir.create(sp_dir, showWarnings = FALSE, recursive = TRUE)

  # RDS of the full annotation list
  saveRDS(annotations[[sp]], file.path(sp_dir, paste0(sp, "_genome_annotation.rds")))

  # BED per feature
  bed_features <- c("exons", "introns", "5UTR", "3UTR",
                    "TE", "TE_exonic", "TE_intronic", "TE_intergenic",
                    "miRNA_mature", "miRNA_precursor", "lncRNA", "intergenic")
  for (feat in bed_features) {
    gr <- annotations[[sp]][[feat]]
    if (length(gr) > 0) {
      gr_to_bed(gr, file.path(sp_dir, paste0(sp, "_", feat, ".bed")))
    }
  }
  message("Saved annotation files for ", sp, " to ", sp_dir)
}
```

# Classify each miRNA as intergenic or intragenic

A miRNA is intragenic if its precursor (hairpin) overlaps any gene transcript body; otherwise it is intergenic. We use the precursor coordinates (not the mature ~22 nt) because the precursor is the relevant proxy unit for genomic context, and better reflects where the pri-miRNA lives relative to gene structure.

```{r classify-mirna}
classify_mirna <- function(sp, annot) {
  prec <- annot$miRNA_precursor
  tx   <- annot$transcripts

  if (length(prec) == 0) {
    return(data.frame(species = sp, miRNA_id = character(0),
                      chr = character(0), start = integer(0), end = integer(0),
                      strand = character(0), genomic_class = character(0),
                      host_gene_id = character(0), stringsAsFactors = FALSE))
  }

  # miRNA IDs from the gff3 ID attribute (ShortStack stores as "Cluster_XXXX")
  # gffToGRanges parses GFF3 attributes into mcols columns
  prec_id <- if ("ID" %in% colnames(mcols(prec))) as.character(prec$ID) else paste0("miRNA_", seq_along(prec))

  # Find overlaps between precursors and transcript bodies
  ol <- findOverlaps(prec, tx, ignore.strand = TRUE)
  intragenic_idx <- unique(queryHits(ol))

  # For each intragenic precursor, record the host gene (first hit)
  # gffToGRanges parses GTF attributes into mcols (gene_id, transcript_id, etc.)
  tx_gene_id <- if ("gene_id" %in% colnames(mcols(tx))) as.character(tx$gene_id) else as.character(seq_along(tx))
  host_map <- tibble::tibble(
    prec_idx   = queryHits(ol),
    host_gene  = tx_gene_id[subjectHits(ol)]
  ) %>%
    dplyr::group_by(prec_idx) %>%
    dplyr::summarise(host_gene_id = paste(unique(host_gene), collapse = ";"),
                     .groups = "drop")

  class_df <- data.frame(
    species     = sp,
    miRNA_id    = prec_id,
    chr         = as.character(seqnames(prec)),
    start       = start(prec),
    end         = end(prec),
    strand      = as.character(strand(prec)),
    stringsAsFactors = FALSE
  )
  class_df$genomic_class <- ifelse(seq_len(nrow(class_df)) %in% intragenic_idx,
                                   "intragenic", "intergenic")

  # Attach host gene IDs for intragenic miRNAs
  class_df$host_gene_id <- NA_character_
  if (nrow(host_map) > 0) {
    class_df$host_gene_id[host_map$prec_idx] <- host_map$host_gene_id
  }

  class_df
}

mirna_class <- do.call(rbind, lapply(names(annotations),
                                     function(sp) classify_mirna(sp, annotations[[sp]])))

fwrite(mirna_class, file.path("../output/22.1-genomic-context-annotations", "miRNA_genomic_class.csv"))

cat("\nmiRNA genomic classification:\n")
mirna_class %>%
  group_by(species, genomic_class) %>%
  summarise(n = n(), .groups = "drop") 
```

# Define promoter regions

Promoters are defined strand-aware as 1500 bp upstream and 400 bp downstream of the TSS. The TSS is the 5' end of the feature:

- `+` strand: TSS = `start`; promoter = `[start - 1500, start + 400]`
- `-` strand: TSS = `end`;   promoter = `[end - 400, end + 1500]`

All coordinates are clamped to `[1, chromosome_length]` to avoid running off chromosome ends.

## Helper: TSS and promoter from a GRanges

```{r promoter-helper}
#' Build promoter GRanges from a set of feature coordinates
#'
#' @param gr GRanges with strand info (features whose TSS defines the promoter)
#' @param upstream bp upstream of TSS (on the feature's strand)
#' @param downstream bp downstream of TSS (on the feature's strand)
#' @param seqlengths named integer vector of chromosome lengths, for clamping
make_promoters <- function(gr, upstream = PROM_UPSTREAM, downstream = PROM_DOWNSTREAM,
                           seqlengths = NULL) {
  # promoters() from GenomicRanges is strand-aware by default:
  #   + strand: [start - upstream, start + downstream - 1]
  #   - strand: [end   - downstream + 1, end + upstream]
  prom <- promoters(gr, upstream = upstream, downstream = downstream)

  # Clamp to chromosome bounds (start >= 1)
  start(prom) <- pmax(start(prom), 1L)

  # Clamp end to seqlengths if provided
  if (!is.null(seqlengths)) {
    sl <- seqlengths[as.character(seqnames(prom))]
    end(prom) <- pmin(end(prom), as.integer(sl))
  }
  prom
}
```

## lncRNA promoters

TSS = 5' end of each lncRNA transcript (from the lncRNA GTF).

```{r lncrna-promoters}
lncrna_promoters <- list()

for (sp in names(annotations)) {
  annot <- annotations[[sp]]
  lnc   <- annot$lncRNA
  if (length(lnc) == 0) next

  # Use lncRNA gene_id as the name (gffToGRanges parses GTF attributes into mcols)
  lnc_names <- if ("gene_id" %in% colnames(mcols(lnc))) as.character(lnc$gene_id) else paste0("lncRNA_", seq_along(lnc))

  prom <- make_promoters(lnc,
                         upstream   = PROM_UPSTREAM,
                         downstream = PROM_DOWNSTREAM,
                         seqlengths = annot$seqlengths)
  # Strip inherited mcols for consistent structure across species
  mcols(prom) <- NULL
  names(prom) <- lnc_names
  prom$feature_id <- lnc_names
  prom$feature_type <- "lncRNA"
  prom$species <- sp

  lncrna_promoters[[sp]] <- prom

  sp_dir <- file.path("../output/22.1-genomic-context-annotations", sp)
  gr_to_bed(prom, file.path(sp_dir, paste0(sp, "_lncRNA_promoters.bed")))
  message("  ", sp, ": ", length(prom), " lncRNA promoters defined")
}

lncrna_promoters_all <- do.call(c, unname(lncrna_promoters))
```

## miRNA promoters

TSS depends on genomic class:

- Intergenic miRNA: TSS = 5' end of the precursor (hairpin) itself - the precursor is the transcribed unit.
- Intragenic miRNA: TSS = 5' end of the host gene it sits inside - the miRNA is processed from an intron (or exon) of the host transcript, so the host's TSS is the relevant regulatory point.

```{r mirna-promoters}
mirna_promoters <- list()

for (sp in names(annotations)) {
  annot   <- annotations[[sp]]
  prec    <- annot$miRNA_precursor
  tx      <- annot$transcripts
  class_df <- mirna_class %>% dplyr::filter(species == sp)

  if (length(prec) == 0) next

  prec_id <- if ("ID" %in% colnames(mcols(prec))) as.character(prec$ID) else paste0("miRNA_", seq_along(prec))

  # For each precursor, decide which TSS to use based on genomic class
  prom_list <- list()
  for (i in seq_along(prec)) {
    pid   <- prec_id[i]
    p_gr  <- prec[i]
    row   <- class_df %>% dplyr::filter(miRNA_id == pid)

    if (nrow(row) == 0) next
    gclass <- row$genomic_class[1]

    if (gclass == "intragenic") {
      # Use the host gene's TSS. Find the overlapping transcript.
      ol <- findOverlaps(p_gr, tx, ignore.strand = TRUE)
      if (length(ol) == 0) next
      host_idx <- subjectHits(ol)[1]
      host_tx  <- tx[host_idx]
      tss_gr   <- host_tx
      # Use gene_id from mcols (gffToGRanges parses GTF attributes)
      host_id  <- if ("gene_id" %in% colnames(mcols(tx))) as.character(tx$gene_id[host_idx]) else as.character(host_idx)
    } else {
      # Intergenic: use the precursor's own 5' end
      tss_gr  <- p_gr
      host_id <- NA_character_
    }

    prom <- make_promoters(tss_gr,
                           upstream   = PROM_UPSTREAM,
                           downstream = PROM_DOWNSTREAM,
                           seqlengths = annot$seqlengths)
    # Strip inherited mcols so all promoters have a consistent structure
    # (intragenic promoters come from transcripts, intergenic from precursors -
    # they carry different attribute columns that would break c())
    mcols(prom) <- NULL
    prom$feature_id   <- pid
    prom$feature_type <- "miRNA"
    prom$genomic_class <- gclass
    prom$host_gene_id  <- host_id
    prom$species       <- sp
    prom_list[[pid]] <- prom
  }

  if (length(prom_list) > 0) {
    sp_prom <- do.call(c, unname(prom_list))
    mirna_promoters[[sp]] <- sp_prom

    sp_dir <- file.path("../output/22.1-genomic-context-annotations", sp)
    gr_to_bed(sp_prom, file.path(sp_dir, paste0(sp, "_miRNA_promoters.bed")))
    message("  ", sp, ": ", length(sp_prom), " miRNA promoters defined")
  }
}

mirna_promoters_all <- do.call(c, unname(mirna_promoters))
```

# Compiled promoter table

A tidy table of every promoter defined above, with the source feature, genomic class, host gene (for intragenic miRNAs), and promoter coordinates.

```{r promoter-table}
promoter_table <- do.call(rbind, lapply(names(annotations), function(sp) {
  rows <- list()
  if (sp %in% names(lncrna_promoters)) {
    p <- lncrna_promoters[[sp]]
    rows[[length(rows) + 1]] <- data.frame(
      species       = sp,
      feature_type  = "lncRNA",
      feature_id    = p$feature_id,
      genomic_class = "lncRNA",
      host_gene_id  = NA_character_,
      chr           = as.character(seqnames(p)),
      promoter_start = start(p),
      promoter_end   = end(p),
      strand         = as.character(strand(p)),
      stringsAsFactors = FALSE
    )
  }
  if (sp %in% names(mirna_promoters)) {
    p <- mirna_promoters[[sp]]
    rows[[length(rows) + 1]] <- data.frame(
      species       = sp,
      feature_type  = "miRNA",
      feature_id    = p$feature_id,
      genomic_class = p$genomic_class,
      host_gene_id  = p$host_gene_id,
      chr           = as.character(seqnames(p)),
      promoter_start = start(p),
      promoter_end   = end(p),
      strand         = as.character(strand(p)),
      stringsAsFactors = FALSE
    )
  }
  do.call(rbind, rows)
}))

fwrite(promoter_table, file.path("../output/22.1-genomic-context-annotations", "promoter_regions.csv"))

cat("\nPromoter counts by species and feature type:\n")
promoter_table %>%
  group_by(species, feature_type, genomic_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  knitr::kable()
```

# Summary

```{r summary}
cat("Genomic context annotation files written to:\n  ", normalizePath("../output/22.1-genomic-context-annotations"), "\n\n")

cat("Per-species feature counts:\n")
count_summary %>%
  filter(!feature %in% c("genome")) %>%
  pivot_wider(names_from = feature, values_from = n) %>%
  knitr::kable()

cat("\nmiRNA genomic classification:\n")
mirna_class %>%
  count(species, genomic_class, name = "n_miRNAs") %>%
  knitr::kable()

cat("\nPromoter regions defined:\n")
promoter_table %>%
  count(species, feature_type, name = "n_promoters") %>%
  knitr::kable()
```

## Output files

For each species (`Apul`, `Peve`, `Ptuh`), the following are written to `M-multi-species/output/22.1-genomic-context-annotations/<species>/`:

- `<species>_genome_annotation.rds` - RDS of the full `GRanges` annotation list (for direct loading in downstream Rmds)
- `<species>_exons.bed`, `<species>_introns.bed` - gene-model partitions
- `<species>_5UTR.bed`, `<species>_3UTR.bed` - UTR coordinates
- `<species>_TE.bed` - all RepeatMasker intervals
- `<species>_TE_exonic.bed`, `<species>_TE_intronic.bed`, `<species>_TE_intergenic.bed` - TE by gene-context
- `<species>_miRNA_mature.bed`, `<species>_miRNA_precursor.bed` - ShortStack miRNA coordinates
- `<species>_lncRNA.bed` - lncRNA transcript coordinates
- `<species>_intergenic.bed` - genome minus all annotated features
- `<species>_lncRNA_promoters.bed` - lncRNA promoter regions (TSS -1500 / +400)
- `<species>_miRNA_promoters.bed` - miRNA promoter regions (TSS -1500 / +400)

And at the top level of `M-multi-species/output/22.1-genomic-context-annotations/`:

- `miRNA_genomic_class.csv` - intergenic vs intragenic classification per miRNA, with host gene IDs
- `promoter_regions.csv` - tidy table of all promoter coordinates

# Session info

```{r}
sessionInfo()
```
