---
title: "Apul MixOmics"
output: html_document
date: "2024-10-01"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Script to analyzed relationships between RNA-sRNA-WGBS using mixOmics.  

# Set up 

Load libraries 
```{r}
library(tidyverse)
library(ggplot2)
library(mixOmics)
```

# Read in files 

Read in count matrices.  

mrna is missing sample 150, removing from all 
```{r}
mrna<-read.csv("D-Apul/output/07-Apul-Hisat/Apul-gene_count_matrix.csv")%>%rename("sample140"=RNA.ACR.140, "sample145"=RNA.ACR.145, "sample173"=RNA.ACR.173, "sample178"=RNA.ACR.178)

rownames(mrna)<-mrna$gene_id

mrna<-mrna%>%dplyr::select(!gene_id)

srna<-read.table("D-Apul/output/03.1-Apul-sRNA-summary/Apul_counts_sRNA_normalized.txt")%>%dplyr::select(!sample150)
sirna<-read.table("D-Apul/output/03.1-Apul-sRNA-summary/Apul_counts_siRNA_normalized.txt")%>%dplyr::select(!sample150)

metadata <- data.frame(
  sample = c("sample145", "sample140", "sample173", "sample178"),
  species = "Apul"
)
```

Follow the mixOmics tutorial for conducting correlations between these two data sets. 

All data sets have dimensions of 4 samples. We can add additional species to this as we go.  

# Conduct mixOmics  

Set matrices. 

```{r}
mrna<-t(mrna)
srna<-t(srna)
sirna<-t(sirna)

dim(mrna)
dim(srna)
dim(sirna)
dim(metadata)
```

Remove columns with 0's in each matrix or with 0 variance (all values identical) 

```{r}
mrna <- mrna[, colSums(mrna) != 0]

mrna<-as.data.frame(mrna)
mrna <- mrna[, sapply(mrna, function(x) length(unique(x)) > 1)]
dim(mrna)

srna <- srna[, colSums(srna) != 0]
dim(srna)

sirna <- sirna[, colSums(sirna) != 0]
dim(sirna)
```


## Conduct preliminary PCA  

```{r}
pca.mrna <- pca(mrna, ncomp = 3, center = TRUE, scale = TRUE)
pca.srna <- pca(srna, ncomp = 3, center = TRUE, scale = TRUE)
pca.sirna <- pca(sirna, ncomp = 3, center = TRUE, scale = TRUE)

plot(pca.mrna)
plot(pca.srna)
plot(pca.sirna)
```

Similar variance explained because we are dealing with individual samples within a species. 

```{r}
plotIndiv(pca.mrna, comp = c(1, 2), 
          group = metadata$species, 
          ind.names = metadata$sample, 
          legend = TRUE, title = 'mRNA, PCA comp 1 - 2')

plotIndiv(pca.srna, comp = c(1, 2), 
          group = metadata$species, 
          ind.names = metadata$sample, 
          legend = TRUE, title = 'sRNA, PCA comp 1 - 2')

plotIndiv(pca.sirna, comp = c(1, 2), 
          group = metadata$species, 
          ind.names = metadata$sample, 
          legend = TRUE, title = 'siRNA, PCA comp 1 - 2')
```

# Multi block PLS with DIABLO 

```{r}
X1 <- mrna
X2 <- srna 
X3 <- sirna
X <- list(mrna = X1, srna = X2, sirna = X3)

Y <-metadata$sample
```

```{r}
result.diablo <- block.plsda(X, Y) # run the method
plotIndiv(result.diablo) # plot the samples
plotVar(result.diablo) # plot the variables
```

sparse 
```{r}
list.keepX = list(mrna = c(50, 4), srna = c(50,4), sirna = c(50, 4)) 

# run the method
result.sparse.diablo <-  block.splsda(X, Y, keepX = list.keepX) 

# plot the contributions of each feature to each dimension
plotLoadings(result.sparse.diablo, ncomp = 1) 
plotIndiv(result.sparse.diablo) # plot the samples
plotVar(result.sparse.diablo) # plot the variables
```

```{r}
color.edge <- color.GreenRed(50)  # set the colours of the connecting lines

network(result.sparse.diablo, 
        cutoff = 0.99, # only show connections with a correlation above threshold
        shape.node = c("rectangle", "circle"),
        color.node = c("cyan", "pink"),
        color.edge = color.edge, 
        interactive = FALSE,
        lwd.edge = 2)
```

```{r}
circosPlot(result.sparse.diablo, cutoff=0.90)
```
