---
title: "mirTarRnaSeq"
author: "Jill Ashey"
date: "2024-10-01"
output: html_document
---

This script will use the R package [mirTarRnaSeq](https://bioconductor.org/packages/3.13/bioc/html/mirTarRnaSeq.html), which an be used for interactive mRNA miRNA sequencing statistical analysis, to assess coexpression between mRNAs and miRNAs. 

I will be using Part 1 of their workflow (see [vignette](https://bioconductor.org/packages/3.13/bioc/vignettes/mirTarRnaSeq/inst/doc/mirTarRnaSeq.pdf)).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(mirTarRnaSeq)
```

Using the test data as an initial pass to understand Part 1 of the package. Part 1 

I will use the `miRNA` df as miRNA counts input and `mRNA` df as mRNA counts input; this data is a part of the package. I will also use `miRanda` for interaction data. 

Select only mRNAs that were identified as targets 
```{r}
mRNASub <- miRanComp(mRNA, miRanda)
```

Select specific miRNA. Not sure why this needs to be done...
```{r}
miRNA_select<-rownames(miRNA)
```

Combine miRNA and mRNA files; make list of files to be used in model 
```{r}
combine <- combiner(mRNASub, miRNA, miRNA_select)
gene_var <- geneVari(combine, miRNA_select)
```

Run Gausian model 
```{r}
test <- runModels(combine, geneVariant, miRNA_select, family = glm_gaussian(), scale = 100)

par(oma=c(2,2,2,2))
par(mfrow=c(2,3),mar=c(4,3,3,2))
plot(test[["all_models"]][["BHLF1"]])
plot(modelData(test[["all_models"]][["BHLF1"]]))
#To test AIC model performance
G <- do.call(rbind.data.frame, test[["AICvalues"]])
names(G) <- c("AIC_G")
#Low values seems like a reasonable model
plot(density(G$AIC_G))
```

Run all combos
```{r}
vmiRNA <- rownames(miRNA)

all_miRNA_models <- runAllMirnaModels(mirnas = vmiRNA[1:6], DiffExpmRNA = mRNA, DiffExpmiRNA = miRNA, miranda_data = miRanda, prob = 0.75, cutoff = 0.05, fdr_cutoff = 0.1, method = "fdr", family = glm_multi(), scale = 2, mode = "multi")

hasgenes <- lapply(all_miRNA_models, function(x) nrow(x$SigFDRGenes)) > 0
all_miRNA_models <- all_miRNA_models[hasgenes]
```

Look at data 
```{r}
all_miRNA_models$`ebv-mir-bart1-3p and ebv-mir-bart1-5p`
```

Need to play with test data more 

