---
title: "mRNA-miRNA interactions functional enrichment"
author: "Kathleen Durkin"
date: "2025-07-09"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: ../../references.bib
link-citations: true
---

This script will use topGO to analyze functional enrichment of miRNA targets for Apul in the 5UTR and CDS regions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(topGO)
library(tidyverse)
```

Code used below was created by `Jill Ashey`, modified for use with A.pulchra datasets by `Kathleen Durkin`

# Format topGO files

## Read in and format annotation files

```{r}
# Read in Apul annotations
annot_locations <- read.delim("../output/02-Apul-reference-annotation/Apulcra-genome-mRNA-IDmapping-2024_12_12.tab")
# Remove unneeded columns 
annot_locations <- annot_locations %>% dplyr::select(-X, -V13)
# Ensure there are no duplicate rows
annot_locations <- annot_locations %>% distinct()

head(annot_locations)
# Looks good!
```

This file shows each gene as it's genomic location. We want to use gene IDs to associate genes, so add gene IDs to this annotation table

Read in file that associates each mRNA genomic location with corresponding gene ID
```{r}
mRNA_FUNids <- read.table("../output/15-Apul-annotate-UTRs/Apul-mRNA-FUNids.txt", header=FALSE, col.names=c("location", "type", "mRNA_ID", "gene_ID", "product"), sep="\t")

# Remove unwanted text from parent column
mRNA_FUNids$gene_ID <- gsub("Parent=", "", mRNA_FUNids$gene_ID)
# Only need to keep mRNA location and gene ID
mRNA_FUNids <- mRNA_FUNids %>% dplyr::select(location, gene_ID)
```

join with annotation file
```{r}
# join
annot <- left_join(annot_locations, mRNA_FUNids, by = c("V1" = "location"))

# ensure there are no duplicate rows
annot <- annot %>% distinct()
```


## Set up gene2GO object

Want to isolate a list of GO terms per gene
```{r}
gene2go <- annot %>% filter(!is.na(Gene.Ontology.IDs)) %>% dplyr::select(gene_ID, Gene.Ontology.IDs)
gene2go <- gene2go %>% dplyr::rename(GO.ID = Gene.Ontology.IDs)

gene2go_list <- setNames(
  strsplit(as.character(gene2go$GO.ID), ";"), 
  gene2go$gene_ID
)
```

Note: I think this means genes that had a Uniprot ID but no GO terms are excluded from this analysis

## Define reference set

Define reference set of genes. This should be all genes *found in our samples*, NOT all genes in the A.pulchra genome. Some genes (e.g., reproduction pathways) may not be found/expected in our samples for valid biological reasons.
```{r}
# Read in counts matrix
Apul_counts <- read.csv("../output/07-Apul-Hisat/Apul-gene_count_matrix.csv")
# Exclude genes with all 0 counts
Apul_counts <- Apul_counts[rowSums(Apul_counts[, 2:6]) != 0, ]

# Select gene IDs of the genes present in our samples
all_genes <- Apul_counts$gene_id
length(all_genes)
```
So we have a reference set of 33624 genes present in our samples.

# CDS

## Read in PCC/miranda data

This is a table of all putative miRNA-mRNA binding predicted by miRanda, plus Pearsons correlation coefficients for coexpression of each putative binding pair.
```{r}
data <- read.csv("../output/09.01-Apul-mRNA-miRNA-interactions-CDS_5UTR/miRanda_PCC_miRNA_CDS.csv")
data <- rename(data,
               miRNA = V1,
               mRNA = V4.y)
head(data)
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```


# FA of all miRNA targets

Functional annotation of all putative miRNA targets
```{r}
cor_bind_FA <- left_join(data, annot, by = c("mRNA" = "gene_ID")) %>% distinct()

nrow(cor_bind_FA)
nrow(cor_bind_FA[!is.na(cor_bind_FA$Gene.Ontology.IDs),])
```

Of the 38152 putative miRNA targets (CDS region) predicted by miRanda, 12181 have available annotations

```{r}
sig_cor_bind_FA <- cor_bind_FA[cor_bind_FA$p_value < 0.05,]

nrow(sig_cor_bind_FA)
nrow(sig_cor_bind_FA[!is.na(sig_cor_bind_FA$Gene.Ontology.IDs),])
```

Of the 4075 putative CDS miRNA targets predicted by miRanda that are also have significantly correlated expression, only 540 have available annotations. This is an average of 7 annotated targets per miRNA. This unfortunately means **functional enrichment analysis will likely be uninformative for most groups of significantly coexpressed targets.**

Save
```{r}
write.csv(cor_bind_FA, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_targets_FA.csv")
write.csv(sig_cor_bind_FA, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_sig_cor_targets_FA.csv")
```



# FE of specific miRNA's targets (all targets)

Create topGO function for use with miRNA names
```{r}

miRNA_topGO_FE <- function(miRNA.name, input_interactions) {
  
  #Isolate genes in our input module of interest
  interacting_genes <- input_interactions %>%
    filter(miRNA == miRNA.name) %>%
    pull(mRNA)
  
  if (length(interacting_genes) > 0) {
    # Create factor for all reference genes, where 1 represents module membership and 0 means the gene is not in module of interest
    gene_list <- factor(as.integer(all_genes %in% interacting_genes))
    names(gene_list) <- all_genes
    str(gene_list)
    
    ## Biological Process ##
    # Create topGO object
    GO_BP <- new("topGOdata", 
                ontology = "BP", # Biological Process
                allGenes = gene_list,
                annot = annFUN.gene2GO, 
                gene2GO = gene2go_list,
                geneSel=topDiffGenes)
    
    # Run GO enrichment test
    GO_BP_FE <- runTest(GO_BP, algorithm = "weight01", statistic = "fisher")
    # View the results
    GO_BP_results <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
    # Filter by significant results
    GO_BP_results$Fisher<-as.numeric(GO_BP_results$Fisher)
    GO_BP_results_sig<-GO_BP_results[GO_BP_results$Fisher<0.05,]
    
    
    ## Molecular Function ##
    # Create topGO object
    GO_MF <- new("topGOdata", 
                ontology = "MF", # Molecular Function
                allGenes = gene_list,
                annot = annFUN.gene2GO, 
                gene2GO = gene2go_list,
                geneSel=topDiffGenes)
    
    # Run GO enrichment test
    GO_MF_FE <- runTest(GO_MF, algorithm = "weight01", statistic = "fisher")
    # View the results
    GO_MF_results <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
    # Filter by significant results
    GO_MF_results$Fisher<-as.numeric(GO_MF_results$Fisher)
    GO_MF_results_sig<-GO_MF_results[GO_MF_results$Fisher<0.05,]
  
    # Return
    # Add type column only if results exist
    if (nrow(GO_BP_results_sig) > 0) {
      GO_BP_results_sig$type <- "Biological.Process"
    }
    if (nrow(GO_MF_results_sig) > 0) {
      GO_MF_results_sig$type <- "Molecular.Function"
    }
    GO_results <- rbind(GO_BP_results_sig, GO_MF_results_sig)
    print(GO_results)
  }
}

miRNA_topGO_FE("Cluster_10051", cor_bind_FA)
```

Loop through all miRNA and run functional enrichment on the miRNA's targets (all predicted targets)
```{r}

interacting_miRNAs <- unique(cor_bind_FA$miRNA) %>% na.omit
results_all_targets <- NULL  # initialize empty df

for(miRNA in interacting_miRNAs) {
  
  # Run topGO enrichment function
  miRNA_results <- miRNA_topGO_FE(miRNA, cor_bind_FA)
  
  # Only keep results if not empty
  if (nrow(miRNA_results) > 0) {
    
    # Add the miRNA source column
    miRNA_results$miRNA <- miRNA

    # Bind to the accumulating results data frame
    results_all_targets <- rbind(results_all_targets, miRNA_results)
  }
}

head(results_all_targets)

```


Save results
```{r}
write.csv(results_all_targets, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_all_targets_topGO_FE.csv")
```


# FE of specific miRNA's targets (significant cor targets)

Loop through all miRNA and run functional enrichment on the miRNA's significantly correlated targets
```{r}

interacting_miRNAs_sig <- unique(sig_cor_bind_FA$miRNA) %>% na.omit
results_sig_cor_targets <- NULL  # initialize empty df

for(miRNA in interacting_miRNAs_sig) {
  
  # Run topGO enrichment function
  miRNA_results <- miRNA_topGO_FE(miRNA, sig_cor_bind_FA)
  
  # Only keep results if not empty
  if (nrow(miRNA_results) > 0) {
    
    # Add the miRNA source column
    miRNA_results$miRNA <- miRNA

    # Bind to the accumulating results data frame
    results_sig_cor_targets <- rbind(results_sig_cor_targets, miRNA_results)
  }
}

head(results_sig_cor_targets)

```


Save results
```{r}
write.csv(results_sig_cor_targets, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_sig_cor_targets_topGO_FE.csv")
```

# FE of all targets negatively correlated with miRNAs (regardless of correlation significance)

Filter PCC miranda data
```{r}
# Filter so that only negative correlations remain 
neg_corr_data <- data %>%
  filter(PCC.cor < 0)
```

Make list of target genes for input to topGO
```{r}
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(neg_corr_data$mRNA))
```

```{r}
# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs, regardless of correlation significance. 

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", 
            ontology="BP", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
neg_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
neg_cor_gene2go$GO.ID <- trimws(neg_cor_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- neg_cor_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- GO_BP_En_sig_gene %>% distinct()
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", 
            ontology="MF", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
neg_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
neg_cor_gene2go$GO.ID <- trimws(neg_cor_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- neg_cor_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_neg_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_neg_corr_df`
```{r}
test <- GO_neg_corr_df %>%
  inner_join(neg_corr_data, by = c("gene_ID" = "mRNA")) #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_neg_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_neg_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_neg_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_top10GO_neg_corr_target_enrichment.pdf", plot, width = 20, height = 25, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_top10GO_neg_corr_target_enrichment.png", plot, width = 20, height = 25, dpi = 300)
```


# FE of all targets positively correlated with miRNAs (regardless of correlation significance)

Filter PCC miranda data
```{r}
# Filter so that only positive correlations remain 
pos_corr_data <- data %>%
  filter(PCC.cor > 0)

# Make list of target genes for input to topGO
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(pos_corr_data$mRNA))

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs with positively correlated expression, regardless of correlation significance. 


### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", 
            ontology="BP", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
pos_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
pos_cor_gene2go$GO.ID <- trimws(pos_cor_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- pos_cor_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", 
            ontology="MF", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
pos_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
pos_cor_gene2go$GO.ID <- trimws(pos_cor_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- pos_cor_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_pos_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_pos_corr_df`
```{r}
test <- GO_pos_corr_df %>%
  inner_join(pos_corr_data, by = c("gene_ID" = "mRNA")) #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_pos_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_pos_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_pos_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_top10GO_pos_corr_target_enrichment.pdf", plot, width = 20, height = 25, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_top10GO_pos_corr_target_enrichment.png", plot, width = 20, height = 25, dpi = 300)
```



# FE of all targets significantly negatively correlated with miRNAs

Filter PCC miranda data
```{r}
# Filter so that only significant negative correlations remain 
sig_neg_corr_data <- data %>%
  filter(PCC.cor < 0) %>%
  filter(p_value < 0.05)

# Make list of target genes for input to topGO
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(sig_neg_corr_data$mRNA))

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```


The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs with significantly negatively correlated expression.

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", 
            ontology="BP", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
sig_neg_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
sig_neg_cor_gene2go$GO.ID <- trimws(sig_neg_cor_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- sig_neg_cor_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", 
            ontology="MF", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
sig_neg_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
sig_neg_cor_gene2go$GO.ID <- trimws(sig_neg_cor_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- sig_neg_cor_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_sig_neg_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_sig_neg_corr_df`
```{r}
test <- GO_sig_neg_corr_df %>%
  inner_join(sig_neg_corr_data, by = c("gene_ID" = "mRNA")) #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_sig_neg_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_sig_neg_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_top10GO_sig_neg_corr_target_enrichment.pdf", plot, width = 20, height = 25, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_top10GO_sig_neg_corr_target_enrichment.png", plot, width = 20, height = 25, dpi = 300)
```


# FE of all targets significantly positively correlated with miRNAs

Filter PCC miranda data
```{r}
# Filter so that only significant negative correlations remain 
sig_pos_corr_data <- data %>%
  filter(PCC.cor > 0) %>%
  filter(p_value < 0.05)

# Make list of target genes for input to topGO
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(sig_neg_corr_data$mRNA))

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs with significantly positively correlated expression.

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", 
            ontology="BP", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
sig_pos_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
sig_pos_cor_gene2go$GO.ID <- trimws(sig_pos_cor_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- sig_pos_cor_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", 
            ontology="MF", 
            gene2GO=gene2go_list, 
            allGenes=GeneList, 
            annot = annFUN.gene2GO, 
            geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
sig_pos_cor_gene2go <- gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
sig_pos_cor_gene2go$GO.ID <- trimws(sig_pos_cor_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- sig_pos_cor_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_sig_pos_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_pos_corr_df`
```{r}
test <- GO_sig_pos_corr_df %>%
  inner_join(sig_pos_corr_data, by = c("gene_ID" = "mRNA")) #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_sig_pos_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_sig_pos_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/miRNA_CDS_topGO_sig_pos_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/top10GO_sig_pos_corr_target_enrichment.pdf", last_plot(), width = 20, height = 25, dpi = 300)
ggsave("../output/09.11-Apul-mRNA-miRNA-interactions-FE-CDS_5UTR/top10GO_sig_pos_corr_target_enrichment.png", last_plot(), width = 20, height = 25, dpi = 300)
```

