---
title: "10-Apul lncRNA"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`"  
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```



# Run HiSat on RNA-seq

Will end up with 5 sorted bam files. 

## Grab Trimmed RNA-seq Reads

```{r, engine='bash'}
cd ../data/fastq/
echo "*.fq" >> .gitignore
```

```{r, engine='bash'}
wget -r \
--no-directories --no-parent \
-P ../data/fastq/ \
-A "*fastq.gz" https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/

```




## Genome

```{r, engine='bash'}
cd ../data

wget -O Apulcra-genome.fa "https://osf.io/download/kn96u/"
```


```{r, engine='bash'}
head ../data/Apulcra-genome.fa
```





## HiSat

```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2_extract_exons.py \
../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
> ../output/05.32-lncRNA-discovery/m_exon.tab
```

```{r, engine='bash', eval=TRUE}
head ../output/05.32-lncRNA-discovery/m_exon.tab
wc -l ../output/05.32-lncRNA-discovery/m_exon.tab
```

```{r, engine='bash'}

/home/shared/hisat2-2.2.1/hisat2_extract_splice_sites.py \
../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
> ../output/05.32-lncRNA-discovery/m_splice_sites.tab

```


```{r, engine='bash', eval=TRUE}
head ../output/05.32-lncRNA-discovery/m_splice_sites.tab
```


NO GTF 




```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2-build \
../data/GCF_013753865.1_Amil_v2.1_genomic.fna \
../output/05.32-lncRNA-discovery/GCF_013753865.1_Amil_v2.1.index \
--exon ../output/05.32-lncRNA-discovery/m_exon.tab \
--ss ../output/05.32-lncRNA-discovery/m_splice_sites.tab \
-p 24 \
../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
2> ../output/05.32-lncRNA-discovery/hisat2-build_stats.txt
```


RUN THIS CODE 




```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2-build \
../data/Apulcra-genome.fa \
../output/07-Apul-Hisat/Apulcra-genome.index \
#--exon ../output/05.32-lncRNA-discovery/m_exon.tab \
#--ss ../output/05.32-lncRNA-discovery/m_splice_sites.tab \
-p 24 \
#../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
2> ../output/07-Apul-Hisat/hisat2-build_stats.txt
```

```{r, engine='bash', eval=TRUE}
cat ../output/07-Apul-Hisat/hisat2-build_stats.txt
```


```{r, engine='bash', eval=TRUE}
find ../data/fastq/*gz
```

```{r, engine='bash', eval=TRUE}
find ../data/fastq/*R2_001.fastp-trim.20230519.fastq.gz \
| xargs basename -s -S1-TP2_R2_001.fastp-trim.20230519.fastq.gz | xargs -I{} \
echo {}
```


```{r, engine='bash'}
cd ../output/07-Apul-Hisat/
echo "*sam" >> .gitignore
```


```{r, engine='bash'}
find ../data/fastq/*R2_001.fastp-trim.20230519.fastq.gz \
| xargs basename -s -S1-TP2_R2_001.fastp-trim.20230519.fastq.gz | xargs -I{} \
/home/shared/hisat2-2.2.1/hisat2 \
-x ../output/07-Apul-Hisat/Apulcra-genome.index \
-p 48 \
-1 ../data/fastq/{}-S1-TP2_R1_001.fastp-trim.20230519.fastq.gz \
-2 ../data/fastq/{}-S1-TP2_R2_001.fastp-trim.20230519.fastq.gz \
-S ../output/07-Apul-Hisat/{}.sam \
2> ../output/07-Apul-Hisat/hisat.out
```

```{r, engine='bash', eval=TRUE}
cat ../output/07-Apul-Hisat/hisat.out
```

## convert to bams

```{r, engine='bash'}
for samfile in ../output/07-Apul-Hisat/*.sam; do
  bamfile="${samfile%.sam}.bam"
  sorted_bamfile="${samfile%.sam}.sorted.bam"
  
  # Convert SAM to BAM
  /home/shared/samtools-1.12/samtools view -bS -@ 20 "$samfile" > "$bamfile"
  
  # Sort BAM
  /home/shared/samtools-1.12/samtools sort -@ 20 "$bamfile" -o "$sorted_bamfile"
  
  # Index sorted BAM
  /home/shared/samtools-1.12/samtools index -@ 20 "$sorted_bamfile"
done
```





# StringTie
StringTie uses the sorted BAM files to assemble transcripts for each sample, outputting them as GTF (Gene Transfer Format) files. And then merges all individual GTF assemblies into a single merged GTF file. This step extracts transcript information and merges GTFs from all samples--an important step in creating a canonical list of lncRNAs across all samples included in the pipeline.

Getting gff

```{r, engine='bash'}

cd ../data
wget -O Apulcra-genome.gff "https://osf.io/download/f9dbr/"
```


```{r, engine='bash'}

head ../data/Apulcra-genome.gff

```

```{r, engine='bash'}
find ../output/07-Apul-Hisat/*sorted.bam \
| xargs basename -s .sorted.bam | xargs -I{} \
/home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
-p 32 \
-eB \
-G ../data/Apulcra-genome.gff \
-o ../output/07-Apul-Hisat/{}.gtf \
../output/07-Apul-Hisat/{}.sorted.bam
```


```{r, engine='bash', eval=TRUE}
wc -l ../output/07-Apul-Hisat/RNA*.gtf
ls ../output/07-Apul-Hisat/RNA*.gtf
#head ../output/07-Apul-Hisat/RNA*.gtf
```



# Count Matrix

list format

RNA-ACR-140 ../output/15-Apul-hisat/RNA-ACR-140.gtf
RNA-ACR-145 ../output/15-Apul-hisat/RNA-ACR-145.gtf
RNA-ACR-173 ../output/15-Apul-hisat/RNA-ACR-173.gtf
RNA-ACR-178 ../output/15-Apul-hisat/RNA-ACR-178.gtf


```{r, engine='bash'}
head ../output/07-Apul-Hisat/list01.txt
```



```{r, engine='bash'}
python /home/shared/stringtie-2.2.1.Linux_x86_64/prepDE.py \
-i ../output/07-Apul-Hisat/list01.txt \
-g ../output/07-Apul-Hisat/gene_count_matrix.csv \
-t ../output/07-Apul-Hisat/transcript_count_matrix.csv

head ../output/07-Apul-Hisat/*matrix.csv

```






# Getting FASTA for anno


$ bedtools getfasta [OPTIONS] -fi <input FASTA> -bed <BED/GFF/VCF>



grep -w "mRNA" annotation.gff > mRNA_only.gff
```{r, engine='bash'}
grep -w "mRNA" ../data/Apulcra-genome.gff > ../data/Apulcra-genome-mRNA_only.gff
```

awk '{if ($0 !~ /^#/ && $3 != "") {feature[$3]++}} END {for (f in feature) print f, feature[f]}' ../data/Apulcra-genome.gff

```{r, engine='bash'}
awk '{if ($0 !~ /^#/ && $3 != "") {feature[$3]++}} END {for (f in feature) print f, feature[f]}' ../data/Apulcra-genome.gff
```



```{r, engine='bash'}
/home/shared/bedtools2/bin/fastaFromBed \
-fi ../data/Apulcra-genome.fa \
-bed ../data/Apulcra-genome-mRNA_only.gff \
-fo ../output/07-Apul-Hisat/genes.fasta \
-name -split
```


```{r, engine='bash'}
head ../output/07-Apul-Hisat/genes.fasta 
grep -c ">" ../output/07-Apul-Hisat/genes.fasta
```


# SP Download
```{r, engine='bash'}

cd ../data/blast_dbs

curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
mv uniprot_sprot.fasta.gz uniprot_sprot_r2024_05.fasta.gz
gunzip -k uniprot_sprot_r2024_05.fasta.gz

head ../../data/blast_dbs/uniprot_sprot_r2024_05.fasta
echo "Number of Sequences"
grep -c ">" ../../data/blast_dbs/uniprot_sprot_r2024_05.fasta

/home/shared/ncbi-blast-2.15.0+/bin/makeblastdb \
-in ../../data/blast_dbs/uniprot_sprot_r2024_05.fasta \
-dbtype prot \
-out ../../data/blast_dbs/uniprot_sprot_r2024_05
```


# Blastp
```{r, engine='bash'}

fasta="../output/07-Apul-Hisat/genes.fasta"

/home/shared/ncbi-blast-2.15.0+/bin/blastx \
-query $fasta \
-db ../data/blast_dbs/uniprot_sprot_r2024_05 \
-out ../output/07-Apul-Hisat/blastx_out.tab \
-evalue 1E-05 \
-num_threads 48 \
-max_target_seqs 1 \
-max_hsps 1 \
-outfmt 6

wc -l ../output/07-Apul-Hisat/blastx_out.tab

tr '|' '\t' < ../output/07-Apul-Hisat/blastx_out.tab \
> ../output/07-Apul-Hisat/blastx_out_sep.tab

head -1 ../output/07-Apul-Hisat/blastx_out_sep.tab
```



# Download GO info


```{bash}
cd ../output/07-Apul-Hisat

curl -H "Accept: text/plain; format=tsv" "https://rest.uniprot.org/uniprotkb/stream?fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgo_p%2Cgo_c%2Cgo%2Cgo_f%2Cgo_id&format=tsv&query=%28*%29+AND+%28reviewed%3Atrue%29" -o SwissProt-Annot-GO.tsv

wc -l ../output/07-Apul-Hisat/SwissProt-Annot-GO.tsv 
```



Join blast with GO info


```{r}
bltabl <- read.csv("./output/07-Apul-Hisat/blastx_out_sep.tab", sep = '\t', header = FALSE)

spgo <- read.csv("../output/07-Apul-Hisat/SwissProt-Annot-GO.tsv", sep = '\t', header = TRUE)

annot_tab <- left_join(bltabl, spgo, by = c("V3" = "Entry")) %>%
  select(
    query = V1,
    blast_hit = V3,
    evalue = V13,
    ProteinNames = Protein.names,
    BiologicalProcess = Gene.Ontology..biological.process.,
    GeneOntologyIDs = Gene.Ontology.IDs
  )

head(annot_tab)

write.table(annot_tab, 
            file = "../output/07-Apul-Hisat/Apul-gene-annot-GO.tsv", 
            sep = "\t", 
            row.names = FALSE, 
            quote = FALSE)

system("head ../output/07-Apul-Hisat/Apul-gene-annot-GO.tsv", intern = TRUE )

```