---
title: "11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome"
author: "Kathleen Durkin"
date: "2024-10-02"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
bibliography: ../../references.bib
link-citations: true
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)
library(reticulate)
library(ggplot2)
library(stringr)
library(ggvenn)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

Use [ShortStack](https://github.com/MikeAxtell/ShortStack) [@axtell2013a; @shahid2014; @johnson2016a]to perform alignment of sRNAseq data and annotation of sRNA-producing genes.

This is the same ShortStack analysis as performed in the `deeep-dive` project However, we now have a genome for our actual study species, *A. pulchra*, so we can rerun sRNA discovery with the actual *pulchra* genome for reference, instead of using *A. millipora*. Additionally, ShortStack has an updated version, [ShortStack 4.1.0](https://github.com/MikeAxtell/ShortStack?tab=readme-ov-file#shortstack-version-4-major-changes), which provides much faster analysis times *and* additional functionality for visualizing miRNA hairpin structures and generating genome-browser-ready quantitative coverage tracks of aligned small RNAs. 

As in `deep-dive`, we will also include a customized miRBase database, utilizing cnidarian miRNAs curated by Jill Ashley, which includes published cnidarian miRNAs:

- [`cnidarian-mirbase-mature-v22.1.fasta`](../../data/cnidarian-mirbase-mature-v22.1.fasta)

------------------------------------------------------------------------

Inputs:

-   Requires trimmed sRNAseq files generated in `deep-dive` [08.2-Dapul-sRNAseq-trimming-31bp-fastp-merged.Rmd](https://github.com/urol-e5/deep-dive/blob/main/D-Apul/code/08.2-Dapul-sRNAseq-trimming-31bp-fastp-merged.Rmd)

    -   Filenames formatted: `*fastp-adapters-polyG-31bp-merged.fq.gz`

-   *A.pulchra* genome FastA. Not currently publicly available (still being annotated by colaborators)

Outputs:

-   See [ShortStack outputs documentation](https://github.com/MikeAxtell/ShortStack#outputs) for full list and detailed descriptions.

Software requirements:

-   Utilizes a [ShortStack](https://github.com/MikeAxtell/ShortStack#installation) Conda/Mamba environment, per the installation instructions.

Replace with name of your ShortStack environment and the path to the corresponding conda installation (find this *after* you've activated the environment).

E.g.

``` bash
# Activate environment
conda activate ShortStack4_env

# Find conda path
which conda
```

------------------------------------------------------------------------

# Set R variables

```{r R-variables, eval=TRUE}
shortstack_conda_env_name <- c("ShortStack-4.1.0_env")
shortstack_cond_path <- c("/home/sam/programs/mambaforge/condabin/conda")
```

# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""

echo "# Trimmed FastQ naming pattern"
echo "export trimmed_fastqs_pattern='*fastp-adapters-polyG-31bp-merged.fq.gz'"

echo "# Data directories"
echo 'export expression_dir=/home/shared/8TB_HDD_02/shedurkin/deep-dive-expression'
echo 'export expression_data_dir="${expression_dir}/data"'
echo 'export output_dir_top=${expression_dir}/D-Apul/output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome'
echo ""

echo "# Input/Output files"
echo 'export genome_fasta_dir=${expression_dir}/D-Apul/data'
echo 'export genome_fasta_name="Apulchra-genome.fa"'
echo 'export shortstack_genome_fasta_name="Apulchra-genome.fa"'
echo 'export trimmed_fastqs_dir="${genome_fasta_dir}/sRNA-trimmed-reads"'

echo 'export mirbase_mature_fasta_version=cnidarian-mirbase-mature-v22.1.fasta'
echo 'export genome_fasta="${genome_fasta_dir}/${shortstack_genome_fasta_name}"'
echo ""

echo "# Set number of CPUs to use"
echo 'export threads=40'
echo ""

echo "# Initialize arrays"
echo 'export trimmed_fastqs_array=()'


} > .bashvars

cat .bashvars
```

# Load [ShortStack](https://github.com/MikeAxtell/ShortStack) conda environment

If this is successful, the first line of output should show that the Python being used is the one in your [ShortStack](<https://github.com/MikeAxtell/ShortStack> conda environment path.

E.g.

`python:         /home/sam/programs/mambaforge/envs/mirmachine_env/bin/python`

```{r load-shortstack-conda-env, eval=TRUE}
use_condaenv(condaenv = shortstack_conda_env_name, conda = shortstack_cond_path)
py_config()
```
Note: I sometimes get an error "failed to initialize requested version of Python," which seems to stem from the `reticulate` package default loading a python environment. I've been able to fix this by manually uninstalling the `reticulate` package, then restarting R and reinstalling `reticulate` before rerunning this code document.
# Download reference files

## A.pulchra genome

```{r download-genome, engine='bash', eval=FALSE}
# Load bash variables into memory
source .bashvars

wget -O ${genome_fasta_dir}/${shortstack_genome_fasta_name} "https://osf.io/download/kn96u/"
```

## Cnidarian+miRBase database

Available n `deep-dive` repo, [here](https://github.com/urol-e5/deep-dive/blob/main/data/cnidarian-mirbase-mature-v22.1.fasta)

```{r download-db, engine='bash', eval=FALSE}
# Load bash variables into memory
source .bashvars

wget -O ${expression_data_dir}/"${mirbase_mature_fasta_version}" "https://raw.githubusercontent.com/urol-e5/deep-dive/refs/heads/main/data/cnidarian-mirbase-mature-v22.1.fasta"
```

```{r check-db-file, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

head -5 ${expression_data_dir}/"${mirbase_mature_fasta_version}"
```

## Trimmed sRNA-seq reads

Trimmed in `deep-dive`, [08.2-Dapul-sRNAseq-trimming-31bp-fastp-merged.Rmd](https://github.com/urol-e5/deep-dive/blob/main/D-Apul/code/08.2-Dapul-sRNAseq-trimming-31bp-fastp-merged.Rmd)

```{r download-trimmed-reads, engine='bash', eval=FALSE}
# Load bash variables into memory
source .bashvars

wget \
--directory-prefix ${trimmed_fastqs_dir} \
--recursive \
--no-check-certificate \
--continue \
--cut-dirs 7 \
--no-host-directories \
--no-parent \
--quiet \
--accept ${trimmed_fastqs_pattern} \
"https://gannet.fish.washington.edu/Atumefaciens/gitrepos/deep-dive/D-Apul/output/08.2-Apul-sRNAseq-trimming-31bp-fastp-merged/trimmed-reads/"
```

# Run ShortStack

## Modify genome filename for ShortStack compatability

```{r rename-genome-filename, engine='bash', cache=TRUE, eval=TRUE}
# Load bash variables into memory
source .bashvars

# Check for FastA file first
# Then create rename file if doesn't exist
if [ -f "${genome_fasta_dir}/${shortstack_genome_fasta_name}" ]; then
  echo "${genome_fasta_dir}/${shortstack_genome_fasta_name}"
  echo ""
  echo "Already exists. Nothing to do."
  echo ""
else

  # Copy genome FastA to ShortStack-compatible filename (ending with .fa)
  cp ${genome_fasta_dir}/${genome_fasta_name} ${genome_fasta_dir}/${shortstack_genome_fasta_name}
fi

# Confirm
ls -lh ${genome_fasta_dir}/${shortstack_genome_fasta_name}
```


## Excecute ShortStack command

Uses the `--dn_mirna` option to identify miRNAs in the genome, without relying on the `--known_miRNAs`.

This part of the code redirects the output of `time` to the end of `shortstack.log` file.

-   `; } \ 2>> ${output_dir_top}/shortstack.log`



```{r shortstack, engine='bash', cache=TRUE}
# Load bash variables into memory
source .bashvars

# Make output directory, if it doesn't exist
mkdir --parents "${output_dir_top}"

# Create array of trimmed FastQs
trimmed_fastqs_array=(${trimmed_fastqs_dir}/${trimmed_fastqs_pattern})


# Pass array contents to new variable as space-delimited list
trimmed_fastqs_list=$(echo "${trimmed_fastqs_array[*]}")


###### Run ShortStack ######
{ time \
ShortStack \
--genomefile "${genome_fasta}" \
--readfile ${trimmed_fastqs_list} \
--known_miRNAs ${expression_data_dir}/${mirbase_mature_fasta_version} \
--dn_mirna \
--threads ${threads} \
--outdir ${output_dir_top}/ShortStack_out \
&> ${output_dir_top}/shortstack.log ; } \
2>> ${output_dir_top}/shortstack.log

```

## Check runtime

```{r engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

tail -n 3 ${output_dir_top}/shortstack.log \
| grep "real" \
| awk '{print "ShortStack runtime:" "\t" $2}'

```


# Results

## ShortStack synopsis

```{r shortstack-synopsis, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

tail -n 25 ${output_dir_top}/shortstack.log
```

ShortStack identified 39 miRNAs. (This is in comparison to running ShortStack using the *A. millipora* genome, which yielded 38 miRNAs)

## Inspect `Results.txt`

```{r results-txt-file, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

head ${output_dir_top}/ShortStack_out/Results.txt

echo ""
echo "----------------------------------------------------------"
echo ""

echo "Nummber of potential loci:"
awk '(NR>1)' ${output_dir_top}/ShortStack_out/Results.txt | wc -l
```

Column 20 of the `Results.txt` file identifies if a cluster is a miRNA or not (`Y` or `N`).

```{r results-txt-miRNAs, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

echo "Number of loci characterized as miRNA:"
awk '$20=="Y" {print $0}' ${output_dir_top}/ShortStack_out/Results.txt \
| wc -l
echo ""

echo "----------------------------------------------------------"

echo ""
echo "Number of loci _not_ characterized as miRNA:"
awk '$20=="N" {print $0}' ${output_dir_top}/ShortStack_out/Results.txt \
| wc -l

```

Column 21 of the `Results.txt` file identifies if a cluster aligned to a known miRNA (miRBase) or not (`Y` or `NA`).

Since there are no miRNAs, the following code will *not* print any output.

The `echo` command after the `awk` command is simply there to prove that the chunk executed.

```{r results-txt-miRBase-miRNAs, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

echo "Number of loci matching miRBase miRNAs:"
awk '$21!="NA" {print $0}' ${output_dir_top}/ShortStack_out/Results.txt \
| wc -l
echo ""

echo "----------------------------------------------------------"

echo ""
echo "Number of loci _not_ matching miRBase miRNAs:"
awk '$21=="NA" {print $0}' ${output_dir_top}/ShortStack_out/Results.txt \
| wc -l

```

Although there are loci with matches to miRBase miRNAs, ShortStack did *not* annotate these clusters as miRNAs likely [because they do not *also* match secondary structure criteria](https://github.com/MikeAxtell/ShortStack#mirna-annotation).

### Directory tree of all ShortStack outputs

Many of these are large (by GitHub standards) BAM files, so will not be added to the repo.

Additionally, it's unlikely we'll utilize most of the other files (bigwig) generated by ShortStack.

```{r shortstack-directory-tree, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

tree -h ${output_dir_top}/

```


## Visualize

We noticed that a) not all of the identified miRNAs have database matches, and b) some reads have a match in the database but are *not* classified as miRNAs. Let's look at this in more depth.

```{r load-results, eval=TRUE}
Apul_shortstack_results <- read.csv("../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Results.txt", sep="\t")
```

```{r generate-plots, eval=TRUE}
# Reads identified as miRNAs (but not necessarily known)
Apul_shortstack_results %>% 
  filter(MIRNA == "Y") %>%
  mutate(known_miRNAs = str_sub(known_miRNAs, 1, 40)) %>%
  mutate(Locus = str_sub(Locus, 20, 40)) %>%
  ggplot(aes(x = reorder(Locus, Reads), y = Reads, fill = known_miRNAs)) +
  geom_bar(stat = "identity") +
   geom_text(aes(label = Reads), vjust = 0.5, hjust = 0, color = "black", size = 2.5, angle = 90) +
  labs(x = "miRNA", y = "Read count", 
       title = "Reads identified by ShortStack as miRNAs",
       fill = "Annotation") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
  
ggsave("../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/figures/Apul_ShortStack_miRNA_histogram.png", width = 12, height = 7, units = "in")


# Reads matched in the reference db (but not necessarily identified as miRNA)
Apul_shortstack_results %>% 
  filter(!is.na(known_miRNAs)) %>%
  mutate(known_miRNAs = str_sub(known_miRNAs, 1, 40)) %>%
  mutate(Locus = str_sub(Locus, 20, 40)) %>%
  ggplot(aes(x = reorder(Locus, Reads), y = Reads, fill = MIRNA)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Reads), vjust = 0.5, hjust = 0, color = "black", size = 2.5, angle = 90) +
  labs(x = "miRNA", y = "Read count", 
       title = "Reads with miRBase+cnidarian database matches",
       fill = "Identified as miRNA?") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

ggsave("../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/figures/Apul_ShortStack_dbmatch_histogram.png", width = 12, height = 7, units = "in")
```

There's two miRNAs with very high read counts, and it's making visualization of the rest difficult. Let's remove them and retry visualizing the rest.

```{r generate-plots-reduced, eval=TRUE}
# Reads identified as miRNAs (but not necessarily known)
Apul_shortstack_results %>% 
  filter(MIRNA == "Y") %>%
  filter(Reads < 200000) %>%
  mutate(known_miRNAs = str_sub(known_miRNAs, 1, 40)) %>%
  mutate(Locus = str_sub(Locus, 20, 40)) %>%
  ggplot(aes(x = reorder(Locus, Reads), y = Reads, fill = known_miRNAs)) +
  geom_bar(stat = "identity") +
   geom_text(aes(label = Reads), vjust = 0.5, hjust = 0, color = "black", size = 2.5, angle = 90) +
  labs(x = "miRNA", y = "Read count", 
       title = "Reads identified by ShortStack as miRNAs",
       fill = "Annotation") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
  
ggsave("../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/figures/Apul_ShortStack_miRNA_histogram_reduced.png", width = 12, height = 7, units = "in")


# Reads matched in the reference db (but not necessarily identified as miRNA)
Apul_shortstack_results %>% 
  filter(!is.na(known_miRNAs)) %>%
  filter(Reads < 200000) %>%
  mutate(known_miRNAs = str_sub(known_miRNAs, 1, 40)) %>%
  mutate(Locus = str_sub(Locus, 20, 40)) %>%
  ggplot(aes(x = reorder(Locus, Reads), y = Reads, fill = MIRNA)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Reads), vjust = 0.5, hjust = 0, color = "black", size = 2.5, angle = 90) +
  labs(x = "miRNA", y = "Read count", 
       title = "Reads with miRBase+cnidarian database matches",
       fill = "Identified as miRNA?") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

ggsave("../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/figures/Apul_ShortStack_dbmatch_histogram_reduced.png", width = 12, height = 7, units = "in")
```

```{r venn-diagram, eval=TRUE}
# Make list
mirnas <- Apul_shortstack_results %>% filter(MIRNA == "Y") %>% pull(Locus)
matches <- Apul_shortstack_results %>% filter(!is.na(known_miRNAs)) %>% pull(Locus)

Apul_shortstack_vennlist <- list(
  "Identified as miRNA" = mirnas,
  "Database match" = matches
)

# Make venn diagrams
ggvenn(Apul_shortstack_vennlist)

ggsave("../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/figures/Apul_ShortStack_venn.png", width = 12, height = 7, units = "in")
```

------------------------------------------------------------------------

# Citations