---
title: "24-Apul-miRanda-input-comparisons"
author: "Kathleen Durkin"
date: "2024-02-12"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: ../../references.bib
link-citations: true
---

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

Up to this point, our miRNA target prediction has been primarily performed using the tool miRanda, which requires only seed binding, with 3'UTR regions as input. However, there is growing evidence that cnidarian miRNA binding functions similarly to plants, requiring near-full complementarity and to act primarily through target cleavage, which can occur at any location in the mRNA sequence (not just the 3'UTR). This would mean our current miRanda target predictions are both a) not sufficiently restricted to fll complementarity, and b) incorrectly limited to the 3'UTR.

Let's see what happens when we adjust those parameters

# miRNA and full mRNA coding sequences

Inputs:

-   FASTA of A.pulchra mRNA coding sequences

-   FASTA of A.pulchra mature miRNAs

## Get mRNA fasta

```{r, engine='bash', eval=FALSE}
/home/shared/bedtools2/bin/bedtools getfasta \
-fi "../data/Apulchra-genome.fa" \
-bed "../data/Apulcra-genome-mRNA_only.gff" \
-fo "../data/24-Apul-miRanda-input-comparisons/Apul_mRNA_full.fa"

```

## Run miRanda

```{r, engine='bash', eval=FALSE}

# Same settings we've been using: 
# score cutoff >100
# energy cutoff <-10
# strict binding

/home/shared/miRanda-3.3a/src/miranda \
../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature.fasta \
../data/24-Apul-miRanda-input-comparisons/Apul_mRNA_full.fa \
-sc 100 \
-en -10 \
-strict \
-out ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict_all.tab
```

## Summarize results

Let's look at the output

```{r, engine='bash', eval=FALSE}

echo "miranda run finished!"
echo "Counting number of interacting miRNA-lncRNA pairs"

zgrep -c "Performing Scan" ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict_all.tab

echo "Parsing output"
grep -A 1 "Scores for this hit:" ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict_all.tab | sort | grep '>' > ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt

echo "counting number of putative interactions predicted (can include multiple interactions between single miRNA-lncRNA pair)"
wc -l ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict_all.tab

```

```{r, engine='bash'}
echo "number of putative interactions:"
wc -l ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt
```

This is a lot of putative interactions! Note though, that miRanda only requires complementarity of a 8bp seed region of the miRNA. We instead want to look for binding with full or near-full complementarity. Let's look at how many putative interactions are predicted for a binding length of at least 21 nucleotides (the length of our smallest mature miRNA).

```{r, engine='bash'}
echo "number of putative interactions of at least 21 nucleotides"
awk -F'\t' '$7 >= 21' ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt | wc -l
echo ""
echo "check some:"
awk -F'\t' '$7 >= 21' ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt | head -5
```

The header for this output is formatted as:

mirna Target Score Energy-Kcal/Mol Query-Aln(start-end) Subject-Al(Start-End) Al-Len Subject-Identity Query-Identity

We can see from the percent identities (last 2 entries) that this number includes alignments with multiple mismatches. Let's filter again to reduce the number of permissible mismatches. Let's say we want no more than 3 mismatches (a gap is counted as a mismatch). For an alignment of 21 nucleotides, this would be an percent identity of (21-3)/21 = 85.7%. The miRNA is our "subject", so we will filter by column 8.

```{r, engine='bash'}
echo "number of putative interactions of at least 21 nucleotides, with at most 3 mismatches"
awk -F'\t' '$7 >= 21' ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt | awk -F'\t' '$8 >= 85' | wc -l
echo ""
echo "check some:"
awk -F'\t' '$7 >= 21' ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt | awk -F'\t' '$8 >= 85' | head -5
```

So filtering for full or near-full complementarity reduced the number of putative interactions between miRNA and full mRNA sequences from 617982 to 143.

# miRNA and mRNA 5'UTRs

We've also created a gff of 1kb 5'UTR regions (using the same method used to define 1kb 3'UTRs, in `15-Apul-annotate-UTRs`). Let's try running this through miRanda as well, since it's possible (though not necessarily expected) that miRNAs will bind here

## Get 5'UTR fasta

```{r, engine='bash', eval=FALSE}
/home/shared/bedtools2/bin/bedtools getfasta \
-fi "../data/Apulchra-genome.fa" \
-bed "../output/15-Apul-annotate-UTRs/Apul.GFFannotation.5UTR_1kb_corrected.gff" \
-fo "../data/24-Apul-miRanda-input-comparisons/Apul_5UTR_1kb_corrected.fa"

```

## Run miRanda

```{r, engine='bash', eval=FALSE}

# Same settings we've been using: 
# score cutoff >100
# energy cutoff <-10
# strict binding

/home/shared/miRanda-3.3a/src/miranda \
../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature.fasta \
../data/24-Apul-miRanda-input-comparisons/Apul_5UTR_1kb_corrected.fa \
-sc 100 \
-en -10 \
-strict \
-out ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict_all.tab
```

## Summarize results

Let's look at the output

```{r, engine='bash', eval=FALSE}

echo "Number of interacting miRNA-lncRNA pairs"
zgrep -c "Performing Scan" ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict_all.tab

echo "Parsing output"
grep -A 1 "Scores for this hit:" ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict_all.tab | sort | grep '>' > ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict-parsed.txt

echo "Number of putative interactions predicted (can include multiple interactions between single miRNA-lncRNA pair)"
wc -l ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict_all.tab

```

How many putative interactions are predicted for a binding length of at least 21 nucleotides (the length of our smallest mature miRNA)?

```{r, engine='bash'}
echo "number of putative interactions:"
wc -l ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict-parsed.txt
echo ""
echo "number of putative interactions of at least 21 nucleotides"
awk -F'\t' '$7 >= 21' ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict-parsed.txt | wc -l
echo ""
echo "number of putative interactions of at least 21 nucleotides, with at most 3 mismatches"
awk -F'\t' '$7 >= 21' ../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict-parsed.txt | awk -F'\t' '$8 >= 85' | wc -l
```

So filtering for full or near-full complementarity reduced the number of putative interactions between miRNA and 5'UTR sequences from 87830 to 12

# miRNA and mRNA 3'UTRs

Now let's see how filtering changes the outputs of miRanda run with only the 3'UTR mRNA region (the input we have been using up till now)

```{r, engine='bash'}
echo "total number of putative interactions:"
wc -l ../output/09-Apul-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_apul_updated.txt

echo "number of putative interactions of at least 21 nucleotides"
awk -F'\t' '$7 >= 21' ../output/09-Apul-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_apul_updated.txt | wc -l

echo "number of putative interactions of at least 21 nucleotides, with at most 3 mismatches"
awk -F'\t' '$7 >= 21' ../output/09-Apul-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_apul_updated.txt | awk -F'\t' '$8 >= 85' | wc -l
```

When only mRNA 3'UTR regions are used as input, filtering for full/near-full complementarity reduces the number of putative interactions from 6109 to 13.

# Examine coexpression

Now that we've found putative interactions with high complementarity, we need to validate miRNA function by examining patterns of coexpression. We'd expect a putatively-interacting miRNA-mRNA pair to be highly coexpressed, and we'd expect a negative relationship to indicate target cleavage.

## Obtain Pearson's coefficient correlation values

This has already been run by Jill in `09-Apul-mRNA-miRNA-interactions`, but I think the csv stored in this repo has been filtered in some way, because it contains fewer entries than expected:

```{r, engine='bash'}
wc -l ../output/09-Apul-mRNA-miRNA-interactions/Apul-PCC_miRNA_mRNA.csv
```

Read in, format, and normalize data

```{r}
mRNA_counts <- read_csv("../../D-Apul/output/07-Apul-Hisat/gene_count_matrix.csv")
mRNA_counts <- as.data.frame(mRNA_counts)
rownames(mRNA_counts) <- mRNA_counts[,1] #set first column that contains gene names as rownames
mRNA_counts <- mRNA_counts[,-1] # remove column w/ gene names 

# Remove any genes with 0 counts across samples 
mRNA_counts<-mRNA_counts %>%
    mutate(Total = rowSums(.[, 1:5]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)

miRNA_counts <- read.delim("../output/03.1-Apul-sRNA-summary/Apul_miRNA_ShortStack_counts_formatted.txt")

# Remove any miRNAs with 0 for all samples 
miRNA_counts <- miRNA_counts %>%
    mutate(Total = rowSums(.[, 1:5]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)

# Rename gene count cols to match miRNA count cols
colnames(mRNA_counts) <- c("sample140", "sample145", "sample150", "sample173", "sample178")

# Function to normalize counts (simple RPM normalization)
normalize_counts <- function(counts) {
  rpm <- t(t(counts) / colSums(counts)) * 1e6
  return(rpm)
}

miRNA_norm <- normalize_counts(miRNA_counts)
mRNA_norm <- normalize_counts(mRNA_counts)

```

```{r,}
nrow(miRNA_norm)*nrow(mRNA_norm)
```

We'd expect 1311336 PCC values (after filtering out genes/miRNAs with 0 counts), but there are only 916384 in Jill's PCC csv. I'll rerun the PCC just to make sure I have correlation values for all miRNA-mRNA pairs of interest

```{r, eval=FALSE}
library(energy)

# Function to calculate PCC and p-value for a pair of vectors
calc_pcc <- function(x, y) {
  result <- cor.test(x, y, method = "pearson")
  return(c(PCC = result$estimate, p_value = result$p.value))
}

# Create a data frame of all miRNA-mRNA pairs
pairs <- expand.grid(miRNA = rownames(miRNA_norm), mRNA = rownames(mRNA_norm))

# Calculate PCC and p-value for each pair
pcc_results <- pairs %>%
  rowwise() %>%
  mutate(
    pcc_stats = list(calc_pcc(miRNA_norm[miRNA,], mRNA_norm[mRNA,]))
  ) %>%
  unnest_wider(pcc_stats)

# Adjust p-values for FDR
pcc_results <- pcc_results %>%
  mutate(adjusted_p_value = p.adjust(p_value, method = "fdr"))

# Save as csv
write.csv(pcc_results, "../output/24-Apul-miRanda-input-comparisons/Apul-PCC_miRNA_mRNA-full.csv")
```

Check

```{r}
# Read in results 
# pcc_results <- read.csv("../output/24-Apul-miRanda-input-comparisons/Apul-PCC_miRNA_mRNA-full.csv")

# Use this code to download the PCC results if needed 
pcc_results <- read.csv("https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/24-Apul-miRanda-input-comparisons/Apul-PCC_miRNA_mRNA-full.csv")

nrow(pcc_results)
nrow(pcc_results%>% filter(PCC.cor > 0.90))
nrow(pcc_results %>% filter(p_value < 0.05))
nrow(pcc_results %>% filter(p_value < 0.05 & PCC.cor > 0.90))
```

of the 1,311,336 possible miRNA-mRNA interactions, 35,445 have a Pearson's correlation coefficient of above 0.9, and 66,154 have a significant correlation (pval\<0.05). All of the coefficients \>0.9 are significant.

## full mRNA

```{r}
# miRNA-mRNA_full miRanda output
miRNA_mRNA_miRanda <- read_delim("../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-mRNA_full-strict-parsed.txt", col_names=FALSE)
colnames(miRNA_mRNA_miRanda) <- c("mirna", "Target", "Score", "Energy_Kcal_Mol", "Query_Aln", "Subject_Aln", "Al_Len", "Subject_Identity", "Query_Identity")

# format miRNA and mRNA names
geneIDs <- read_delim("../output/15-Apul-annotate-UTRs/Apul-mRNA-FUNids.txt", col_names=FALSE)
geneIDs$X4 <- gsub("Parent=", "", geneIDs$X4)

miRNA_mRNA_miRanda$mirna <- gsub(">", "", miRNA_mRNA_miRanda$mirna)
miRNA_mRNA_miRanda$mirna <- gsub("\\..*", "", miRNA_mRNA_miRanda$mirna)

miRNA_mRNA_miRanda <- left_join(miRNA_mRNA_miRanda, geneIDs, by=c("Target" = "X1"))
miRNA_mRNA_miRanda <- select(miRNA_mRNA_miRanda, -X2,-X3,-X5)

# Finally, create a column that conatins both the miRNA and interacting mRNA
pcc_results$interaction <- paste(pcc_results$miRNA, "_", pcc_results$mRNA)
miRNA_mRNA_miRanda$interaction <- paste(miRNA_mRNA_miRanda$mirna, "_", miRNA_mRNA_miRanda$X4)

# Annotate w PCC info 
miRNA_mRNA_miRanda <- left_join(miRNA_mRNA_miRanda, pcc_results, by="interaction")
```

```{r}
# Filter to high complementarity putative targets
target_21bp <- miRNA_mRNA_miRanda[miRNA_mRNA_miRanda$Al_Len > 20,]
target_21bp_3mis <- target_21bp[target_21bp$Subject_Identity>85,]

# How many w significant correlation?
nrow(miRNA_mRNA_miRanda)
nrow(miRNA_mRNA_miRanda %>% filter(p_value < 0.05))
nrow(target_21bp %>% filter(p_value < 0.05))
nrow(target_21bp_3mis %>% filter(p_value < 0.05))
```

For miRNA binding to the mRNA CDS, miRanda predicts 617,982 putative interactions. Of these, 24,936 have significant PCCs; 4,300 are \>21bp and have signficant PCCs; and 7 are \>21bp with \<=3 mismatches and have significant PCCs.

## 3'UTR

```{r}
# Read in data

# miRNA-mRNA_full miRanda output
miRNA_3UTR_miRanda <- read_delim("../output/09-Apul-mRNA-miRNA-interactions/miranda_strict_all_1kb_parsed_apul_updated.txt", col_names=FALSE)
colnames(miRNA_3UTR_miRanda) <- c("mirna", "Target", "Score", "Energy_Kcal_Mol", "Query_Aln", "Subject_Aln", "Al_Len", "Subject_Identity", "Query_Identity")

miRNA_3UTR_miRanda$mirna <- gsub(">", "", miRNA_3UTR_miRanda$mirna)
miRNA_3UTR_miRanda$mirna <- gsub("\\..*", "", miRNA_3UTR_miRanda$mirna)
miRNA_3UTR_miRanda$Target <- gsub("::.*", "", miRNA_3UTR_miRanda$Target)

# Finally, create a column that conatins both the miRNA and interacting mRNA
miRNA_3UTR_miRanda$interaction <- paste(miRNA_3UTR_miRanda$mirna, "_", miRNA_3UTR_miRanda$Target)

# Annotate w PCC info 
miRNA_3UTR_miRanda <- left_join(miRNA_3UTR_miRanda, pcc_results, by="interaction")
```

```{r}
# Filter to high complementarity putative targets
target_3UTR_21bp <- miRNA_3UTR_miRanda[miRNA_3UTR_miRanda$Al_Len > 20,]
target_3UTR_21bp_3mis <- target_3UTR_21bp[target_3UTR_21bp$Subject_Identity>85,]

# How many w significant correlation?
nrow(miRNA_3UTR_miRanda)
nrow(miRNA_3UTR_miRanda %>% filter(p_value < 0.05))
nrow(target_3UTR_21bp %>% filter(p_value < 0.05))
nrow(target_3UTR_21bp_3mis %>% filter(p_value < 0.05))
```

For miRNA binding to the 3'UTR, miRanda predicts 6,109 putative interactions. Of these, 266 have significant PCCs; 86 are \>21bp and have signficant PCCs; and 1 is \>21bp with \<=3 mismatches and has a significant PCC.

## 5'UTR

```{r}
# miRNA-5'UTR miRanda output
miRNA_5UTR_miRanda <- read_delim("../output/24-Apul-miRanda-input-comparisons/Apul-miRanda-5UTR_1kb-strict-parsed.txt", col_names=FALSE)
colnames(miRNA_5UTR_miRanda) <- c("mirna", "Target", "Score", "Energy_Kcal_Mol", "Query_Aln", "Subject_Aln", "Al_Len", "Subject_Identity", "Query_Identity")

# format miRNA and mRNA names
geneIDs <- read_delim("../output/15-Apul-annotate-UTRs/Apul-5UTR-FUNids.txt", col_names=FALSE)
geneIDs$X4 <- gsub("Parent=", "", geneIDs$X4)

miRNA_5UTR_miRanda$mirna <- gsub(">", "", miRNA_5UTR_miRanda$mirna)
miRNA_5UTR_miRanda$mirna <- gsub("\\..*", "", miRNA_5UTR_miRanda$mirna)

miRNA_5UTR_miRanda <- left_join(miRNA_5UTR_miRanda, geneIDs, by=c("Target" = "X1"))
miRNA_5UTR_miRanda <- select(miRNA_5UTR_miRanda, -X2,-X3,-X5)

# Somehow ended up with a many-to-many relationship, so remove duplicate rows
miRNA_5UTR_miRanda <- miRNA_5UTR_miRanda[!duplicated(miRNA_5UTR_miRanda), ]

# Finally, create a column that contains both the miRNA and interacting mRNA
pcc_results$interaction <- paste(pcc_results$miRNA, "_", pcc_results$mRNA)
miRNA_5UTR_miRanda$interaction <- paste(miRNA_5UTR_miRanda$mirna, "_", miRNA_5UTR_miRanda$X4)

# Annotate w PCC info 
miRNA_5UTR_miRanda <- left_join(miRNA_5UTR_miRanda, pcc_results, by="interaction")
```

```{r}
# Filter to high complementarity putative targets
target_5UTR_21bp <- miRNA_5UTR_miRanda[miRNA_5UTR_miRanda$Al_Len > 20,]
target_5UTR_21bp_3mis <- target_5UTR_21bp[target_5UTR_21bp$Subject_Identity>85,]

# How many w significant correlation?
nrow(miRNA_5UTR_miRanda)
nrow(miRNA_5UTR_miRanda %>% filter(p_value < 0.05))
nrow(target_5UTR_21bp %>% filter(p_value < 0.05))
nrow(target_5UTR_21bp_3mis %>% filter(p_value < 0.05))
```

For miRNA binding to the 5'UTR, miRanda predicts 87,830 putative interactions. Of these, 3,426 have significant PCCs; 596 are \>21bp and have signficant PCCs; and 0 are \>21bp with \<=3 mismatches and have significant PCCs.

# Summary

How does different input and/or complementarity filtering affect \# putative interactions:

| Input     | unfiltered | filtered for complementarity | \% retained |
|-----------|------------|------------------------------|-------------|
| 3'UTR     | 6109       | 13                           | 0.213 %     |
| full mRNA | 617982     | 143                          | 0.0231 %    |
| 5'UTR     | 87830      | 12                           | 0.0137 %    |

For different filters, how many putative interactions ***also show significant coexpression*** (PCC pval \< 0.05)?

| Input     | All   | 21bp | 21bp, \>=3 mismatch |
|-----------|-------|------|---------------------|
| 3'UTR     | 266   | 86   | 1                   |
| full mRNA | 24936 | 4300 | 7                   |
| 5'UTR     | 3426  | 596  | 0                   |

Note that, in general, only \~1/3 of significant coexpressions have a *negative* relationship (which would support functional target cleavage or translational repression)

Note also that some putative interactions indicated by miRanda are not present in the counts data (i.e. the miRNA and/or mRNA had 0 counts in our RNAseq data), and are thus excluded from the PCC-filtered data


Is there a clear "cutoff" for what complementarity parameters are most associated with significant coexpression?

```{r}

miRNA_mRNA_miRanda <- miRNA_mRNA_miRanda %>% mutate(Source = "mRNA")
miRNA_3UTR_miRanda <- miRNA_3UTR_miRanda %>% mutate(Source = "3UTR")
miRNA_5UTR_miRanda <- miRNA_5UTR_miRanda %>% mutate(Source = "5UTR")

# Combine
combined_df <- bind_rows(miRNA_mRNA_miRanda, miRNA_3UTR_miRanda, miRNA_5UTR_miRanda)

# Convert Source to a factor for discrete shapes
combined_df$Source <- factor(combined_df$Source)

# Convert p-value condition into a categorical variable for coloring
combined_df$Significance <- ifelse(combined_df$p_value < 0.05, "p < 0.05", "p ≥ 0.05")
significant_df <- combined_df %>% filter(p_value < 0.05)

# Plot with jitter (significant coexpression only)
ggplot(significant_df, aes(x = Al_Len, 
                           y = as.numeric(gsub("%", "", Subject_Identity)), 
                           color = Source)) +
  geom_jitter(size = 1.5, width = 1, height = 1, alpha = 0.1) +
  scale_color_manual(values = c("mRNA" = "red", "3UTR" = "blue", "5UTR" = "green")) +
  facet_wrap(~ Source) +
  labs(x = "Alignment Length (Al_Len)", 
       y = "Subject Identity (%)", 
       color = "Dataset", 
       title = "Significant (p < 0.05) Alignment Length vs. Subject Identity") +
  theme_minimal()
```

Interesting... there are signifcant coexpressions happening across both metrics of complementarity. Also interesting, there are notably more significant coexpressions from using 5'UTR inputs than using 3'UTRs -- we were expecting more in the 3UTr because that would be the binding location for translational repression.

Next steps:

-   discuss w team -- should we look for both miRNAs that function through target cleavage (full complementary, full coding sequence) *and* miRNAs that function through "canonical" translational silencing (seed complementarity, 3'UTR)?

-   Run same comparison using BLAST and/or RNAhybrid as the tool. I'm still not 100% clear on what sequence features miRanda takes into consideration, and any that *are* included will have been based on mammalian miRNAs. It may be more appropriate to use a different tool

-   if we decide to shift target prediction requirements, need to rerun all target prediction done thus far :(

```{r}
# Save putative interactions with significantly correlated coexpression for visualization
# (e.g. creating an interaction network plot in Cytoscape)
write.csv(combined_df[combined_df$p_value < 0.05,], "../output/24-Apul-miRanda-input-comparisons/miRanda-PCC-significant-mRNA_3UTR_5UTR.csv", row.names = FALSE)

```

