---
title: "Long non-coding RNA discovery pipeline"
author: "Steven Roberts"
date: "`r format(Sys.time(), '%d %B, %Y')`"  
format:
  html:
    toc: true
    toc-depth: 2
    html-math-method: katex
    css: styles.css
    theme: sandstone
editor: 
  markdown: 
    wrap: 72
---

Steven's attempt


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center", # Align plots to the center
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```


#Variables
```{r setup, include=FALSE}
# Global R options
knitr::opts_chunk$set(echo = TRUE)

# Define key paths and tool directories
 
DATA_DIR <- "../data/25-lncRNA"
OUTPUT_DIR <- "../output/25-lncRNA"
THREADS <- "14"
  
FASTQ_SOURCE <- "https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/"
FASTQ_SUFFIX <- "fastq.gz"
GENOME_SOURCE <- "https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/D-Apul/data/Apulcra-genome.fa"


GTF_SOURCE <- "https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/D-Apul/data/Apulchra-genome.gtf"
GFF_SOURCE <- "https://gannet.fish.washington.edu/seashell/bu-github/deep-dive-expression/D-Apul/data/Apulcra-genome.gff"


#RAVEN
# HISAT2_DIR <- "/home/shared/hisat2-2.2.1/"
# SAMTOOLS_DIR <- "/home/shared/samtools-1.12/"
# STRINGTIE_DIR <- "/home/shared/stringtie-2.2.1.Linux_x86_64"
# GFFCOMPARE_DIR <- "/home/shared/gffcompare-0.12.6.Linux_x86_64"
# BEDTOOLS_DIR <- "/home/shared/bedtools2/bin"
# CPC2_DIR <- "/home/shared/CPC2_standalone-1.0.1"
# CONDA_PATH <- "/opt/anaconda/anaconda3/bin/conda"

#KLONE
HISAT2_DIR <- ""
SAMTOOLS_DIR <- ""
STRINGTIE_DIR <- ""
GFFCOMPARE_DIR <- ""
BEDTOOLS_DIR <- ""
CPC2_DIR <- ""
CONDA_PATH <- ""

GENOME_FASTA <- file.path(DATA_DIR, "genome.fasta")
GENOME_GTF <- file.path(DATA_DIR, "genome.gtf")
GENOME_GFF <- file.path(DATA_DIR, "genome.gff")
FASTQ_DIR <- file.path(DATA_DIR, "fastq")
GENOME_INDEX <- file.path(OUTPUT_DIR, "genome.index")

# Export these as environment variables for bash chunks.
Sys.setenv(
  THREADS = THREADS,
  DATA_DIR = DATA_DIR,
  FASTQ_SOURCE = FASTQ_SOURCE,
  FASTQ_SUFFIX = FASTQ_SUFFIX,
  OUTPUT_DIR = OUTPUT_DIR,
  GENOME_SOURCE = GENOME_SOURCE,
  GTF_SOURCE = GTF_SOURCE,
  GFF_SOURCE = GFF_SOURCE,
  HISAT2_DIR = HISAT2_DIR,
  SAMTOOLS_DIR = SAMTOOLS_DIR,
  STRINGTIE_DIR = STRINGTIE_DIR,
  GFFCOMPARE_DIR = GFFCOMPARE_DIR,
  BEDTOOLS_DIR = BEDTOOLS_DIR,
  CPC2_DIR = CPC2_DIR,
  CONDA_PATH = CONDA_PATH,
  GENOME_FASTA = GENOME_FASTA,
  GENOME_GTF = GENOME_GTF,
  GENOME_GFF = GENOME_GFF,
  FASTQ_DIR = FASTQ_DIR,
  GENOME_INDEX = GENOME_INDEX
)
```


```{bash}
mkdir -p "${DATA_DIR}"
mkdir -p "${OUTPUT_DIR}"
```


# Download Genome and Reads for Hisat

```{r, engine='bash'}

wget -nv -r \
--no-directories --no-parent \
-P ${FASTQ_DIR} \
-A "*${FASTQ_SUFFIX}" ${FASTQ_SOURCE}
```

```{r, engine='bash'}
ls ${FASTQ_DIR}
```


```{r, engine='bash'}

curl -o "${GENOME_FASTA}" "${GENOME_SOURCE}"
```

```{r, engine='bash'}


curl -o "${GENOME_GTF}" "${GTF_SOURCE}"
```


```{r, engine='bash'}


curl -o "${GENOME_GFF}" "${GFF_SOURCE}"
```

```{bash}
output_fasta=$(head -1 "${GENOME_FASTA}")
output_gff=$(head -2 "${GENOME_GFF}")
output_gtf=$(head -1 "${GENOME_GTF}")

if [[ "$output_fasta" == *html* || "$output_gff" == *html* || "$output_gtf" == *html* ]]; then
    echo "FAIL"
else
    echo "$output_fasta"
    echo "$output_gff"
    echo "$output_gtf"
fi
```


# HISAT

```{r, engine='bash'}
"${HISAT2_DIR}hisat2_extract_exons.py" "${GENOME_GTF}" > "${OUTPUT_DIR}/exon.txt"

"${HISAT2_DIR}hisat2_extract_splice_sites.py" "${GENOME_GTF}" > "${OUTPUT_DIR}/splice_sites.txt"

```



```{r, engine='bash'}

"${HISAT2_DIR}hisat2-build" \
  -p "${THREADS}" \
  "${GENOME_FASTA}" \
  "${GENOME_INDEX}" \
  --exon "${OUTPUT_DIR}/exon.txt" \
  --ss "${OUTPUT_DIR}/splice_sites.txt" \
  2> "${OUTPUT_DIR}/hisat2-build_stats.txt"

```




```{bash}
# Loop over every file ending in .fastq.gz that contains "_R2_"
for r2 in "${FASTQ_DIR}"/*_R2_*."${FASTQ_SUFFIX}"; do
    # Get the basename (filename without path)
    base=$(basename "$r2")
    
    # Derive a sample name by taking everything before "_R2_"
    sample="${base%%_R2_*}"
    
    # Construct the corresponding R1 filename by replacing "_R2_" with "_R1_"
    r1="${r2/_R2_/_R1_}"
    
    # Define the output SAM file name using the sample name
    output="${OUTPUT_DIR}/${sample}.sam"
    
    # Run hisat2 with the paired-end files
    "${HISAT2_DIR}hisat2" \
      -x "${GENOME_INDEX}" \
      -p "${THREADS}" \
      -1 "$r1" \
      -2 "$r2" \
      -S "$output"
done
```

## convert SAM to BAM


```{bash}
for samfile in "${OUTPUT_DIR}/${sample}"*.sam; do
  bamfile="${samfile%.sam}.bam"
  sorted_bamfile="${samfile%.sam}.sorted.bam"
  
  # Convert SAM to BAM
  "${SAMTOOLS_DIR}samtools" view -bS -@ "${THREADS}" "$samfile" > "$bamfile"
  
  # Sort BAM
  "${SAMTOOLS_DIR}samtools" sort -@ "${THREADS}" "$bamfile" -o "$sorted_bamfile"
  
  # Index sorted BAM
  "${SAMTOOLS_DIR}samtools" index -@ "${THREADS}" "$sorted_bamfile"
done
```





# StringTie

StringTie uses the sorted BAM files to assemble transcripts for each sample, outputting them as GTF (Gene Transfer Format) files. And then merges all individual GTF assemblies into a single merged GTF file. This step extracts transcript information and merges GTFs from all samples–an important step in creating a canonical list of lncRNAs across all samples included in the pipeline.

```{bash}
find "${OUTPUT_DIR}" -name "*sorted.bam" \
| xargs -n 1 basename -s .sorted.bam | xargs -I{} \
"${STRINGTIE_DIR}stringtie" \
-p "${THREADS}" \
-G "${GENOME_GFF}" \
-o "${OUTPUT_DIR}/{}.gtf" \
"${OUTPUT_DIR}/{}.sorted.bam"
```


```{bash}
"${STRINGTIE_DIR}stringtie" \
--merge \
-G "${GENOME_GFF}" \
-o "${OUTPUT_DIR}/stringtie_merged.gtf" \
"${OUTPUT_DIR}/"*.gtf
```

#GFFCOMPARE

![](http://gannet.fish.washington.edu/seashell/snaps/Monosnap_687474703a2f2f67616e6e65742e666973682e77617368696e67746f6e2e6564752f7365617368656c6c2f736e6170732f323032332d31312d30335f30392d3_2024-12-20_04-02-37.png)





