---
title: "26-Apul-siRNA-mRNA-interactions"
author: "Kathleen Durkin"
date: "2024-02-27"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: ../../references.bib
link-citations: true
---

@gresova_small_2022, @luck_sirna-finder_2019

ShortStack outputs a gff of genomic regions with accumulating sRNAs (`Results.gff3`) and annotates those regions as "unkown sRNA", "miRNA", or "siRNA". However, these regions are generally several hundred basepairs long, since they are "accumulation regions", not the ~21nt sRNA molecules themselves. While ShortStack also identifies and outputs fastas for the miRNAs, it doesn't do that for siRNAs. This means we need to ID the precise siRNA sequences ourselves.

Approach:

1. Extract regions of accumulating siRNA from ShortStack output

2. Convert gff of siRNA regions to .bed

3. Convert sample sRNAseq bam files to .bed

4. Use bedtools `multiintersect` to find sRNA sequences that appear within the ShortStack siRNA regions and in our sample sRNAseq reads.



# Extract siRNA regions and convert to .bed

```{r, engine='bash'}
awk '$3 ~ /siRNA/' ../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Results.gff3 > ../output/26-Apul-siRNA-mRNA-interactions/Apul-ShortStack-siRNA_regions.gff3

awk '$3 ~ /siRNA/ {print $1 "\t" $4-1 "\t" $5}' ../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/Results.gff3 > ../output/26-Apul-siRNA-mRNA-interactions/Apul-ShortStack-siRNA_regions.bed
```

How many siRNA regions are there? This should be our total number of siRNA at the end of this process

```{r, engine='bash'}
wc -l ../output/26-Apul-siRNA-mRNA-interactions/Apul-ShortStack-siRNA_regions.bed
```

# Convert sample sRNAseq bam files to .bed

```{r, engine='bash'}
reads_dir="../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out"
output_dir="../output/26-Apul-siRNA-mRNA-interactions"
/home/shared/bedtools2/bin/bamToBed \
-i ${reads_dir}/sRNA-ACR-140-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bam \
> ${output_dir}/sRNA-ACR-140-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed

/home/shared/bedtools2/bin/bamToBed \
-i ${reads_dir}/sRNA-ACR-145-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bam \
> ${output_dir}/sRNA-ACR-145-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed

/home/shared/bedtools2/bin/bamToBed \
-i ${reads_dir}/sRNA-ACR-150-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bam \
> ${output_dir}/sRNA-ACR-150-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed

/home/shared/bedtools2/bin/bamToBed \
-i ${reads_dir}/sRNA-ACR-173-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bam \
> ${output_dir}/sRNA-ACR-173-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed

/home/shared/bedtools2/bin/bamToBed \
-i ${reads_dir}/sRNA-ACR-178-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bam \
> ${output_dir}/sRNA-ACR-178-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed
```


# Use bedtools `multiintersect`

```{r, engine='bash'}
output_dir="../output/26-Apul-siRNA-mRNA-interactions"

/home/shared/bedtools2/bin/bedtools multiinter -i \
${output_dir}/Apul-ShortStack-siRNA_regions.bed \
${output_dir}/sRNA-ACR-140-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed \
${output_dir}/sRNA-ACR-145-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed \
${output_dir}/sRNA-ACR-150-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed \
${output_dir}/sRNA-ACR-173-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed \
${output_dir}/sRNA-ACR-178-S1-TP2-fastp-adapters-polyG-31bp-merged_condensed.bed \
-header \
> ${output_dir}/Apul-shared-siRNA_clusters.bed

```


How many outputs?

```{r, engine='bash'}
wc -l ../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters.bed
```

Ignoring the header line, there are 2899993 intersecting regions.

First, we'll only keep regions identified as siRNA regions by ShortStack (present in the first input file) and present in at least two of the samples

```{r, engine='bash'}
awk '$4 >= 3 && $6 > 0' ../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters.bed > ../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters-filtered.bed

wc -l ../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters-filtered.bed
```

There are more than 142  shared clusters here, so most of these are not par of the true siRNA. Let's use intersectBed to annotate each shared segment with the ShortStack siRNA accumulation region it falls within.

```{r, engine='bash'}
/home/shared/bedtools2/bin/bedtools intersect -wa -wb -a ../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters-filtered.bed -b ../output/26-Apul-siRNA-mRNA-interactions/Apul-ShortStack-siRNA_regions.bed > ../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters-annotated.bed
```

Now read in to R and annotate with additional information on each siRNA region (provided in the ShortStack gff3)

```{r}
library(dplyr)
library(tidyr)

# Read in data
intersectBed <- read.table("../output/26-Apul-siRNA-mRNA-interactions/Apul-shared-siRNA_clusters-annotated.bed", sep="\t", header=FALSE)
siRNA_gff3 <- read.table("../output/26-Apul-siRNA-mRNA-interactions/Apul-ShortStack-siRNA_regions.gff3", sep="\t", header=FALSE)

# Create a column that contains full genomic location info
intersectBed$region <- paste(intersectBed$V12, "_", intersectBed$V13, "_", intersectBed$V14)
# For siRNA_gff, need to first convert genomic loction columns to bed format
siRNA_gff3$V4 <-siRNA_gff3$V4 - 1
siRNA_gff3$region <- paste(siRNA_gff3$V1, "_", siRNA_gff3$V4, "_", siRNA_gff3$V5)
# Remove unnecessary columns and rename columns
colnames(siRNA_gff3) <- c("seqid", "source", "type", "start", "end", "score", "strand", "phase", "attributes", "region")

# Annotate intersectBed results with the gff info
intersectBed <- left_join(intersectBed, siRNA_gff3, by="region")

# Extract cluster and dicer call info from "attributes" column
intersectBed <- intersectBed %>% separate(attributes, into = c("ID", "DicerCall", "MIRNA"), sep = ";") %>%
  mutate(ID = gsub("ID=Cluster_", "Cluster_", ID), 
         DicerCall = gsub("DicerCall=", "", DicerCall)) %>%
  select(-MIRNA)

# Calculate lengths of intersecting regions
intersectBed$length <- intersectBed$V3 - intersectBed$V2
```

```{r}
# Keep only intersections with lengths of at least the lowest DicerCall value (21)
intersectBed_filt <- intersectBed[intersectBed$length >= 21,]

# Remove intersections with lengths >=30. These may be menaingful (e.g. a precursor), but no siRNA should be longer than ~25nt.
intersectBed_filt <- intersectBed_filt[intersectBed_filt$length <= 30,]

nrow(intersectBed)
nrow(intersectBed_filt)

length(unique(intersectBed$ID))
length(unique(intersectBed_filt$ID))
```
We've filtered 490 intersecting regions to 155, and from coverage of 130 ShortStack siRNA "Clusters" to 116 (out of the 142 output by ShortStack)

In cases where there are multiple long intersections within a single ShortStack siRNA region, I can't think of a robust way to decide which intersecting region is more likely to constitute a genuine siRNA. Instead, I'll just label both as diffrent siRNAs of the same cluster for now. 

```{r}

# Initialize updated_ID col
intersectBed_filt$updated_ID <- intersectBed_filt$ID

# Append _a, _b, etc. to the IDs that are repeated
for(i in 1:nrow(intersectBed_filt)) {
  # Find all rows with the same ID
  matching_rows <- which(intersectBed_filt$ID == intersectBed_filt$ID[i])
  
  # If the ID is repeated, append _a, _b, etc.
  if(length(matching_rows) > 1) {
    # Find the position of the current row within the matching rows
    position <- which(matching_rows == i)
    
    # Append the corresponding letter (_a, _b, etc.) to the ID
    intersectBed_filt$updated_ID[i] <- paste0(intersectBed_filt$ID[i], "_", letters[position])
  }
}
```


Ok, we now have a set of putative siRNA loci. Let's save them as a gff and extract fastas for them

```{r}
intersectBed_filt$attributes <- paste0("siRNALociID=", intersectBed_filt$updated_ID, ";DicerCall=", intersectBed_filt$DicerCall, ";Length=", intersectBed_filt$length)
intersectBed_filt_gff <- intersectBed_filt %>% select(V1, source, type, V2, V3, score, strand, phase, attributes)

#convert genomic location to gff3 format
intersectBed_filt_gff$V2 <- intersectBed_filt_gff$V2 + 1

write.table(intersectBed_filt_gff, "../output/26-Apul-siRNA-mRNA-interactions/Apul-siRNA_loci.gff3", sep="\t", quote=FALSE, row.names=FALSE, col.names=FALSE)
```


```{r, engine='bash'}
/home/shared/bedtools2/bin/bedtools getfasta -fi ../data/Apulchra-genome.fa -bed ../output/26-Apul-siRNA-mRNA-interactions/Apul-siRNA_loci.gff3 -fo ../output/26-Apul-siRNA-mRNA-interactions/Apul-siRNA_loci.fa
```










# Prep for BLASTs

## Isolate the siRNA sequences

```{r extract-mirna-seqs, engine='bash'}
full_mirna_fasta="../output/11-Apul-sRNA-ShortStack_4.1.0-pulchra_genome/ShortStack_out/mir.fasta"
premirna_fasta="../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_precursor.fasta"
mature_mirna_fasta="../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature.fasta"
star_mirna_fasta="../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_star.fasta"

# Pull out all sequences that DON'T contain "mature" or "star" in sequence name
# Note the pre-miRNAs have sequences for both strands
awk '
    # If the line starts with ">", check the header
    /^>/ {
        if ($0 ~ /mature/ || $0 ~ /star/) {
            print_seq = 0  # Skip sequences with "mature" or "star" in the header
        } else {
            print_seq = 1  # Mark sequences for printing
        }
    }
    # Print the header and the next two lines if marked for printing
    print_seq {
        print
        if (!/^>/) { getline; print }  # Capture second sequence line
    }
' "$full_mirna_fasta" > "$premirna_fasta"

# Pull out all sequences that contain "mature" in sequence name
grep -A 1 "mature" $full_mirna_fasta | grep -v "^--$" > $mature_mirna_fasta

# Pull out all sequences that contain "star" in sequence name
grep -A 1 "star" $full_mirna_fasta | grep -v "^--$" > $star_mirna_fasta
```

```{r, engine='bash', eval=TRUE}
premirna_fasta="../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_precursor.fasta"
mature_mirna_fasta="../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature.fasta"
star_mirna_fasta="../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_star.fasta"

# Check we have appropriate headers, same number of sequences in each
grep "^>" $premirna_fasta | head -2
echo ""
grep "^>" $mature_mirna_fasta | head -2
echo ""
grep "^>" $star_mirna_fasta | head -2
echo ""
grep "^>" $premirna_fasta | wc -l
echo ""
grep "^>" $mature_mirna_fasta | wc -l
echo ""
grep "^>" $star_mirna_fasta | wc -l
echo ""
```

## Check miRNA lengths

```{r, engine='bash'}
# Extract sequence lengths for precursors
awk '/^>/ {if (seqlen){print seqlen}; printf $0" " ;seqlen=0;next; } { seqlen += length($0)}END{print seqlen}' ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_precursor.fasta > ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_precursor_lengths.txt

# Sequence lengths for matures
awk '/^>/ {if (seqlen){print seqlen}; printf $0" " ;seqlen=0;next; } { seqlen += length($0)}END{print seqlen}' ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature.fasta > ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature_lengths.txt
```

```{r, eval=TRUE}
# Summary stats of precursor and mature lengths

precursor_lengths <- read.table("../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_precursor_lengths.txt", sep = " ", header = FALSE, col.names = c("seqID", "length"))
mature_lengths <- read.table("../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature_lengths.txt", sep = " ", header = FALSE, col.names = c("seqID", "length"))

cat("Average pre-miRNA length: ", mean(precursor_lengths$length))
cat("\n")
cat("Range of pre-miRNA lengths: ", range(precursor_lengths$length))
cat("\n")
cat("Average mature miRNA length: ", mean(mature_lengths$length))
cat("\n")
cat("Range of mature miRNA lengths: ", range(mature_lengths$length))
```

## check lncRNAs

LncRNAs were identified from Apul RNA-seq data in `deep-dive-expression/D-Apul/code/10-Apul-lncRNA` -- see details there.
Fasta of Apul lncRNAs stored at `deep-dive-expression/D-Apul/output/10-Apul-lncRNA/Apul_lncRNA.fasta`


```{r check-lncRNAs, engine='bash', eval=TRUE}
echo "Number of lncRNAs:"
grep "^>" ../output/10-Apul-lncRNA/Apul_lncRNA.fasta | wc -l
```

```{r, engine='bash'}
# Extract sequence lengths for precursors
awk '/^>/ {if (seqlen){print seqlen}; printf $0" " ;seqlen=0;next; } { seqlen += length($0)}END{print seqlen}' ../output/10-Apul-lncRNA/Apul_lncRNA.fasta > ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_lncRNA_lengths.txt

```

```{r, eval=TRUE}
# Summary stats of lncRNA lengths

lncRNA_lengths <- read.table("../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_lncRNA_lengths.txt", sep = " ", header = FALSE, col.names = c("seqID", "length"))

cat("Average lncRNA length: ", mean(lncRNA_lengths$length))
cat("\n")
cat("Range of lncRNA lengths: ", range(lncRNA_lengths$length))

ggplot(lncRNA_lengths, aes(x = length)) +
  geom_histogram(binwidth = 500) +
  labs(title = "A. pulchra lncRNA sequence lengths",
       x = "Sequence Length [nucleotides]",
       y = "Frequency") +
  xlim(200, 50000) +
  ylim(0, 800) +
  theme_minimal()
```



# BLASTs

## Make databases

Database of pre-miRNAs:

```{r make-premirna-databse, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_precursor.fasta \
-dbtype nucl \
-out ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/Apul-db/Apul_ShortStack_4.1.0_precursor
```

Database of mature miRNAs:

```{r make-mature-mirna-databse, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/Apul_ShortStack_4.1.0_mature.fasta \
-dbtype nucl \
-out ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/Apul-db/Apul_ShortStack_4.1.0_mature
```


## Run BLASTn

Generate a list of blast results. It seems plausible that a single lncRNA, which would be hundreds or thousands of nucleotides long, could interact with multiple miRNAs, so I will allow up to 10 hits (~25% of Apul miRNAs) for each lncRNA. I want to see the top hits no matter how poor the match is, so I will not filter by e-value at this stage. I’ll also include the “-word_size 4” option, which reduces the required length of the initial match.


Full pre-miRNAs:

```{r blastn-premirna, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/blastn \
-task blastn \
-query ../output/10-Apul-lncRNA/Apul_lncRNA.fasta \
-db ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/Apul-db/Apul_ShortStack_4.1.0_precursor \
-out ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_precursor_blastn.tab \
-num_threads 40 \
-word_size 4 \
-max_target_seqs 10 \
-max_hsps 1 \
-outfmt 6

wc -l ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_precursor_blastn.tab
```
Note we have less than 241830 (10 * [# of lncRNAs]) output alignments because, while I did not set an evalue threshold, the default evalue threshold of evalue=10 is still in place. That means extremely poor matches were still excluded by default. 


Mature miRNAs:

Note that I'm using the blastn-short option here because all of our mature miRNAs are less than 30 nucleotides long (recommended by [BLAST user manual](https://www.ncbi.nlm.nih.gov/books/NBK279684/table/appendices.T.blastn_application_options/))

```{r blastn-mature-mirna, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/blastn \
-task blastn \
-query ../output/10-Apul-lncRNA/Apul_lncRNA.fasta \
-db ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/Apul-db/Apul_ShortStack_4.1.0_mature \
-out ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_mature_blastn.tab \
-num_threads 40 \
-word_size 4 \
-max_target_seqs 10 \
-max_hsps 1 \
-outfmt 6

wc -l ../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_mature_blastn.tab
```


# Examine BLAST tables

Read into R and assign informative column labels
```{r read-in-blast-tables, eval=TRUE}
precursor_lncRNA_BLASTn <- read.table("../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_precursor_blastn.tab", sep="\t", header=FALSE)
mature_lncRNA_BLASTn <- read.table("../output/17-Apul-miRNA-lncRNA-BLASTs-RNAhybrid/blasts/lncRNA_to_mature_blastn.tab", sep="\t", header=FALSE)

colnames(precursor_lncRNA_BLASTn) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
colnames(mature_lncRNA_BLASTn) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
```





