---
title: "30.00-Apul-transcriptome-GOslims"
author: "Sam White"
date: "2025-06-17"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

# BACKGROUND

This notebook will perform annotation of expressed genes, as previously determined by [`07-Apul-Hisat.qmd`](https://github.com/urol-e5/deep-dive-expression/blob/1af5c4c898b1770e7ec5f2e4f3a07f8a5ffb72bb/D-Apul/code/07-Apul-Hisat.qmd) (GitHub).

Briefly, the notebook will perform the following tasks:

1.  Extract all genes from the genome, as GFF and FastA.

2.  Create a subset of only *expressed* genes, based on gene count matrix.

3.  BLASTx expressed genes against SwissProt database.

4.  Get gene ontology.

5.  Map gene ontology to GOslims and get counts.

::: callout-important
Expressed genes were defined as those genes having at least one count across all samples.
:::

## INPUTS

-   Gene count matrix

-   Genome FastA

-   Genome GFF

## OUTPUTS

-   Genes BED

-   Genes FastA

-   Expressed genes FastA

-   Expressed genes SwissProt IDs only file.

-   Expressed genes to SwissProt IDs mapping file.

-   Expressed genes to SwissProtIDs and GO mapping file.

-   Counts file of expressed genes GOslims.

## SOFTWARE

-   [DIAMOND BLAST](https://github.com/bbuchfink/diamond) [@buchfink2021]

-   [bedtools](https://github.com/arq5x/bedtools2) [@quinlan2010]

-   [samtools](https://github.com/samtools/samtools) [@danecek2021]

-   [Biostrings](https://bioconductor.org/packages/release/bioc/html/Biostrings.html) [@h.pagès2017] (Bioconductor R package)

-   [GO.db](https://www.bioconductor.org/packages/release/data/annotation/html/GO.db.html) [@carlson2017] (Bioconductor R package)

-   [GSEABase](https://www.bioconductor.org/packages/release/bioc/html/GSEABase.html) [@martinmorgan2017] (Bioconductor R package)

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(Biostrings)
library(GSEABase)
library(GO.db)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# VARIABLES

```{r variables}
# PROGRAMS
blastdbs_dir <- file.path("..", "..", "M-multi-species", "data", "blastdbs")
programs_dir <- file.path("", "home", "shared")
bedtools_dir <- file.path(programs_dir, "bedtools-v2.30.0", "bin")
blast_dir <- file.path(programs_dir, "ncbi-blast-2.15.0+", "bin")
diamond <- file.path(programs_dir, "diamond-2.1.8")
fastaFromBed <- file.path(bedtools_dir, "fastaFromBed")
samtools_dir <- file.path(programs_dir, "samtools-1.12")
samtools <- file.path(samtools_dir, "samtools")



# FILES
count_matrix <- "../output/07-Apul-Hisat/Apul-gene_count_matrix.csv"
diamond_db <- "20250618-diamond"
diamond_output <- file.path("..", "output", "30.00-Apul-transcriptome-GOslims", "Apulchra-expressed-genes.blastx.outfmt6")
genome_fasta <- file.path("..", "data", "Apulcra-genome.fa")
genes_fasta <- file.path("..", "output", "30.00-Apul-transcriptome-GOslims", "Apulchra-genes.fasta")
genes_fasta_index <- file.path("..", "output", "30.00-Apul-transcriptome-GOslims", "Apulchra-genes.fasta.fai")
genes_subset_fasta <- file.path("..", "output", "30.00-Apul-transcriptome-GOslims", "Apulchra-subset-genes.fasta")
genes_subset_fasta_index <- file.path("..", "output", "30.00-Apul-transcriptome-GOslims", "Apulchra-subset-genes.fasta.fai")
genes_bed <- file.path("..", "output", "30.00-Apul-transcriptome-GOslims", "Apulchra-genes.bed")
og_genome_gff <- file.path("..", "data", "Apulcra-genome.gff")

# THREADS
threads <- "40"

##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"

# FORMATTING
line <- "-----------------------------------------------"

# Export these as environment variables for bash chunks.
Sys.setenv(
  blastdbs_dir = blastdbs_dir,
  count_matrix = count_matrix,
  diamond = diamond,
  diamond_db = diamond_db,
  diamond_output = diamond_output,
  fastaFromBed = fastaFromBed,
  genes_bed = genes_bed,
  genes_fasta = genes_fasta,
  genes_fasta_index = genes_fasta_index,
  genes_subset_fasta = genes_subset_fasta,
  genes_subset_fasta_index = genes_subset_fasta_index,
  genome_fasta = genome_fasta,
  og_genome_gff = og_genome_gff,
  line = line,
  samtools = samtools,
  threads = threads,
  transcripts_fasta = transcripts_fasta,
  transcripts_gtf = transcripts_gtf
)
```

# Extract genes as FastA

## Extract genes from GFF

```{r extract-genes-from-gff}

# Read GFF, skipping comment lines
genome_gff <- readr::read_tsv(
  og_genome_gff,
  comment = "#",
  col_names = c("seqid", "source", "type", "start", "end", "score", "strand", "phase", "attributes")
)

# Filter for gene features
genes <- as.data.frame(genome_gff) %>%
  filter(type == "gene") %>%
  mutate(
    chrom = seqid,
    start = start - 1,  # BED is 0-based
    end = end,
    gene_id = sub("ID=([^;]+);?.*", "\\1", attributes)
  ) %>%
  select(chrom, start, end, gene_id)

str(genes)
```

### Write genes to BED

```{r write-genes-to-bed}
# Write to BED
write_tsv(genes, genes_bed, col_names = FALSE)
```

## Create genes FastA

```{r create-genes-FastA, engine='bash'}

"${fastaFromBed}" -fi "${genome_fasta}" -bed "${genes_bed}" -nameOnly > "${genes_fasta}"

# Create FastA index
"${samtools}" faidx "${genes_fasta}"

head "${genes_fasta}"

echo ""
echo ""

head "${genes_fasta_index}"
```

# Subset Expressed Genes

Only those with at least *one* read in each sample

### Peek at count matrix

```{r check-count-matrix, engine=bash}
head "${count_matrix}" | column -t -s","

echo ""
echo ""

wc -l "${count_matrix}"
```

## Import count matrix

```{r import-count-matrix}

# Read the data into a data frame
count_matrix_df <- read.csv(count_matrix, header = TRUE)

str(count_matrix_df)
```

## Only genes with at least one read per sample

```{r subset-count-matrix}
# Filter rows where all values are greater than 0
filtered_count_matrix_df <- count_matrix_df[apply(count_matrix_df > 0, 1, all), ]


str(filtered_count_matrix_df)
```

## Subset genes FastA

Only expressed genes

```{r subset-genes-FastA}
# Get the row names (gene_ids) of the filtered data frame
filtered_gene_ids <- filtered_count_matrix_df$gene_id

fasta <- readDNAStringSet(genes_fasta)

subset_fasta <- fasta[names(fasta) %in% filtered_gene_ids]

```

```{r write-genes-subset-fasta}
writeXStringSet(subset_fasta, genes_subset_fasta)
```

### Peek at genes subset FastA

```{r check-genes-subset-fasta, engine=bash}
head "${genes_subset_fasta}"

echo ""
grep "^>" --count "${genes_subset_fasta}"
```

# BLASTx

## Download SwissProt

```{r download-sp, engine=bash}
cd "${blastdbs_dir}"

curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz

mv uniprot_sprot.fasta.gz 20250618-uniprot_sprot.fasta.gz

gunzip --keep 20250618-uniprot_sprot.fasta.gz
```

## Create BLASTdb

```{r create-BLASTdb, engine=bash}
cd "${blastdbs_dir}"

"${diamond}" makedb \
--in 20250618-uniprot_sprot.fasta.gz \
--db "${diamond_db}" \
--quiet \
--threads "${threads}"
```

## Run DIAMOND BLASTx

```{r diamond-blast, engine=bash}
"${diamond}" blastx \
--db "${blastdbs_dir}"/"${diamond_db}.dmnd" \
--query "${genes_subset_fasta}" \
--out "${diamond_output}" \
--outfmt 6  \
--sensitive \
--evalue 1e-10 \
--max-target-seqs 1 \
--block-size 15.0 \
--index-chunks 4 \
--threads "${threads}" \
2> ../output/30.00-Apul-transcriptome-GOslims/diamond-blastx.log

head "${diamond_output}"

echo ""

wc -l "${diamond_output}"
```

# GENE ONTOLOGY

## Get gene IDs and SwissProt IDs

```{r gene-and-SPIDs, engine=bash}
awk -F"|" '{print $1"\t"$2}' "${diamond_output}" \
| awk '{print $1"\t"$3}' \
> ../output/30.00-Apul-transcriptome-GOslims/gene-SPIDs.txt

head ../output/30.00-Apul-transcriptome-GOslims/gene-SPIDs.txt

echo ""
echo ""

wc -l ../output/30.00-Apul-transcriptome-GOslims/gene-SPIDs.txt
```

## Get SwissProt IDs

```{r get-SPIDs, engine=bash}
awk -F"|" '{print $2}' "${diamond_output}" \
| sort --unique \
> ../output/30.00-Apul-transcriptome-GOslims/SPIDs.txt

head ../output/30.00-Apul-transcriptome-GOslims/SPIDs.txt

echo ""
echo ""

wc -l ../output/30.00-Apul-transcriptome-GOslims/SPIDs.txt
```

## Retrieve UniProt records

A difference in number of records could be due to retrieval from only "reviewed" records, while BLAST may have included both "reviewed" and "unreviewed." SwissProt records

```{r uniprot-records, engine=bash}
python3 \
../../M-multi-species/code/uniprot-retrieval.py \
../output/30.00-Apul-transcriptome-GOslims/SPIDs.txt \
../output/30.00-Apul-transcriptome-GOslims/

gunzip ../output/30.00-Apul-transcriptome-GOslims/uniprot-retrieval.tsv.gz

echo ""
echo ""

wc -l ../output/30.00-Apul-transcriptome-GOslims/uniprot-retrieval.tsv
```

## Map GO to Genes

```{r merge-genes-GOIDs}
# Read gene-SPIDs.txt
gene_spids <- read.delim("../output/30.00-Apul-transcriptome-GOslims/gene-SPIDs.txt", header = FALSE, stringsAsFactors = FALSE)
colnames(gene_spids) <- c("GeneID", "Entry")

# Read uniprot-retrieval.tsv
uniprot <- read.delim("../output/30.00-Apul-transcriptome-GOslims/uniprot-retrieval.tsv", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)

# Merge on Entry
gene_SPID_GOID_merged <- merge(gene_spids, uniprot[, c("Entry", "Gene Ontology IDs")], by = "Entry", all.x = TRUE)

# View result
str(gene_SPID_GOID_merged)
```

### Write merged to file

```{r write-merged-gene-SP-GO}
# Optionally, write to file
write.table(gene_SPID_GOID_merged, "../output/30.00-Apul-transcriptome-GOslims/gene-SPIDs-GOIDs.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

## Clean up merged

```{r remove-NA-and-uniprotIDs, eval=TRUE}

full.gene.df <- as.data.frame(gene_SPID_GOID_merged)

# Clean whitespace, filter NA/empty rows, select columns, and split GO terms using column name variables
gene.GO.df <- full.gene.df %>%
  mutate(!!"Gene Ontology IDs" := str_replace_all(.data[["Gene Ontology IDs"]], "\\s*;\\s*", ";")) %>% # Clean up spaces around ";"
  filter(!is.na(.data[["GeneID"]]) & !is.na(.data[["Gene Ontology IDs"]]) & .data[["Gene Ontology IDs"]] != "") %>% 
  dplyr::select(all_of(c("GeneID", "Gene Ontology IDs")))


str(gene.GO.df)
```

## Flatten gene GOID file

```{r flatten-gene-and-GO-IDs, eval=TRUE}
flat.gene.GO.df <- gene.GO.df %>% separate_rows(!!sym("Gene Ontology IDs"), sep = ";")

str(flat.gene.GO.df)
```

## Group by GOID

```{r group-by-GO, eval=TRUE}
grouped.gene.GO.df <- flat.gene.GO.df %>%
  group_by(!!sym("Gene Ontology IDs")) %>%
  summarise(!!"GeneID" := paste(.data[["GeneID"]], collapse = ","))

str(grouped.gene.GO.df)
```

## Vectorize GOIDs

```{r vectorize-GOIDs, eval=TRUE}
# Vector of GO IDs
go_ids <- grouped.gene.GO.df[["Gene Ontology IDs"]]

str(go_ids)
```

## Prepare GOslim OBO

```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destintation <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destintation)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

## GOslims from OBO

```{r extract-GOslims-from-OBO, eval=TRUE}

# Create GSEAbase GOCollection using `go_ids`
myCollection <- GOCollection(go_ids)

# Retrieve GOslims from GO OBO file set
slim <- getOBOCollection(gseabase_files)

str(slim)
```

## Biological Process GOslims

```{r retrieve-BP-GOslims, eval=TRUE}
# Retrieve Biological Process (BP) GOslims
slimdf <- goSlim(myCollection, slim, "BP", verbose)

str(slimdf)
```

## Map GO to GOslims

```{r map-GO-to-GOslims, eval=TRUE}
# List of GOslims and all GO IDs from `go_ids`
gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

# Maps `go_ids` to matching GOslims
mapped <- lapply(gomap, intersect, ids(myCollection))

# Append all mapped GO IDs to `slimdf`
# `sapply` needed to apply paste() to create semi-colon delimited values
slimdf$GO.IDs <- sapply(lapply(gomap, intersect, ids(myCollection)), paste, collapse=";")

# Remove "character(0) string from "GO.IDs" column
slimdf$GO.IDs[slimdf$GO.IDs == "character(0)"] <- ""

# Add self-matching GOIDs to "GO.IDs" column, if not present
for (go_id in go_ids) {
  # Check if the go_id is present in the row names
  if (go_id %in% rownames(slimdf)) {
    # Check if the go_id is not present in the GO.IDs column
    # Also removes white space "trimws()" and converts all to upper case to handle
    # any weird, "invisible" formatting issues.
    if (!go_id %in% trimws(toupper(strsplit(slimdf[go_id, "GO.IDs"], ";")[[1]]))) {
      # Append the go_id to the GO.IDs column with a semi-colon separator
      if (length(slimdf$GO.IDs) > 0 && nchar(slimdf$GO.IDs[nrow(slimdf)]) > 0) {
        slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
      } else {
        slimdf[go_id, "GO.IDs"] <- go_id
      }
    }
  }
}

str(slimdf)
```

## Flatten GOslims

```{r flatten-GOslims-file, eval=TRUE}
# "Flatten" file so each row is single GO ID with corresponding GOslim
# rownames_to_column needed to retain row name info
slimdf_separated <- as.data.frame(slimdf %>%
  rownames_to_column('GOslim') %>%
  separate_rows(GO.IDs, sep = ";"))

# Group by unique GO ID
grouped_slimdf <- slimdf_separated %>%
  filter(!is.na(GO.IDs) & GO.IDs != "") %>%
  group_by(GO.IDs) %>%
  summarize(GOslim = paste(GOslim, collapse = ";"),
            Term = paste(Term, collapse = ";"))


str(grouped_slimdf)
```

## Counts of GOslims

```{r sort-and-select-slimdf-counts, eval=TRUE}

slimdf.sorted <- slimdf %>% arrange(desc(Count))

slim.count.df <- slimdf.sorted %>% 
  dplyr::select(Term, Count, Percent)

str(slim.count.df)
```

### Write GOslims to file

Need to create a column name for GOslimIDs from data frame rownames.

```{r write-goslim-counts-to-file}
# Create header vector
header <- c("GOslimID", colnames(slim.count.df))

# Write header to file
writeLines(paste(header, collapse = "\t"), "../output/30.00-Apul-transcriptome-GOslims/GOslim-counts.tsv")

# Append data frame contents to existing file, which contains header info
write.table(
  slim.count.df,
  "../output/30.00-Apul-transcriptome-GOslims/GOslim-counts.tsv",
  sep = "\t",
  row.names = TRUE,
  quote = FALSE,
  col.names = FALSE,
  append = TRUE
)
```

# REFERENCES
