---
title: "07.2- A pulcra HiSat2 - with splice"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`"  
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```



# Run HiSat on RNA-seq

Will end up with 5 sorted bam files. 

## Grab Trimmed RNA-seq Reads

```{r, engine='bash'}
cd ../data/fastq/
echo "*.fq" >> .gitignore
```

```{r, engine='bash'}
wget -r \
--no-directories --no-parent \
-P ../data/fastq/ \
-A "*fastq.gz" https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/

```




## Genome

```{r, engine='bash'}
cd ../data

wget -O Apulcra-genome.fa "https://osf.io/download/kn96u/"
```


```{r, engine='bash'}
head ../data/Apulcra-genome.fa
```


## Grab GTF 

```{r, engine='bash'}
cd ../data

wget -O Apulchra-genome.gtf "https://raw.githubusercontent.com/urol-e5/timeseries_molecular/refs/heads/main/D-Apul/data/Apulchra-genome.gtf"
```



## HiSat

```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2_extract_exons.py \
../data/Apulchra-genome.gtf \
> ../output/07.2-Apul-Hisat/m_exon.tab
```

```{r, engine='bash', eval=TRUE}
head ../output/07.2-Apul-Hisat/m_exon.tab
wc -l ../output/07.2-Apul-Hisat/m_exon.tab
```

```{r, engine='bash'}

/home/shared/hisat2-2.2.1/hisat2_extract_splice_sites.py \
../data/Apulchra-genome.gtf \
> ../output/07.2-Apul-Hisat/m_splice_sites.tab

```


```{r, engine='bash', eval=TRUE}
head ../output/07.2-Apul-Hisat/m_splice_sites.tab
```



```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2-build \
../data/Apulcra-genome.fa \
../output/07.2-Apul-Hisat/Apulcra-genome.index \
--exon ../output/07.2-Apul-Hisat/m_exon.tab \
--ss ../output/07.2-Apul-Hisat/m_splice_sites.tab \
-p 48 \
../data/Apulcra-genome.gtf \
2> ../output/07.2-Apul-Hisat/hisat2-build_stats.txt
```



```{r, engine='bash', eval=TRUE}
tail ../output/07.2-Apul-Hisat/hisat2-build_stats.txt
```


```{r, engine='bash'}
cd ../output/07-Apul-Hisat/
echo "*sam" >> .gitignore
```


```{r, engine='bash'}
find ../data/fastq/*R2_001.fastp-trim.20230519.fastq.gz \
| xargs basename -s -S1-TP2_R2_001.fastp-trim.20230519.fastq.gz | xargs -I{} \
/home/shared/hisat2-2.2.1/hisat2 \
-x ../output/07.2-Apul-Hisat/Apulcra-genome.index \
-p 48 \
-1 ../data/fastq/{}-S1-TP2_R1_001.fastp-trim.20230519.fastq.gz \
-2 ../data/fastq/{}-S1-TP2_R2_001.fastp-trim.20230519.fastq.gz \
-S ../output/07.2-Apul-Hisat/{}.sam \
2> ../output/07.2-Apul-Hisat/hisat.out
```

```{r, engine='bash', eval=TRUE}
cat ../output/07.2-Apul-Hisat/hisat.out
```

## convert to bams

```{r, engine='bash'}
for samfile in ../output/07.2-Apul-Hisat/*.sam; do
  bamfile="${samfile%.sam}.bam"
  sorted_bamfile="${samfile%.sam}.sorted.bam"
  
  # Convert SAM to BAM
  /home/shared/samtools-1.12/samtools view -bS -@ 20 "$samfile" > "$bamfile"
  
  # Sort BAM
  /home/shared/samtools-1.12/samtools sort -@ 20 "$bamfile" -o "$sorted_bamfile"
  
  # Index sorted BAM
  /home/shared/samtools-1.12/samtools index -@ 20 "$sorted_bamfile"
done
```



dd

# StringTie
StringTie uses the sorted BAM files to assemble transcripts for each sample, outputting them as GTF (Gene Transfer Format) files. And then merges all individual GTF assemblies into a single merged GTF file. This step extracts transcript information and merges GTFs from all samples--an important step in creating a canonical list of lncRNAs across all samples included in the pipeline.

Getting gff

```{r, engine='bash'}

<!-- cd ../data -->
<!-- wget -O Apulcra-genome.gff "https://osf.io/download/f9dbr/" -->
```


```{r, engine='bash'}

head ../data/Apulcra-genome.gff

```


```
-eB \
-e: Only estimate expression for transcripts that are present in the reference annotation provided with the -G option.
-B: Enables the creation of a ballgown output table, useful for downstream analysis with the R ballgown package.
```


```{r, engine='bash'}
find ../output/07.2-Apul-Hisat/*sorted.bam \
| xargs basename -s .sorted.bam | xargs -I{} \
/home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
-p 32 \
-eB \
-G ../data/Apulcra-genome.gff \
-o ../output/07.2-Apul-Hisat/{}.gtf \
../output/07.2-Apul-Hisat/{}.sorted.bam
```




```{r, engine='bash', eval=TRUE}
wc -l ../output/07.2-Apul-Hisat/RNA*.gtf
ls ../output/07.2-Apul-Hisat/RNA*.gtf
#head ../output/07.2-Apul-Hisat/RNA*.gtf
```



# Count Matrix

list format

RNA-ACR-140 ../output/15-Apul-hisat/RNA-ACR-140.gtf
RNA-ACR-145 ../output/15-Apul-hisat/RNA-ACR-145.gtf
RNA-ACR-173 ../output/15-Apul-hisat/RNA-ACR-173.gtf
RNA-ACR-178 ../output/15-Apul-hisat/RNA-ACR-178.gtf


```{r, engine='bash'}
cat ../output/07.2-Apul-Hisat/list01.txt
```



```{r, engine='bash'}
python /home/shared/stringtie-2.2.1.Linux_x86_64/prepDE.py \
-i ../output/07.2-Apul-Hisat/list01.txt \
-g ../output/07.2-Apul-Hisat/gene_count_matrix.csv \
-t ../output/07.2-Apul-Hisat/transcript_count_matrix.csv

head ../output/07.2-Apul-Hisat/*matrix.csv

```


```{bash}
head ../output/07.2-Apul-Hisat/*matrix.csv
```


