---
title: "Apul MixOmics"
output: html_document
date: "2024-10-01"
editor_options: 
  chunk_output_type: console
---

Things are very highly correlated - mixOmics is not a good framework for us to use within species. This will maybe be more helpful for between species analyses.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Script to analyzed relationships between RNA-sRNA-lncRNA using mixOmics.  

# Set up 

Load libraries 
```{r}
library(tidyverse)
library(ggplot2)
library(mixOmics)
```

# Read in files 

Read in count matrices.  

mrna is missing sample 150, removing from all for now, will revisit and add this in after analysis is re done (SR to do) 
```{r}
mrna<-read.csv("F-Ptuh/output/06-Ptuh-Hisat/Ptuh-gene_count_matrix.csv")%>%rename("sample47"=RNA.POC.47, "sample48"=RNA.POC.48, "sample50"=RNA.POC.50, "sample53"=RNA.POC.53, "sample57"=RNA.POC.57)

rownames(mrna)<-mrna$gene_id

mrna<-mrna%>%dplyr::select(!gene_id)

srna<-read.table("F-Ptuh/output/03.1-Ptuh-sRNA-summary/Ptuh_counts_sRNA_normalized.txt")

lncrna<-read.table("F-Ptuh/output/08-Ptuh-lncRNA-matrix/Peve-lncRNA-counts.txt", header=T)%>%rename("sample47"=`...output.06.Ptuh.Hisat.RNA.POC.47.sorted.bam`, "sample48"=`...output.06.Ptuh.Hisat.RNA.POC.48.sorted.bam`, "sample50"=`...output.06.Ptuh.Hisat.RNA.POC.50.sorted.bam`, "sample53"=`...output.06.Ptuh.Hisat.RNA.POC.53.sorted.bam`, "sample57"=`...output.06.Ptuh.Hisat.RNA.POC.57.sorted.bam` )%>%dplyr::select(Geneid, sample47, sample48, sample50, sample53, sample57)

rownames(lncrna)<-lncrna$Geneid

lncrna<-lncrna%>%dplyr::select(!Geneid)

metadata <- data.frame(
  sample = c("sample48", "sample50", "sample53", "sample57", "sample47"),
  species = "Ptuh"
)
```

Follow the mixOmics tutorial for conducting correlations between these two data sets. 

All data sets have dimensions of 4 samples. We can add additional species to this as we go.  

# Conduct mixOmics  

Set matrices. 

```{r}
mrna<-t(mrna)
srna<-t(srna)
lncrna<-t(lncrna)

dim(mrna)
dim(srna)
dim(lncrna)
dim(metadata)
```

Remove columns with 0's in each matrix or with 0 variance (all values identical) 

```{r}
mrna <- mrna[, colSums(mrna) != 0]

mrna<-as.data.frame(mrna)
mrna <- mrna[, sapply(mrna, function(x) length(unique(x)) > 1)]
dim(mrna)

srna <- srna[, colSums(srna) != 0]
dim(srna)

srna<-as.data.frame(srna)
srna <- srna[, sapply(srna, function(x) length(unique(x)) > 1)]
dim(srna)

lncrna <- lncrna[, colSums(lncrna) != 0]
dim(lncrna)

lncrna<-as.data.frame(lncrna)
lncrna <- lncrna[, sapply(lncrna, function(x) length(unique(x)) > 1)]
dim(lncrna)
```

No lncRNAs in Ptuh

## Conduct preliminary PCA  

```{r}
pca.mrna <- pca(mrna, ncomp = 2, center = TRUE, scale = TRUE)
pca.srna <- pca(srna, ncomp = 2, center = TRUE, scale = TRUE)
#pca.lncrna <- pca(lncrna, ncomp = 2, center = TRUE, scale = TRUE)

plot(pca.mrna)
plot(pca.srna)
#plot(pca.lncrna)
```

Similar variance explained because we are dealing with individual samples within a species. 

```{r}
plotIndiv(pca.mrna, comp = c(1, 2), 
          group = metadata$species, 
          ind.names = metadata$sample, 
          legend = TRUE, title = 'mRNA, PCA comp 1 - 2')

plotIndiv(pca.srna, comp = c(1, 2), 
          group = metadata$species, 
          ind.names = metadata$sample, 
          legend = TRUE, title = 'sRNA, PCA comp 1 - 2')

#plotIndiv(pca.lncrna, comp = c(1, 2), 
#          group = metadata$species, 
#          ind.names = metadata$sample, 
#          legend = TRUE, title = 'lncRNA, PCA comp 1 - 2')
```

# Multi block PLS with DIABLO 

```{r}
X1 <- mrna
X2 <- srna 
#X3 <- lncrna
X <- list(mrna = X1, srna = X2)

Y <-metadata$sample
```

```{r}
result.diablo <- block.plsda(X, Y) # run the method
plotIndiv(result.diablo) # plot the samples
plotVar(result.diablo) # plot the variables
```

sparse 
```{r}
list.keepX = list(mrna = c(20, 2), srna = c(20,2)) 

# run the method
result.sparse.diablo <-  block.splsda(X, Y, keepX = list.keepX) 

# plot the contributions of each feature to each dimension
plotLoadings(result.sparse.diablo, ncomp = 1) 
plotIndiv(result.sparse.diablo) # plot the samples
plotVar(result.sparse.diablo) # plot the variables
```

```{r}
color.edge <- color.GreenRed(50)  # set the colours of the connecting lines

network(result.sparse.diablo, 
        cutoff = 0.99, # only show connections with a correlation above threshold
        shape.node = c("rectangle", "circle"),
        color.node = c("cyan", "pink"),
        color.edge = color.edge, 
        interactive = FALSE,
        lwd.edge = 2)
```

```{r}
circosPlot(result.sparse.diablo, cutoff=0.99, comp=1:2)
```
