---
title: "mRNA-miRNA interactions functional enrichment"
author: "Jill Ashey"
date: "2025-02-07"
output: html_document
---

This script will use topGO to analyze functional enrichment of miRNA targets for Ptuh 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(topGO)
library(tidyverse)
```

Set up gene2go file 
```{r}
# Read in Apul annotations
annot <- read.delim("../../F-Ptuh/output/02-Ptuh-reference-annotation/Pocillopora_meandrina_HIv1-IDmapping-2024_09_04.tab")

# Format data
annot_go <- annot %>%
  select(V1, Gene.Ontology.IDs) %>%
  separate_rows(Gene.Ontology.IDs, sep = "; ") %>%
  group_by(V1) %>%
  summarize(GO_terms = paste(Gene.Ontology.IDs, collapse = "; "))

# Save as gene2go file name 
write_tsv(annot_go, "../../M-multi-species/data/Pmea_gene2go.tab")
```

## Functional enrichment of all targets, as predicted by miranda

TO DO

## Functional enrichment of targets negatively correlated with miRNAs (regardless of correlation significance)

Read in PCC miranda data
```{r}
data <- read.csv("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-miranda_PCC_miRNA_mRNA.csv")
head(data)

# Filter so that only negative correlations remain 
neg_corr_data <- data %>%
  filter(PCC.cor < 0)
```

Read in gene2go information 
```{r}
ptuh_gene2go <- read.delim("../../M-multi-species/data/Pmea_gene2go.tab", sep = "\t")
colnames(ptuh_gene2go) <- c("mRNA", "GO.ID")
```

Make list of genes for input to topGO
```{r}
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(neg_corr_data$mRNA))

# All genes 
all_genes <- as.character(ptuh_gene2go$mRNA)

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs, regardless of correlation significance. 

Read in gene-to-go-mappings
```{r}
gene2go_topgo <- readMappings("../../M-multi-species/data/Pmea_gene2go.tab", IDsep = ";", sep = "\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_neg_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_neg_corr_df`
```{r}
test <- GO_neg_corr_df %>%
  inner_join(neg_corr_data, by = "mRNA") #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_neg_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_neg_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_neg_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_neg_corr_target_enrichment.pdf", plot, width = 20, height = 25, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_neg_corr_target_enrichment.png", plot, width = 20, height = 25, dpi = 300)
```

## Functional enrichment of targets positively correlated with miRNAs (regardless of correlation significance)

Read in PCC miranda data
```{r}
data <- read.csv("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-miranda_PCC_miRNA_mRNA.csv")
head(data)

# Filter so that only negative correlations remain 
pos_corr_data <- data %>%
  filter(PCC.cor > 0)
```

Read in gene2go information 
```{r}
ptuh_gene2go <- read.delim("../../M-multi-species/data/Pmea_gene2go.tab", sep = "\t")
colnames(ptuh_gene2go) <- c("mRNA", "GO.ID")
```

Make list of genes for input to topGO
```{r}
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(pos_corr_data$mRNA))

# All genes 
all_genes <- as.character(ptuh_gene2go$mRNA)

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs, regardless of correlation significance. 

Read in gene-to-go-mappings
```{r}
gene2go_topgo <- readMappings("../../M-multi-species/data/Pmea_gene2go.tab", IDsep = ";", sep = "\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_pos_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_pos_corr_df`
```{r}
test <- GO_pos_corr_df %>%
  inner_join(pos_corr_data, by = "mRNA") #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_pos_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_pos_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_pos_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_pos_corr_target_enrichment.pdf", plot, width = 20, height = 25, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_pos_corr_target_enrichment.png", plot, width = 20, height = 25, dpi = 300)
```

## Functional enrichment of targets significantly negatively correlated with miRNAs (regardless of correlation significance)

Read in PCC miranda data
```{r}
data <- read.csv("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-miranda_PCC_miRNA_mRNA.csv")
head(data)

# Filter so that only negative correlations remain 
sig_neg_corr_data <- data %>%
  filter(PCC.cor < 0) %>%
  filter(p_value < 0.05)
```

Read in gene2go information 
```{r}
ptuh_gene2go <- read.delim("../../M-multi-species/data/Pmea_gene2go.tab", sep = "\t")
colnames(ptuh_gene2go) <- c("mRNA", "GO.ID")
```

Make list of genes for input to topGO
```{r}
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(sig_neg_corr_data$mRNA))

# All genes 
all_genes <- as.character(ptuh_gene2go$mRNA)

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs, regardless of correlation significance. 

Read in gene-to-go-mappings
```{r}
gene2go_topgo <- readMappings("../../M-multi-species/data/Pmea_gene2go.tab", IDsep = ";", sep = "\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_sig_neg_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_sig_neg_corr_df`
```{r}
test <- GO_sig_neg_corr_df %>%
  inner_join(sig_neg_corr_data, by = "mRNA") #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_sig_neg_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_sig_neg_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_sig_neg_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_sig_neg_corr_target_enrichment.pdf", plot, width = 20, height = 25, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_sig_neg_corr_target_enrichment.png", plot, width = 20, height = 25, dpi = 300)
```

## Functional enrichment of targets positively correlated with miRNAs (regardless of correlation significance)

Read in PCC miranda data
```{r}
data <- read.csv("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-miranda_PCC_miRNA_mRNA.csv")
head(data)

# Filter so that only negative correlations remain 
sig_pos_corr_data <- data %>%
  filter(PCC.cor > 0) %>%
  filter(p_value < 0.05)
```

Read in gene2go information 
```{r}
ptuh_gene2go <- read.delim("../../M-multi-species/data/Pmea_gene2go.tab", sep = "\t")
colnames(ptuh_gene2go) <- c("mRNA", "GO.ID")
```

Make list of genes for input to topGO
```{r}
# Genes of interest - ie those targeted by miRNAs
target_genes <- as.character(unique(sig_pos_corr_data$mRNA))

# All genes 
all_genes <- as.character(ptuh_gene2go$mRNA)

# Apply 1 or 0 if gene is gene of interest 
GeneList <- factor(as.integer(all_genes %in% target_genes))
names(GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs, regardless of correlation significance. 

Read in gene-to-go-mappings
```{r}
gene2go_topgo <- readMappings("../../M-multi-species/data/Pmea_gene2go.tab", IDsep = ";", sep = "\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_BP_FE <- runTest(GO_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_BP_En <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_BP_En$Fisher<-as.numeric(GO_BP_En$Fisher)
GO_BP_En_sig<-GO_BP_En[GO_BP_En$Fisher<0.05,]
```

Merge `GO_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_BP_En_sig$GO.ID <- trimws(GO_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_BP_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_BP_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_BP_En_sig_gene$ontology <- "Biological Processes"

# Keep only unique rows 
GO_BP_En_sig_gene <- unique(GO_BP_En_sig_gene)
```

### Cellular Components

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_MF_FE <- runTest(GO_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_MF_En <- GenTable(GO_MF, Fisher = GO_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_MF_En$Fisher<-as.numeric(GO_MF_En$Fisher)
GO_MF_En_sig<-GO_MF_En[GO_MF_En$Fisher<0.05,]
```

Merge `GO_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
ptuh_gene2go <- ptuh_gene2go %>%
  separate_rows(GO.ID, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
ptuh_gene2go$GO.ID <- trimws(ptuh_gene2go$GO.ID)
GO_MF_En_sig$GO.ID <- trimws(GO_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_MF_En_sig_gene <- ptuh_gene2go %>%
  left_join(GO_MF_En_sig, by = "GO.ID") %>%
  na.omit()

# Add ontology column 
GO_MF_En_sig_gene$ontology <- "Molecular Functions"

# Keep only unique rows 
GO_MF_En_sig_gene <- unique(GO_MF_En_sig_gene)
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_sig_pos_corr_df <- rbind(GO_BP_En_sig_gene, GO_MF_En_sig_gene)
```

Merge with `GO_pos_corr_df`
```{r}
test <- GO_sig_pos_corr_df %>%
  inner_join(sig_pos_corr_data, by = "mRNA") #%>%
  #mutate(direction = ifelse(PCC.cor > 0, "Positive", "Negative")) #%>%
  #filter(ontology != "Cellular Components") #%>%
  #filter(p_value < 0.1)

# Save as csv 
write.csv(test, "../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_sig_pos_corr_target_enrichment.csv")
```

Plot!
```{r}
plot<-ggplot(test, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(shape = 21, size = 5) + 
  #scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 24) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_sig_pos_corr_target_enrichment.pdf", plot, width = 20, height = 35, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-topGO_sig_pos_corr_target_enrichment.png", plot, width = 20, height = 35, dpi = 300)
```

Examine the top 10 most significant GO terms for BP and MF
```{r}
# Function to get top 5 unique terms
get_top_10_unique <- function(data, ontology_type) {
  data %>%
    filter(ontology == ontology_type) %>%
    arrange(Fisher) %>%
    distinct(Term, .keep_all = TRUE) %>%
    slice_head(n = 10)
}

# Get top 5 unique Biological Processes
top_10_BP <- get_top_10_unique(test, "Biological Processes")

# Get top 5 unique Molecular Functions
top_10_MF <- get_top_10_unique(test, "Molecular Functions")

# Combine results
top_10_combined <- bind_rows(top_10_BP, top_10_MF)
unique(top_10_combined$Term)

# Plot 
top_10_combined <- top_10_combined %>%
  arrange(desc(Fisher)) %>%
  mutate(Term = factor(Term, levels = unique(Term)))

plot<-ggplot(top_10_combined, aes(x = Term, y = Fisher, fill = p_value)) +
  #ylim(0, 1) +
  #geom_hline(yintercept = 0.05, linetype = "solid", color = "black", linewidth = 1)+
  #geom_hline(yintercept = 0.05, color = "black", linetype = "solid", linewidth = 0.5) +  # Add line at 0.05 
  geom_point(size = 10, color = "black") + 
  scale_size(range = c(2, 20)) + 
  xlab('') + 
  ylab("Fisher p-value") +
  theme_bw(base_size = 35) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); plot

# Save plot 
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_sig_pos_corr_target_enrichment.pdf", last_plot(), width = 20, height = 25, dpi = 300)
ggsave("../../F-Ptuh/output/11-Ptuh-mRNA-miRNA-interactions/Ptuh-top10GO_sig_pos_corr_target_enrichment.png", last_plot(), width = 20, height = 25, dpi = 300)
```




