---
title: "32-Ptuh-ncRNA-machinery-BLAST"
author: "Kathleen Durkin"
date: "2026-04-15"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: false 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: ../../references.bib
link-citations: true
---


```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(DT)
library(Biostrings)
library(tm)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center", # Align plots to the center
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Proteins 

Will be using P. meandrina protein sequences here: 
https://gannet.fish.washington.edu/seashell/snaps/Pocillopora_meandrina_HIv1.genes.pep.faa


Download if necessary
```{bash}
cd ../data
curl -o Pocillopora_meandrina_HIv1.genes.pep.faa https://gannet.fish.washington.edu/seashell/snaps/Pocillopora_meandrina_HIv1.genes.pep.faa

```


```{bash}
head ../data/Pocillopora_meandrina_HIv1.genes.pep.faa
```

# Make BLAST db

```{bash}

/home/shared/ncbi-blast-2.15.0+/bin/makeblastdb \
-in ../data/Pocillopora_meandrina_HIv1.genes.pep.faa \
-dbtype prot \
-out ../output/32-Ptuh-ncRNA-machinery-BLAST/Ptuh-proteins
```

```{bash}
head ../../data/ncRNA_machinery.fasta
```

# BLASTp

```{bash}
fasta="../../data/ncRNA_machinery.fasta"

/home/shared/ncbi-blast-2.15.0+/bin/blastp \
-query $fasta \
-db ../output/32-Ptuh-ncRNA-machinery-BLAST/Ptuh-proteins \
-out ../output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_out.tab \
-evalue 1E-05 \
-num_threads 48 \
-max_target_seqs 1 \
-max_hsps 1 \
-outfmt 6
```

```{r, engine='bash', eval=TRUE}       
wc -l ../output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_out.tab
```

```{r, engine='bash', eval=TRUE}        
head ../output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_out.tab
```

# Parse and annotate

This .tab file associates A. pulchra proteins with ncRNA machinery proteins found in the reference db. However, the names are just taken straight from the fasta headers, which ends up being mostly just the Uniprot accession number. Let's parse this into a more interpretable db.

parse the reference ncRNA machinery fasta headers

```{r, engine='bash'}
cd ../../data/
  
grep "^>" ncRNA_machinery.fasta | \
sed 's/^>//' | \
awk -F'|' '{
  # Get accession
  acc = $2
  
  # Everything after the third pipe
  rest = $3
  
  # Strip entry_name (first token before space)
  sub(/^[^ ]+ /, "", rest)
  
  # Parse fields using OS=, OX=, GN=, CGN=, PE=
  protein = rest; sub(/ CGN=.*/, "", protein); sub(/ OS=.*/, "", protein)
  
  species = rest; sub(/.*OS=/, "", species); sub(/ OX=.*/, "", species)
  
  if (rest ~ /GN=/) {
    gene = rest; sub(/.*GN=/, "", gene); sub(/ PE=.*/, "", gene)
  } else {
    gene = "NA"
  }
  
  if (rest ~ /CGN=/) {
    cgn = rest; sub(/.*CGN=/, "", cgn); sub(/ OS=.*/, "", cgn)
  } else {
    cgn = "NA"
  }
  
  # Quote protein and species in case of commas
  gsub(/"/, "\"\"", protein)
  gsub(/"/, "\"\"", species)
  
  print acc "," "\"" protein "\"" "," gene "," cgn "," "\"" species "\""
}' | \
sed '1i accession,protein_name,gene_name,std_gene_name,species' \
> ncRNA_machinery_reference_table.csv
```

Now parse blastp output rows to reduce to accession numbers

```{bash}
cd ../output/32-Ptuh-ncRNA-machinery-BLAST

awk -F'\t' 'BEGIN{OFS="\t"} {split($1,a,"|"); $1=a[2]; print}' ncRNAmach-blastp-Ptuh_out.tab > ncRNAmach-blastp-Ptuh_parsed.tab
```

Now read in and match with reference db, so that blast results are associated with protein/gene names (not just accession numbers)

(Note that I can't rely on just the gene names, because most of the Uniprot entries used non-standard gene names)

```{r}
library(dplyr)

# Read reference table
ref <- read.csv("../../data/ncRNA_machinery_reference_table.csv", stringsAsFactors = FALSE)

# Read parsed BLAST output
blast <- read.delim("../output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_parsed.tab", header = FALSE,
                    col.names = c("accession", "target", "pident", "length",
                                  "mismatch", "gapopen", "qstart", "qend",
                                  "sstart", "send", "evalue", "bitscore"))

# Join
blast_annotated <- blast %>%
  left_join(ref, by = "accession") %>%
  dplyr::select(target, protein_name, gene_name, std_gene_name, species, accession, pident, length, mismatch, gapopen, qstart, qend, sstart, send, evalue, bitscore)

write.csv(blast_annotated, "../output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_annotated.csv", row.names = FALSE)
```

Finally, let's reduce this to unique associations. In the ncNRA machinery db I compiled, I sometimes included multiple sequences of a given protein sequenced in different species. This was partly due to convenience, but also because (a) I wanted "backups" in case one of the sequences was mis-annotated or had other problems (these are all unreviewed entries), and (b) in case a protein is highly divergent, and only a subset of its entries matched our species. 

Now that I've done the blast, though, I only really need to know the [p.evermanni mRNA] - [standardized gene name] association. Let's make this reduced db.

```{r}
blast_reduc <- blast_annotated %>% dplyr::select(target, std_gene_name) %>% distinct()

write.csv(blast_reduc, "../output/32-Ptuh-ncRNA-machinery-BLAST/ncRNAmach-blastp-Ptuh_annotated_reduced.csv", row.names = FALSE)
```

